### 2022-10-18 code/Matrinem data analysis.Rmd
Uploaded data and code to repo

---
title: "Results section"
author: "Kaare"
date: "6/16/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r loading libraries}
library(readxl)
library(tidyverse)
library(broom)
library(ggpubr)
#library(AICcmodavg)
library(visdat)
library(rstatix)
library(gt)
library(gtsummary)
library(patchwork)
```

# Introduction

|           | PreExp 1 | PreExp 2 | PreExp 3 | NEC-FvT |
|-----------|----------|----------|----------|---------|
| **Aim**   |          |          |          |         |
| Protocols |          |          |          |         |
| N         |          |          |          |         |

## Pre-Experiment 1 (PreExp 1)

Experiment 1 was a pilot to test whether it was possible to feed 3-day old mice with the PICC. The experimental setup followed the standard setup. Antibiotics was administered according to the description in Chen et al., though with a lower dose based on previous experiments. Our primary study objective was survival. Secondary objective was macroscopic evaluation of intestines for inflammation, intestinal distention and similar NEC-like symptoms. One preliminary hypothesis was a reduction in weight of offspring from antibiotics-treated mothers.

#### Weight development

# Experiment 2 - weight at birth.Summary: Offspring from antibiotics-treated mice were smaller at birth

```{r}
data <- read_excel("../Data/CleanData/weight_development.xlsx") %>% 
  select(sample_id:cage, bodyweight_birthday_group_average, bodyweight_birthday_plus_24H_group_average, bodyweight_birthday_plus_48H_group_average, bodyweight_baseline, bodyweight_baseline_plus_24h, bodyweight_baseline_plus_48h) %>% 
  rename(day_0 = "bodyweight_birthday_group_average",
         day_1 = "bodyweight_birthday_plus_24H_group_average", 
         day_2 = "bodyweight_birthday_plus_48H_group_average",
         day_3 = bodyweight_baseline, 
         day_4 = `bodyweight_baseline_plus_24h`, 
         day_5 = `bodyweight_baseline_plus_48h`,) %>% 
  pivot_longer(., cols = c(day_0:day_5),
                names_to = "time",
                values_to = "bodyweight") %>% 
  mutate(received_antibiotics = recode_factor(received_antibiotics, 
                                       yes = "MAT",
                                       no = "CON"))
  #%>%
  # #mutate(across(where(is_character),as_factor))

exp2.bw.density <- data %>% 
  filter(exp_number == "2") %>% 
  filter(time == 'day_0') %>% 
  ggplot(aes(x = bodyweight, color = received_antibiotics)) +
    geom_density() +
  geom_vline(data = data %>% 
              filter(exp_number == "2") %>% 
  filter(time == 'day_0') %>% 
  group_by(received_antibiotics) %>% 
  summarise(grp.mean = mean(bodyweight)), aes(xintercept=grp.mean, color=received_antibiotics),
             linetype="dashed") +
  guides(fill = guide_legend(title = ""),
         colour = guide_legend(title = " ")) + 
  labs(title = "", y = "n", x = "Bodyweight (g)") +
  theme_minimal()

exp2.bw.col <- data %>% 
  filter(exp_number == "2") %>% 
  filter(time == 'day_0') %>% 
  ggplot(aes(x = received_antibiotics, y = bodyweight, fill = received_antibiotics)) +
  geom_bar(stat = "summary",
           fun = "mean") +
  labs(title = "", y = "Mean body weight", x = "") +
  theme_minimal() +  
  stat_compare_means(
    method = "t.test", 
    label.y = 1.6, 
    label.x.npc = 0.4) +
  theme(legend.position="none")

birth_weight_patch <- (exp2.bw.density + exp2.bw.col) + plot_annotation(
  title = 'Effect of Maternal Antibiotics Treatment on birth weight of offspring',
  subtitle = 'Experimental round 2',
  tag_levels = 'A',
  caption = 'MAT = Maternal Antibiotics Treatment, CON = Water Controls \n 
  Statistics: non-paired two.sided t-test.'
)
birth_weight_patch
#ggsave("../Results/Weight/Figure_1_Birthweights.png", birth_weight_patch)

#add gt table with group results

```

```{r eval=FALSE, include=FALSE}
### Weight development during experiments   ###########

bw_development_feed <- data %>% 
filter(exp_number %in% c("1","3","4")) %>% 
filter(time %in% c("day_3", "day_5")) %>%
ggplot(aes(x = time ,
           y = bodyweight, 
           fill = type_of_feed)) +
  facet_wrap(~ exp_number + type_of_feed, 
             ncol = 2) + 
  geom_bar(stat = "identity",
           position = "dodge") + 
  guides(fill = guide_legend(title = " ")) +
  labs(
      caption = "Experimental round 2 not shown due to no formula feeding.",
    y = "Bodyweight (g)"
  ) +
  theme_bw()

bw_development_ab <- data %>% 
filter(exp_number %in% c("1","2","3","4")) %>% 
filter(time %in% c("day_3", "day_5")) %>%
ggplot(aes(x = time ,
           y = bodyweight, 
           fill = received_antibiotics)) +
  facet_wrap(~ exp_number + received_antibiotics, 
             ncol = 2) + 
  geom_bar(stat = "identity",
           position = "dodge") + 
  guides(fill = guide_legend(title = " ")) +
  labs(caption = "All mice recieved antibiotics in experimental round 4.",
    y = "Bodyweight (g)"
  ) +
  theme_bw()

bodyweight_develop_patch <- (bw_development_feed + bw_development_ab) + plot_annotation (
  title = 'Effect of Maternal Antibiotics Treatment on body weight development of offspring',
  subtitle = 'Faceted by experiment and Feeding or MAT-status',
  tag_levels = 'A',
  caption = 'Form = formular feed, Milk = breastfeeding.  
      MAT = Maternal Antibiotics Treatment, CON = Water Controls.')

bodyweight_develop_patch
#ggsave("../Results/Weight/Figure_2_bodyweight_development.png", bodyweight_develop_patch)

```

### Research Question 1: Body weight of offspring from water and antibiotic-treated mice at baseline

```{r style baseline bodyweight}
get_box_stats <- function(y, upper_limit = max(data$Bodyweight[data$Time == "Day2"]) * 1.0) {
  return(data.frame(
    y = 0.9 * upper_limit,
    label = paste(
      "n =", length(y), "\n",
      "Mean =", round(mean(y), 2), "\n" #,
      #"Median =", round(median(y), 2), "\n"
    )
  ))
}

my_comparisons <- list( c("Water", "AB"))

data %>% 
  filter(Time == "Day2") %>% 
  ggplot(aes(
    received_antibiotic, Bodyweight
  )) + 
  geom_boxplot() +
  facet_wrap(~experiment_no, 
             nrow = 1) + 
  geom_dotplot(
     binaxis = "y", stackdir = "center", dotsize = 0.5) + 
  stat_summary(
    fun.data = get_box_stats, geom = "text", hjust = 0.5, vjust = 0.9) +
  labs(
    title = "Body weight of offspring from antibiotic-treated vs water-treated mice",
    subtitle = "Results shown from each of the 4 experimental rounds",
    caption = "In Exp. 4 all mice received antibiotoics, hence no control.
    Comparison method: Student T-test",
    x = "Maternal Treatment",
    y = "Bodyweight at end of study (g)"
  ) +
  theme_bw() + 
  stat_compare_means(
    method = "t.test", 
    label.y = 3, 
    comparisons = my_comparisons
    )
```

```{r weightgain plots}
############# Effect of Formula vs. Breastfeeding on Weight Gain" ###################

bodyweight_gain_BFvsFORM <- data %>% 
  mutate(weightgain = weight_end - weight_baseline) %>% 
  ggplot(aes(
    type_of_feed, weightgain
  )) + 
  geom_boxplot() +
  facet_wrap(~experiment_no, 
             nrow = 1) +
geom_dotplot(
     binaxis = "y", stackdir = "center", dotsize = 0.5) + 
  stat_summary(
    fun.data = get_box_stats, geom = "text", hjust = 0.2, vjust = 0.9) +
  labs(
    title = "Effect of Formula vs. Breastfeeding on Weight Gain",
    subtitle = "Results shown from each of the 4 experimental rounds",
    caption = "In Exp. 2 all mice were breastfed.
    Comparison method: Student T-test",
    x = "Type of feed",
    y = "Body weight increase  (g)"
  ) +
  theme_bw() + 
  stat_compare_means(
    method = "t.test", 
    label.y = 3, 
    comparisons = my_comparisons_2
    )
my_comparisons_2 <- list( c("Breastmilk", "Formula"))
bodyweight_gain_BFvsFORM
```

```{r weightgain plot 2}
############# Effect of  Antibiotocis on weightgain in Breastfed offspring" ################

bodyweight_gain_WATER_BF_vs_AB_BF <- 
  data %>% 
  mutate(weightgain = weight_end - weight_baseline) %>% 
  filter(type_of_feed == "Breastmilk") %>% 
  ggplot(aes(
    received_antibiotic, weightgain#, colour = mother
    )) + 
  geom_boxplot()+
  geom_point()+
  facet_wrap(~experiment_no, 
             nrow = 1) +
  labs(
    title = "Breastfed offspring +/- Maternal Antiobiotics Treatment"
  )
bodyweight_gain_WATER_BF_vs_AB_BF
```

#### Histological analysis

#### Cytokine expression

boxpltos

```{r}
boxplots_all_genes <- delta_gene_expr %>% 
  ggplot(
    aes(x = groups,
        y = relative_fold_change,
        fill = groups)
  ) + 
  geom_boxplot() +
  facet_wrap(~genes,
             ncol = 5, 
             scales = "free") +
  labs(title = "Gene expression fold change",
       y = "Gene expression fold change",
       x = "Groups"
      )
boxplots_all_genes

```

#### Gut microbiome analysis

### Experiment 2 (E2)

#### Weight development

#### Histological analysis

#### Intestinal permeability

#### Cytokine expression

#### Gene expression

#### Gut microbiome analysis

### Experiment 4 (NEC-FvT)

The PreExps are now over and it is time to conduct the study. Having determined the optimal experimental protocol in PreExp 1-3 we are ready with the design.

#### Overview - Experimental Setup

#### Intestinal permeability - FITC

```{r FITC load data, include=FALSE}
fitc <- read_excel("../Data/CleanData/fitc.xlsx") %>% 
  rename(values = `result_µg/ml_undiluted`) %>% 
  drop_na(values)
```

Intestinal permeability is assessed with the FITC-Dextran assay. First I inspect the distribution of results.

```{r FITC check for normality}
fitc %>% 
  filter(exp_number == 4) %>% 
  group_by(group) %>% 
  shapiro_test(values)
```

As results and highly skewed I try transforming the FITC results to log10.

```{r FITC try log transforming data}
fitc <- fitc %>% 
  mutate(log_10_values = log10(values))  # Logtransform results

```

```{r FITC check for normality log transform}
fitc %>% 
  filter(exp_number == 4) %>% 
  group_by(group) %>% 
  shapiro_test(log_10_values)
```

```{r FITC visual inspections of data distribution}
#Density plots non-log
density <- fitc %>% 
  filter(exp_number == 4) %>% 
  ggplot(aes(x = values, fill = group)) + 
  geom_density() +
  facet_wrap(~ group,
             scales = "free")

#Jitter plots non-log
jitter <- fitc %>% 
  filter(exp_number == 4) %>% 
  ggplot(aes(x = group, y = values)) + 
  geom_jitter() +
  facet_wrap(~ group,
             scales = "free")

#Density plots log
density <- fitc %>% 
  filter(exp_number == 4) %>% 
  ggplot(aes(x = , fill = group)) + 
  geom_density() +
  facet_wrap(~ group,
             scales = "free")

#Jitter plots non-log
jitter <- fitc %>% 
  filter(exp_number == 4) %>% 
  ggplot(aes(x = group, y = values)) + 
  geom_jitter() +
  facet_wrap(~ group,
             scales = "free")

```

#### Cytokine expression

#### Gene expression

#### Histological analysis

#### Gut microbiome analysis

### 2022-10-18 code/Round 1/Cytokines adjusted.Rmd
Uploaded data and code to repo

---
title: "Cytokine analysis script - round 1 adjusted"
author: "Kaare"
date: "11th October 2022"    
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

```{r load packages, include=FALSE}
library(readxl) 
library(tidyverse)
library(gt)
library(gtsummary)
library(broom)
library(rstatix)
library(knitr)
library(patchwork)
library(ggpubr)
```
# Introduction to script 
This script is useful for analysing cytokine expression from MSD Mesoscale assays. 
It is currently optimized for the 'Pro-inflammatory Vplex 10 Mouse kit', analysing the cytokines INFg, IL_10, IL_12p70, IL_1b, IL_2, IL_4,IL_5, IL_6, KCGRO, and TNFa.

Required data structure is a wide tibble (column headers = variables, samples and values = rows).

This script is per default set to handle ileum and serum expression, although any tissue will work. This will just require resetting of the "tissue"-variable.

# Cytokines for experiment 1. - Tissue weight adjusted.

### Basic data structure and results table.

First you load the data and check the structure. Make sure cytokine concentrations are all numeric (<dbl>). If not, mutate non-numeric columns with mutate_if(is_character).  
```{r load data, echo=FALSE}
# Import dataset
cytokines <- read_excel("../../Data/RawData/cytokines_round_1_r.xlsm")
glimpse(cytokines)

cytokines <- cytokines %>% 
  pivot_longer(.,
               cols = INFg:TNFa ,
               names_to = "variables",
               values_to = "values"
  ) %>% 
  drop_na(
    values) %>% 
  mutate(MAT = ifelse(str_detect(group,'AB'),'AB',
                                  ifelse(str_detect(group,'CON'),'CON', "miss")),
         FEED = ifelse(str_detect(group,'FF'),'FF',
                                  ifelse(str_detect(group,'BF'),'BF', "miss"))) %>% 
  mutate(
    group = recode_factor(  #Change the order of the groups to desired plot order
      group,
      "CON-BF"= "CON-BF",
      "CON-FF" = "CON-FF",
      "AB-BF"= "AB-BF",
      "AB-FF"= "AB-FF"
      )) %>% 
  mutate(values = if_else(tissue == "Illeum", values/tissue_weight_mg, values)) %>% 
  pivot_wider(id_cols = c(sample,group,tissue,tissue_weight_mg, MAT,FEED), names_from = variables, values_from = values)
```

Set the tissue type to whatever tissue your data are extracted from, or that you want to analyze first.
```{r load data and set tissue type, include=FALSE}
#set tissue type
sample_tissue <- "Illeum"
```

Below is a table of the mean levels of cytokines in `r sample_tissue` tissue for each of the experimental groups.

The values will be displayed as mean + sd if normally distributed or median + IQR if not. An appropriate statistical test for differences will be performed (student-t for two, normally distributed groups, Anova for more than 2 groups, Kruskal wallis for non-normally distrubuted results.)
```{r}
tbl1 <- cytokines %>% 
  filter(tissue == sample_tissue) %>% 
  select(group,INFg:TNFa) %>% 
  tbl_summary(by = group,
              digits = everything() ~ c(2,1,1),
              missing = "no",
              include = !INFg) %>%
  add_p() %>% 
  modify_header(
    label ~ "**Groups**")

tbl1
# Save and export files
# tbl1 %>%  #Save as excel file.
#   as_hux_table() %>%
#   huxtable::quick_xlsx(., file = "../Results/Round 1/ileum-cytokines.xlsx")
# 
# tbl1 %>% #Save as image file. 
# as_flex_table() %>%
#   flextable::save_as_image(path = "../../Results/Round 1/ileum_cytokines_table.png",webshot = "webshot")
```
This table can be exported with the following code either to word, excel, pdf, html or image files. See the syntax of huxtable and flextable (?flextable). 
```{r ileum cytokine export, echo=TRUE, warning=FALSE}
tbl1_p <- tbl1 %>% 
  filter_p() %>% 
  as_tibble()
```

### Plots and regression models
First the data must be transformed to a long format (pivot longer). Mutate any new columns  with group treatments needed for the regression model (in this case MAT and FEED (experimental factors)). 
```{r long format, echo=FALSE}
cytokines_long <- cytokines %>% 
  pivot_longer(.,
               cols = INFg:TNFa ,
               names_to = "variables",
               values_to = "values"
  ) %>% 
  drop_na(
    values) %>% 
  mutate(MAT = ifelse(str_detect(group,'AB'),'AB',
                                  ifelse(str_detect(group,'CON'),'CON', "miss")),
         FEED = ifelse(str_detect(group,'FF'),'FF',
                                  ifelse(str_detect(group,'BF'),'BF', "miss"))) %>% 
  mutate(
    group = recode_factor(  #Change the order of the groups to desired plot order
      group,
      "CON-BF"= "CON-BF",
      "CON-FF" = "CON-FF",
      "AB-BF"= "AB-BF",
      "AB-FF"= "AB-FF"
      ))

head(cytokines_long)

```

To see all the cytokine leves plotted run this code: 
```{r plots all cytokines, warning=FALSE}
plot_all_cytokines <- cytokines_long %>%
  filter(tissue == sample_tissue) %>% 
  filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot
ggplot(aes(x = group, 
           y = values, 
           fill = group)) + 
  geom_boxplot() +
  facet_wrap(~variables, 
             scales = 'free', 
             nrow = 3,
             ncol = 3) +
  labs(title = "Levels of pro-inflammatory cytokines in the ileum",  # Change title, and axis names
       x = "",
       y = "Tissue conc. (pg/mg)",
       fill = "Groups") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1")

all_cytokines_patch <- plot_all_cytokines + plot_annotation (
  title = 'Cytokine expression',
   subtitle = 'All cytokines measured',
  tag_levels = 'A',
  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

all_cytokines_patch
#ggsave('../../Results/Round 1/All Cytokines-boxplot.png',all_cytokines_patch)
```

Statistically significant group differences were observed for IL-6 (`r inline_text(tbl1, column = "p.value", variable = "IL_6")`), KCGRO (`r inline_text(tbl1, column = "p.value", variable = "KCGRO")`) and TNFa (`r inline_text(tbl1, column = "p.value", variable = "TNFa")`).

These cytokines are plottet below:
```{r}
#tbl1_p contains cytokines with significant changes

signif_cytokines <- cytokines_long %>%
  filter(tissue == sample_tissue) %>% 
  filter(variables %in% tbl1_p$`**Groups**`) %>%  #Exclude any cytokines not wanted in the plot
ggplot(aes(x = group, 
           y = values, 
           fill = group)) + 
  geom_boxplot() +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.55) +
  stat_compare_means(
    aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
    label.y.npc = 0.9,
    label.x.npc = 0.5) +
  facet_wrap(~variables, 
             scales = 'free', 
             nrow = 3,
             ncol = 3) +
  labs(title = "Levels of pro-inflammatory cytokines in the ileum", # Change title, and axis names
       #subtitle = "Statistically significant changes", 
       x = "",
       y = "Tissue conc. (pg/mg)",
       fill = "Groups") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1")

signif_cytokines_patch <- signif_cytokines + plot_annotation (
  title = 'Cytokine expression',
  subtitle = 'Statistically significant changes',
  tag_levels = 'A',
  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding. \n
  Pairwise comparisons with CON-BF using wilcoxon test',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))
signif_cytokines_patch
#ggsave('../../Results/Round 1/Significant Cytokines-boxplot.png',signif_cytokines_patch)
```

A regression model of the effects of type of feed and antibiotics treatment
```{r echo=FALSE, warning=FALSE}
#### Make regression model on long format
linear_reg_model <- cytokines_long %>% 
  filter(tissue == sample_tissue) %>% 
  group_by(variables) %>% 
  do(lm(data = ., values ~ MAT+FEED+MAT*FEED) %>% tidy)  %>% filter(term!='(Intercept)')

linear_reg_model %>%  
  select(-statistic) %>% 
  kable(., caption = 'Effect of treatment factors on cytokine expression levels', digits = 3)

#### Make regression model on one variable
# reg_model_1 <- cytokines_long %>% 
#   filter(tissue == sample_tissue) %>% 
#   filter(variables %in% "TNFa") %>% 
#   lm(data =., values ~ FEED + MAT)
# summary(reg_model_1)
```

**The conclusion** is that many of the cytokines show an interaction effect of the two treatment parameters

### 2022-10-18 code/Round 1/Cytokines unadjusted.Rmd
Uploaded data and code to repo

--- 2022-10-18 code/Round 1/Cytokines adjusted.Rmd
@@
 ---
-title: "Cytokine analysis script - round 1 adjusted"
+title: "Cytokine analysis script - round 1"
 author: "Kaare"
@@
 output: html_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 
-# Cytokines for experiment 1. - Tissue weight adjusted.
+# Cytokines for experiment 1.
 
@@
 # Import dataset
-cytokines <- read_excel("../../Data/RawData/cytokines_round_1_r.xlsm")
+#cytokines <- read_excel("../../Data/RawData/cytokines_round_1_r.xlsm")
+cytokines <- read_excel("../../Data/CleanData/cytokines.xlsx")
 glimpse(cytokines)
-
-cytokines <- cytokines %>% 
-  pivot_longer(.,
-               cols = INFg:TNFa ,
-               names_to = "variables",
-               values_to = "values"
-  ) %>% 
-  drop_na(
-    values) %>% 
-  mutate(MAT = ifelse(str_detect(group,'AB'),'AB',
-                                  ifelse(str_detect(group,'CON'),'CON', "miss")),
-         FEED = ifelse(str_detect(group,'FF'),'FF',
-                                  ifelse(str_detect(group,'BF'),'BF', "miss"))) %>% 
-  mutate(
-    group = recode_factor(  #Change the order of the groups to desired plot order
-      group,
-      "CON-BF"= "CON-BF",
-      "CON-FF" = "CON-FF",
-      "AB-BF"= "AB-BF",
-      "AB-FF"= "AB-FF"
-      )) %>% 
-  mutate(values = if_else(tissue == "Illeum", values/tissue_weight_mg, values)) %>% 
-  pivot_wider(id_cols = c(sample,group,tissue,tissue_weight_mg, MAT,FEED), names_from = variables, values_from = values)
 ```
@@
 #set tissue type
-sample_tissue <- "Illeum"
+sample_tissue <- "ileum"
+study_round <- "1"
 ```
@@
 tbl1 <- cytokines %>% 
+  filter(exp_number == study_round) %>% 
   filter(tissue == sample_tissue) %>% 
@@
               missing = "no",
-              include = !INFg) %>%
+              include = !"INFg") %>%
   add_p() %>% 
   modify_header(
-    label ~ "**Groups**")
+    label ~ "Cytokines") %>% 
+  modify_spanning_header((c("stat_1", "stat_2", "stat_3", "stat_4") ~ "Treatment groups (N total = {N})"))
 
-tbl1
+  tbl1_gt <- tbl1 %>% 
+  as_gt() %>%
+  tab_header(title = md("**Proinflammatory cytokines**")) %>%
+  opt_align_table_header("center") %>% 
+  tab_options(heading.title.font.size = 20,
+              heading.background.color = "lightgrey",
+              ) %>% 
+  tab_source_note(
+    source_note = " * P-value indicates overall group differences"
+  ) %>%
+  tab_source_note(
+    source_note = md("**Matrinem round 1**")
+  ) %>%
+  cols_width(label ~ px(100),
+             stat_1 ~ px(150),
+             stat_2 ~ px(150),
+             stat_3 ~ px(150),
+             stat_4 ~ px(150)) 
+tbl1_gt
 # Save and export files
@@
 cytokines_long <- cytokines %>% 
+  filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
   pivot_longer(.,
@@
 plot_all_cytokines <- cytokines_long %>%
-  filter(tissue == sample_tissue) %>% 
   filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot
@@
 signif_cytokines <- cytokines_long %>%
-  filter(tissue == sample_tissue) %>% 
   filter(variables %in% tbl1_p$`**Groups**`) %>%  #Exclude any cytokines not wanted in the plot
@@
 linear_reg_model <- cytokines_long %>% 
-  filter(tissue == sample_tissue) %>% 
   group_by(variables) %>% 

### 2022-10-18 code/Round 2/Cytokines round 2.Rmd
Uploaded data and code to repo

---
title: "Cytokine analysis script - round 2"
author: "Kaare"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

```{r load packages, include=FALSE}
library(readxl) 
library(tidyverse)
library(gt)
library(gtsummary)
library(broom)
library(rstatix)
library(knitr)
library(patchwork)
library(ggpubr)
```

# Introduction to script

This script is useful for analyzing cytokine expression from MSD Mesoscale assays. It is currently optimized for the 'Pro-inflammatory Vplex 10 Mouse kit', analysing the cytokines INFg, IL_10, IL_12p70, IL_1b, IL_2, IL_4,IL_5, IL_6, KCGRO, and TNFa.

Required data structure is a wide tibble (column headers = variables, samples and values = rows).

This script is per default set to handle ileum and serum expression, although any tissue will work. This will just require resetting of the "tissue"-variable.

# Cytokines for experiment 2.

### Basic data structure and results table.

First you load the data and check the structure. Make sure cytokine concentrations are all numeric (<dbl>). If not, mutate non-numeric columns with mutate_if(is_character).

```{r load data, echo=FALSE}
# Import dataset
#cytokines <- read_excel("../../Data/RawData/cytokines_round_1_r.xlsm")
cytokines <- read_excel("../../Data/CleanData/cytokines.xlsx")
glimpse(cytokines)
```

Set the tissue type to whatever tissue your data are extracted from, or that you want to analyze first.

```{r load data and set tissue type, include=FALSE}
#set tissue type
sample_tissue <- "ileum"
study_round <- "2"
```

Below is a table of the mean levels of cytokines in `r sample_tissue` tissue for each of the experimental groups.

The values will be displayed as mean + sd if normally distributed or median + IQR if not. An appropriate statistical test for differences will be performed (student-t for two, normally distributed groups, Anova for more than 2 groups, Kruskal wallis for non-normally distrubuted results.)

```{r}
tbl1 <- cytokines %>% 
  filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>% 
  select(group,INFg:TNFa) %>% 
  tbl_summary(by = group,
              digits = everything() ~ c(2,1,1),
              missing = "no") %>%
  add_p() %>% 
  modify_header(
    label ~ "Cytokines") %>% 
  modify_spanning_header((c("stat_1", "stat_2", "stat_3") ~ "Treatment groups (N total = {N})"))

tbl1_gt <- tbl1 %>% 
 as_gt() %>%
  tab_header(title = md("**Proinflammatory cytokines**")) %>%
  opt_align_table_header("center") %>% 
  tab_options(heading.title.font.size = 20,
              heading.background.color = "lightgrey",
              ) %>% 
  tab_source_note(
    source_note = " * P-value indicates overall group differences"
  ) %>%
  tab_source_note(
    source_note = md("**Matrinem round 2**")
  ) %>%
  cols_width(label ~ px(100),
             stat_1 ~ px(150),
             stat_2 ~ px(150),
             stat_3 ~ px(150)) 
tbl1_gt
# Save and export files
# tbl1 %>%  #Save as excel file.
#   as_hux_table() %>%
#   huxtable::quick_xlsx(., file = "../Results/Round 1/ileum-cytokines.xlsx")
# 
# tbl1 %>% #Save as image file.
# as_flex_table() %>%
#   flextable::save_as_image(path = "../../Results/Round 2/ileum_cytokines_table.png",webshot = "webshot")
```

This table can be exported with the following code either to word, excel, pdf, html or image files. See the syntax of huxtable and flextable (?flextable).

```{r ileum cytokine export, echo=TRUE, warning=FALSE}
tbl1_p <- tbl1 %>%
  filter_p() %>% 
  as_tibble()
```

### Plots and regression models

First the data must be transformed to a long format (pivot longer). Mutate any new columns with group treatments needed for the regression model (in this case MAT and FEED (experimental factors)).

```{r long format, echo=FALSE}
cytokines_long <- cytokines %>% 
  filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  pivot_longer(.,
               cols = INFg:TNFa ,
               names_to = "variables",
               values_to = "values"
  ) %>% 
  drop_na(
    values) %>% 
  # mutate(MAT = ifelse(str_detect(group,'AB'),'AB',
  #                                 ifelse(str_detect(group,'CON'),'CON', "miss")),
  #        FEED = ifelse(str_detect(group,'FF'),'FF',
  #                                 ifelse(str_detect(group,'BF'),'BF', "miss"))) %>% 
  mutate(
    group = recode_factor(  #Change the order of the groups to desired plot order
      group,
      "CON-BF"= "CON-BF",
      "CON-FF" = "CON-FF",
      "AB-BF"= "AB-BF",
      "AB-FF"= "AB-FF"
      ))

head(cytokines_long)

```

To see all the cytokine leves plotted run this code:

```{r plots all cytokines, warning=FALSE}
plot_all_cytokines <- cytokines_long %>%
  #filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot
ggplot(aes(x = group, 
           y = values, 
           fill = group)) + 
  geom_boxplot() +
  facet_wrap(~variables, 
             scales = 'free', 
             nrow = 5,
             ncol = 2) +
  labs(title = "Levels of pro-inflammatory cytokines in the ileum",  # Change title, and axis names
       x = "",
       y = "Tissue conc.",
       fill = "Groups",
       caption = "Matrinem round 2") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="lightgrey"), # Change legend box background color
    plot.caption = element_text(color = "darkgrey", size=5)) + 
  scale_fill_brewer(palette="Set1")

all_cytokines_patch <- plot_all_cytokines + plot_annotation (
  title = 'Cytokine expression',
   subtitle = 'All cytokines measured',
  tag_levels = 'A',
  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

all_cytokines_patch
#ggsave('../../Results/Round 1/All Cytokines-boxplot.png',all_cytokines_patch)
```

Statistically significant group differences were observed for \`r ifelse([tbl1_p$Cytokines] > 0, tbl1_1$Cytokines, "no cytokines)\`\`

These cytokines are plottet below:

```{r}
#tbl1_p contains cytokines with significant changes

signif_cytokines <- cytokines_long %>%
  filter(variables %in% tbl1_p$Cytokines) %>%  #Exclude any cytokines not wanted in the plot
ggplot(aes(x = group, 
           y = values, 
           fill = group)) + 
  geom_boxplot() +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.55) +
  stat_compare_means(
    aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
    label.y.npc = 0.9,
    label.x.npc = 0.5) +
  facet_wrap(~variables, 
             scales = 'free', 
             nrow = 3,
             ncol = 3) +
  labs(title = "Levels of pro-inflammatory cytokines in the ileum", # Change title, and axis names
       #subtitle = "Statistically significant changes", 
       x = "",
       y = "Tissue conc. (pg/mg)",
       fill = "Groups",
       caption = "Matrinem round 2") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1")

signif_cytokines_patch <- signif_cytokines + plot_annotation (
  title = 'Cytokine expression',
  subtitle = 'Statistically significant changes',
  tag_levels = 'A',
  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding. \n
  Pairwise comparisons with CON-BF using wilcoxon test',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))
signif_cytokines_patch
#ggsave('../../Results/Round 1/Significant Cytokines-boxplot.png',signif_cytokines_patch)
```

A regression model of the effects of type of feed and antibiotics treatment

```{r echo=FALSE, warning=FALSE}
#### Make regression model on long format
linear_reg_model <- cytokines_long %>% 
  group_by(variables) %>% 
  do(lm(data = ., values ~ MAT+FEED+MAT*FEED) %>% tidy)  %>% filter(term!='(Intercept)')

linear_reg_model %>%  
  select(-statistic) %>% 
  kable(., caption = 'Effect of treatment factors on cytokine expression levels', digits = 3)

#### Make regression model on one variable
# reg_model_1 <- cytokines_long %>% 
#   filter(tissue == sample_tissue) %>% 
#   filter(variables %in% "TNFa") %>% 
#   lm(data =., values ~ FEED + MAT)
# summary(reg_model_1)
```

**The conclusion** is that many of the cytokines show an interaction effect of the two treatment parameters

### 2022-10-18 code/Round 2/FITC round 2.Rmd
Uploaded data and code to repo

---
title: "FITC analysis script - round 2"
author: "Kaare"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

```{r load packages, include=FALSE}
library(readxl) 
library(tidyverse)
library(gt)
library(gtsummary)
library(broom)
library(rstatix)
library(knitr)
library(patchwork)
library(ggpubr)
```

# Introduction to script

This script is useful for analyzing intestinal permeability from FITC-Dextran assays.

Required data structure is a wide tibble (column headers = variables, samples and values = rows).

This script is per default set to handle ileum and serum expression, although any tissue will work. This will just require resetting of the "tissue"-variable.

# FITC for experiment 2.

### Basic data structure and results table.

First you load the data and check the structure. Make sure cytokine concentrations are all numeric (<dbl>). If not, mutate non-numeric columns with mutate_if(is_character).

```{r load data, echo=FALSE}
# Import dataset
fitc <- read_excel("../../Data/CleanData/fitc.xlsx")
glimpse(fitc)
```

Set the tissue type to whatever tissue your data are extracted from, or that you want to analyze first.

```{r load data and set tissue type, include=FALSE}
#set tissue type
study_round <- "2"

fitc<- fitc %>% 
  rename("µg/ml" = `result_µg/ml_undiluted`)
```

```{r data distribution}
no_outliers <- fitc %>% 
  filter(exp_number == 2) %>% 
  group_by(group2) %>% 
  mutate(fitc_level = ifelse(fitc_level > 10, mean(fitc_level), fitc_level))

  no_outliers %>% 
qplot(x = group2, y = fitc_level, data = ., geom = "jitter",
    color = group2)

```

Below is a table of the levels of intestinal permeability measurable as the concentration of fitc-dextran in serum for each of the experimental groups.

The values will be displayed as mean + sd if normally distributed or median + IQR if not. An appropriate statistical test for differences will be performed (student-t for two, normally distributed groups, Anova for more than 2 groups, Kruskal wallis for non-normally distrubuted results.)

```{r}
tbl1 <- fitc %>% 
  filter(exp_number == study_round) %>% 
  select(group2, "µg/ml") %>% 
  tbl_summary(by = group2,
              digits = everything() ~ c(2,1,1),
              missing = "no") %>%
  add_p() %>% 
  modify_header(
    label ~ "FITC Conc.",
    stat_1 = "**CON-BF**  \n (n = {n})",
    stat_2 = "**GENTA-BF**  \n (n = {n})",
    stat_3 = "**VANCO-BF**  \n (n = {n})") %>% 
  modify_spanning_header((c("stat_1", "stat_2", "stat_3") ~ "Treatment groups (N total = {N})"))

tbl1_gt <- tbl1 %>% 
 as_gt() %>%
  tab_header(title = md("**Intestinal permeability**")) %>%
  opt_align_table_header("center") %>% 
  tab_options(heading.title.font.size = 20,
              heading.background.color = "lightgrey",
              ) %>% 
  tab_source_note(
    source_note = " * P-value indicates overall group differences"
  ) %>%
  tab_source_note(
    source_note = md("**Matrinem round 2**")
  ) %>%
  cols_width(label ~ px(100),
             stat_1 ~ px(150),
             stat_2 ~ px(150),
             stat_3 ~ px(150)) 
tbl1_gt
# Save and export files
# tbl1 %>%  #Save as excel file.
#   as_hux_table() %>%
#   huxtable::quick_xlsx(., file = "../Results/Round 1/ileum-cytokines.xlsx")
# 
# tbl1 %>% #Save as image file.
# as_flex_table() %>%
#   flextable::save_as_image(path = "../../Results/Round 2/ileum_cytokines_table.png",webshot = "webshot")
```

This table can be exported with the following code either to word, excel, pdf, html or image files. See the syntax of huxtable and flextable (?flextable).

```{r ileum cytokine export, echo=TRUE, warning=FALSE}
tbl1_p <- tbl1 %>%
  filter_p() %>% 
  as_tibble()
```

### Plots and regression models

First the data must be transformed to a long format (pivot longer). Mutate any new columns with group treatments needed for the regression model (in this case MAT and FEED (experimental factors)).

```{r long format, echo=FALSE}
fitc_long <- fitc %>% 
  filter(exp_number == study_round) %>% 
  rename(values = "µg/ml") %>% 
  drop_na(
    values) %>% 
  mutate(
    group = recode_factor(  #Change the order of the groups to desired plot order
      group,
      "CON-BF"= "CON-BF",
      "GENTA-BF" = "GENTA-BF",
      "VANCO-BF"= "VANCO-BF"
      ))

head(fitc_long)

```

Create a comparison vector to see pariwise comparisons

```{r plots all fitc, warning=FALSE}
my_comparisons <- list( c("CON-BF", "GENTA-BF"), c("CON-BF", "VANCO-BF"), c("GENTA-BF", "VANCO-BF"))

plot_all_fitc <- fitc_long %>%
ggplot(aes(x = group2, 
           y = values, 
           fill = group2)) + 
  geom_boxplot() +
  geom_point() +
  #coord_cartesian(ylim = c(1.5, 16)) + 
  stat_compare_means(comparisons = my_comparisons, label.y = c(12, 10, 8), method = "wilcox.test")+
  stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
    label.y.npc = 0.75,
    label.x.npc = 0.55) +
  labs( # Change title, and axis names
       x = "",
       y = "Serum FITC conc. (µg/ml)",
       fill = "Groups",
       caption = "Matrinem round 2") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    #axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="lightgrey"),#Change legend box background color
    plot.caption = element_text(color = "darkgrey", size=5)) + 
  scale_fill_brewer(palette="Set1")

all_fitc_patch <- plot_all_fitc + plot_annotation (
  title = 'Intestinal permeability',
   #subtitle = 'All fitc measured',
  #tag_levels = 'A',
  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
  Overall difference= Kruskal-Wallis \n
  Pariwise comparisons = Wilcoxon test.',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

all_fitc_patch
#ggsave('../../Results/Round 1/All fitc-boxplot.png',all_fitc_patch)
```

### 2022-10-18 code/Round 3/Cytokines round 2+3.Rmd
Uploaded data and code to repo

--- 2022-10-18 code/Round 2/Cytokines round 2.Rmd
@@
 ---
-title: "Cytokine analysis script - round 2"
+title: "Cytokine analysis script - round 3"
 author: "Kaare"
@@
 sample_tissue <- "ileum"
-study_round <- "2"
+study_round <- c("2","3")
 ```
@@
 tbl1 <- cytokines %>% 
-  filter(exp_number == study_round) %>% 
+  filter(exp_number %in% c("2","3")) %>% 
   filter(tissue == sample_tissue) %>% 
@@
     label ~ "Cytokines") %>% 
-  modify_spanning_header((c("stat_1", "stat_2", "stat_3") ~ "Treatment groups (N total = {N})"))
+  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Treatment groups (N total = {N})"))
 
@@
   tab_source_note(
-    source_note = md("**Matrinem round 2**")
+    source_note = md("**Matrinem round 2 and 3**")
   ) %>%
@@
              stat_2 ~ px(150),
-             stat_3 ~ px(150)) 
+             stat_3 ~ px(150),
+             stat_4 ~ px(150),
+             stat_5 ~ px(150),
+             stat_6 ~ px(150)) 
 tbl1_gt
@@
 cytokines_long <- cytokines %>% 
-  filter(exp_number == study_round) %>% 
+  filter(exp_number %in% c("2","3")) %>% 
   filter(tissue == sample_tissue) %>%
@@
       "CON-FF" = "CON-FF",
-      "AB-BF"= "AB-BF",
-      "AB-FF"= "AB-FF"
+      "GENTA-BF"= "GENTA-BF",
+      "GENTA-FF"= "GENTA-FF",
+      "VANCO-BF"= "VANCO-BF",
+      "VANCO-FF"= "VANCO-FF",
       ))
@@
        fill = "Groups",
-       caption = "Matrinem round 2") +
+       caption = "Matrinem round 2 & 3") +
   theme_bw()+
@@
   tag_levels = 'A',
-  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin',
+  caption = 'CON = Water, GENTA = Gentamicin,  VANCO = Vancomycin  \n
+   BF = Breast-Feeding, FF = Formula-Feeding',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
@@
   stat_compare_means(
-    aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
-    label.y.npc = 0.9,
+    aes(label = paste0("Global p = ", ..p.format.., "\n(Kruskal-Wallis)")),
+    label.y.npc = 0.85,
     label.x.npc = 0.5) +
@@
        x = "",
-       y = "Tissue conc. (pg/mg)",
+       y = "Tissue conc.",
        fill = "Groups",
-       caption = "Matrinem round 2") +
+       caption = "Matrinem round 2 & 3") +
   theme_bw()+
@@
   caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding. \n
-  Pairwise comparisons with CON-BF using wilcoxon test',
+  Pairwise comparisons with CON-BF using wilcoxon test. NS = Not Significant, * p<0.05, ** p<0.01, *** p<0.001',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),

### 2022-10-18 code/Round 3/Cytokines round 3.Rmd
Uploaded data and code to repo

--- 2022-10-18 code/Round 3/Cytokines round 2+3.Rmd
@@
 sample_tissue <- "ileum"
-study_round <- c("2","3")
+study_round <- "3"
 ```
@@
 tbl1 <- cytokines %>% 
-  filter(exp_number %in% c("2","3")) %>% 
+  filter(exp_number == study_round) %>% 
   filter(tissue == sample_tissue) %>% 
@@
   tab_source_note(
-    source_note = md("**Matrinem round 2 and 3**")
+    source_note = md("**Matrinem round 3**")
   ) %>%
@@
 cytokines_long <- cytokines %>% 
-  filter(exp_number %in% c("2","3")) %>% 
+  filter(exp_number == study_round) %>% 
   filter(tissue == sample_tissue) %>%
@@
        fill = "Groups",
-       caption = "Matrinem round 2 & 3") +
+       caption = "Matrinem round 3") +
   theme_bw()+
@@
        y = "Tissue conc.",
-       fill = "Groups",
-       caption = "Matrinem round 2 & 3") +
+       fill = "Groups") +
   theme_bw()+

### 2022-10-18 code/Round 3/FITC round 2 and 3 copy.Rmd
Uploaded data and code to repo

---
title: "FITC analysis script - round 2 and 3"
author: "Kaare"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

```{r load packages, include=FALSE}
library(readxl) 
library(tidyverse)
library(gt)
library(gtsummary)
library(broom)
library(rstatix)
library(knitr)
library(patchwork)
library(ggpubr)
```

# Introduction to script

This script is useful for analyzing intestinal permeability from FITC-Dextran assays.

Required data structure is a wide tibble (column headers = variables, samples and values = rows).

This script is per default set to handle ileum and serum expression, although any tissue will work. This will just require resetting of the "tissue"-variable.

# FITC for experiment 3.

### Basic data structure and results table.

First you load the data and check the structure. Make sure cytokine concentrations are all numeric (<dbl>). If not, mutate non-numeric columns with mutate_if(is_character).

```{r load data, echo=FALSE}
# Import dataset
fitc <- read_excel("../../Data/CleanData/fitc.xlsx") %>% 
  rename(values = `result_µg/ml_undiluted`) %>% 
  drop_na(values)

glimpse(fitc)
```

Check for data distribution.

```{r load data and set tissue type, include=FALSE}
#set tissue type

normality_check <- fitc %>% 
  filter(exp_number == 2 | exp_number == 3) %>% 
  group_by(group2) %>% 
  shapiro_test(values)

normality_check %>% 
  filter(p < 0.05)

# fitc<- fitc %>% 
#   mutate(values = log10(values))     # Log transforming doesn't help.

#Density plots. Works
fitc %>% 
  filter(exp_number == 2 | exp_number == 3) %>% 
  ggplot(aes(x = values, fill = group2)) + 
  geom_density() +
  facet_wrap(~ group2,
             scales = "free")

#Jitter plots. Works
fitc %>% 
  filter(exp_number == 2 | exp_number == 3) %>% 
  ggplot(aes(x = group2, y = values)) + 
  geom_jitter() +
  facet_wrap(~ group2,
             scales = "free")

```

Exclude outliers
```{r data distribution}
outliers <- fitc %>% 
  filter(exp_number == study_round) %>% 
  group_by(group2) %>% 
  mutate(values = ifelse(values > 50, mean(values), values)) %>% 
qplot(x = group2, y = values, data = ., geom = "jitter",
    color = group2)

```

Below is a table of the levels of intestinal permeability measurable as the concentration of fitc-dextran in serum for each of the experimental groups.

The values will be displayed as mean + sd if normally distributed or median + IQR if not. An appropriate statistical test for differences will be performed (student-t for two, normally distributed groups, Anova for more than 2 groups, Kruskal wallis for non-normally distrubuted results.)

```{r}
tbl1 <- fitc %>% 
  filter(exp_number == 2 | exp_number == 3) %>% 
  mutate(Log10_values = log10(values)) %>% 
  select(group2, values, Log10_values) %>% 
  tbl_summary(by = group2,
              digits = everything() ~ c(2,1,1),
              missing = "no",
              type = list(values ~ 'continuous2', Log10_values ~ 'continuous2'),
              statistic = all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
              label = list(values ~ "µg/mL", Log10_values ~ "Log transformed")) %>%
  add_p(values ~ "aov") %>% 
  modify_header(
    label = "FITC Conc.",
    stat_1 = "**CON-BF**  \n (n = {n})",
    stat_2 = "**CON-FF**  \n (n = {n})",
    stat_3 = "**GENTA-BF**  \n (n = {n})",
    stat_4 = "**GENTA-FF**  \n (n = {n})",
    stat_5 = "**VANCO-BF**  \n (n = {n})",
    stat_6 = "**VANCO-FF**  \n (n = {n})") %>% 
  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Treatment groups (N total = {N})")) %>% 
  bold_labels()

tbl1_gt <- tbl1 %>% 
 as_gt() %>%
  tab_header(title = md("**Intestinal permeability**")) %>%
  opt_align_table_header("center") %>% 
  tab_options(heading.title.font.size = 20,
              heading.background.color = "lightgrey"
              ) %>% 
  tab_source_note(
    source_note = " * P-value indicates overall group differences"
  ) %>%
  tab_source_note(
    source_note = md("**Matrinem round 3**")
  ) %>%
  cols_width(label ~ px(100),
             stat_1 ~ px(150),
             stat_2 ~ px(150),
             stat_3 ~ px(150),
             stat_4 ~ px(150),
             stat_5 ~ px(150),
             stat_6 ~ px(150))

tbl1_gt
tbl1_gt %>% 
  gtsave("../../Results/Round 3/FITC table round 3 TEST.png", expand = 50) #use .tex for latex files
```

This table can be exported with the following code either to word, excel, pdf, html or image files. See the syntax of huxtable and flextable (?flextable).

```{r ileum cytokine export, echo=TRUE, warning=FALSE}
tbl1_p <- tbl1 %>%
  filter_p() %>% 
  as_tibble()
```

### Plots and regression models

First the data must be transformed to a long format (pivot longer). Mutate any new columns with group treatments needed for the regression model (in this case MAT and FEED (experimental factors)).

```{r long format, echo=FALSE}
fitc_long <- fitc %>% 
  filter(exp_number == 2 | exp_number == 3)%>% 
  drop_na(
    values) %>% 
    mutate(Log10_values = log10(values)) %>% 
  mutate(
    group = recode_factor(  #Change the order of the groups to desired plot order
      group,
      "CON-BF"= "CON-BF",
      "CON-FF" = "CON-FF",
      "GENTA-BF"= "GENTA-BF",
      "GENTA-FF"= "GENTA-FF",
      "VANCO-BF"= "VANCO-BF",
      "VANCO-FF"= "VANCO-FF"
      ))

head(fitc_long)

```

Create a comparison vector to see pariwise comparisons.
```{r plots all fitc, warning=FALSE}
my_comparisons <- list( c("CON-BF", "GENTA-BF"), c("CON-BF", "VANCO-BF"), c("GENTA-BF", "VANCO-BF"))
```

Boxplots with jitter - normal values
```{r plots all fitc, warning=FALSE}
plot_all_fitc <- fitc_long %>%
ggplot(aes(x = group2, 
           y = values, 
           fill = group2)) + 
  geom_boxplot() +
  geom_point() +
  #coord_cartesian(ylim = c(1.5, 16)) + 
  # stat_compare_means(comparisons = my_comparisons, label.y = c(2.5, 2.2, 2), method = "wilcox.test")+
  stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
    label.y.npc = 0.75,
    label.x.npc = 0.55) +
  labs( # Change title, and axis names
       x = "",
       y = "Serum FITC conc. (µg/ml)",
       fill = "Groups",
       caption = "Matrinem round 2 and 3") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    #axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="lightgrey"),#Change legend box background color
    plot.caption = element_text(color = "darkgrey", size=5)) + 
  scale_fill_brewer(palette="Set1")

all_fitc_patch <- plot_all_fitc + plot_annotation (
  title = 'Intestinal permeability',
   #subtitle = 'All fitc measured',
  #tag_levels = 'A',
  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
  Overall difference= Kruskal-Wallis.',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

all_fitc_patch
#ggsave('../../Results/Round 3/FITC boxplots.png',all_fitc_patch)
```

Log transformed values
```{r plots all fitc, warning=FALSE}

plot_all_fitc_log10 <- fitc_long %>%
ggplot(aes(x = group2, 
           y = Log10_values, 
           fill = group2)) + 
  geom_boxplot() +
  geom_point() +
  #coord_cartesian(ylim = c(1.5, 16)) + 
  # stat_compare_means(comparisons = my_comparisons, label.y = c(2.5, 2.2, 2), method = "wilcox.test")+
  stat_compare_means(aes(label = paste0("Anova \n p = ", ..p.format..)),
    label.y.npc = 0.75,
    label.x.npc = 0.55, method = "aov") +
  labs( # Change title, and axis names
       x = "",
       y = "Serum FITC conc. log10 transformed",
       fill = "Groups",
       caption = "Matrinem round 2 and 3") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    #axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="lightgrey"),#Change legend box background color
    plot.caption = element_text(color = "darkgrey", size=5)) + 
  scale_fill_brewer(palette="Set1")

all_fitc_patch_log10 <- plot_all_fitc_log10 + plot_annotation (
  title = 'Intestinal permeability',
   #subtitle = 'All fitc measured',
  #tag_levels = 'A',
  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
  Overall difference= Anova',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

all_fitc_patch_log10
ggsave('../../Results/Round 3/FITC boxplot log 10.png',all_fitc_patch)
```

```{r}
FITC_PATCH <- all_fitc_patch + all_fitc_patch_log10
```

### 2022-10-18 code/Round 3/FITC round 3.Rmd
Uploaded data and code to repo

--- 2022-10-18 code/Round 2/FITC round 2.Rmd
@@
 ---
-title: "FITC analysis script - round 2"
+title: "FITC analysis script - round 3"
 author: "Kaare"
@@
 
-# FITC for experiment 2.
+# FITC for experiment 3.
 
@@
 # Import dataset
-fitc <- read_excel("../../Data/CleanData/fitc.xlsx")
+fitc <- read_excel("../../Data/CleanData/fitc.xlsx") %>% 
+  rename(values = `result_µg/ml_undiluted`) %>% 
+  drop_na(values)
+
 glimpse(fitc)
@@
 
-Set the tissue type to whatever tissue your data are extracted from, or that you want to analyze first.
+Check for data distribution.
 
@@
 #set tissue type
-study_round <- "2"
 
-fitc<- fitc %>% 
-  rename("µg/ml" = `result_µg/ml_undiluted`)
+normality_check <- fitc %>% 
+  filter(exp_number == 3) %>% 
+  group_by(group2) %>% 
+  shapiro_test(values)
+
+# fitc<- fitc %>% 
+#   mutate(values = log10(values))     # Log transforming doesn't help.
+
+#Density plots. Works
+fitc %>% 
+  filter(exp_number == 2 | exp_number == 3) %>% 
+  ggplot(aes(x = values, fill = group2)) + 
+  geom_density() +
+  facet_wrap(~ group2,
+             scales = "free")
+
+#Jitter plots. Works
+fitc %>% 
+  filter(exp_number == 2 | exp_number == 3) %>% 
+  ggplot(aes(x = group2, y = values)) + 
+  geom_jitter() +
+  facet_wrap(~ group2,
+             scales = "free")
+
 ```
 
+Exclude outliers
 ```{r data distribution}
-no_outliers <- fitc %>% 
-  filter(exp_number == 2) %>% 
+outliers <- fitc %>% 
+  filter(exp_number == study_round) %>% 
   group_by(group2) %>% 
-  mutate(fitc_level = ifelse(fitc_level > 10, mean(fitc_level), fitc_level))
-
-  no_outliers %>% 
-qplot(x = group2, y = fitc_level, data = ., geom = "jitter",
+  mutate(values = ifelse(values > 50, mean(values), values)) %>% 
+qplot(x = group2, y = values, data = ., geom = "jitter",
     color = group2)
@@
 tbl1 <- fitc %>% 
-  filter(exp_number == study_round) %>% 
-  select(group2, "µg/ml") %>% 
+  filter(exp_number == 2 | exp_number == 3) %>% 
+  mutate(Log10_values = log10(values)) %>% 
+  select(group2, values, Log10_values) %>% 
   tbl_summary(by = group2,
               digits = everything() ~ c(2,1,1),
-              missing = "no") %>%
-  add_p() %>% 
+              missing = "no",
+              statistic = all_continuous() ~ "{mean} ({sd})") %>%
+  add_p(values ~ "aov") %>% 
   modify_header(
-    label ~ "FITC Conc.",
+    label = "FITC Conc.",
     stat_1 = "**CON-BF**  \n (n = {n})",
-    stat_2 = "**GENTA-BF**  \n (n = {n})",
-    stat_3 = "**VANCO-BF**  \n (n = {n})") %>% 
-  modify_spanning_header((c("stat_1", "stat_2", "stat_3") ~ "Treatment groups (N total = {N})"))
+    stat_2 = "**CON-FF**  \n (n = {n})",
+    stat_3 = "**GENTA-BF**  \n (n = {n})",
+    stat_4 = "**GENTA-FF**  \n (n = {n})",
+    stat_5 = "**VANCO-BF**  \n (n = {n})",
+    stat_6 = "**VANCO-FF**  \n (n = {n})") %>% 
+  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Treatment groups (N total = {N})"))
 
@@
   tab_source_note(
-    source_note = md("**Matrinem round 2**")
+    source_note = md("**Matrinem round 3**")
   ) %>%
@@
              stat_2 ~ px(150),
-             stat_3 ~ px(150)) 
+             stat_3 ~ px(150),
+             stat_4 ~ px(150),
+             stat_5 ~ px(150),
+             stat_6 ~ px(150))
+
 tbl1_gt
@@
   filter(exp_number == study_round) %>% 
-  rename(values = "µg/ml") %>% 
   drop_na(
@@
       "CON-BF"= "CON-BF",
-      "GENTA-BF" = "GENTA-BF",
-      "VANCO-BF"= "VANCO-BF"
+      "CON-FF" = "CON-FF",
+      "GENTA-BF"= "GENTA-BF",
+      "GENTA-FF"= "GENTA-FF",
+      "VANCO-BF"= "VANCO-BF",
+      "VANCO-FF"= "VANCO-FF"
       ))
@@
   #coord_cartesian(ylim = c(1.5, 16)) + 
-  stat_compare_means(comparisons = my_comparisons, label.y = c(12, 10, 8), method = "wilcox.test")+
+  # stat_compare_means(comparisons = my_comparisons, label.y = c(2.5, 2.2, 2), method = "wilcox.test")+
   stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),

### 2022-10-18 code/Round 4/Cytokines round 4.Rmd
Uploaded data and code to repo

--- 2022-10-18 code/Round 2/Cytokines round 2.Rmd
@@
 ---
-title: "Cytokine analysis script - round 2"
+title: "Cytokine analysis script - round 4"
 author: "Kaare"
@@
 sample_tissue <- "ileum"
-study_round <- "2"
+study_round <- "4"
+
 ```
@@
   modify_header(
-    label ~ "Cytokines") %>% 
+    label ~ "Cytokines",
+    stat_1 = "**VANCO-BF (CON)**\n(n = {n})",
+    stat_2 = "**VANCO-FF (FVT)**\n(n = {n})",
+    stat_3 = "**VANCO-FF (SM)**\n(n = {n})") %>% 
   modify_spanning_header((c("stat_1", "stat_2", "stat_3") ~ "Treatment groups (N total = {N})"))
@@
  as_gt() %>%
-  tab_header(title = md("**Proinflammatory cytokines**")) %>%
+  tab_header(title = md("**Proinflammatory cytokines (pg/mL)**")) %>%
   opt_align_table_header("center") %>% 
@@
   tab_source_note(
-    source_note = md("**Matrinem round 2**")
+    source_note = md("**Matrinem round 4**")
   ) %>%
@@
   mutate(
-    group = recode_factor(  #Change the order of the groups to desired plot order
+    group = recode_factor(#Change the order of the groups to desired plot order
       group,
-      "CON-BF"= "CON-BF",
-      "CON-FF" = "CON-FF",
-      "AB-BF"= "AB-BF",
-      "AB-FF"= "AB-FF"
+      "VANCO-BF"= "VANCO-BF",
+      "VANCO-FORM+FVT" = "VANCO-FF (FVT)",
+      "VANCO-FORM+SM" = "VANCO-FF (SM)"
       ))
@@
        fill = "Groups",
-       caption = "Matrinem round 2") +
+       caption = "Matrinem round 4") +
   theme_bw()+
@@
   tag_levels = 'A',
-  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin',
+  caption = 'VANCO = Vancomycin, BF = Breast-Feeding, FF = Formula-Feeding\n
+  FVT = Fecal Virome Transfer, SM = SM-buffer',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
@@
 ```
-
-Statistically significant group differences were observed for \`r ifelse([tbl1_p$Cytokines] > 0, tbl1_1$Cytokines, "no cytokines)\`\`
 
@@
   tag_levels = 'A',
-  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding. \n
-  Pairwise comparisons with CON-BF using wilcoxon test',
+  caption = 'VANCO = Vancomycin, BF = Breast-Feeding, FF = Formula-Feeding\n
+  FVT = Fecal Virome Transfer, SM = SM-buffer',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),

### 2022-10-18 code/Round 4/FITC round 2+3 and 4 comparison.Rmd
Uploaded data and code to repo

--- 2022-10-18 code/Round 3/FITC round 2 and 3 copy.Rmd
@@
 ---
-title: "FITC analysis script - round 2 and 3"
+title: "FITC analysis script - round 2,3 4"
 author: "Kaare"
@@
 
-# FITC for experiment 3.
+# FITC for experiments 2,3 and 4.
 
@@
 normality_check <- fitc %>% 
-  filter(exp_number == 2 | exp_number == 3) %>% 
   group_by(group2) %>% 
@@
 fitc %>% 
-  filter(exp_number == 2 | exp_number == 3) %>% 
   ggplot(aes(x = values, fill = group2)) + 
@@
 fitc %>% 
-  filter(exp_number == 2 | exp_number == 3) %>% 
   ggplot(aes(x = group2, y = values)) + 
   geom_jitter() +
-  facet_wrap(~ group2,
+  facet_wrap(~ group2 + exp_number,
              scales = "free")
@@
 tbl1 <- fitc %>% 
-  filter(exp_number == 2 | exp_number == 3) %>% 
+  filter(exp_number == 4) %>% 
   mutate(Log10_values = log10(values)) %>% 
-  select(group2, values, Log10_values) %>% 
-  tbl_summary(by = group2,
+  select(group, values, Log10_values) %>% 
+  tbl_summary(by = group,
               digits = everything() ~ c(2,1,1),
@@
     label = "FITC Conc.",
-    stat_1 = "**CON-BF**  \n (n = {n})",
-    stat_2 = "**CON-FF**  \n (n = {n})",
-    stat_3 = "**GENTA-BF**  \n (n = {n})",
-    stat_4 = "**GENTA-FF**  \n (n = {n})",
-    stat_5 = "**VANCO-BF**  \n (n = {n})",
-    stat_6 = "**VANCO-FF**  \n (n = {n})") %>% 
-  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Treatment groups (N total = {N})")) %>% 
+    stat_1 = "**VANCO-BF (CON)**\n(n = {n})",
+    stat_2 = "**VANCO-FF (FVT)**\n(n = {n})",
+    stat_3 = "**VANCO-FF (SM)**\n(n = {n})") %>% 
+  modify_spanning_header((c("stat_1", "stat_2", "stat_3") ~ "Treatment groups (N total = {N})")) %>%  
   bold_labels()
@@
   tab_source_note(
-    source_note = md("**Matrinem round 3**")
+    source_note = md("**Matrinem round 4**")
   ) %>%
@@
              stat_2 ~ px(150),
-             stat_3 ~ px(150),
-             stat_4 ~ px(150),
-             stat_5 ~ px(150),
-             stat_6 ~ px(150))
-
+             stat_3 ~ px(150))
 tbl1_gt
 tbl1_gt %>% 
-  gtsave("../../Results/Round 3/FITC table round 3 TEST.png", expand = 50) #use .tex for latex files
+  gtsave("../../Results/Round 4/FITC table round 3 TEST.png", expand = 50) #use .tex for latex files
 ```
@@
 fitc_long <- fitc %>% 
-  filter(exp_number == 2 | exp_number == 3)%>% 
+  filter(exp_number == 4)%>% 
   drop_na(
@@
   mutate(
-    group = recode_factor(  #Change the order of the groups to desired plot order
+    group = recode_factor(#Change the order of the groups to desired plot order
       group,
-      "CON-BF"= "CON-BF",
-      "CON-FF" = "CON-FF",
-      "GENTA-BF"= "GENTA-BF",
-      "GENTA-FF"= "GENTA-FF",
-      "VANCO-BF"= "VANCO-BF",
-      "VANCO-FF"= "VANCO-FF"
+      "CON-BF"= "VANCO-BF",
+      "FVT-FORM" = "VANCO-FF (FVT)",
+      "SM-FORM" = "VANCO-FF (SM)"
       ))
@@
 ```{r plots all fitc, warning=FALSE}
-my_comparisons <- list( c("CON-BF", "GENTA-BF"), c("CON-BF", "VANCO-BF"), c("GENTA-BF", "VANCO-BF"))
+my_comparisons <- list(c("VANCO-BF", "VANCO-FF (FVT)"), c("VANCO-BF", "VANCO-FF (SM)"), c("VANCO-FF (FVT)", "VANCO-FF (SM)"))
 ```
@@
 plot_all_fitc <- fitc_long %>%
-ggplot(aes(x = group2, 
+ggplot(aes(x = group, 
            y = values, 
-           fill = group2)) + 
+           fill = group)) + 
   geom_boxplot() +
@@
   #coord_cartesian(ylim = c(1.5, 16)) + 
-  # stat_compare_means(comparisons = my_comparisons, label.y = c(2.5, 2.2, 2), method = "wilcox.test")+
+  stat_compare_means(comparisons = my_comparisons, label.y = c(60, 50, 45), method = "wilcox.test") +
   stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
@@
        fill = "Groups",
-       caption = "Matrinem round 2 and 3") +
+       caption = "Matrinem round 4") +
   theme_bw()+
@@
   #tag_levels = 'A',
-  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
-  Overall difference= Kruskal-Wallis.',
+  caption = 'VANCO = Vancomycin, BF = Breast-Feeding, FF = Formula-Feeding\n
+  FVT = Fecal Virome Transfer, SM = SM-buffer',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
@@
 plot_all_fitc_log10 <- fitc_long %>%
-ggplot(aes(x = group2, 
+ggplot(aes(x = group, 
            y = Log10_values, 
-           fill = group2)) + 
+           fill = group)) + 
   geom_boxplot() +
@@
   #coord_cartesian(ylim = c(1.5, 16)) + 
-  # stat_compare_means(comparisons = my_comparisons, label.y = c(2.5, 2.2, 2), method = "wilcox.test")+
+  stat_compare_means(comparisons = my_comparisons, label.y = c(2.5, 2.2, 2), method = "wilcox.test")+
   stat_compare_means(aes(label = paste0("Anova \n p = ", ..p.format..)),
@@
   #tag_levels = 'A',
-  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
-  Overall difference= Anova',
+  caption = 'VANCO = Vancomycin, BF = Breast-Feeding, FF = Formula-Feeding\n
+  FVT = Fecal Virome Transfer, SM = SM-buffer',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
@@
 all_fitc_patch_log10
-ggsave('../../Results/Round 3/FITC boxplot log 10.png',all_fitc_patch)
+ggsave('../../Results/Round 4/FITC boxplot log 10.png',all_fitc_patch_log10)
 ```

### 2022-10-18 code/Round 4/FITC round 4.Rmd
Uploaded data and code to repo

--- 2022-10-18 code/Round 4/FITC round 2+3 and 4 comparison.Rmd
@@
 ---
-title: "FITC analysis script - round 2,3 4"
+title: "FITC analysis script - round 4"
 author: "Kaare"
@@
 
-# FITC for experiments 2,3 and 4.
+# FITC for experiment 4.
 
@@
 
-```{r load data and set tissue type, include=FALSE}
+```{r load data and set tissue type, eval=FALSE, include=FALSE}
 #set tissue type
@@
 normality_check <- fitc %>% 
-  group_by(group2) %>% 
+  filter(exp_number == 4) %>% 
+  group_by(group) %>% 
   shapiro_test(values)
@@
 
-# fitc<- fitc %>% 
-#   mutate(values = log10(values))     # Log transforming doesn't help.
+fitc<- fitc %>% 
+  mutate(values = log10(values))     # Log transforming doesn't help.
 
@@
 fitc %>% 
-  ggplot(aes(x = values, fill = group2)) + 
+  filter(exp_number == 4) %>% 
+  ggplot(aes(x = values, fill = group)) + 
   geom_density() +
-  facet_wrap(~ group2,
+  facet_wrap(~ group,
              scales = "free")
@@
 fitc %>% 
-  ggplot(aes(x = group2, y = values)) + 
+  filter(exp_number == 4) %>% 
+  ggplot(aes(x = group, y = values)) + 
   geom_jitter() +
-  facet_wrap(~ group2 + exp_number,
+  facet_wrap(~ group,
              scales = "free") 
-
 ```
@@
 Exclude outliers
-```{r data distribution}
+```{r data distribution, eval=FALSE, include=FALSE}
 outliers <- fitc %>% 
@@
               missing = "no",
-              type = list(values ~ 'continuous2', Log10_values ~ 'continuous2'),
-              statistic = all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
+              #type = list(values ~ 'continuous2', Log10_values ~ 'continuous2'),
+              statistic = list(values ~ "{median} ({p25}, {p75})", Log10_values ~"{median} ({p25}, {p75})"),
               label = list(values ~ "µg/mL", Log10_values ~ "Log transformed")) %>%
-  add_p(values ~ "aov") %>% 
+  add_p(list(values ~ "kruskal.test", Log10_values ~ "kruskal.test")) %>% 
+  # modify_table_styling(
+  #   columns = label,
+  #   rows = label == "µg/mL",
+  #   footnote = "Non-parametric distribution"
+  # ) %>% 
   modify_header(
@@
 tbl1_gt <- tbl1 %>% 
- as_gt() %>%
+ as_gt(include = -tab_footnote) %>%
+  tab_footnote( 
+    footnote = md("Log transformation did not normalize distribution.  \n  Non-parametric test (Kruskall-Wallis) used for both datasets."),
+    locations = cells_column_labels(columns = c(p.value))
+  ) %>% 
   tab_header(title = md("**Intestinal permeability**")) %>%
@@
               ) %>% 
-  tab_source_note(
-    source_note = " * P-value indicates overall group differences"
-  ) %>%
+  # tab_source_note(
+  #   source_note = " µg/mL values"
+  # ) %>%
   tab_source_note(
@@
 tbl1_gt
-tbl1_gt %>% 
-  gtsave("../../Results/Round 4/FITC table round 3 TEST.png", expand = 50) #use .tex for latex files
+# tbl1_gt %>% 
+#   gtsave("../../Results/Round 4/FITC table round 3 TEST.png", expand = 50) #use .tex for latex files
 ```
@@
 Create a comparison vector to see pariwise comparisons.
-```{r plots all fitc, warning=FALSE}
+```{r comparison list, warning=FALSE}
 my_comparisons <- list(c("VANCO-BF", "VANCO-FF (FVT)"), c("VANCO-BF", "VANCO-FF (SM)"), c("VANCO-FF (FVT)", "VANCO-FF (SM)"))
@@
 Boxplots with jitter - normal values
-```{r plots all fitc, warning=FALSE}
+```{r plots all fitc normal values, warning=FALSE}
 plot_all_fitc <- fitc_long %>%
@@
 Log transformed values
-```{r plots all fitc, warning=FALSE}
+```{r plots all fitc log transformed, warning=FALSE}
 
@@
   stat_compare_means(comparisons = my_comparisons, label.y = c(2.5, 2.2, 2), method = "wilcox.test")+
-  stat_compare_means(aes(label = paste0("Anova \n p = ", ..p.format..)),
+  stat_compare_means(aes(label = paste0("\n p = ", ..p.format..)),
     label.y.npc = 0.75,
-    label.x.npc = 0.55, method = "aov") +
+    label.x.npc = 0.55) +
   labs( # Change title, and axis names
@@
 
-```{r}
+```{r patch output}
 FITC_PATCH <- all_fitc_patch + all_fitc_patch_log10
+FITC_PATCH
 ```

### 2022-10-18 code/Round 4/Gene expression.Rmd
Uploaded data and code to repo

---
title: "Gene expression"
author: "Kaare D. Tranaes"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readxl)
library(tidyverse)
library(broom)
library(ggpubr)
library(AICcmodavg)
library(visdat)
library(rstatix)
library(gt)
library(gtsummary)
library(patchwork)
```

#### Gene expression

barcharts
```{r}
delta_gene_expr <- read_excel("../Data/CleanData/gene_expression_fold_changes.xlsx")

bar_charts_all_genes <- delta_gene_expr %>% 
  mutate(
    groups = recode_factor(  #Change the order of the groups to desired plot order
      groups,
      "con-bf"= "CON-BF",
      "fvt-form" = "VANCO-FF (FVT)",
      "sm-form" = "VANCO-FF (SM)"
      )) %>% 
  group_by(genes, groups) %>% 
  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
  ggplot(
    aes(x = groups,
        y = mean_fold_change,
        fill = groups)
  ) + 
  geom_col() +
  facet_wrap(~genes,
             ncol = 5, 
             scales = "free") +
  labs(title = "Level of gene expression in ileum tissue",
       y = "Gene expression relative fold change",
       x = "",
       fill = "Groups",
       caption = "Matrinem round 4") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="lightgrey"), # Change legend box background color
    plot.caption = element_text(color = "darkgrey", size=5)) + 
  scale_fill_brewer(palette="Set1")

bar_charts_all_genes

bar_charts_selected_genes <- delta_gene_expr %>% 
  filter(genes %in% c("TNFa","IL1b", "REG3G")) %>% 
  mutate(
    groups = recode_factor(  #Change the order of the groups to desired plot order
      groups,
      "con-bf"= "CON-BF",
      "fvt-form" = "VANCO-FF (FVT)",
      "sm-form" = "VANCO-FF (SM)"
      )) %>% 
  group_by(genes, groups) %>% 
  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
  ggplot(
    aes(x = groups,
        y = mean_fold_change,
        fill = groups)
  ) + 
  geom_col() +
  facet_wrap(~genes,
             ncol = 5, 
             scales = "free") +
  labs(title = "Level of gene expression in ileum tissue - significant changes",
       y = "Gene expression relative fold change",
       x = "",
       fill = "Groups",
       caption = "Matrinem round 4") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="lightgrey"), # Change legend box background color
    plot.caption = element_text(color = "darkgrey", size=5)) + 
  scale_fill_brewer(palette="Set1")

bar_charts_selected_genes
```

### 2022-10-18 code/Round 4/Histology all rounds.Rmd
Uploaded data and code to repo

---
title: "HISTO1"
author: "Kaare"
date: "6/27/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MATRINEM histo-analysis. 
AB-PAS stains quantified as % of total tissue. Experiments 2-4 collected samples for histology, but not experiment 1. 
Samples are of ileum (small intestine) and colon collected from mouse pups 6 days postnatally.

```{r load packages, include=FALSE}
# Packages required
library(readxl)
library(dplyr)
library(tidyverse)
```
```{r}
data <- read_excel("Data_files/Histo analysis .xlsx")
allrounds <- read_excel("Data_files/bodyweight_round_1234.xlsx")

```

```{r}
clean_data <- data %>% 
  mutate(ID = str_remove(ID, "-C") %>% str_remove(.,"-I") %>% str_remove(.," I")) %>% 
  mutate(ID = as.numeric(ID))

filter_joined_df <- semi_join(allrounds, clean_data, by = c("sample_id" = "ID"))

histo <- filter_joined_df %>% 
  inner_join(clean_data, by = c("sample_id" = "ID"))

histo <- histo %>% 
  pivot_longer(., `PAS-RELATIVE-TO-TOTAL(%)`:`AB-RELATIVE-TO-TOTAL(%)`, 
               names_to = "AB-PAS",
               values_to = "% of total tissue stained")

```

```{r}
histo %>%  ### Try to wrap histo longer to have all stain values in 1 column and facet-wrap by PAS + AB
  filter(TISSUE == "ileum") %>% 
  ggplot(aes(x = received_antibiotic, y = `% of total tissue stained`)) +
  geom_boxplot() + 
  facet_wrap(~ experiment_no) +
  labs(title =  "Ileum AB-PAS staining % of total tissue", subtitle = "Faceted by experimental round (2-4)",
       x = "Maternal Treatment") + 
  theme_bw()

```

```{r}
histo %>% 
  filter(TISSUE == "colon") %>% 
   ggplot(aes(x = received_antibiotic, y = `% of total tissue stained`)) +
  geom_boxplot() + 
  facet_wrap(~ experiment_no) +
  labs(title =  "COLON AB-PAS staining % of total tissue", subtitle = "Faceted by experimental round (2-4)",
       x = "Maternal Treatment") + 
  theme_bw()

```
Visually no effect.

### 2022-10-18 code/Round 4/Histology round 4 copy.Rmd
Uploaded data and code to repo

---
title: "HISTO1"
author: "Kaare"
date: "6/27/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## MATRINEM histo-analysis. 
AB-PAS stains quantified as % of total tissue. Experiments 2-4 collected samples for histology, but not experiment 1. 
Samples are of ileum (small intestine) and colon collected from mouse pups 6 days postnatally.

```{r load packages, include=FALSE}
# Packages required
library(readxl)
library(dplyr)
library(tidyverse)
```

```{r}
data <- read_excel("../../Data/RawData/Histo_analysis .xlsx")
chars <- read_excel("../../Data/CleanData/study_id_info.xlsx")

```

```{r}
histo <- data %>% 
  mutate(ID = str_remove(ID, "-C") %>% str_remove(.,"-I") %>% str_remove(.," I")) %>% 
  mutate(ID = as.numeric(ID)) %>% 
  rename(sample_id = ID)

histo <- histo %>%  
  left_join(., chars, by = "sample_id") %>% 
  select(sample_id:`AB-RELATIVE-TO-TOTAL(%)`,group2, exp_number) %>% 
  relocate(., exp_number, group2, .after = sample_id)

```

Ileum plots are shown below for round 4

```{r}

```

```{r}
tissue <- "ileum"
round <- "4"

####Bar char 
histo %>%  
  filter(TISSUE == tissue & exp_number == round) %>% 
  group_by(group2) %>% 
  summarise(mean = mean(`AB-RELATIVE-TO-TOTAL(%)`)) %>% 
  ggplot(
    aes(x = group2 
        ,y = mean 
        ,fill = group2
        )) +
  geom_bar(stat = "identity") + 
  labs(title =  paste(tissue ,"AB staining % of total tissue"), subtitle = "No subtitle",
       y = "% of tissue AB stained (mucin thickness?)",
       x = "Treatment groups") + 
  theme_bw()

####Boxplot char 
histo %>%  
  filter(TISSUE == tissue & exp_number == round) %>% 
  ggplot(
    aes(x = group2 
        ,y = `AB-RELATIVE-TO-TOTAL(%)` 
        ,fill = group2
        )) +
  geom_boxplot() + 
  labs(title =  paste(tissue ,"AB staining % of total tissue"), subtitle = "No subtitle",
       y = "% of tissue AB stained (mucin thickness?)",
       x = "Treatment groups") + 
  theme_bw()

```

```{r}
tissue <- "colon"
round <- "4"

####Bar char 
histo %>%  
  filter(TISSUE == tissue & exp_number == round) %>% 
  group_by(group2) %>% 
  summarise(mean = mean(`AB-RELATIVE-TO-TOTAL(%)`)) %>% 
  ggplot(
    aes(x = group2 
        ,y = mean 
        ,fill = group2
        )) +
  geom_bar(stat = "identity") + 
  labs(title =  paste(tissue ,"AB staining % of total tissue"), subtitle = tissue,
       y = "% of tissue AB stained (mucin thickness?)",
       x = "Treatment groups") + 
  theme_bw()

####Boxplot char 
histo %>%  
  filter(TISSUE == tissue & exp_number == round) %>% 
  ggplot(
    aes(x = group2 
        ,y = `AB-RELATIVE-TO-TOTAL(%)` 
        ,fill = group2
        )) +
  geom_boxplot() + 
  labs(title =  paste(tissue ,"AB staining % of total tissue"), subtitle = "No subtitle",
       y = "% of tissue AB stained (mucin thickness?)",
       x = "Treatment groups") + 
  theme_bw()

```

### 2022-10-18 code/cleaning_raw_data NEW.Rmd
Uploaded data and code to repo

---
title: "document_title"
author: "Kaare D. Tranaes"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, error = FALSE)
```

# Raw data cleaning document

::: {#Project Description}
This is a documentation of generating the working files from the raw data files. Every step of the process is explained below.  
The raw data files can be accessed in the RawData folder associated with this project.  
Once cleaned up, a new data file for further analysis is created and stored in CleanData folder.
:::

```{r load required libraries}
library(readxl) 
library(tidyverse)
library(writexl)
library(broom)
```
## Weight development data
```{r load data}
import_round1 <- read_excel("../Data/RawData/bodyweight_round_1.xlsx"
        #                     , col_types = c("text","text", "text", "text", 
        # "text", "text", "text", "text", "numeric", 
        # "numeric", "numeric", "text")
        ) %>% 
  select(-`Pup ID`) %>% 
  rename(sample_id = `Sample ID`,
         exp_number = Round,
         received_antibiotics = `Treatment with AB`, 
         maternal_treatment = `Type of AB`,
         type_of_feed = `Feeding`,
         group = Group,
         cage = Mother,
         bodyweight_baseline = `Baseline weight`,
         bodyweight_baseline_plus_24h = `Middle of study weight`,
         bodyweight_baseline_plus_48h = `Weight End`)

import_round2 <- read_excel("../Data/RawData/bodyweight_round_2.xlsx"
        #                     ,col_types = c("text", "text", "text", 
        # "text", "text", "text", "text", "numeric", 
        # "numeric", "numeric", "numeric","numeric", "numeric")
        ) %>% 
  rename(sample_id = `ID`,
         exp_number = Round,
         received_antibiotics = `Treatment with AntiBiotic`, 
         maternal_treatment = `Type of AB`,
         type_of_feed = `Feed`,
         group = Groups,
         cage = Cage,
         bodyweight_birthday_group_average = birthweight,
         bodyweight_birthday_plus_24H_group_average = `birth+1day`,
         bodyweight_birthday_plus_48H_group_average = `birth+2days (group mean)`,
         bodyweight_baseline = `Baseline`,
         bodyweight_baseline_plus_24h = `Study day 1`,
         bodyweight_baseline_plus_48h = `End of study`)

import_round3 <- read_excel("../Data/RawData/bodyweight_round_3.xlsx"
        ) %>% 
  select(-id) %>% 
  rename(sample_id = `Sample ID`,
         exp_number = Exp_number,
         received_antibiotics = `Antibiotic yes/no`, 
         maternal_treatment = `Type of AB`,
         type_of_feed = `Type of feeding`,
         group = Group,
         bodyweight_baseline = `Start`,
         bodyweight_baseline_plus_24h = `Day 2`,
         bodyweight_baseline_plus_48h = `End weiight`) %>% 
  mutate(sample_id = as.numeric(sample_id),
          exp_number = as.numeric(exp_number))

import_round4 <- read_excel("../Data/RawData/bodyweight_round_4.xlsx"
        #                     ,col_types = c("text", "text", "text", 
        # "text", "text", "text", "text", "text", 
        # "numeric", "numeric", "numeric")
        ) %>% 
  select(-mark) %>% 
  rename(sample_id = `Sample ID`, 
         group = group,
         bodyweight_baseline = `weight 0H`,
         bodyweight_baseline_plus_24h = `Weight 24H`,
         bodyweight_baseline_plus_48h = `Weight 48H`) %>% 
  mutate(cage = as.numeric(cage))

weights_all_rounds <- bind_rows(import_round1, import_round2, import_round3, import_round4) %>% 
  relocate(comments, .after = bodyweight_birthday_plus_48H_group_average) %>% 
  mutate(received_antibiotics = received_antibiotics %>% trimws(),
         maternal_treatment = maternal_treatment %>% trimws()) %>% 
  mutate(type_of_feed = if_else(type_of_feed == "Formula", "FF",
                                      if_else(type_of_feed == "Breastmilk", "BF",
                                              if_else(type_of_feed == "Breastfeeding", "BF", type_of_feed)))) %>% 
  #mutate_if(is.character,tolower) %>% 
  mutate(maternal_treatment = if_else(maternal_treatment == "Vanco", "VANCO",
                                      if_else(maternal_treatment == "Vancomycin", "VANCO",
                                      if_else(maternal_treatment == "Genta", "GENTA",
                                              if_else(maternal_treatment == "Gentamycin", "GENTA",
                                                      if_else(maternal_treatment == "Water", "CON",
                                                              if_else(maternal_treatment == "Mix", "AB",
                                                              maternal_treatment))))))) %>% 
  unite(group2, maternal_treatment:type_of_feed, sep= "-", remove = FALSE)

View(weights_all_rounds)
rio::export(weights_all_rounds, file = "../Data/CleanData/weight_development.xlsx", overwrite = TRUE)
```

## Study IDs 
```{r}
study_chars <- weights_all_rounds %>% 
  select(1:7)

rio::export(study_chars, file = "../Data/CleanData/study_id_info.xlsx", overwrite = T)

```

## Cytokines from mesoscale assays
```{r}
study_chars <- read_excel("../Data/CleanData/study_id_info.xlsx")

mesoscale_round_1 <- read_excel("../Data/RawData/cytokines_round_1_r.xlsm")

mesoscale_round_2_and_3 <- read_excel("../Data/RawData/cytokines_round_2_and_3.xlsx") %>% 
  rename("tissue_weight_mg" = "tissue_weight_(mg)", "exp_number" = "exp_no")

mesoscale_round_4 <- read_excel("../Data/RawData/cytokines_round_4.xlsx") %>%
  rename("group" = "Group", "exp_number" = "exp_no")

mesoscale_all_rounds <- bind_rows(mesoscale_round_1, mesoscale_round_2_and_3, mesoscale_round_4) %>% 
  select(-group, -treatment, - exp_number)

cytokines <- left_join(study_chars, mesoscale_all_rounds, by = c("sample_id" = "sample")) %>% 
  relocate(tissue:tissue_weight_mg, .after= type_of_feed)

#rio::export(cytokines, file = "../Data/CleanData/cytokines.xlsx")
```

## Intestinal permeability - FITC-Dextran assays
```{r}
fitc_round_2_and_3 <- read_excel("../Data/RawData/FITC_round_2-3.xlsx") %>% 
  rename(exp_number = exp_no) %>% 
  mutate("result_µg/ml_undiluted" = `result_µg/ml`*dilution_factor)

fitc_round_4 <- read_excel("../Data/RawData/FITC_round_4.xlsx") %>% 
  mutate("result_µg/ml_undiluted" = `result_µg/ml`*dilution_factor)

fitc_all_rounds <- bind_rows(fitc_round_2_and_3, fitc_round_4) %>% 
  select(-group, - exp_number, -cage)

fitc <- left_join(study_chars, fitc_all_rounds, by = c("sample_id" = "sample")) %>%  
  filter(exp_number != "1")  # FITC-Dextran not performed in experimental round 1.

rio::export(fitc, file = "../Data/CleanData/fitc.xlsx")
```

## Gene expression - qPCR results
```{r}
### Start here. Or do it in excel.
study_chars <- read_excel("../Data/CleanData/study_id_info.xlsx")
qpcr <- read_excel("../Data/RawData/all_genes_ct_values.xlsx")

qpcr <- qpcr %>% 
  mutate(mean_cqt = (Cq1 + Cq2)/2) %>% 
  pivot_wider(id_cols = Sample, 
              names_from = Gene,
              values_from = mean_cqt)

gene_expr <- study_chars %>%   #### Maybe export this to excel and try to create double delta ct analysis there. 
  filter(sample_id %in% 89:112) %>% 
  left_join(., qpcr, by = c("sample_id" = "Sample"))

delta_gene_expr <- gene_expr %>%
  rename(ct_value_house_gene_HRPT = `House gene (HRPT)`) %>% 
  pivot_longer(cols = CD8:VEGFa,
               names_to = "genes",
               values_to = "ct_value") %>%
  select(sample_id, groups, genes, ct_value, ct_value_house_gene_HRPT) %>% 
    filter(!is.na(ct_value)) %>%
  mutate(dct = ct_value-ct_value_house_gene_HRPT) %>% 
  group_by(genes, groups) %>% 
  mutate(avg_group_dct_val = mean(dct)) %>% 
  group_by(genes) %>% 
  mutate(ddct = dct - (avg_group_dct_val[groups=="con-bf"])) %>% 
  mutate("2^-(ddct)" = 2^-ddct) %>% 
  group_by(genes, groups) %>% 
  mutate(avg_group_2_exp_ddct_val = mean(`2^-(ddct)`)) %>%
  group_by(genes) %>% 
  mutate(relative_fold_change = `2^-(ddct)`/(avg_group_2_exp_ddct_val[groups=="con-bf"]))

# delta_gene_expr %>% 
#   group_by(genes, groups) %>% 
#   summarise(mean_fold_change = mean(relative_fold_change))

#rio::export(delta_gene_expr, file = "../Data/CleanData/gene_expression_fold_changes.xlsx")

```

```{r}

bar_charts <- delta_gene_expr %>% 
  group_by(genes, groups) %>% 
  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
  ggplot(
    aes(x = groups,
        y = mean_fold_change,
        fill = groups)
  ) + 
  geom_col() +
  facet_wrap(~genes,
             ncol = 5) +
  labs(title = "Gene expression fold change",
       y = "Gene expression fold change",
       x = "Groups"
      )
```

```{r}
delta_gene_expr %>%
  filter(genes %in% "CD8") %>% 
  anova(relative_fold_change ~ groups, data = .)
```

### 2022-10-18 code/lin_reg_vanco_interaction.Rmd
Uploaded data and code to repo

---
title: "Untitled"
author: "Kaare D. Tranaes"
date: "`r Sys.Date()`"
output: html_document
---
```{r}
library(readxl)
library(tidyverse)
library(broom)
library(jtools)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#_______Read in data______________________________________
```{r}
fitc <- read_excel("../Data/CleanData/fitc.xlsx") %>% 
  rename(values = `result_µg/ml_undiluted`) %>% 
  drop_na(values)
```
#_________________

#_______Test effect of antibiotics and formula feeding_______ 
```{r}
fitc %>% 
  filter(exp_number == 3) %>% 
  lm(values ~ received_antibiotics + type_of_feed, data = .) %>% summary()  
```
#Turns out not to be significant: Multiple R-squared:  0.09071, p-value: 0.1894
#______________

#_______Test effect of interaction between antibiotics and formula feeding_______
```{r}
fitc %>% 
  filter(exp_number == 3) %>% 
  lm(values ~ received_antibiotics*type_of_feed, data = .) %>% summary()  
```
#Turns out not to be significant: Multiple R-squared:  0.1244, p-value: 0.205
#______________

#_______Test effect of type of antibiotics________________
```{r }
lm_vanco <- fitc %>% 
  filter(exp_number == 3) %>% 
  lm(values ~ maternal_treatment*type_of_feed, data = .)
```
#Significant effect of VANCO-treatment: "maternal_treatmentVANCO:type_of_feedFF   Estimate(58.538)     SD(33.151)   (tvalu)1.766    P =0.087  
#Multiple R-squared:  0.1244, p-value: 0.008879
#______________

```{r results = 'asis'}
export_summs(lm_vanco)
```

### 2022-10-18 code/Matrinem data analysis.Rmd
Created new result script

@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE)
 ```
@@
 library(patchwork)
+library(jtools)
 ```
 
-# Introduction
+# MATRINEM - Attempts at creating a NEC mouse model
 
+## Preface
+
+## Introduction
+
+This thesis is about...
+
+## Background
+
+## Methods and Materials
+
+Prior to results are presented a short description of the studies conducted needs to be made.
+
+Experiment no. 4 was the goal, but before we could test our overall hypothesis we needed to find out which combination of antibiotics and feeding scheme would results in the most NEC-like phenotype.
+
+These research questions were investigated in 3 pre-experiments (PreExp 1-3), each with their own research questions.
+
+### **Overview of all experiments**
+
++-----------+-------------------------------------------------------------------------------+----------+----------+---------+
 |           | PreExp 1                                                                      | PreExp 2 | PreExp 3 | NEC-FvT |
-|-----------|----------|----------|----------|---------|
-| **Aim**   |          |          |          |         |
-| Protocols |          |          |          |         |
-| N         |          |          |          |         |
++===========+===============================================================================+==========+==========+=========+
+| **Aim**   | To test feasibility of formula feeding with PICC line                         |          |          |         |
++-----------+-------------------------------------------------------------------------------+----------+----------+---------+
+| Protocols | Testing a broad spectrum antibiotics mix in combination with formula feeding. |          |          |         |
+|           |                                                                               |          |          |         |
+|           | 4 groups                                                                      |          |          |         |
+|           |                                                                               |          |          |         |
+|           | 1.  CON-BF                                                                    |          |          |         |
+|           | 2.  CON-FF                                                                    |          |          |         |
+|           | 3.  AB-BF                                                                     |          |          |         |
+|           | 4.  AB-FF                                                                     |          |          |         |
++-----------+-------------------------------------------------------------------------------+----------+----------+---------+
+| N         | 38 (32 survived)                                                              |          |          |         |
++-----------+-------------------------------------------------------------------------------+----------+----------+---------+
 
-## Pre-Experiment 1 (PreExp 1)
+### PreExp 1 - Feasibility Study
+
+#### **Rationale:**
+
+*Chen et al.*, showed that Maternal Antibiotics Treatment (**MAT**) resulted in mild NEC-like symptoms of mice offspring.
+We believed that MAT combined with formula feeding (**FF**) of offspring would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+
+#### **Aim:**
+
+To test if it was possible to keep mice separated from the mother alive from postnatal day 3 and 48 hours onwards.
+
+#### **Research Questions:**
+
+-   Can we keep offspring alive with formula feeding?
+-   Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.
 
@@
 
-#### Weight development
+**Results**: `r inline_text(tbl1, {N})` mice were born to 6 different mothers. Of these 32 survived and were included in the analysis. Cytokines
+
+```{r}
+cytokines_1 <- read_excel(path = "data/processed/cytokines.xlsx") %>%  #Load in cytokine data for all experiments
+  filter(exp_number == 1) %>% 
+  filter(tissue == "ileum") %>% 
+  mutate(
+    group = recode_factor(  #Change the order of the groups to desired plot order
+      group,
+      "CON-BF"= "CON-BF",
+      "CON-FF" = "CON-FF",
+      "AB-BF"= "AB-BF",
+      "AB-FF"= "AB-FF"
+      ))
+
+cytokines_1_long <- cytokines_1 %>% 
+  pivot_longer(cols = INFg:TNFa, 
+               names_to = "cytokines", 
+               values_to = "cytokine_concentration") %>% 
+  drop_na(cytokine_concentration)
+
+normality_check <- cytokines_1_long %>% 
+  group_by(cytokines) %>% 
+  shapiro_test(cytokine_concentration) %>% 
+  select(cytokines, p) %>% 
+  filter(p < 0.05)
+
+normality_plot <- cytokines_1_long %>%
+  filter(cytokines %in% normality_check$cytokines ) %>% 
+  ggplot(aes(x = group, 
+             y = cytokine_concentration,
+             color = group)) + 
+  geom_jitter() +
+  facet_wrap(~ cytokines,
+             scales = "free") + 
+  labs(title = "Non-normally distributed cytokines (Shapiro-Wilk Normality Test, p < 0.05)",
+       y = "Concentration of cytokine in tissue (pg/mL)",
+       x = "") +
+  theme_bw()
+
+normality_plot
+
+normality_result_1 <- normality_check %>% 
+  knitr::kable(caption = "Non-normally distributed cytokines (Shapiro-Wilk Normality Test, p < 0.05)")
+normality_result_1
+```
+
+```{r}
+tbl1 <- cytokines_1 %>% 
+  select(group,INFg:TNFa) %>% 
+  tbl_summary(by = group,
+              digits = everything() ~ c(2,1,1),
+              missing = "no",
+              include = !"INFg") %>%
+  add_p() %>% 
+  modify_header(
+    label ~ "Cytokines") %>% 
+  modify_spanning_header((c("stat_1", "stat_2", "stat_3", "stat_4") ~ "Total number of mice analyzed = {N}"))
+
+  tbl1_gt <- tbl1 %>% 
+  as_gt() %>%
+  tab_header(title = md("**Proinflammatory cytokines**")) %>%
+  opt_align_table_header("center") %>% 
+  tab_options(heading.title.font.size = 20,
+              heading.background.color = "lightgrey",
+              ) %>% 
+  tab_source_note(
+    source_note = " * P-value indicates overall group differences"
+  ) %>%
+  tab_source_note(
+    source_note = md("**Matrinem round 1**")
+  ) %>%
+  cols_width(label ~ px(100),
+             stat_1 ~ px(150),
+             stat_2 ~ px(150),
+             stat_3 ~ px(150),
+             stat_4 ~ px(150)) 
+tbl1_gt
+# Save and export files
+# tbl1 %>%  #Save as excel file.
+#   as_hux_table() %>%
+#   huxtable::quick_xlsx(., file = "../Results/Round 1/ileum-cytokines.xlsx")
+# 
+# tbl1 %>% #Save as image file.
+# as_flex_table() %>%
+#   flextable::save_as_image(path = "../../Results/Round 1/ileum_cytokines_table.png",webshot = "webshot")
+```
+
+```{r plots all cytokines, warning=FALSE}
+plot_all_cytokines <- cytokines_1_long %>%
+  #filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot
+ggplot(aes(x = group, 
+           y = cytokine_concentration, 
+           fill = group)) + 
+  geom_boxplot() +
+  facet_wrap(~cytokines, 
+             scales = 'free', 
+             nrow = 3,
+             ncol = 4) +
+  labs(title = "Levels of pro-inflammatory cytokines in the ileum",  # Change title, and axis names
+       x = "",
+       y = "Tissue conc. (pg/mL)",
+       fill = "Groups") +
+  theme_bw() +
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="white")) + # Change legend box background color
+  scale_fill_brewer(palette="Set1")
+
+all_cytokines_patch <- plot_all_cytokines + plot_annotation (
+  title = 'Cytokine expression',
+   subtitle = 'All cytokines measured',
+  tag_levels = 'A',
+  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding',
+  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
+                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
+                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))
+
+all_cytokines_patch
+#ggsave('../../Results/Round 1/All Cytokines-boxplot.png',all_cytokines_patch)
+```
+
+```{r}
+#tbl1_p contains cytokines with significant changes
+tbl1_p <- tbl1 %>% 
+  filter_p() %>% 
+  as_tibble()
+
+signif_cytokines <- cytokines_1_long %>%
+  filter(cytokines %in% tbl1_p$Cytokines) %>%  #Exclude any cytokines not wanted in the plot
+ggplot(aes(x = group, 
+           y = cytokine_concentration, 
+           fill = group)) + 
+  geom_boxplot() +
+  stat_compare_means(aes(label = ..p.signif..),
+                   method = "wilcox.test", ref.group = "CON-BF",
+                   label.y.npc = 0.55) +
+  stat_compare_means(
+    aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
+    label.y.npc = 0.9,
+    label.x.npc = 0.5) +
+  facet_wrap(~cytokines, 
+             scales = 'free', 
+             nrow = 3,
+             ncol = 3) +
+  labs(title = "Levels of pro-inflammatory cytokines in the ileum", # Change title, and axis names
+       #subtitle = "Statistically significant changes", 
+       x = "",
+       y = "Tissue conc. (pg/mg)",
+       fill = "Groups") +
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="white")) + # Change legend box background color
+  scale_fill_brewer(palette="Set1")
+
+signif_cytokines_patch <- signif_cytokines + plot_annotation (
+  title = 'Cytokine expression',
+  subtitle = 'Statistically significant changes',
+  tag_levels = 'A',
+  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding. \n
+  Pairwise comparisons with CON-BF using wilcoxon test',
+  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
+                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
+                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))
+signif_cytokines_patch
+#ggsave('../../Results/Round 1/Significant Cytokines-boxplot.png',signif_cytokines_patch)
+```
+
+```{r echo=FALSE, warning=FALSE}
+#### Make regression model on long format
+reg_mod_cyt_1 <- cytokines_1_long %>% 
+  group_by(cytokines) %>% 
+  do(lm(data = ., cytokine_concentration ~ maternal_treatment*type_of_feed) %>% tidy)  %>% filter(term!='(Intercept)') %>% 
+  ungroup() %>% 
+  filter(p.value < 0.05)
+
+reg_mod_cyt_1 %>% 
+  select(-statistic) %>% 
+  knitr::kable(., caption = 'Significant effectors on cytokine expression', digits = 3)
+```
+
+Statistically significant group differences were observed for IL-6 (`r inline_text(tbl1, column = "p.value", variable = "IL_6")`), KCGRO (`r inline_text(tbl1, column = "p.value", variable = "KCGRO")`) and TNFa (`r inline_text(tbl1, column = "p.value", variable = "TNFa")`).
+
+**The conclusion** is that many of the cytokines show an interaction effect of the two treatment parameters
+
+#### Summary of PreExp 1
+
+-   Can we keep offspring alive with formula feeding?
+
+    -   After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours.
+    -   The formulafed animals gained weight - albeit less than their breastfed peers - during the 48 hours.
+    -   MAT did not affect growth. CON-BF and AB-BF gained weight at a similar pace.
+
+-
+
+    ## Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.
+
+### PreExp 2 - Selectively targeting gram+ or gram- bacteria
 

### 2022-10-19 submission/manuscript.Rmd
Getting reading for manuscript drafting

```
output: html_document

```

## Introduction

## Results

## Discussion

### 2022-10-19 code/Result_section.Rmd
Updated file names

--- 2022-10-18 code/Matrinem data analysis.Rmd
@@
 author: "Kaare"
-date: "6/16/2022"
+date: "18/10/2022"
 output: html_document
@@
 
-+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
++-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
 |             | PreExp 1                                                                      | PreExp 2                                                       | PreExp 3                                                               | NEC-FvT                                                         |
-+===========+===============================================================================+==========+==========+=========+
-| **Aim**   | To test feasibility of formula feeding with PICC line                         |          |          |         |
-+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
-| Protocols | Testing a broad spectrum antibiotics mix in combination with formula feeding. |          |          |         |
++=============+===============================================================================+================================================================+========================================================================+=================================================================+
+| **Aim**     | To test feasibility of formula feeding with PICC line                         | Selectively targeting gram positive vs.
+gram negative bacteria | Testing if there is an interaction-effect with feeding and antibiotics | Combining reults from PreExp 1-3 into final experimental setup.
+|
++-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
+| Protocols   | Testing a broad spectrum antibiotics mix in combination with formula feeding.
+| Groups                                                         | Groups                                                                 | Groups                                                          |
 |             |                                                                               |                                                                |                                                                        |                                                                 |
@@
 |             |                                                                               |                                                                |                                                                        |                                                                 |
-|           | 1.  CON-BF                                                                    |          |          |         |
-|           | 2.  CON-FF                                                                    |          |          |         |
-|           | 3.  AB-BF                                                                     |          |          |         |
-|           | 4.  AB-FF                                                                     |          |          |         |
-+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
-| N         | 38 (32 survived)                                                              |          |          |         |
-+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
+|             | 1.
+CON-BF                                                                    |                                                                |                                                                        |                                                                 |
+|             | 2.
+CON-FF                                                                    |                                                                |                                                                        |                                                                 |
+|             | 3.
+AB-BF                                                                     |                                                                |                                                                        |                                                                 |
+|             | 4.
+AB-FF                                                                     |                                                                |                                                                        |                                                                 |
++-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
+| N           | 38 (32 survived)                                                              | 18                                                             | 32                                                                     | 26                                                              |
++-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
 
@@
 
-**Results**: `r inline_text(tbl1, {N})` mice were born to 6 different mothers. Of these 32 survived and were included in the analysis. Cytokines
+**Results**:
+
+```{r include=FALSE}
+preexp_1 <- read_excel("../data/processed/weight_development.xlsx") %>% 
+  filter(exp_number == "1")
+
+```
 
 ```{r}
-cytokines_1 <- read_excel(path = "data/processed/cytokines.xlsx") %>%  #Load in cytokine data for all experiments
+#__________Load in cytokine data for all experiments________
+cytokines_1 <- read_excel(path = "../data/processed/cytokines.xlsx") %>%  
   filter(exp_number == 1) %>% 
@@
       ))
+#___________________________________________________________
 
+#__________ Pivot longer  ____________
 cytokines_1_long <- cytokines_1 %>% 
@@
   drop_na(cytokine_concentration)
+#___________________________________________________________
+```
 
+```{r}
+#___________ Check cytokine distribution for normality, filter significant ones _____________________
 normality_check <- cytokines_1_long %>% 
@@
   shapiro_test(cytokine_concentration) %>% 
-  select(cytokines, p) %>% 
+  select(cytokines, p)
+#___________________________________________________________
+
+non_normal_cytokines <- normality_check %>% 
   filter(p < 0.05)
 
+#___________ Jitterplot of all non-normally distributed cytokines _____________________
 normality_plot <- cytokines_1_long %>%
-  filter(cytokines %in% normality_check$cytokines ) %>% 
+  filter(cytokines %in% non_normal_cytokines$cytokines ) %>% 
   ggplot(aes(x = group, 
@@
   theme_bw()
+#___________________________________________________________
 
+#___________ Jitterplot of all non-normally distributed cytokines _____________________
+normality_result_1 <- normality_check %>% 
+  knitr::kable(caption = "Normallity check of cytokine distribution (Shapiro-Wilk Normality Test, p < 0.05)")
+#___________________________________________________________
+
+names_of_cytokines <-unique(cytokines_1_long$cytokines)
+```
+
+`r nrow(preexp_1)` mice were born to 6 different mothers. Of these 32 survived and were included in the analysis.
+
+`r nrow(cytokines_1)` samples were available for analysis of concentrations of the proinflammatory cytokines `r names_of_cytokines`.
+Concentrations were measured in 1-cm sections of proximal/distal ileum tissue using MSD Mesoscale Kits.
+Distributions were assessed for normality with Shapiro-Wilk Normality Test (see table below), and plot of data were made to assess distributions visually:
+
+```{r echo=FALSE, fig.cap='Table:}
+normality_result_1
+```
+
+```{r Normality plot, echo=FALSE, fig.cap='Non-normally distributed plots'}
 normality_plot
-
-normality_result_1 <- normality_check %>% 
-  knitr::kable(caption = "Non-normally distributed cytokines (Shapiro-Wilk Normality Test, p < 0.05)")
-normality_result_1
 ```
@@
              stat_4 ~ px(150))
-tbl1_gt
 # Save and export files
@@
 #   flextable::save_as_image(path = "../../Results/Round 1/ileum_cytokines_table.png",webshot = "webshot")
+
+```
+
+The median concentration of the various cytokines are shown in the table below:
+
+```{r, echo=FALSE}
+tbl1_gt
 ```
@@
 
-all_cytokines_patch
+#all_cytokines_patch
 #ggsave('../../Results/Round 1/All Cytokines-boxplot.png',all_cytokines_patch)
@@
                 plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))
-signif_cytokines_patch
+#signif_cytokines_patch
 #ggsave('../../Results/Round 1/Significant Cytokines-boxplot.png',signif_cytokines_patch)
 ```
+
+The cytokines that were significantly different between groups are shown below:
+
+```{r, echo= FALSE}
+signif_cytokines_patch
+```
+
+A linear regression analysis calculates the factors responsible for the changes observed.
 
@@
   filter(p.value < 0.05)
+```
 
+```{r echo=FALSE, warning=FALSE}
 reg_mod_cyt_1 %>% 
@@
 ```
-
-Statistically significant group differences were observed for IL-6 (`r inline_text(tbl1, column = "p.value", variable = "IL_6")`), KCGRO (`r inline_text(tbl1, column = "p.value", variable = "KCGRO")`) and TNFa (`r inline_text(tbl1, column = "p.value", variable = "TNFa")`).
 
@@
 
--
-
-    ## Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.
-
 ### PreExp 2 - Selectively targeting gram+ or gram- bacteria
 
-# Experiment 2 - weight at birth.Summary: Offspring from antibiotics-treated mice were smaller at birth
-
-```{r}
-data <- read_excel("../Data/CleanData/weight_development.xlsx") %>% 
-  select(sample_id:cage, bodyweight_birthday_group_average, bodyweight_birthday_plus_24H_group_average, bodyweight_birthday_plus_48H_group_average, bodyweight_baseline, bodyweight_baseline_plus_24h, bodyweight_baseline_plus_48h) %>%
-  rename(day_0 = "bodyweight_birthday_group_average",
-         day_1 = "bodyweight_birthday_plus_24H_group_average", 
-         day_2 = "bodyweight_birthday_plus_48H_group_average",
-         day_3 = bodyweight_baseline, 
-         day_4 = `bodyweight_baseline_plus_24h`, 
-         day_5 = `bodyweight_baseline_plus_48h`,) %>% 
-  pivot_longer(., cols = c(day_0:day_5),
-                names_to = "time",
-                values_to = "bodyweight") %>% 
-  mutate(received_antibiotics = recode_factor(received_antibiotics, 
-                                       yes = "MAT",
-                                       no = "CON"))
-  #%>%
-  # #mutate(across(where(is_character),as_factor))
-
-exp2.bw.density <- data %>% 
-  filter(exp_number == "2") %>% 
-  filter(time == 'day_0') %>% 
-  ggplot(aes(x = bodyweight, color = received_antibiotics)) +
-    geom_density() +
-  geom_vline(data = data %>% 
-              filter(exp_number == "2") %>% 
-  filter(time == 'day_0') %>% 
-  group_by(received_antibiotics) %>% 
-  summarise(grp.mean = mean(bodyweight)), aes(xintercept=grp.mean, color=received_antibiotics),
-             linetype="dashed") +
-  guides(fill = guide_legend(title = ""),
-         colour = guide_legend(title = " ")) + 
-  labs(title = "", y = "n", x = "Bodyweight (g)") +
-  theme_minimal()
-
-exp2.bw.col <- data %>% 
-  filter(exp_number == "2") %>% 
-  filter(time == 'day_0') %>% 
-  ggplot(aes(x = received_antibiotics, y = bodyweight, fill = received_antibiotics)) +
-  geom_bar(stat = "summary",
-           fun = "mean") +
-  labs(title = "", y = "Mean body weight", x = "") +
-  theme_minimal() +  
-  stat_compare_means(
-    method = "t.test", 
-    label.y = 1.6, 
-    label.x.npc = 0.4) +
-  theme(legend.position="none")
-
-birth_weight_patch <- (exp2.bw.density + exp2.bw.col) + plot_annotation(
-  title = 'Effect of Maternal Antibiotics Treatment on birth weight of offspring',
-  subtitle = 'Experimental round 2',
-  tag_levels = 'A',
-  caption = 'MAT = Maternal Antibiotics Treatment, CON = Water Controls \n 
-  Statistics: non-paired two.sided t-test.'
-)
-birth_weight_patch
-#ggsave("../Results/Weight/Figure_1_Birthweights.png", birth_weight_patch)
-
-#add gt table with group results
-
-```
-
-```{r eval=FALSE, include=FALSE}
-### Weight development during experiments   ###########
-
-bw_development_feed <- data %>% 
-filter(exp_number %in% c("1","3","4")) %>% 
-filter(time %in% c("day_3", "day_5")) %>%
-ggplot(aes(x = time ,
-           y = bodyweight, 
-           fill = type_of_feed)) +
-  facet_wrap(~ exp_number + type_of_feed, 
-             ncol = 2) + 
-  geom_bar(stat = "identity",
-           position = "dodge") + 
-  guides(fill = guide_legend(title = " ")) +
-  labs(
-      caption = "Experimental round 2 not shown due to no formula feeding.",
-    y = "Bodyweight (g)"
-  ) +
-  theme_bw()
-
-bw_development_ab <- data %>% 
-filter(exp_number %in% c("1","2","3","4")) %>% 
-filter(time %in% c("day_3", "day_5")) %>%
-ggplot(aes(x = time ,
-           y = bodyweight, 
-           fill = received_antibiotics)) +
-  facet_wrap(~ exp_number + received_antibiotics, 
-             ncol = 2) + 
-  geom_bar(stat = "identity",
-           position = "dodge") + 
-  guides(fill = guide_legend(title = " ")) +
-  labs(caption = "All mice recieved antibiotics in experimental round 4.",
-    y = "Bodyweight (g)"
-  ) +
-  theme_bw()
-
-bodyweight_develop_patch <- (bw_development_feed + bw_development_ab) + plot_annotation (
-  title = 'Effect of Maternal Antibiotics Treatment on body weight development of offspring',
-  subtitle = 'Faceted by experiment and Feeding or MAT-status',
-  tag_levels = 'A',
-  caption = 'Form = formular feed, Milk = breastfeeding.  
-      MAT = Maternal Antibiotics Treatment, CON = Water Controls.')
-
-bodyweight_develop_patch
-#ggsave("../Results/Weight/Figure_2_bodyweight_development.png", bodyweight_develop_patch)
-
-```
-
-### Research Question 1: Body weight of offspring from water and antibiotic-treated mice at baseline
-
-```{r style baseline bodyweight}
-get_box_stats <- function(y, upper_limit = max(data$Bodyweight[data$Time == "Day2"]) * 1.0) {
-  return(data.frame(
-    y = 0.9 * upper_limit,
-    label = paste(
-      "n =", length(y), "\n",
-      "Mean =", round(mean(y), 2), "\n" #,
-      #"Median =", round(median(y), 2), "\n"
-    )
-  ))
-}
-
-my_comparisons <- list( c("Water", "AB"))
-
-data %>% 
-  filter(Time == "Day2") %>% 
-  ggplot(aes(
-    received_antibiotic, Bodyweight
-  )) + 
-  geom_boxplot() +
-  facet_wrap(~experiment_no, 
-             nrow = 1) + 
-  geom_dotplot(
-     binaxis = "y", stackdir = "center", dotsize = 0.5) + 
-  stat_summary(
-    fun.data = get_box_stats, geom = "text", hjust = 0.5, vjust = 0.9) +
-  labs(
-    title = "Body weight of offspring from antibiotic-treated vs water-treated mice",
-    subtitle = "Results shown from each of the 4 experimental rounds",
-    caption = "In Exp. 4 all mice received antibiotoics, hence no control.
-    Comparison method: Student T-test",
-    x = "Maternal Treatment",
-    y = "Bodyweight at end of study (g)"
-  ) +
-  theme_bw() + 
-  stat_compare_means(
-    method = "t.test", 
-    label.y = 3, 
-    comparisons = my_comparisons
-    )
-```
-
-```{r weightgain plots}
-############# Effect of Formula vs. Breastfeeding on Weight Gain" ###################
-
-bodyweight_gain_BFvsFORM <- data %>% 
-  mutate(weightgain = weight_end - weight_baseline) %>% 
-  ggplot(aes(
-    type_of_feed, weightgain
-  )) + 
-  geom_boxplot() +
-  facet_wrap(~experiment_no, 
-             nrow = 1) +
-geom_dotplot(
-     binaxis = "y", stackdir = "center", dotsize = 0.5) + 
-  stat_summary(
-    fun.data = get_box_stats, geom = "text", hjust = 0.2, vjust = 0.9) +
-  labs(
-    title = "Effect of Formula vs. Breastfeeding on Weight Gain",
-    subtitle = "Results shown from each of the 4 experimental rounds",
-    caption = "In Exp. 2 all mice were breastfed.
-    Comparison method: Student T-test",
-    x = "Type of feed",
-    y = "Body weight increase  (g)"
-  ) +
-  theme_bw() + 
-  stat_compare_means(
-    method = "t.test", 
-    label.y = 3, 
-    comparisons = my_comparisons_2
-    )
-my_comparisons_2 <- list( c("Breastmilk", "Formula"))
-bodyweight_gain_BFvsFORM
-```
-
-```{r weightgain plot 2}
-############# Effect of  Antibiotocis on weightgain in Breastfed offspring" ################
-
-bodyweight_gain_WATER_BF_vs_AB_BF <- 
-  data %>% 
-  mutate(weightgain = weight_end - weight_baseline) %>% 
-  filter(type_of_feed == "Breastmilk") %>% 
-  ggplot(aes(
-    received_antibiotic, weightgain#, colour = mother
-    )) + 
-  geom_boxplot()+
-  geom_point()+
-  facet_wrap(~experiment_no, 
-             nrow = 1) +
-  labs(
-    title = "Breastfed offspring +/- Maternal Antiobiotics Treatment"
-  )
-bodyweight_gain_WATER_BF_vs_AB_BF
-```
-
-#### Histological analysis
-
-#### Cytokine expression
-
-boxpltos
-
-```{r}
-boxplots_all_genes <- delta_gene_expr %>% 
-  ggplot(
-    aes(x = groups,
-        y = relative_fold_change,
-        fill = groups)
-  ) + 
-  geom_boxplot() +
-  facet_wrap(~genes,
-             ncol = 5, 
-             scales = "free") +
-  labs(title = "Gene expression fold change",
-       y = "Gene expression fold change",
-       x = "Groups"
-      )
-boxplots_all_genes
-
-```
-
-#### Gut microbiome analysis
-
-### Experiment 2 (E2)
-
-#### Weight development
-
-#### Histological analysis
-
-#### Intestinal permeability
-
-#### Cytokine expression
-
-#### Gene expression
-
-#### Gut microbiome analysis
-
-### Experiment 4 (NEC-FvT)
-
-The PreExps are now over and it is time to conduct the study. Having determined the optimal experimental protocol in PreExp 1-3 we are ready with the design.
-
-#### Overview - Experimental Setup
-
-#### Intestinal permeability - FITC
-
-```{r FITC load data, include=FALSE}
-fitc <- read_excel("../Data/CleanData/fitc.xlsx") %>% 
-  rename(values = `result_µg/ml_undiluted`) %>% 
-  drop_na(values)
-```
-
-Intestinal permeability is assessed with the FITC-Dextran assay. First I inspect the distribution of results.
-
-```{r FITC check for normality}
-fitc %>% 
-  filter(exp_number == 4) %>% 
-  group_by(group) %>% 
-  shapiro_test(values)
-```
-
-As results and highly skewed I try transforming the FITC results to log10.
-
-```{r FITC try log transforming data}
-fitc <- fitc %>% 
-  mutate(log_10_values = log10(values))  # Logtransform results
-
-```
-
-```{r FITC check for normality log transform}
-fitc %>% 
-  filter(exp_number == 4) %>% 
-  group_by(group) %>% 
-  shapiro_test(log_10_values)
-```
-
-```{r FITC visual inspections of data distribution}
-#Density plots non-log
-density <- fitc %>% 
-  filter(exp_number == 4) %>% 
-  ggplot(aes(x = values, fill = group)) + 
-  geom_density() +
-  facet_wrap(~ group,
-             scales = "free")
-
-#Jitter plots non-log
-jitter <- fitc %>% 
-  filter(exp_number == 4) %>% 
-  ggplot(aes(x = group, y = values)) + 
-  geom_jitter() +
-  facet_wrap(~ group,
-             scales = "free")
-
-#Density plots log
-density <- fitc %>% 
-  filter(exp_number == 4) %>% 
-  ggplot(aes(x = , fill = group)) + 
-  geom_density() +
-  facet_wrap(~ group,
-             scales = "free")
-
-#Jitter plots non-log
-jitter <- fitc %>% 
-  filter(exp_number == 4) %>% 
-  ggplot(aes(x = group, y = values)) + 
-  geom_jitter() +
-  facet_wrap(~ group,
-             scales = "free")
-
-```
-
-#### Cytokine expression
-
-#### Gene expression
-
-#### Histological analysis
-
-#### Gut microbiome analysis
+-   Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.

### 2022-10-19 code/weight_analysis.Rmd
Made progress on data analysis script

--- 2022-10-18 code/Matrinem data analysis.Rmd
@@
 ---
-title: "Results section"
-author: "Kaare"
-date: "6/16/2022"
+title: "Weight analysis"
+author: "Kaare D. Tranaes"
+date: "`r Sys.Date()`"
 output: html_document
-editor_options: 
-  chunk_output_type: console
 ---
@@
 
-```{r loading libraries}
-library(readxl)
-library(tidyverse)
-library(broom)
-library(ggpubr)
-#library(AICcmodavg)
-library(visdat)
-library(rstatix)
-library(gt)
-library(gtsummary)
-library(patchwork)
-```
-
-# Introduction
-
-|           | PreExp 1 | PreExp 2 | PreExp 3 | NEC-FvT |
-|-----------|----------|----------|----------|---------|
-| **Aim**   |          |          |          |         |
-| Protocols |          |          |          |         |
-| N         |          |          |          |         |
-
-## Pre-Experiment 1 (PreExp 1)
-
-Experiment 1 was a pilot to test whether it was possible to feed 3-day old mice with the PICC.
-The experimental setup followed the standard setup.
-Antibiotics was administered according to the description in Chen et al., though with a lower dose based on previous experiments.
-Our primary study objective was survival.
-Secondary objective was macroscopic evaluation of intestines for inflammation, intestinal distention and similar NEC-like symptoms.
-One preliminary hypothesis was a reduction in weight of offspring from antibiotics-treated mothers.
-
-#### Weight development
-
-# Experiment 2 - weight at birth.Summary: Offspring from antibiotics-treated mice were smaller at birth
-
+# Analysis of the weight development in the MARTRINEM study 
 ```{r}
-data <- read_excel("../Data/CleanData/weight_development.xlsx") %>% 
+data <- read_excel("data/processed/weight_development.xlsx") %>% 
   select(sample_id:cage, bodyweight_birthday_group_average, bodyweight_birthday_plus_24H_group_average, bodyweight_birthday_plus_48H_group_average, bodyweight_baseline, bodyweight_baseline_plus_24h, bodyweight_baseline_plus_48h) %>%
@@
 ```
-
-#### Histological analysis
-
-#### Cytokine expression
-
-boxpltos
-
-```{r}
-boxplots_all_genes <- delta_gene_expr %>% 
-  ggplot(
-    aes(x = groups,
-        y = relative_fold_change,
-        fill = groups)
-  ) + 
-  geom_boxplot() +
-  facet_wrap(~genes,
-             ncol = 5, 
-             scales = "free") +
-  labs(title = "Gene expression fold change",
-       y = "Gene expression fold change",
-       x = "Groups"
-      )
-boxplots_all_genes
-
-```
-
-#### Gut microbiome analysis
-
-### Experiment 2 (E2)
-
-#### Weight development
-
-#### Histological analysis
-
-#### Intestinal permeability
-
-#### Cytokine expression
-
-#### Gene expression
-
-#### Gut microbiome analysis
-
-### Experiment 4 (NEC-FvT)
-
-The PreExps are now over and it is time to conduct the study. Having determined the optimal experimental protocol in PreExp 1-3 we are ready with the design.
-
-#### Overview - Experimental Setup
-
-#### Intestinal permeability - FITC
-
-```{r FITC load data, include=FALSE}
-fitc <- read_excel("../Data/CleanData/fitc.xlsx") %>% 
-  rename(values = `result_µg/ml_undiluted`) %>% 
-  drop_na(values)
-```
-
-Intestinal permeability is assessed with the FITC-Dextran assay. First I inspect the distribution of results.
-
-```{r FITC check for normality}
-fitc %>% 
-  filter(exp_number == 4) %>% 
-  group_by(group) %>% 
-  shapiro_test(values)
-```
-
-As results and highly skewed I try transforming the FITC results to log10.
-
-```{r FITC try log transforming data}
-fitc <- fitc %>% 
-  mutate(log_10_values = log10(values))  # Logtransform results
-
-```
-
-```{r FITC check for normality log transform}
-fitc %>% 
-  filter(exp_number == 4) %>% 
-  group_by(group) %>% 
-  shapiro_test(log_10_values)
-```
-
-```{r FITC visual inspections of data distribution}
-#Density plots non-log
-density <- fitc %>% 
-  filter(exp_number == 4) %>% 
-  ggplot(aes(x = values, fill = group)) + 
-  geom_density() +
-  facet_wrap(~ group,
-             scales = "free")
-
-#Jitter plots non-log
-jitter <- fitc %>% 
-  filter(exp_number == 4) %>% 
-  ggplot(aes(x = group, y = values)) + 
-  geom_jitter() +
-  facet_wrap(~ group,
-             scales = "free")
-
-#Density plots log
-density <- fitc %>% 
-  filter(exp_number == 4) %>% 
-  ggplot(aes(x = , fill = group)) + 
-  geom_density() +
-  facet_wrap(~ group,
-             scales = "free")
-
-#Jitter plots non-log
-jitter <- fitc %>% 
-  filter(exp_number == 4) %>% 
-  ggplot(aes(x = group, y = values)) + 
-  geom_jitter() +
-  facet_wrap(~ group,
-             scales = "free")
-
-```
-
-#### Cytokine expression
-
-#### Gene expression
-
-#### Histological analysis
-
-#### Gut microbiome analysis

### 2022-10-19 code/Matrinem_data_analysis.Rmd

--- 2022-10-19 code/Result_section.Rmd
@@
 author: "Kaare"
-date: "18/10/2022"
-output: html_document
+date: "6/16/2022"
+output:
+  pdf_document: default
+  html_document: default
 editor_options:
@@
 knitr::opts_chunk$set(echo = FALSE)
+#setwd(dir = "../../2022_matrinem_thesis/")
 ```
 
-```{r loading libraries}
+```{r loading libraries, message=FALSE}
 library(readxl)
@@
 
-## Preface
-
-## Introduction
-
-This thesis is about...
-
-## Background
-
-## Methods and Materials
-
-Prior to results are presented a short description of the studies conducted needs to be made.
+## The research questions
 
@@
 
-+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
++-----------+-------------------------------------------------------------------------------+----------+----------+---------+
 |           | PreExp 1                                                                      | PreExp 2 | PreExp 3 | NEC-FvT |
-+=============+===============================================================================+================================================================+========================================================================+=================================================================+
-| **Aim**     | To test feasibility of formula feeding with PICC line                         | Selectively targeting gram positive vs.
-gram negative bacteria | Testing if there is an interaction-effect with feeding and antibiotics | Combining reults from PreExp 1-3 into final experimental setup.
-|
-+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
-| Protocols   | Testing a broad spectrum antibiotics mix in combination with formula feeding.
-| Groups                                                         | Groups                                                                 | Groups                                                          |
++===========+===============================================================================+==========+==========+=========+
+| **Aim**   | To test feasibility of formula feeding with PICC line                         |          |          |         |
++-----------+-------------------------------------------------------------------------------+----------+----------+---------+
+| Protocols | Testing a broad spectrum antibiotics mix in combination with formula feeding. |          |          |         |
 |           |                                                                               |          |          |         |
@@
 |           |                                                                               |          |          |         |
-|             | 1.
-CON-BF                                                                    |                                                                |                                                                        |                                                                 |
-|             | 2.
-CON-FF                                                                    |                                                                |                                                                        |                                                                 |
-|             | 3.
-AB-BF                                                                     |                                                                |                                                                        |                                                                 |
-|             | 4.
-AB-FF                                                                     |                                                                |                                                                        |                                                                 |
-+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
-| N           | 38 (32 survived)                                                              | 18                                                             | 32                                                                     | 26                                                              |
-+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
+|           | 1.  CON-BF                                                                    |          |          |         |
+|           | 2.  CON-FF                                                                    |          |          |         |
+|           | 3.  AB-BF                                                                     |          |          |         |
+|           | 4.  AB-FF                                                                     |          |          |         |
++-----------+-------------------------------------------------------------------------------+----------+----------+---------+
+| N         | 38 (32 survived)                                                              |          |          |         |
++-----------+-------------------------------------------------------------------------------+----------+----------+---------+
 
@@
 
-#### **Rationale:**
-
-*Chen et al.*, showed that Maternal Antibiotics Treatment (**MAT**) resulted in mild NEC-like symptoms of mice offspring.
-We believed that MAT combined with formula feeding (**FF**) of offspring would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
-
-#### **Aim:**
-
-To test if it was possible to keep mice separated from the mother alive from postnatal day 3 and 48 hours onwards.
-
-#### **Research Questions:**
-
--   Can we keep offspring alive with formula feeding?
--   Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.
-
-Experiment 1 was a pilot to test whether it was possible to feed 3-day old mice with the PICC.
-The experimental setup followed the standard setup.
-Antibiotics was administered according to the description in Chen et al., though with a lower dose based on previous experiments.
-Our primary study objective was survival.
-Secondary objective was macroscopic evaluation of intestines for inflammation, intestinal distention and similar NEC-like symptoms.
-One preliminary hypothesis was a reduction in weight of offspring from antibiotics-treated mothers.
-
-**Results**:
+#Load in the data
 
 ```{r include=FALSE}
-preexp_1 <- read_excel("../data/processed/weight_development.xlsx") %>% 
+preexp_1 <- read_excel("data/processed/weight_development.xlsx") %>% 
   filter(exp_number == "1") 
-
 ```
@@
 #__________Load in cytokine data for all experiments________
-cytokines_1 <- read_excel(path = "../data/processed/cytokines.xlsx") %>%  
+cytokines_1 <- read_excel(path = "data/processed/cytokines.xlsx") %>%  
   filter(exp_number == 1) %>% 
@@
 ```
+
+#### **Rationale:**
+
+*Chen et al.*, showed that Maternal Antibiotics Treatment (**MAT**) resulted in mild NEC-like symptoms of mice offspring.
+We believed that MAT combined with formula feeding (**FF**) of offspring would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+
+#### **Aim:**
+
+To test if it was possible to keep mice separated from the mother alive from postnatal day 3 and 48 hours onwards.
+
+#### **Research Questions:**
+
+##### Can we keep offspring alive with formula feeding?
+
+`r nrow(preexp_1)` mice were born to 6 different mothers.
+Of these 32 survived and were included in the analysis.
+After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours for 2 days (48 hours).
+
+##### Will MAT and FF affect levels of inflammatory cytokines in ileum tissue?
 
@@
 
-`r nrow(preexp_1)` mice were born to 6 different mothers. Of these 32 survived and were included in the analysis.
-
 `r nrow(cytokines_1)` samples were available for analysis of concentrations of the proinflammatory cytokines `r names_of_cytokines`.
 Concentrations were measured in 1-cm sections of proximal/distal ileum tissue using MSD Mesoscale Kits.
-Distributions were assessed for normality with Shapiro-Wilk Normality Test (see table below), and plot of data were made to assess distributions visually:
-
-```{r echo=FALSE, fig.cap='Table:}
-normality_result_1
-```
-
-```{r Normality plot, echo=FALSE, fig.cap='Non-normally distributed plots'}
-normality_plot
-```
 
@@
              stat_4 ~ px(150))
-# Save and export files
-# tbl1 %>%  #Save as excel file.
-#   as_hux_table() %>%
-#   huxtable::quick_xlsx(., file = "../Results/Round 1/ileum-cytokines.xlsx")
-# 
-# tbl1 %>% #Save as image file.
-# as_flex_table() %>%
-#   flextable::save_as_image(path = "../../Results/Round 1/ileum_cytokines_table.png",webshot = "webshot")
-
 ```
 
-The median concentration of the various cytokines are shown in the table below:
+The concentrations of the various cytokines are shown in the table below:
 
@@
 tbl1_gt
+
 ```
@@
 #ggsave('../../Results/Round 1/All Cytokines-boxplot.png',all_cytokines_patch)
+```
+
+Cytokine levels are shown below:
+
+```{r echo=FALSE}
+all_cytokines_patch
+```
+
+Distributions were assessed for normality with Shapiro-Wilk Normality Test (see table below), and plot of data were made to assess distributions visually:
+
+```{r echo=FALSE, fig.cap='Table of test results:'}
+normality_result_1
+```
+
+Most measured cytokines wore not normally disrtibuted. Consider logtransformation
+
+```{r Normality plot, echo=FALSE, fig.cap='Non-normally distributed plots'}
+normality_plot
 ```
@@
 
+The plot above shows that the cytokines significantly affected / difference were `r tbl1_p$Cytokines`.
+
 A linear regression analysis calculates the factors responsible for the changes observed.
@@
   select(-statistic) %>% 
-  knitr::kable(., caption = 'Significant effectors on cytokine expression', digits = 3)
+  knitr::kable(., caption = 'Significant effectors on cytokine expression', digits = 5)
+
+ff_affected_cytokines <- reg_mod_cyt_1 %>% 
+  filter(term %in% "type_of_feedFF") %>% 
+  pull(cytokines)
+
+ff_affected_cytokines_pvals <- reg_mod_cyt_1 %>% 
+  filter(term %in% "type_of_feedFF") %>% 
+  pull(p.value)
 ```
 
-**The conclusion** is that many of the cytokines show an interaction effect of the two treatment parameters
+Formula-feeding looks to be the most determining factor for changes to cytokine expression.
+
+**The conclusion** is that many of the cytokines show an interaction effect of the two treatment parameters.
 
@@
 
--   Can we keep offspring alive with formula feeding?
+-   We could keep them alive with proper feeding technique.
 
-    -   After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours.
-    -   The formulafed animals gained weight - albeit less than their breastfed peers - during the 48 hours.
-    -   MAT did not affect growth. CON-BF and AB-BF gained weight at a similar pace.
+-   Formula feeding looks to be the most determining factor for `r ff_affected_cytokines` (p-values `r ff_affected_cytokines_pvals`, respectively).
 
@@
 
--   Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.
+#### **Rationale** 
+- Based on the results from Preexp 1 we knew that we could perform the experimental setup without too many casualties. 
+- We had seen that formula-feeding did affect the expression of cytokines, but visual inspection of the intestines showed that we had not managed to create a NEC-like phenotype. 
+- After considering the results we realized that our assumptions were based on a dysbiotics microbiome in the mother leading to less immunizing milk.
+However, we had likely not created a dysbiosis as much as an overall depletion by administering a broad-spectered antibiotic.
+We therefore hypothezised that targeting specific stains of bacteria would lead to a more dysbiotic microbiome and allow for greater disbiosis in the offspring.
+
+- In **PreExp 2** we therefore tested whether selectively targeting gram+ or gram- bacteria would lead to different and more pronounced NEC-like symptoms.
+
+#### Research questions: 
+##### Will selectively targering gram+ or gram- bacteria with Vancomycin and Gentamicin respectively lead to more pronounced NEC-like symptoms compared with no treatment on the background of breastfeeding.
+
+### Experiment 4 (NEC-FvT)
+
+The PreExps are now over and it is time to conduct the study. Having determined the optimal experimental protocol in PreExp 1-3 we are ready with the design.
+
+#### Overview - Experimental Setup
+
+#### Intestinal permeability - FITC
+
+```{r FITC load data, include=FALSE}
+fitc <- read_excel("data/processed/fitc.xlsx") %>% 
+  rename(values = `result_µg/ml_undiluted`) %>% 
+  drop_na(values)
+```
+
+Intestinal permeability is assessed with the FITC-Dextran assay. First I inspect the distribution of results.
+
+```{r FITC check for normality}
+fitc %>% 
+  filter(exp_number == 4) %>% 
+  group_by(group) %>% 
+  shapiro_test(values)
+```
+
+As results and highly skewed I try transforming the FITC results to log10.
+
+```{r FITC try log transforming data}
+fitc <- fitc %>% 
+  mutate(log_10_values = log10(values))  # Logtransform results
+
+```
+
+```{r FITC check for normality log transform}
+fitc %>% 
+  filter(exp_number == 4) %>% 
+  group_by(group) %>% 
+  shapiro_test(log_10_values)
+```
+
+```{r FITC visual inspections of data distribution}
+#Density plots non-log
+density <- fitc %>% 
+  filter(exp_number == 4) %>% 
+  ggplot(aes(x = values, fill = group)) + 
+  geom_density() +
+  facet_wrap(~ group,
+             scales = "free")
+
+#Jitter plots non-log
+jitter <- fitc %>% 
+  filter(exp_number == 4) %>% 
+  ggplot(aes(x = group, y = values)) + 
+  geom_jitter() +
+  facet_wrap(~ group,
+             scales = "free")
+
+#Density plots log
+density <- fitc %>% 
+  filter(exp_number == 4) %>% 
+  ggplot(aes(x = , fill = group)) + 
+  geom_density() +
+  facet_wrap(~ group,
+             scales = "free")
+
+#Jitter plots non-log
+jitter <- fitc %>% 
+  filter(exp_number == 4) %>% 
+  ggplot(aes(x = group, y = values)) + 
+  geom_jitter() +
+  facet_wrap(~ group,
+             scales = "free")
+
+```

### 2022-10-19 submission/manuscript.Rmd
Wrote first draft for results

@@
 
-## Results
+- NEC is caused by bacterial dysbiosis - we think. 
+- (AND) high-osmolar formula feed is hard on intestinal lining. 
+- (AND) Chen et al. saw NEC-like symptoms with MAT.
+- (BUT) No enough on its own plus too old mice. 
+- (Therefore) We want to see if MAT + FF will result in more severe NEC-like symptoms. (AIM)
+- (AND) FvT are currently being studied as potential remidies to NEC - thought to be mediated
+via stabilizing effect on microbiome. 
+- (Therefore) We want to see if A) we manage to induce NEC, are we able to "cure" it, or B) in 
+case of no NEC, is FvT on it's own safe or associated with any adverse effects?
+
+## Results (decide on which voice to use, active or passive. Read other papers.)
+- We treated XX mothers with Vancomycin (1 g/l) from pregnancy-day 15 until delivery, where vancomycin-mixed drinking water was replaced with regular water.
+- 2 days after birth xx of the 26 pups were separated from the mother and received XX µl FvT solution or an equal amount of SM buffer (control treatment).
+The remaning xx of 26 pups remained with the mother for breastfeeding thoughout the study.
+- Pups separated from their mother were formula fed (FF) with a solution of XXXXX milk formulation (see materials) every 3 hours for 48 hours. (futher setup description see materials and methods).
+- All 26 pups survived the 48 hour study. On the final day of the study pups fasted for XX hours before given xx µl FITC Dextran (as per materials protocol). 
+- 2 hours after FITC administtion pups were euthanized by decapitation and blood collected in eppendorfs for analysis of serum FITC concentration.
+Intestinal samples were collected for gene expression assays, histological assessment, pro-inflmmatory cytokine analysis and gut microbiome sequencing.
+
+* Gut permeability: 
+- We found the lowest concentration of serum FITC Dextran in the group treated with FF and Fvt (XXXX Xg/mL).
+This was lower than even the breastfed controls, which had the highest concentration of the three groups (**FIGURE**).
+
+* Cytokine expression
+- High levels of XX cytokine was foudn in group XX.
+
+* Histological analysis
+- We found no significant changes in the mucin layer between the treatment group (Fvt) Fvt-Control (SM) or formula-feed control (VANCO-BF). The thickness of the stain was the same for all groups.
+* This supports the impression we got from the visual inspection of the intestines during sampling.
+
+* Gene expression 
+We standardized the gene expression to the control group and the house gene.
+However, we found statistically significant difference in expression of the house-gene, making the delta-delta measure bad.
+Comparing (on what other basis) in stead.
+
+* Gut microbiome
+Differneces?
 
 ## Discussion
+
+## Materials and methods
+Describe Pre-Exp 1-3 and knowledge gained from them.

### 2022-10-20 submission/manuscript.Rmd
Progressed on manuscript.

@@
 
+#### Rationale:
+
+*Chen et al.*, showed that Maternal Antibiotics Treatment (**MAT**) resulted in mild NEC-like symptoms of mice offspring.
+We believed that MAT combined with formula feeding (**FF**) of offspring would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+
 ## Results (decide on which voice to use, active or passive. Read other papers.)
+
+Experiment no. 4 was the goal, but before we could test our overall hypothesis we needed to find out which combination of antibiotics and feeding scheme would results in the most NEC-like phenotype.
+These research questions were investigated in 3 pre-experiments (PreExp 1-3), each with their own research questions.
+
+#### Model feasibility
+- Our first experiement (PreExp 1) was designed to test wheteher  it was possible to keep mice separated from the mother alive from postnatal day 3 and 48 hours onwards. 
+- `r nrow(preexp_1)` mice were born to 6 different mothers.
+Of these 32 survived and were included in the analysis.
+After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours for 2 days (48 hours).
+(Conclusion) - Visual inspection of intestines during sampling concluded that we had not created a NEC phenotype in any treatment group.
+(BUT) - We found that AB-FF pups consistently looked different that CON-BF in the level of cytokine expression in the ileum, with statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
+- A multiple linear regression analysis, evaluating the effects of MAT (AB or CON) and type of feed (BF of FF) separately and interactively found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val$IL-6.
+
+We therefore decided to test different types of antibiotics.
+
+#### Microbiome dysbiosis
+Based on the results from Preexp 1 we knew that we could perform the experimental setup without too many casualties. 
+- We had seen that formula-feeding did affect the expression of cytokines, but visual inspection of the intestines showed that we had not managed to create a NEC-like phenotype. 
+- After considering the results we realized that our assumptions were based on a dysbiotics microbiome in the mother leading to less immunizing milk.
+However, we had likely not created a dysbiosis as much as an overall depletion by administering a broad-spectered antibiotic.
+We therefore hypothezised that targeting specific stains of bacteria would lead to a more dysbiotic microbiome and allow for greater disbiosis in the offspring.
+
+### Round 4
 - We treated XX mothers with Vancomycin (1 g/l) from pregnancy-day 15 until delivery, where vancomycin-mixed drinking water was replaced with regular water.

### 2022-10-20 submission/manuscript.Rmd
Finished first ROUGH draft of result section

@@
 
-*Chen et al.*, showed that Maternal Antibiotics Treatment (**MAT**) resulted in mild NEC-like symptoms of mice offspring.
-We believed that MAT combined with formula feeding (**FF**) of offspring would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+*Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
+We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
 
@@
 Experiment no. 4 was the goal, but before we could test our overall hypothesis we needed to find out which combination of antibiotics and feeding scheme would results in the most NEC-like phenotype.
-These research questions were investigated in 3 pre-experiments (PreExp 1-3), each with their own research questions.
+These research questions were investigated in 3 pre-experiments (PreExp 1-3), each with specific research questions.
 
 #### Model feasibility
-- Our first experiement (PreExp 1) was designed to test wheteher  it was possible to keep mice separated from the mother alive from postnatal day 3 and 48 hours onwards. 
+- Our first experiement (PreExp 1) was designed to test wheteher it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
+This required feeding wit a venous catheter inserted orally into the stomach.
 - `r nrow(preexp_1)` mice were born to 6 different mothers.
@@
 
-We therefore decided to test different types of antibiotics.
-
-#### Microbiome dysbiosis
+#### Effects of various types of antibiotics - (Microbiome dysbiosis)
 Based on the results from Preexp 1 we knew that we could perform the experimental setup without too many casualties. 
@@
 However, we had likely not created a dysbiosis as much as an overall depletion by administering a broad-spectered antibiotic.
-We therefore hypothezised that targeting specific stains of bacteria would lead to a more dysbiotic microbiome and allow for greater disbiosis in the offspring.
+We therefore hypothezised that targeting specific strains of bacteria would lead to a more dysbiotic microbiome and allow for greater disbiosis in the offspring.
+
+We tested 2 different types, all on a background of breastfeeding. 
+Contrary to our hypothesis, these mice had lower levels of intestinal permeability - both relative to controls and in absolute terms compared with FF-pups from PreExp 1.
+
+#### Interaction effects between antibiotics treatment and formula feeding
+- In round 3 we repeated the experimental setup with regards to MAT, adding either Vancomycin, Gentamicin or sterile water to the drinking water of pregnant "females" (use same word throughout(mothers)).
+We repeated the group-setup from PreExp  1 and  separated pups into breastfeeding or formula feeding, for a total of 6 groups (see overview table).
+- All pups survived the experiment and were included in the final analysis. 
+- 2 hours after FITC administtion pups were euthanized by decapitation and blood collected in eppendorfs for analysis of serum FITC concentration.
+Intestinal samples were collected for gene expression assays, histological assessment, pro-inflammatory cytokine analysis and gut microbiome sequencing.
+
+* Gut permeability: 
+- Formula-fed pups consistently showed higher levels of Serum FITC concentration compared to breastfed pups, irrespective of MAT status.
+VANCO-FF and VANCO-BF had the highest and lowest levels of FITC, respectively.
+- VANCO-BF was the group with the lowest FITC in PreExp 2.
+- This is indicative of an interaction-effect with vancomycin and formula feeding. (**FIGURE**).
+
+* Cytokine expression
+- We found significant differences in concentrations of XXSignificantSCYTOKINESXX.
+Linear regression analysis indicated the biggest detminant of differneces in outcomes was formula feeding, with the type of antibiotic being of lesser importance compared.
+Directionally the trend was the same for all cytokines.
+- IL-2: significantly lower concentrations in all formula-fed pups compared with CON-BF.
+- IL-6: Significantly greater levels in all formula-fed pups compared with CON-BF.
+- INFg: significantly lower concentrations in all formula-fed pups compared with CON-BF.
+- KCGRO: Significantly greater levels in all formula-fed pups compared with CON-BF.
 
@@
 - 2 hours after FITC administtion pups were euthanized by decapitation and blood collected in eppendorfs for analysis of serum FITC concentration.
-Intestinal samples were collected for gene expression assays, histological assessment, pro-inflmmatory cytokine analysis and gut microbiome sequencing.
+Intestinal samples were collected for gene expression assays, histological assessment, pro-inflammatory cytokine analysis and gut microbiome sequencing.
 
@@
 This was lower than even the breastfed controls, which had the highest concentration of the three groups (**FIGURE**).
+- The VANCO-FF (Fvt) compared with VANCO-FF from PreExp 3 this change was significant/non-significant. 
+- It is reasonable/unreasonable to compare the rounds, given the similarity/lack of similarity in response of the VANCO-BF groups from Exp3(XXX) and NEC-FvT (XXXX)
 
 * Cytokine expression
-- High levels of XX cytokine was foudn in group XX.
+- Only TNFa was significantly different in Exp 4, with VANCO-FF (FvT) having the highest level (XXX), signifiacntly greater than both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively). 
+- The VANCO-FF (FvT) had lower/higer absolute levels of cytokine concentration compared to both VANCO-BF and VANCO-FF from Exp3. 
+- THis comparison is reasonable/unreasonable given the similarity/lack of similarity in response of the VANCO-BF groups from Exp3(XXX) and VANCO-FF (SM) from Exp 4(XXXX).
 
-* Histological analysis
+### Post experimental analyses. 
+* Histological analysis - Mention and Discuss this after finishing the result section. 
+- Histological image analysis was performed after all trials had finished. 
+- We saw no degradation of mucin layer thickness in any of the pups. This supports the visial inspections of intestines that was made during sampling. 
 - We found no significant changes in the mucin layer between the treatment group (Fvt) Fvt-Control (SM) or formula-feed control (VANCO-BF). The thickness of the stain was the same for all groups.
-* This supports the impression we got from the visual inspection of the intestines during sampling.
 
@@
 * Gut microbiome
+
 Differneces?

### 2022-10-22 submission/manuscript.Rmd
New manuscript progress

@@
 - We treated XX mothers with Vancomycin (1 g/l) from pregnancy-day 15 until delivery, where vancomycin-mixed drinking water was replaced with regular water.
-- 2 days after birth xx of the 26 pups were separated from the mother and received XX µl FvT solution or an equal amount of SM buffer (control treatment).
+- 2 days after birth XX of the 26 pups were separated from the mother and received XX µl FvT solution or an equal amount of SM buffer (control treatment).
 The remaning xx of 26 pups remained with the mother for breastfeeding thoughout the study.
@@
 
-### Post experimental analyses. 
+### Post experimental analyses and Discussion of results 
 * Histological analysis - Mention and Discuss this after finishing the result section. 
@@
 * Gene expression
-We standardized the gene expression to the control group and the house gene.
-However, we found statistically significant difference in expression of the house-gene, making the delta-delta measure bad.
-Comparing (on what other basis) in stead.
+
+We standardized the gene expression to the control group and the house gene. However, we found statistically significant difference in expression of the house-gene, making the delta-delta measure bad.
 
 * Gut microbiome
-
-Differneces?
 

### 2022-10-26 code/Result_section.Rmd
Updated results manuscript

@@
 
-```{r loading libraries}
+```{r loading libraries, message=FALSE}
 library(readxl)
@@
 # MATRINEM - Attempts at creating a NEC mouse model
-
-## Preface
-
-## Introduction
-
-This thesis is about...
-
-## Background
-
-## Methods and Materials
-
-Prior to results are presented a short description of the studies conducted needs to be made.
-
-Experiment no. 4 was the goal, but before we could test our overall hypothesis we needed to find out which combination of antibiotics and feeding scheme would results in the most NEC-like phenotype.
-
-These research questions were investigated in 3 pre-experiments (PreExp 1-3), each with their own research questions.
 
@@
 
-```{r echo=FALSE, fig.cap='Table:}
+```{r echo=FALSE, fig.cap='Table XX'}
 normality_result_1

### 2022-10-26 submission/manuscript.Rmd
Updated results manuscript

@@
-```
+---
+title: "Results section"
+author: "Kaare"
+date: "`r format(Sys.time(), '%d %B, %Y')`"
 output: html_document
+editor_options: 
+  chunk_output_type: console
+---
 
-```
+# Code
+
+# Manuscript
 
@@
 
+## Results
+
+The main goal of this thesis project was to create a NEC mouse model by combining Maternal Antibiotics Treatment with Formula-Feeding.
+This hypothesis was tested in a series of experiments (Exp 1-4), each successive experiment building on the results of the former.
+The following is a description of each experiment. An abridged version of each experiment is provided, followed by a detailed  description of experimental setup and results.
+
+### Exp 1: Model feasibility
+
 #### Rationale:
-
 *Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
@@
 
-## Results (decide on which voice to use, active or passive. Read other papers.)
+#### Aim: 
+- To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
+- Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
+- Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum
 
-Experiment no. 4 was the goal, but before we could test our overall hypothesis we needed to find out which combination of antibiotics and feeding scheme would results in the most NEC-like phenotype.
-These research questions were investigated in 3 pre-experiments (PreExp 1-3), each with specific research questions.
+#### Conclusion:
+- Survival rate: xx of `r nrow(preexp_1)` (%)
+- Visaul inspection: no clear signs of intestinal inflammation
+- Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
 
-#### Model feasibility
-- Our first experiement (PreExp 1) was designed to test wheteher it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
-This required feeding wit a venous catheter inserted orally into the stomach.
-- `r nrow(preexp_1)` mice were born to 6 different mothers.
-Of these 32 survived and were included in the analysis.
-After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours for 2 days (48 hours).
+#### Detailed description of Exp 1. **Date of exp** 
+*Antimals and experimental setup*
+- Six male and 18 female C57BL/6N mice were co housed for 24 hours in 6 cages (1 male + 3 female each).
+- Two weeks after mating, pregnant females were separated evenly into maternal antibiotic treatment (MAT) and water-only controls (CON). 
+- Antibiotics (Gentamicin, Vancomycin, Neomycin, and Ampicillin (all 0.5 g/L)) was mixed into the drinking water of the MAT mice while controls received water only.
+- Antibiotic treatment commenced on gestational day 15 and continued until delivery. 
+- Each mother was housed with her offspring in separate cages during the first 3 days after birth. 
+- On the 4th day, `r nrow(preexp_1)` mice, XXX (Filter `r nrow(preexp_1)` for feed type) from antibiotics treated mothers and XXX (Filter `r nrow(preexp_1)` for feed type) were divided into four experimental groups: MAT + formula-feeding (AB-FF) MAT + breastfeeding (AB-BF), Water + formula-feeding (CON-FF), and CON + breastfeeding (CON-BF).
+
+*Feeding* **See materials and methods**
+- Formula-feeding was performed by fixating the pup between thumb and forefinger, and inserting 2 cm of a venous catheter, designed for human infants, into the mouth of the pup using pliers. 
+- The PICC was connected to a 1.0 ml syringe with formula. For pups >2g, 0.08 ml formula was administered (0.06 for pups ≤ 2g). 
+- Feeding was done in the same order every 3 hours for the study duration, including night hours to ensuring no overnight fasting.
+
+At 08:00 am on the day of euthanasia FORM pups received a final feeding and BF pups were separated from the mothers (halting breastfeeding).
+At 11:00 (three hours after the last feeding/separation), 50 µL FITC-dextran (0.045 mg/µL) was administered orally using a PICC line and a 1 mL syringe for a final dose of 750 mg/kg.
+Euthanasia and sample collection commenced at noon (12 am).
+
+*Euthanasia and sample collection*
+The head of the pups is cut off, and blood dripping is collected in an Eppendorf tube.
+The blood samples are spun down, and serum is collected and stored at -80 for cytokine measurements.
+Small and large intestines will be removed and evaluated macroscopically (remember photographs) for signs of NEC (intestinal segments of bowel necrosis, intestinal haemorrhage, and perforation).
+In addition, 1 cm of terminal ileum will be removed and stored in 4% paraformaldehyde for histological evaluation.
+The remaining small intestine is preserved in RNAlater for qPCR.
+Cecum and colon are dissected and stored in an Eppendorf tube on dry ice.
+H
+istologic injury score of NEC in neonatal mice
+Initially, the bowel is investigated macroscopically.
+Pictures are taken of the intestines in situ after abdominal incision and abdominal dissection.
+The location and extent of pathological areas are estimated.
+The severity of intestinal lesions will be assessed histopathologically in H&E-stained ileal tissue sections using an intestinal tissue injury score (0–4), where: Grade 0: intact tissue; 1: mild damage, slight submucosal and/or lamina propria separation; 2: moderate-severe separation of the submucosa and/or lamina propria and/or oedema in the submucosal and muscular layers, partial loss of villous sloughing; 3: severe separation of the submucosa and/or lamina propria region, villous sloughing and initial villus necrosis; 4: necrosis and loss of villi structure and/or transmural necrosis.
+Samples with histological scores of 2 or higher will be considered positive for NEC.
+
+Samples to be collected:
+•	From parent mice:
+o	Faecal matter (MAT and CON mice)
+•	From offspring post euthanasia
+o	1-2 cm of ileum immersion-fixed in paraformaldehyde for histological analysis
+o	1-2 cm of ileum snap-frozen for mesoscale + qPCR analysis
+o	Whole caecum for microbiota analysis
+
+ Of these 32 survived and were included in the analysis. After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours for 2 days (48 hours).
 (Conclusion) - Visual inspection of intestines during sampling concluded that we had not created a NEC phenotype in any treatment group.
@@
 
-#### Effects of various types of antibiotics - (Microbiome dysbiosis)
+### Experiment 2: Effects of various types of antibiotics - (Microbiome dysbiosis)
+
+#### Rationale:
+*Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
+We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+
+#### Aim: 
+- To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
+- Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
+- Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum
+
+#### Conclusion:
+- Survival rate: xx of `r nrow(preexp_1)` (%)
+- Visaul inspection: no clear signs of intestinal inflammation
+- Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
+
+#### Detailed description of Exp 2.
 Based on the results from Preexp 1 we knew that we could perform the experimental setup without too many casualties. 
@@
 However, we had likely not created a dysbiosis as much as an overall depletion by administering a broad-spectered antibiotic.
-We therefore hypothezised that targeting specific strains of bacteria would lead to a more dysbiotic microbiome and allow for greater disbiosis in the offspring.
+We therefore hypothesized that targeting specific strains of bacteria would lead to a more dysbiotic microbiome and allow for greater disbiosis in the offspring.
 
@@
 
-#### Interaction effects between antibiotics treatment and formula feeding
+### Experiment 3 - Interaction effects between antibiotics treatment and formula feeding
+
+#### Rationale:
+*Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
+We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+
+#### Aim: 
+- To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
+- Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
+- Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum
+
+#### Conclusion:
+- Survival rate: xx of `r nrow(preexp_1)` (%)
+- Visaul inspection: no clear signs of intestinal inflammation
+- Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
+
+#### Detailed description of Exp 3.
 - In round 3 we repeated the experimental setup with regards to MAT, adding either Vancomycin, Gentamicin or sterile water to the drinking water of pregnant "females" (use same word throughout(mothers)).
@@
 
-### Round 4
+### Experiment 4
+
+#### Rationale:
+*Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
+We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+
+#### Aim: 
+- To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
+- Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
+- Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum
+
+#### Conclusion:
+- Survival rate: xx of `r nrow(preexp_1)` (%)
+- Visaul inspection: no clear signs of intestinal inflammation
+- Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
+
+#### Detailed description of Exp 4.
 - We treated XX mothers with Vancomycin (1 g/l) from pregnancy-day 15 until delivery, where vancomycin-mixed drinking water was replaced with regular water.
@@
 ## Discussion
-
-## Materials and methods
-Describe Pre-Exp 1-3 and knowledge gained from them.

### 2022-10-27 code/Result_section.Rmd
Updated results and wrote more manuscript

@@
 
-### **Overview of all experiments**
-
-+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
-|             | PreExp 1                                                                      | PreExp 2                                                       | PreExp 3                                                               | NEC-FvT                                                         |
-+=============+===============================================================================+================================================================+========================================================================+=================================================================+
-| **Aim**     | To test feasibility of formula feeding with PICC line                         | Selectively targeting gram positive vs.
-gram negative bacteria | Testing if there is an interaction-effect with feeding and antibiotics | Combining reults from PreExp 1-3 into final experimental setup.
-|
-+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
-| Protocols   | Testing a broad spectrum antibiotics mix in combination with formula feeding.
-| Groups                                                         | Groups                                                                 | Groups                                                          |
-|             |                                                                               |                                                                |                                                                        |                                                                 |
-|             | 4 groups                                                                      |                                                                |                                                                        |                                                                 |
-|             |                                                                               |                                                                |                                                                        |                                                                 |
-|             | 1.
-CON-BF                                                                    |                                                                |                                                                        |                                                                 |
-|             | 2.
-CON-FF                                                                    |                                                                |                                                                        |                                                                 |
-|             | 3.
-AB-BF                                                                     |                                                                |                                                                        |                                                                 |
-|             | 4.
-AB-FF                                                                     |                                                                |                                                                        |                                                                 |
-+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
-| N           | 38 (32 survived)                                                              | 18                                                             | 32                                                                     | 26                                                              |
-+-------------+-------------------------------------------------------------------------------+----------------------------------------------------------------+------------------------------------------------------------------------+-----------------------------------------------------------------+
-
-### PreExp 1 - Feasibility Study
-
-#### **Rationale:**
-
-*Chen et al.*, showed that Maternal Antibiotics Treatment (**MAT**) resulted in mild NEC-like symptoms of mice offspring.
-We believed that MAT combined with formula feeding (**FF**) of offspring would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
-
-#### **Aim:**
-
-To test if it was possible to keep mice separated from the mother alive from postnatal day 3 and 48 hours onwards.
-
-#### **Research Questions:**
-
--   Can we keep offspring alive with formula feeding?
--   Will combining MAT with FF result in a more inflammatory intestine as measured by cytokine analysis in the ileum.
-
-Experiment 1 was a pilot to test whether it was possible to feed 3-day old mice with the PICC.
-The experimental setup followed the standard setup.
-Antibiotics was administered according to the description in Chen et al., though with a lower dose based on previous experiments.
-Our primary study objective was survival.
-Secondary objective was macroscopic evaluation of intestines for inflammation, intestinal distention and similar NEC-like symptoms.
-One preliminary hypothesis was a reduction in weight of offspring from antibiotics-treated mothers.
-
-**Results**:
-
 ```{r include=FALSE}
-preexp_1 <- read_excel("../data/processed/weight_development.xlsx") %>% 
+preexp_1 <- read_excel("data/processed/weight_development.xlsx") %>% 
   filter(exp_number == "1")
@@
                values_to = "cytokine_concentration") %>% 
-  drop_na(cytokine_concentration)
+  filter(!cytokine_concentration == 0) %>% 
+  mutate(conc_adjusted4weight = cytokine_concentration/tissue_weight_mg)
 #___________________________________________________________
+
 ```
@@
   group_by(cytokines) %>% 
-  shapiro_test(cytokine_concentration) %>% 
+  shapiro_test(conc_adjusted4weight) %>% 
   select(cytokines, p)
@@
   ggplot(aes(x = group, 
-             y = cytokine_concentration,
+             y = log10(conc_adjusted4weight),
              color = group)) + 
+  geom_boxplot(outlier.shape = NA) + 
   geom_jitter() +
+  #scale_y_continuous(trans='log10') +
   facet_wrap(~ cytokines,
@@
   labs(title = "Non-normally distributed cytokines (Shapiro-Wilk Normality Test, p < 0.05)",
-       y = "Concentration of cytokine in tissue (pg/mL)",
+       y = "Log10 scale - Conc. of cytokine (pg/mL) per mg tissue",
        x = "") +
@@
 ggplot(aes(x = group, 
-           y = cytokine_concentration, 
+           y = conc_adjusted4weight, 
            fill = group)) + 

### 2022-10-27 submission/manuscript.Rmd
Updated results and wrote more manuscript

@@
 
+## Background
+
+- Why we think MAT leads to intestinal dysbiosis?
+    - Less immunizing milk?
+-
+
 ## Results
@@
 #### Detailed description of Exp 1. **Date of exp** 
-*Antimals and experimental setup*
-- Six male and 18 female C57BL/6N mice were co housed for 24 hours in 6 cages (1 male + 3 female each).
+*Antimals and experimental setup* See "Animal experimental setup" in methods.
+- The experimental setup has been described in detail in the methods seciton. 
+- Detials specific for Exp-1 are outlined below. 
+- Briefly, six male and 18 female C57BL/6N mice were co housed for 24 hours in 6 cages (1 male + 3 female each).
 - Two weeks after mating, pregnant females were separated evenly into maternal antibiotic treatment (MAT) and water-only controls (CON). 
@@
 - On the 4th day, `r nrow(preexp_1)` mice, XXX (Filter `r nrow(preexp_1)` for feed type) from antibiotics treated mothers and XXX (Filter `r nrow(preexp_1)` for feed type) were divided into four experimental groups: MAT + formula-feeding (AB-FF) MAT + breastfeeding (AB-BF), Water + formula-feeding (CON-FF), and CON + breastfeeding (CON-BF).
+- Formula-feeding was performed as per METHODS AND MATERIALs.
+In Exp-1 we administered  0.08 ml formula for pups >2g, and 0.06 for pups ≤ 2g every 3 hours throughout the study.
+In later experiments this dose was changed, and feeding intervals were inreased to every 4 hours from midnight til 08 AM.
 
-*Feeding* **See materials and methods**
-- Formula-feeding was performed by fixating the pup between thumb and forefinger, and inserting 2 cm of a venous catheter, designed for human infants, into the mouth of the pup using pliers. 
-- The PICC was connected to a 1.0 ml syringe with formula. For pups >2g, 0.08 ml formula was administered (0.06 for pups ≤ 2g). 
-- Feeding was done in the same order every 3 hours for the study duration, including night hours to ensuring no overnight fasting.
+- `r nrow(preexp_1_survived)` of `r nrow(preexp_1)` pups survived until sampling.
+- Those that did not survive were euthanize because of esophageal perforation during feeding.
+After fine-tuning of the feeding methods no pups were eithanized as a result of reaching humane endpoints, and the pups generally were not very sick.
+- While significantly less than their breastfed controls, all pups increased their heavyweight during the experiment, attesting to their generally reasonable health status.
+- The pups were physically active and reacted to changes in light and to physical sitmuli.
+- Skin color remained bright purple and fur development was visible thoughout the experiment. 
+- No signs of intestinal inflammation were evident from the visual inspection of intestines during sampling.
+Our subjective observation of the intestines were small degrees of bloating in formula-fed pups.
+- Analysis of pro-inflammatory cytokines showed statistically significant changes in the expression of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
 
-At 08:00 am on the day of euthanasia FORM pups received a final feeding and BF pups were separated from the mothers (halting breastfeeding).
-At 11:00 (three hours after the last feeding/separation), 50 µL FITC-dextran (0.045 mg/µL) was administered orally using a PICC line and a 1 mL syringe for a final dose of 750 mg/kg.
-Euthanasia and sample collection commenced at noon (12 am).
+#### Discussion of experimental results - Exp 1.
+- Directionality of change varied, with alternating over- and under-expression of various cytokines.
+As can be seen in [*Cytokine plot of signif results *] pups tended to group based on feeding-scheme, with both formula-fed groups being either suppressed og over-expressed together, compared with breastfed groups, irrespective of antibiotics treatment.
+- This observation is supported by a a multiple linear regression analysis, evaluating the effects of MAT (AB or CON) and type of feed (BF of FF) separately and interactively.
+The analysis found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val$IL-6.
 
-*Euthanasia and sample collection*
-The head of the pups is cut off, and blood dripping is collected in an Eppendorf tube.
-The blood samples are spun down, and serum is collected and stored at -80 for cytokine measurements.
-Small and large intestines will be removed and evaluated macroscopically (remember photographs) for signs of NEC (intestinal segments of bowel necrosis, intestinal haemorrhage, and perforation).
-In addition, 1 cm of terminal ileum will be removed and stored in 4% paraformaldehyde for histological evaluation.
-The remaining small intestine is preserved in RNAlater for qPCR.
-Cecum and colon are dissected and stored in an Eppendorf tube on dry ice.
-H
-istologic injury score of NEC in neonatal mice
-Initially, the bowel is investigated macroscopically.
-Pictures are taken of the intestines in situ after abdominal incision and abdominal dissection.
-The location and extent of pathological areas are estimated.
-The severity of intestinal lesions will be assessed histopathologically in H&E-stained ileal tissue sections using an intestinal tissue injury score (0–4), where: Grade 0: intact tissue; 1: mild damage, slight submucosal and/or lamina propria separation; 2: moderate-severe separation of the submucosa and/or lamina propria and/or oedema in the submucosal and muscular layers, partial loss of villous sloughing; 3: severe separation of the submucosa and/or lamina propria region, villous sloughing and initial villus necrosis; 4: necrosis and loss of villi structure and/or transmural necrosis.
-Samples with histological scores of 2 or higher will be considered positive for NEC.
-
-Samples to be collected:
-•	From parent mice:
-o	Faecal matter (MAT and CON mice)
-•	From offspring post euthanasia
-o	1-2 cm of ileum immersion-fixed in paraformaldehyde for histological analysis
-o	1-2 cm of ileum snap-frozen for mesoscale + qPCR analysis
-o	Whole caecum for microbiota analysis
-
- Of these 32 survived and were included in the analysis. After fine-tuning the feeding method we we able to successfully feed all surviving (32) mice every 3 hours for 2 days (48 hours).
-(Conclusion) - Visual inspection of intestines during sampling concluded that we had not created a NEC phenotype in any treatment group.
-(BUT) - We found that AB-FF pups consistently looked different that CON-BF in the level of cytokine expression in the ileum, with statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
-- A multiple linear regression analysis, evaluating the effects of MAT (AB or CON) and type of feed (BF of FF) separately and interactively found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val$IL-6.
+- In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers. 
+- This is surprising given the findings of Chen et al., who used an identical protocol to our AB-BF vs.
+CON-BF (Maternal antibiotics treatment and weening with the mother for 7 days).
+They claim that MAT during pregnancy resulted in significant alterations in the intestinal microbiota.
+lower body weight at birth and on day 7, visible intestinal injury and inflammation and disruption of the intestinal barrier.
+- Our initial analysis of weight and cytokine expression did not indicate a similar effect of MAT (evaluations of gut permeability, intestinal gene expression, histological analysis and gut microbial composition were conducted in later experiments).
+- One possible explanation is an effect of dosage. While Chen et al., used a concentration of 1 g/l, we went with 0.5 g/l, based on experience from previous studies.
+- We therefore decided to increase the dosage to 1 g/l for subsequent experiments,
+and to include further measures of intestinal dysfunction.
+- We also decided to measure the birth weight of all pups, in addition to body weight at the beginning of the experiment (day 3) and end (day 5).
 
@@
 #### Rationale:
-*Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
-We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+- Exp 1 demonstrated that the type of feed (breastmilk or formula) had a greater effect on cytokine expression than maternal antibiotics treatment. 
+- Central to our hypothesis of NEC development is an intestinal microbiome dysibiosis, and a broad-spectrered antibiotic treatment potentially resulted in an overall depletion rather than the manifisation of a dysbiosis.
+- We therefore hypothesized that targeting specific strains of bacteria in the mother would lead to a more dysbiotic microbiome and allow for greater subsequent dysbiosis in the offspring.
 
 #### Aim: 
-- To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
-- Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
-- Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum
+- To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics, without formula feeding on intestinal permeability and cytokine expression. 
+- To see whether a larger dose of antibiotics (1 g/l) causes decreased body weight at birth and in day 5 compared with non-treated controls.
 
@@
 - Survival rate: xx of `r nrow(preexp_1)` (%)
-- Visaul inspection: no clear signs of intestinal inflammation
+- Visual inspection: no clear signs of intestinal inflammation
 - Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
@@
 #### Detailed description of Exp 2.
-Based on the results from Preexp 1 we knew that we could perform the experimental setup without too many casualties. 
-- We had seen that formula-feeding did affect the expression of cytokines, but visual inspection of the intestines showed that we had not managed to create a NEC-like phenotype. 
-- After considering the results we realized that our assumptions were based on a dysbiotics microbiome in the mother leading to less immunizing milk.
-However, we had likely not created a dysbiosis as much as an overall depletion by administering a broad-spectered antibiotic.
-We therefore hypothesized that targeting specific strains of bacteria would lead to a more dysbiotic microbiome and allow for greater disbiosis in the offspring.
 
-We tested 2 different types, all on a background of breastfeeding. 
-Contrary to our hypothesis, these mice had lower levels of intestinal permeability - both relative to controls and in absolute terms compared with FF-pups from PreExp 1.
+*Antimals and experimental setup*
+- The breeding setup in Exp 2 was identical to Exp 1. 
+- On gestational day 15 vancomycin, gentamicin or control (tab water) was added to the drinking water pregnant females, forming 3 experimental groups; VANCO, GENTA and CON.
+
+*Weight development*
+- We checked the cages twice daily from gestational day 18 and measured the weight of newly-born pups. 
+- VANCO and GENTA mothers gave birth on the 25th of february (getational day 20) with mean offspring body weights of `r mean(bodyweight_birth_vanco` (N = `r nrow(bodyweight_birth_vanco`) and `r mean(bodyweight_birth_genta` (N = `r nrow(bodyweight_birth_vanco`) respectively.
+- The CON mother gave birth on the 26th of february (gestational day 21) with mean offspring body weight of `r mean(bodyweight_birth_con` (N = `r nrow(bodyweight_birth_vanco`). 
+- This difference in body weight at birth was statistically significant (p = `r lm_weight_exp2$p.value`), but adjusting for length of gestation erased the statistical significance (p = `r lm_weight_exp2_adj$p.value`).
+- On day 3 and 5 this difference was in body weight was no longer detectable. *(Figure of body weights in exp 2)*
+
+*Intestinal permeability - FITC-Dextran administration* 
+- On the day of euthanasia pups were separated from the mothers at 07:00 (halting breastfeeding). 
+- At 10:00, 50 µL FITC-dextran (0.045 mg/µL) was administered orally using a PICC line and a 1 mL syringe for a final dose of 750 mg/kg. 
+- Euthanasia and sample collection commenced at noon.
+- Pups were euthanized by cervical dislocation and blood samples were collected and spun in Eppendorf tubes. 
+- FITC molecules were detectable in all groups.
+The highest and lowest concentration was found in the GENTA (`r fitc_exp2_genta` µg/ml) and VANCO (`r fitc_exp2_vanco` µg/ml) groups, respectively, with the CON group in between (`r fitc_exp2_vanco` µg/ml).
+This difference was statistically significant (p = `r anova_fitc_exp2`, one-way analysis of variance).
+
+*Cytokine expression*
+- Tissue weight-adjusted results of cytokine concentration showed significant differences between groups for `r signif_cyto_exp2`. 
+- A statistically significant difference in the average weight of the intestinal biopsy was found ( p = `r tissue_weight_exp2$p.value`), with XXXX having the heaviest (`r tissue_weight_exp2_`) sample weights and XXXX having the lowest.
+- Un-adjusted for tissue weight there were no significant differences between the treatment groups.
+
+#### Discussion of experimental results - Exp 2.
+- Directionality of change varied, with alternating over- and under-expression of various cytokines.
+As can be seen in [*Cytokine plot of signif results *] pups tended to group based on feeding-scheme, with both formula-fed groups being either suppressed og over-expressed together, compared with breastfed groups, irrespective of antibiotics treatment.
+- This observation is supported by a a multiple linear regression analysis, evaluating the effects of MAT (AB or CON) and type of feed (BF of FF) separately and interactively.
+The analysis found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val$IL-6.
+
+- In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers. 
+- This is surprising given the findings of Chen et al., who used an identical protocol to our AB-BF vs.
+CON-BF (Maternal antibiotics treatment and weening with the mother for 7 days).
+They claim that MAT during pregnancy resulted in significant alterations in the intestinal microbiota.
+lower body weight at birth and on day 7, visible intestinal injury and inflammation and disruption of the intestinal barrier.
+- Our initial analysis of weight and cytokine expression did not indicate a similar effect of MAT (evaluations of gut permeability, intestinal gene expression, histological analysis and gut microbial composition were conducted in later experiments).
+- One possible explanation is an effect of dosage. While Chen et al., used a concentration of 1 g/l, we went with 0.5 g/l, based on experience from previous studies.
+- We therefore decided to increase the dosage to 1 g/l for subsequent experiments,
+and to include further measures of intestinal dysfunction.
+- We also decided to measure the birth weight of all pups, in addition to body weight at the beginning of the experiment (day 3) and end (day 5).
 

### 2022-10-27 submission/manuscript.Rmd
Updated manuscript

@@
 #### Discussion of experimental results - Exp 2.
-- Directionality of change varied, with alternating over- and under-expression of various cytokines.
-As can be seen in [*Cytokine plot of signif results *] pups tended to group based on feeding-scheme, with both formula-fed groups being either suppressed og over-expressed together, compared with breastfed groups, irrespective of antibiotics treatment.
-- This observation is supported by a a multiple linear regression analysis, evaluating the effects of MAT (AB or CON) and type of feed (BF of FF) separately and interactively.
-The analysis found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val$IL-6.
+- Weight development similar to Chen. 
+- Whether it was a result of growth restriction or shorter gestational age is difficult to determine. 
+- Chen et al. did not report whether dams were from the same day. 
+- Intestinal permeability response was different depending of type of antibiotics. 
+- Vancomycin decreased intestinal permeability whereas gentamicin increased it.
+  - This has been shown in Winstar rats before (Thulstrup 2015). 
+- Cytokine expression was most affected in the VANCO group.
 
-- In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers. 
-- This is surprising given the findings of Chen et al., who used an identical protocol to our AB-BF vs.
-CON-BF (Maternal antibiotics treatment and weening with the mother for 7 days).
-They claim that MAT during pregnancy resulted in significant alterations in the intestinal microbiota.
-lower body weight at birth and on day 7, visible intestinal injury and inflammation and disruption of the intestinal barrier.
-- Our initial analysis of weight and cytokine expression did not indicate a similar effect of MAT (evaluations of gut permeability, intestinal gene expression, histological analysis and gut microbial composition were conducted in later experiments).
-- One possible explanation is an effect of dosage. While Chen et al., used a concentration of 1 g/l, we went with 0.5 g/l, based on experience from previous studies.
-- We therefore decided to increase the dosage to 1 g/l for subsequent experiments,
-and to include further measures of intestinal dysfunction.
-- We also decided to measure the birth weight of all pups, in addition to body weight at the beginning of the experiment (day 3) and end (day 5).
+- In summary:
 
@@
 #### Rationale:
-*Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
-We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+Based on the results from Exp 2 it was difficult to ascertain which treatment was more likely to manifest in NEC like phenotype. 
+- Gentamicin had increased intestinal permeability the most. However, the greatest effect of cytokine expression was found with the vancomycin treatment. 
+- In addition to this divergence, the interaction effect found in Exp 1 suggested it was likely that combining MAT with formula feeding would yield different results. 
+- Therefore, we decided to increase the breeding population to allow for an experiment that combined both types of antibiotics with either formula-feeding or breastfeeding, for a total of 6 experimental groups.
 
 #### Aim:
-- To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
-- Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
-- Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum
 
 #### Conclusion:
-- Survival rate: xx of `r nrow(preexp_1)` (%)
-- Visaul inspection: no clear signs of intestinal inflammation
-- Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
 
 #### Detailed description of Exp 3.
-- In round 3 we repeated the experimental setup with regards to MAT, adding either Vancomycin, Gentamicin or sterile water to the drinking water of pregnant "females" (use same word throughout(mothers)).
+- In Exp 3 we repeated the experimental setup with regards to MAT, adding either Vancomycin, Gentamicin or sterile water to the drinking water of pregnant femailes.
 We repeated the group-setup from PreExp  1 and  separated pups into breastfeeding or formula feeding, for a total of 6 groups (see overview table).

### 2022-11-01 submission/manuscript.Rmd
progressed on manuscript

@@
 The main goal of this thesis project was to create a NEC mouse model by combining Maternal Antibiotics Treatment with Formula-Feeding.
-This hypothesis was tested in a series of experiments (Exp 1-4), each successive experiment building on the results of the former.
+This hypothesis was tested in a series of experiments (Exp 1-4), each successive experiment building on the findings from the former.
 The following is a description of each experiment. An abridged version of each experiment is provided, followed by a detailed  description of experimental setup and results.
@@
 #### Rationale:
-*Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
-We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+- *Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring. 
+- XXX *[Find paper]* found formula-feeding to be damaging to the intestine. 
+- One of the primary causal factors of NEC is formula-feeding (paper on what is thought to cause NEC).
+- We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
 
@@
 *Antimals and experimental setup* See "Animal experimental setup" in methods.
-- The experimental setup has been described in detail in the methods seciton. 
+- The experimental setup has been described in detail in the methods section. 
 - Detials specific for Exp-1 are outlined below. 
-- Briefly, six male and 18 female C57BL/6N mice were co housed for 24 hours in 6 cages (1 male + 3 female each).
+- Briefly, six male and 18 female C57BL/6N mice were co-housed for 24 hours in 6 cages (1 male + 3 female each).
 - Two weeks after mating, pregnant females were separated evenly into maternal antibiotic treatment (MAT) and water-only controls (CON). 
-- Antibiotics (Gentamicin, Vancomycin, Neomycin, and Ampicillin (all 0.5 g/L)) was mixed into the drinking water of the MAT mice while controls received water only.
+- Antibiotics (Gentamicin, Vancomycin, Neomycin, and Ampicillin (all 0.5 g/L)) were mixed into the drinking water of the MAT mice while controls received water only.
 - Antibiotic treatment commenced on gestational day 15 and continued until delivery. 
 - Each mother was housed with her offspring in separate cages during the first 3 days after birth. 
-- On the 4th day, `r nrow(preexp_1)` mice, XXX (Filter `r nrow(preexp_1)` for feed type) from antibiotics treated mothers and XXX (Filter `r nrow(preexp_1)` for feed type) were divided into four experimental groups: MAT + formula-feeding (AB-FF) MAT + breastfeeding (AB-BF), Water + formula-feeding (CON-FF), and CON + breastfeeding (CON-BF).
+- On day 4, `r nrow(preexp_1)` mice, XXX (Filter `r nrow(preexp_1)` for feed type) from antibiotics treated mothers and XXX (Filter `r nrow(preexp_1)` for feed type) were divided into four experimental groups: MAT + formula-feeding (AB-FF) MAT + breastfeeding (AB-BF), Water + formula-feeding (CON-FF), and CON + breastfeeding (CON-BF).
 - Formula-feeding was performed as per METHODS AND MATERIALs.
-In Exp-1 we administered  0.08 ml formula for pups >2g, and 0.06 for pups ≤ 2g every 3 hours throughout the study.
-In later experiments this dose was changed, and feeding intervals were inreased to every 4 hours from midnight til 08 AM.
+In Exp-1 we administered  0.08 ml formula for pups >2g, and 0.06 for pups ≤ 2g every 3 hours throughout the entire study.
+In later experiments this dose was changed, and feeding intervals were increased to every 4 hours between midnight and 08 AM.
 
-- `r nrow(preexp_1_survived)` of `r nrow(preexp_1)` pups survived until sampling.
+- `r nrow(preexp_1_survived)` of `r nrow(preexp_1)` pups survived the treatments.
 - Those that did not survive were euthanize because of esophageal perforation during feeding. 
-After fine-tuning of the feeding methods no pups were eithanized as a result of reaching humane endpoints, and the pups generally were not very sick.
-- While significantly less than their breastfed controls, all pups increased their heavyweight during the experiment, attesting to their generally reasonable health status.
-- The pups were physically active and reacted to changes in light and to physical sitmuli.
-- Skin color remained bright purple and fur development was visible thoughout the experiment. 
+- After fine-tuning of the feeding methods we were able to feed the pups without further harm.
+- We never had to euthanize a pup that reached our humane endpoints, since all the surviving pups were in generally good shape. 
+- While significantly lighter than their breastfed controls, all pups increased their bodyweight during the experiment, attesting to the efficacy of our feeding regiment. 
+- The pups were physically active and responded to changes in light and physical stimuli.
+- Skin color remained bright purple and fur development was visible throughout the experiment. 
 - No signs of intestinal inflammation were evident from the visual inspection of intestines during sampling.
@@
 - Analysis of pro-inflammatory cytokines showed statistically significant changes in the expression of STATISTICAL-SIGNIF-CYTOKINE-NAMES. 
-
-#### Discussion of experimental results - Exp 1.
 - Directionality of change varied, with alternating over- and under-expression of various cytokines.
@@
 
+#### Discussion of experimental results - Exp 1.
 - In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers. 
-- This is surprising given the findings of Chen et al., who used an identical protocol to our AB-BF vs.
-CON-BF (Maternal antibiotics treatment and weening with the mother for 7 days).
-They claim that MAT during pregnancy resulted in significant alterations in the intestinal microbiota.
+- This is surprising given the findings of Chen et al., who used an identical protocol to our AB-BF vs. CON-BF (Maternal antibiotics treatment and weening with the mother for 7 days).
+- They claim that MAT during pregnancy resulted in significant alterations in the intestinal microbiota.
 lower body weight at birth and on day 7, visible intestinal injury and inflammation and disruption of the intestinal barrier.
-- Our initial analysis of weight and cytokine expression did not indicate a similar effect of MAT (evaluations of gut permeability, intestinal gene expression, histological analysis and gut microbial composition were conducted in later experiments).
-- One possible explanation is an effect of dosage. While Chen et al., used a concentration of 1 g/l, we went with 0.5 g/l, based on experience from previous studies.
-- We therefore decided to increase the dosage to 1 g/l for subsequent experiments,
-and to include further measures of intestinal dysfunction.
+- Our initial analysis of weight and cytokine expression did not indicate a similar effect of MAT (evaluations of gut permeability, intestinal gene expression, histological analysis and gut microbial composition are conducted in later experiments).
+- One possible explanation is an effect of dose.
+While Chen et al., used a concentration of 1 g/l, we, unfortunately, tested 0.5 g/l, based on experience from previous studies, making a direct comparison difficult.
+- We did find significant over-expression of cytokines relevant in inflammation, indicating some effect of the treatments. 
+- Since these biochemical signals were not strong enough to manifest phenotypically, we reasoned that our treatment wasn't sufficiently injurious.
+- We therefore decided to increase the dosage to 1 g/l for subsequent experiments, as well as to increase the osmolality of the formula for the next feeding experiment (Exp 3).
+- With assurance in our experimental setup we decided to include further measures of intestinal dysfunction. 
 - We also decided to measure the birth weight of all pups, in addition to body weight at the beginning of the experiment (day 3) and end (day 5).
@@
 - Central to our hypothesis of NEC development is an intestinal microbiome dysibiosis, and a broad-spectrered antibiotic treatment potentially resulted in an overall depletion rather than the manifisation of a dysbiosis.
-- We therefore hypothesized that targeting specific strains of bacteria in the mother would lead to a more dysbiotic microbiome and allow for greater subsequent dysbiosis in the offspring.
+- We hypothesized that targeting specific strains of bacteria in the mother would lead to a more dysbiotic microbiome and allow for greater subsequent dysbiosis in the offspring.
 
@@
 - To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics, without formula feeding on intestinal permeability and cytokine expression. 
-- To see whether a larger dose of antibiotics (1 g/l) causes decreased body weight at birth and in day 5 compared with non-treated controls.
+- To see whether a larger dose of antibiotics (1 g/l) causes decreased body weight at birth and at day 5 compared with non-treated controls.
 
 #### Conclusion:
-- Survival rate: xx of `r nrow(preexp_1)` (%)
+- Birth weight: `r mean(bodyweight_birth_con` g (CON) vs `r mean(bodyweight_birth_mat` g (MAT)
 - Visual inspection: no clear signs of intestinal inflammation
@@
 - On the day of euthanasia pups were separated from the mothers at 07:00 (halting breastfeeding). 
-- At 10:00, 50 µL FITC-dextran (0.045 mg/µL) was administered orally using a PICC line and a 1 mL syringe for a final dose of 750 mg/kg. 
-- Euthanasia and sample collection commenced at noon.
-- Pups were euthanized by cervical dislocation and blood samples were collected and spun in Eppendorf tubes. 
+- At 10:00, 50 µL FITC-dextran (0.045 mg/µL),for a final dose of 750 mg/kg, was administered orally using a PICC line and a 1 mL syringe.
+- Two hours later euthanasia and sample collection began (12:00).
+- Pups were euthanized by cervical dislocation and blood samples collected and spun in Eppendorf tubes. 
 - FITC molecules were detectable in all groups.
-The highest and lowest concentration was found in the GENTA (`r fitc_exp2_genta` µg/ml) and VANCO (`r fitc_exp2_vanco` µg/ml) groups, respectively, with the CON group in between (`r fitc_exp2_vanco` µg/ml).
-This difference was statistically significant (p = `r anova_fitc_exp2`, one-way analysis of variance).
+- The highest concentration was found in the GENTA group (`r fitc_exp2_genta` µg/ml) and the lowest in the VANCO group (`r fitc_exp2_vanco` µg/ml), respectively, with the CON group in between (`r fitc_exp2_vanco` µg/ml).
+- This difference was statistically significant (p = `r anova_fitc_exp2`, one-way analysis of variance). *(Figure of FITC in exp 2)*
 
 *Cytokine expression*
-- Tissue weight-adjusted results of cytokine concentration showed significant differences between groups for `r signif_cyto_exp2`. 
+- Tissue weight-adjusted results of cytokine concentrations showed significant differences between the 3 groups for `r signif_cyto_exp2`. ### Rerun these experiments.
 - A statistically significant difference in the average weight of the intestinal biopsy was found ( p = `r tissue_weight_exp2$p.value`), with XXXX having the heaviest (`r tissue_weight_exp2_`) sample weights and XXXX having the lowest.
@@
 #### Discussion of experimental results - Exp 2.
-- Weight development similar to Chen. 
+- Weight development was now similar to the findings of Chen et al.
 - Whether it was a result of growth restriction or shorter gestational age is difficult to determine. 
@@
 - In summary: 
+- In experiment 2, Maternal antibiotics treatment did not have an effect.
+- This indicates that the greater effect is of formula-feeding, similar to the findings in exp 1. 
+- The intestinal permeability assay is interesting in that it supports the findings of Thulstrup et al. 2015.
 
@@
 Based on the results from Exp 2 it was difficult to ascertain which treatment was more likely to manifest in NEC like phenotype. 
-- Gentamicin had increased intestinal permeability the most. However, the greatest effect of cytokine expression was found with the vancomycin treatment. 
+- Gentamicin had increased intestinal permeability the most. 
+- However, the greatest effect of cytokine expression was found with the vancomycin treatment. 
 - In addition to this divergence, the interaction effect found in Exp 1 suggested it was likely that combining MAT with formula feeding would yield different results. 
@@
 #### Aim: 
+- Repeat Exp 2 with greater sample sizes, adding the effect of formula feeding. 
+- To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics in combination with formula feeding on intestinal permeability and cytokine expression. 
+-
 
 #### Conclusion:
+- Combining MAT with formula feeding resulted in more pronounced signs of inflammation than MAT alone.
+- VANCO-FF and VANCO-BF had the highest and lowest intestinal permeability scores, respectively.
+- Significantly altered expression of  XXSignificantSCYTOKINESXX. 
+- Greatest determinant of variation was formula-feeding.
 
 #### Detailed description of Exp 3.
-- In Exp 3 we repeated the experimental setup with regards to MAT, adding either Vancomycin, Gentamicin or sterile water to the drinking water of pregnant femailes.
-We repeated the group-setup from PreExp  1 and  separated pups into breastfeeding or formula feeding, for a total of 6 groups (see overview table).
-- All pups survived the experiment and were included in the final analysis. 
-- 2 hours after FITC administtion pups were euthanized by decapitation and blood collected in eppendorfs for analysis of serum FITC concentration.
-Intestinal samples were collected for gene expression assays, histological assessment, pro-inflammatory cytokine analysis and gut microbiome sequencing.
+*Antimals and experimental setup*
+- To increase the number of pups available we doubled the number of mice for breeding, 36 females and 12 males. 
+- Pregnant females were again given either Vancomycin, Gentamicin (1 g/l) or sterile water on gestational day 15.
+- `r nrow(preexp_3` pups were born and separated into breastfeeding or formula feeding, for a total of 6 groups:
+- VANCO-BF, VANCO-FF, GENTA-BF, GENTA-FF, CON-BF and CON-FF.
 
-* Gut permeability: 
+*Intestinal permeability - FITC-Dextran administration* 
+- We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
 - Formula-fed pups consistently showed higher levels of Serum FITC concentration compared to breastfed pups, irrespective of MAT status.
-VANCO-FF and VANCO-BF had the highest and lowest levels of FITC, respectively.
-- VANCO-BF was the group with the lowest FITC in PreExp 2.
-- This is indicative of an interaction-effect with vancomycin and formula feeding. (**FIGURE**).
+- VANCO-FF and VANCO-BF had the highest and lowest levels of FITC, respectively.
 
-* Cytokine expression
+*Cytokine expression*
 - We found significant differences in concentrations of XXSignificantSCYTOKINESXX.
-Linear regression analysis indicated the biggest detminant of differneces in outcomes was formula feeding, with the type of antibiotic being of lesser importance compared.
-Directionally the trend was the same for all cytokines.
+Linear regression analysis indicated the biggest determinant of differences in outcomes was formula feeding, with the type of antibiotic being of lesser importance compared.
+- Directionally the trend was the same for all cytokines. 
+- IL-1b: Non-significant decrease in all antibiotics-treated offspring, irrespective of feeding.
 - IL-2: significantly lower concentrations in all formula-fed pups compared with CON-BF.
@@
 
-### Experiment 4
+#### Discussion of experimental results Exp 3:
+- Intestinal permeability was by far the greatest in VANCO-FF, with levels of FITC-dex close to 10-fold increased in this group. 
+- In contrast, VANCO-BF was the group with the lowest measure of intestinal permeability, similar to the finding in exp 2.
+- Looking at **FIGURE CYTOKINES ROUND 3** shows how formula-fed groups systematically over- or underexpress compared to breastfed analimals, irrespective of type of antibiotic, indicating that the biggest determinant of cytokine variation was again formula feeding.
+
+*Adjusting for weight of tissue*
+- In the previous experiments, cytokine concentrations were adjusted for the weight of the tissue analyzed, to control for biopsy sizes. 
+- We did so because the average weight of tissue samples was not meaningfully different between groups. 
+- This assumption wasn't met in Exp 3, where samples from formula-fed pups weighted on average XX%XX more (p = XXp.valueXX) than breastfed controls. 
+- Vanco-FF had the heaviest ileum tissue samples of all **mean + P (if its true)**.
+- Tissue edema could well be indicative of inflammation, and hence desirable, however it introduces a bias against the formula-fed pups if cytokine concentrations were normalized to weight. 
+- The reported findings are no longer significant **Table of significant cytokines** when adjusting for weight.
+
+- This phenomenon is a common challenge in animal disease models, requiring measure of normalization other that tissue weight, such as total protein content. 
+- Unfortunately this discovery was done after tissue homogenization and analysis, making it too late for further analysis.
+
+*Combining results from exp 2 and exp 3*
+- The protocols for the breastfed groups were identical across experiment 2 and 3, allowing for pooling of these results.
+- While analyzing the results from both groups together increased the group sizes and degrees of statistical freedom, it did not change the clinical conclusions.
+
+### Experiment 4 - FVT Trial
 
 #### Rationale:
-*Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
-We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+
+- Central to the development of NEC is an assumption of a gut microbial dysbiosis, which we tried to mimick by administering antibiotics prenatally. 
+- Results from exp 1-3 indicated that maternal antibiotics treatment had little effect in the context of breastfeeding, but when combined with formula-feeding was able to induce mild symptoms of intestinal inflammation.
+- It was unclear whether this was solely an effect of the formula, or whether an interaction with the maternal antibiotics treatment existed. 
+- If our findings were partly due to a successful manipulation of the gut microbiome, then eliminating the induced disbyosis should also partly mitigate the inflammatory signatures.
+- Relative bacterial abundance / bacterial composition is partly a function of viral (bacteriophage) abundance, and sevaral studies have shown how a FVT can alter the bacterial composition. 
+- We therefore decided to conduct a final experiment, in which we administered an FVT to a vancomycing-treated, formula-fed group (since this combination (VANCO-FF) had the biggest effects in terms of intestinal permeability and proinflammatory cytokine expression in Exp 3).
+- We hypothesized that any dysbiosis created by maternal antibiotics treatment could be mitigated with a follow-up administration of FVT prior to beginnig formula-feeding. 
+- If successful, the FVT-treated group would show fever intestinal injury scores compared with a non-FVT treated group. 
+- Previous experiments had shown that MAT in combination with breastfeeding had no effect on inflammatory markers and VANCO-BF thus acted as a healthy control to both formula-fed treatment groups.
 
 #### Aim: 
-- To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
-- Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
-- Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum
+- To test whether signs of inflammation are reduced in pups treated with an FVT.
+- To test whether any reductions in symptoms can be explained with a manipulation of the microbial composition.
 
 #### Conclusion:
-- Survival rate: xx of `r nrow(preexp_1)` (%)
-- Visaul inspection: no clear signs of intestinal inflammation
-- Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
 
 #### Detailed description of Exp 4.
-- We treated XX mothers with Vancomycin (1 g/l) from pregnancy-day 15 until delivery, where vancomycin-mixed drinking water was replaced with regular water.
+- We treated XX mothers with Vancomycin (1 g/l) from pregnancy-day 15 until delivery.
 - 2 days after birth XX of the 26 pups were separated from the mother and received XX µl FvT solution or an equal amount of SM buffer (control treatment).
-The remaning xx of 26 pups remained with the mother for breastfeeding thoughout the study.
+The remaning xx of 26 pups remained with the mother for breastfeeding throughout the study.
 - Pups separated from their mother were formula fed (FF) with a solution of XXXXX milk formulation (see materials) every 3 hours for 48 hours. (futher setup description see materials and methods).
@@
 
-* Gut permeability: 
-- We found the lowest concentration of serum FITC Dextran in the group treated with FF and Fvt (XXXX Xg/mL).
-This was lower than even the breastfed controls, which had the highest concentration of the three groups (**FIGURE**).
-- The VANCO-FF (Fvt) compared with VANCO-FF from PreExp 3 this change was significant/non-significant. 
-- It is reasonable/unreasonable to compare the rounds, given the similarity/lack of similarity in response of the VANCO-BF groups from Exp3(XXX) and NEC-FvT (XXXX)
+*Gut permeability* 
+- We found the lowest concentration of serum FITC Dextran in the group treated with FVT (XXXX Xg/mL).
+- This was lower than even the breastfed controls, which had the highest concentration of the three groups (**FIGURE**).
 
-* Cytokine expression
-- Only TNFa was significantly different in Exp 4, with VANCO-FF (FvT) having the highest level (XXX), signifiacntly greater than both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively). 
-- The VANCO-FF (FvT) had lower/higer absolute levels of cytokine concentration compared to both VANCO-BF and VANCO-FF from Exp3. 
-- THis comparison is reasonable/unreasonable given the similarity/lack of similarity in response of the VANCO-BF groups from Exp3(XXX) and VANCO-FF (SM) from Exp 4(XXXX).
+*Cytokine expression*
+- Only TNFa was significantly different in Exp 4, with greater level in the FVT-treated group (XXX), compared to both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively).
+
+*Gene expression*
+- We standardized the gene expression to the control group and the house gene. 
+- However, we found statistically significant difference in expression of the house-gene, making the delta-delta measure bad.
+
+#### Discussion of results.
+- FVT and SM showed lower intestinal permeability than breastfed controls.
+- It is possible that something went wrong with the VANCO-BF group, since their the FITC-concentration was `r mean(fitc_level_exp4$bf`, which is 4-5 times higher than VANCO-BF groups on identical protocols in Exp 2 `r mean(fitc_level_exp2$bf` and 3 `r mean(fitc_level_exp3$bf`.
+Compared with theses levels the development looks more biologically reasonable.
+**FIGURE of EXP4 with DIFFERENT CONTROL GROUP**
+- It is also possible that the assay doesn't work.
+
+- Summary
+-
 
@@
 
-* Gene expression
-
-We standardized the gene expression to the control group and the house gene. However, we found statistically significant difference in expression of the house-gene, making the delta-delta measure bad.
-
 * Gut microbiome

### 2022-11-03 submission/manuscript.Rmd
Progressed on manuscript

@@
 
-#### Discussion of experimental results - Exp 1.
+#### Commentary and interpretation of results - Exp 1.
 - In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers. 
@@
 - Tissue weight-adjusted results of cytokine concentrations showed significant differences between the 3 groups for `r signif_cyto_exp2`. ### Rerun these experiments.
-- A statistically significant difference in the average weight of the intestinal biopsy was found ( p = `r tissue_weight_exp2$p.value`), with XXXX having the heaviest (`r tissue_weight_exp2_`) sample weights and XXXX having the lowest.
-- Un-adjusted for tissue weight there were no significant differences between the treatment groups.
 
-#### Discussion of experimental results - Exp 2.
+#### Commentary and interpretation experimental results - Exp 2.
 - Weight development was now similar to the findings of Chen et al.
@@
 
-#### Discussion of experimental results Exp 3:
+#### Commentary and interpretation experimental results Exp 3:
 - Intestinal permeability was by far the greatest in VANCO-FF, with levels of FITC-dex close to 10-fold increased in this group. 
@@
 
-#### Rationale:
+#### Rationale: **What was the actual rationale?**
 
@@
 - Previous experiments had shown that MAT in combination with breastfeeding had no effect on inflammatory markers and VANCO-BF thus acted as a healthy control to both formula-fed treatment groups.
+ **(The problem here is that formula is the only thing with an effect. A breastfed control only shows the effect of the formula)**
 
@@
 #### Conclusion:
+- FVT treated pups showed similar intestinal permeability and pro-inflammatory cytokine concentrations as non-FVT-treated controls. 
+- An `r fold_increase_reg3g`-fold increase in expression of the `r REG3G` gene was found in the FVT group.
 
 #### Detailed description of Exp 4.
-- We treated XX mothers with Vancomycin (1 g/l) from pregnancy-day 15 until delivery.
-- 2 days after birth XX of the 26 pups were separated from the mother and received XX µl FvT solution or an equal amount of SM buffer (control treatment).
-The remaning xx of 26 pups remained with the mother for breastfeeding throughout the study.
-- Pups separated from their mother were formula fed (FF) with a solution of XXXXX milk formulation (see materials) every 3 hours for 48 hours. (futher setup description see materials and methods).
-- All 26 pups survived the 48 hour study. On the final day of the study pups fasted for XX hours before given xx µl FITC Dextran (as per materials protocol). 
+- For the final experiment we repeated the previous breeding setup, assigning 36 females and 12 males for breeding. 
+- Only 5 of the 36 females became pregnant, and all were assigned to vancomycin (1 g/l) from pregnancy-day 15 until delivery, giving birth to `r nrow(preexp_4` pups on Saturday the XXth.
+- 2 days after birth pups were separated into breastfeeding or formula feeding. 
+- The formula-fed pups were further divided into a treatment and control group, receiving either 30 µl fecal virome filtrate (FVT) (0.01 mg farces/µL) or an equal amount of SM buffer (control) 1 hour prior to the first feeding.
+- We repeated the FVT administration 24 hours later (09:00 day 2) for a total dose of 0.6 mg feces material per pup.
+
 - 2 hours after FITC administtion pups were euthanized by decapitation and blood collected in eppendorfs for analysis of serum FITC concentration.
@@
 
-*Gut permeability* 
-- We found the lowest concentration of serum FITC Dextran in the group treated with FVT (XXXX Xg/mL).
-- This was lower than even the breastfed controls, which had the highest concentration of the three groups (**FIGURE**).
+*Intestinal permeability - FITC-Dextran administration* 
+- We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
+- The FVT and FVT-control (SM) had similar permeability scores. 
+- Compared with the vancomycin-treated formula-fed pups from experiment 3, there were no significant differences **Figure X**. 
+- The breastfed pups did, however, have significantly higher levels compared with all other breastfed pups.
 
 *Cytokine expression*
-- Only TNFa was significantly different in Exp 4, with greater level in the FVT-treated group (XXX), compared to both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively).
+- We found significant differences in concentrations of TNFa in Exp 4, with greater level in the FVT-treated group (XXX), compared to both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively). 
+- This is in contrast experiment 3, where differences between VANCO-BF and VANCO-FF found significant differences for XXXCYtokinesXXX.
+OR: - This is similar to the findings of experiment 3, where no differences between VANCO-BF and VANCO-FF were found either.
 
 *Gene expression*
-- We standardized the gene expression to the control group and the house gene. 
-- However, we found statistically significant difference in expression of the house-gene, making the delta-delta measure bad.
+- qPCR analyzis of intestinal samples found statistically significant differences in the level of XXXSIGNIFICANT_GENESXXX expression between groups with the delta-delta method (reformulate).
+- However, a difference in expression of the reference gene between the FVT and breastfed control was observed.
+this difference was statistically significant, invalidating the assumption of *equal baseline expression.*(reformulate)
+- Comparing the groups on the basis of (SOMEHTING ELSE THAN DELTA DELTA) found a `r fold_increase_reg3g`-fold increased expression of the `r REG3G` gene in the FVT groups.
 
-#### Discussion of results.
-- FVT and SM showed lower intestinal permeability than breastfed controls.
-- It is possible that something went wrong with the VANCO-BF group, since their the FITC-concentration was `r mean(fitc_level_exp4$bf`, which is 4-5 times higher than VANCO-BF groups on identical protocols in Exp 2 `r mean(fitc_level_exp2$bf` and 3 `r mean(fitc_level_exp3$bf`.
+#### Commentary and interpretation
+- FVT-treatmed did not seem to change the intestinal permeability. 
+- It is possible that something went wrong with the VANCO-BF group, given FITC-concentration was `r mean(fitc_level_exp4$bf`, which is 4-5 times higher than VANCO-BF groups on identical protocols in Exp 2 `r mean(fitc_level_exp2$bf` and 3 `r mean(fitc_level_exp3$bf`.
 Compared with theses levels the development looks more biologically reasonable.
 **FIGURE of EXP4 with DIFFERENT CONTROL GROUP**
-- It is also possible that the assay doesn't work.
+- The level of pro-inflammatory cytokines were the same in the FVT and SM groups, again indicating a lack of efficacy of the FVT treatment.
 
-- Summary
--
+- Summary: The FVT treatment did not seem to have an effect.
 
 ### Post experimental analyses and Discussion of results
-* Histological analysis - Mention and Discuss this after finishing the result section. 
-- Histological image analysis was performed after all trials had finished. 
-- We saw no degradation of mucin layer thickness in any of the pups. This supports the visial inspections of intestines that was made during sampling. 
-- We found no significant changes in the mucin layer between the treatment group (Fvt) Fvt-Control (SM) or formula-feed control (VANCO-BF). The thickness of the stain was the same for all groups.
 
-* Gut microbiome
+- Histological image analysis was performed after all experiments had finished. 
+- We inspected the morphology of the ileum and colon, and visually assessed thickness of the mucin layers.
+- Analyzing AB-PAS stains for mucin-stained tissue to the total area, showed no differences between any treatment groups. 
+- Formula-fed and breastfed pups were slightly different, irrespective of antibiotics treatment, whereas there were no observable differences between antibiotic treatments. 
+- These findings supports the visual inspections of intestines made during sampling, where no signs of inflammation were detected.
+- *Images of intestines*
+
+* Gut microbiome composition
 

### 2022-11-04 code/Result_section.Rmd
Renamed manuscript file

@@
 date: "18/10/2022"
-output: html_document
+output: pdf_document
 editor_options: 
@@
 #library(AICcmodavg)
-library(visdat)
+#library(visdat)
 library(rstatix)
@@
 library(patchwork)
-library(jtools)
+#library(jtools)
 ```
@@
 
-```{r include=FALSE}
+```{r exp1-weight, include=FALSE}
 preexp_1 <- read_excel("data/processed/weight_development.xlsx") %>% 
@@
 #__________Load in cytokine data for all experiments________
-cytokines_1 <- read_excel(path = "data/processed/cytokines.xlsx") %>%  
-  filter(exp_number == 1) %>% 
-  filter(tissue == "ileum") %>% 
+cytokines <- read_excel(path = "data/processed/cytokines.xlsx") %>% 
   mutate(
@@
       ))
+
+cytokines_1 <- cytokines %>%  
+  filter(exp_number == 1) %>% 
+  filter(tissue == "ileum")
+
 #___________________________________________________________
@@
   filter(!cytokine_concentration == 0) %>% 
-  mutate(conc_adjusted4weight = cytokine_concentration/tissue_weight_mg)
+  mutate(cytokine_conc_adjust_4_weight = cytokine_concentration/tissue_weight_mg)
 #___________________________________________________________
@@
   group_by(cytokines) %>% 
-  shapiro_test(conc_adjusted4weight) %>% 
+  shapiro_test(cytokine_conc_adjust_4_weight) %>% 
   select(cytokines, p)
@@
   ggplot(aes(x = group, 
-             y = log10(conc_adjusted4weight),
+             y = log10(cytokine_conc_adjust_4_weight),
              color = group)) + 
@@
 ggplot(aes(x = group, 
-           y = conc_adjusted4weight, 
+           y = cytokine_conc_adjust_4_weight, 
            fill = group)) + 

### 2022-11-04 submission/results.Rmd
Renamed manuscript file

--- 2022-11-03 submission/manuscript.Rmd
@@
 ---
-title: "Results section"
-author: "Kaare"
-date: "`r format(Sys.time(), '%d %B, %Y')`"
-output: html_document
+title: ''
+output: pdf_document
 editor_options: 
   chunk_output_type: console
+  markdown: 
+    wrap: sentence
 ---
 
-# Code
+<!--# Code -->
 
-# Manuscript
+<!--# Run scripts for plots on another file and save plots on img folder. Then use include_graphics to include plots where relevant -->
 
-## Introduction
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE)
+```
 
-- NEC is caused by bacterial dysbiosis - we think. 
-- (AND) high-osmolar formula feed is hard on intestinal lining. 
-- (AND) Chen et al. saw NEC-like symptoms with MAT.
-- (BUT) No enough on its own plus too old mice. 
-- (Therefore) We want to see if MAT + FF will result in more severe NEC-like symptoms. (AIM)
-- (AND) FvT are currently being studied as potential remidies to NEC - thought to be mediated
-via stabilizing effect on microbiome. 
-- (Therefore) We want to see if A) we manage to induce NEC, are we able to "cure" it, or B) in 
-case of no NEC, is FvT on it's own safe or associated with any adverse effects?
+```{r loading libraries, message=FALSE, warning=FALSE}
+library(readxl)
+#library(tidyverse)
+library(broom)
+library(ggpubr)
+#library(AICcmodavg)
+#library(visdat)
+#library(rstatix)
+#library(gt)
+#library(gtsummary)
+library(patchwork)
+#library(jtools)
+```
 
-## Background
+```{r exp1-weight, include=FALSE}
+#____Load overview of sample IDs______ 
+pups <- read_excel("data/processed/weight_development.xlsx")
 
-- Why we think MAT leads to intestinal dysbiosis?
-    - Less immunizing milk?
--
+#____ Separate into experimental rounds ____#
+preexp_1 <- pups %>% 
+  filter(exp_number == "1")
 
-## Results
+exp1_survived <- preexp_1 %>% 
+  filter(!is.na(bodyweight_baseline_plus_48h)) %>% 
+  nrow()
+#____ ____#
+```
+
+<!-- Manuscript -->
 
@@
 This hypothesis was tested in a series of experiments (Exp 1-4), each successive experiment building on the findings from the former.
-The following is a description of each experiment. An abridged version of each experiment is provided, followed by a detailed  description of experimental setup and results.
+The following is a description of each experiment.
+An abridged version of each experiment is provided, followed by a detailed description of experimental setup and results.
 
@@
 #### Rationale:
+
 -   *Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
@@
 #### Aim:
+
 -   To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
@@
 #### Conclusion:
-- Survival rate: xx of `r nrow(preexp_1)` (%)
+
+-   Survival rate: `r`exp1_survived`of`r nrow(preexp_1)`(`r (exp1_survived / nrow(preexp_1))\*100\`%)
 -   Visual inspection: no clear signs of intestinal inflammation
@@
 #### Detailed description of Exp 1. **Date of exp**
+
 *Antimals and experimental setup* See "Animal experimental setup" in methods.
@@
 - Antibiotic treatment commenced on gestational day 15 and continued until delivery.
-- Each mother was housed with her offspring in separate cages during the first 3 days after birth. 
-- On day 4, `r nrow(preexp_1)` mice, XXX (Filter `r nrow(preexp_1)` for feed type) from antibiotics treated mothers and XXX (Filter `r nrow(preexp_1)` for feed type) were divided into four experimental groups: MAT + formula-feeding (AB-FF) MAT + breastfeeding (AB-BF), Water + formula-feeding (CON-FF), and CON + breastfeeding (CON-BF).
+- Each mother was housed with her offspring in separate cages during the first 2 days after birth.
+
+```{r exp1-groups}
+#____Separate into groups ______ 
+AB_FF <- nrow(preexp_1 %>% 
+  filter(group2 == "AB-FF"))
+
+AB_BF <- nrow(preexp_1 %>% 
+  filter(group2 == "AB-BF"))
+
+CON_FF <- nrow(preexp_1 %>% 
+  filter(group2 == "CON-FF"))
+
+CON_BF <- nrow(preexp_1 %>% 
+  filter(group2 == "CON-BF")) 
+#____ ____#
+
+```
+
+-   On day 3, mice from water or antibiotics treated mothers were separated and assigned to either formula feeding (FF) or breastfeeding (BF) forming four experimental groups: AB + formula-feeding (AB-FF, N = `r AB_FF` ) AB + breastfeeding (AB-BF, N = `r AB_BF` ), Water + formula-feeding (CON-FF, N = `r CON_FF` ), and CON + breastfeeding (CON-BF, N = `r CON_BF` ).
+
 -   Formula-feeding was performed as per METHODS AND MATERIALs.
-In Exp-1 we administered  0.08 ml formula for pups >2g, and 0.06 for pups ≤ 2g every 3 hours throughout the entire study.
+    In Exp-1 we administered 0.08 ml formula for pups \>2g, and 0.06 for pups ≤ 2g every 3 hours throughout the entire study.
     In later experiments this dose was changed, and feeding intervals were increased to every 4 hours between midnight and 08 AM.
 
-- `r nrow(preexp_1_survived)` of `r nrow(preexp_1)` pups survived the treatments.
+-   `r exp1_survived` of `r nrow(preexp_1)` pups survived the treatments.
+
 -   Those that did not survive were euthanize because of esophageal perforation during feeding.
+
 -   After fine-tuning of the feeding methods we were able to feed the pups without further harm.
+
 -   We never had to euthanize a pup that reached our humane endpoints, since all the surviving pups were in generally good shape.
+
 -   While significantly lighter than their breastfed controls, all pups increased their bodyweight during the experiment, attesting to the efficacy of our feeding regiment.
+
 -   The pups were physically active and responded to changes in light and physical stimuli.
+
 -   Skin color remained bright purple and fur development was visible throughout the experiment.
+
 -   No signs of intestinal inflammation were evident from the visual inspection of intestines during sampling.
     Our subjective observation of the intestines were small degrees of bloating in formula-fed pups.
-- Analysis of pro-inflammatory cytokines showed statistically significant changes in the expression of STATISTICAL-SIGNIF-CYTOKINE-NAMES. 
+
+**Cytokines**
+
+`r nrow(cytokines_1)` samples were available for analysis of the proinflammatory cytokines `r names_of_cytokines`.
+Concentrations were measured in 1-cm sections of proximal/distal ileum tissue using MSD Mesoscale Kits.
+Statistically significant changes in the expression of `r names_of_cytokines`.
 - Directionality of change varied, with alternating over- and under-expression of various cytokines.
-As can be seen in [*Cytokine plot of signif results *] pups tended to group based on feeding-scheme, with both formula-fed groups being either suppressed og over-expressed together, compared with breastfed groups, irrespective of antibiotics treatment.
+As can be seen in [*Cytokine plot of signif results* ] pups tended to group based on feeding-scheme, with both formula-fed groups being either suppressed or over-expressed together, compared with breastfed groups, irrespective of antibiotics treatment.
+
 -   This observation is supported by a a multiple linear regression analysis, evaluating the effects of MAT (AB or CON) and type of feed (BF of FF) separately and interactively.
-The analysis found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val$IL-6.
+The analysis found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val\$IL-6.
 
 #### Commentary and interpretation of results - Exp 1.
+
 -   In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers.
@@
 #### Rationale:
+
 -   Exp 1 demonstrated that the type of feed (breastmilk or formula) had a greater effect on cytokine expression than maternal antibiotics treatment.
@@
 #### Aim:
+
 -   To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics, without formula feeding on intestinal permeability and cytokine expression.
@@
 #### Conclusion:
+
 -   Birth weight: `r mean(bodyweight_birth_con` g (CON) vs `r mean(bodyweight_birth_mat` g (MAT)
@@
 
-*Antimals and experimental setup*
-- The breeding setup in Exp 2 was identical to Exp 1. 
+*Antimals and experimental setup* - The breeding setup in Exp 2 was identical to Exp 1.
 - On gestational day 15 vancomycin, gentamicin or control (tab water) was added to the drinking water pregnant females, forming 3 experimental groups; VANCO, GENTA and CON.
 
-*Weight development*
-- We checked the cages twice daily from gestational day 18 and measured the weight of newly-born pups. 
+*Weight development* - We checked the cages twice daily from gestational day 18 and measured the weight of newly-born pups.
 - VANCO and GENTA mothers gave birth on the 25th of february (getational day 20) with mean offspring body weights of `r mean(bodyweight_birth_vanco` (N = `r nrow(bodyweight_birth_vanco`) and `r mean(bodyweight_birth_genta` (N = `r nrow(bodyweight_birth_vanco`) respectively.
@@
 - This difference in body weight at birth was statistically significant (p = `r lm_weight_exp2$p.value`), but adjusting for length of gestation erased the statistical significance (p = `r lm_weight_exp2_adj$p.value`).
-- On day 3 and 5 this difference was in body weight was no longer detectable. *(Figure of body weights in exp 2)*
+- On day 3 and 5 this difference was in body weight was no longer detectable.
+*(Figure of body weights in exp 2)*
 
-*Intestinal permeability - FITC-Dextran administration* 
-- On the day of euthanasia pups were separated from the mothers at 07:00 (halting breastfeeding). 
+*Intestinal permeability - FITC-Dextran administration* - On the day of euthanasia pups were separated from the mothers at 07:00 (halting breastfeeding).
 - At 10:00, 50 µL FITC-dextran (0.045 mg/µL),for a final dose of 750 mg/kg, was administered orally using a PICC line and a 1 mL syringe.
@@
 - The highest concentration was found in the GENTA group (`r fitc_exp2_genta` µg/ml) and the lowest in the VANCO group (`r fitc_exp2_vanco` µg/ml), respectively, with the CON group in between (`r fitc_exp2_vanco` µg/ml).
-- This difference was statistically significant (p = `r anova_fitc_exp2`, one-way analysis of variance). *(Figure of FITC in exp 2)*
+- This difference was statistically significant (p = `r anova_fitc_exp2`, one-way analysis of variance).
+*(Figure of FITC in exp 2)*
 
-*Cytokine expression*
-- Tissue weight-adjusted results of cytokine concentrations showed significant differences between the 3 groups for `r signif_cyto_exp2`. ### Rerun these experiments.
+*Cytokine expression* - Tissue weight-adjusted results of cytokine concentrations showed significant differences between the 3 groups for `r signif_cyto_exp2`.
+### Rerun these experiments.
 
 #### Commentary and interpretation experimental results - Exp 2.
+
 -   Weight development was now similar to the findings of Chen et al.
+
 -   Whether it was a result of growth restriction or shorter gestational age is difficult to determine.
+
 -   Chen et al. did not report whether dams were from the same day.
+
 -   Intestinal permeability response was different depending of type of antibiotics.
+
 -   Vancomycin decreased intestinal permeability whereas gentamicin increased it.
+
     -   This has been shown in Winstar rats before (Thulstrup 2015).
+
 -   Cytokine expression was most affected in the VANCO group.
@@
 -   In summary:
+
 -   In experiment 2, Maternal antibiotics treatment did not have an effect.
+
 -   This indicates that the greater effect is of formula-feeding, similar to the findings in exp 1.
+
 -   The intestinal permeability assay is interesting in that it supports the findings of Thulstrup et al. 2015.
@@
 #### Rationale:
+
 Based on the results from Exp 2 it was difficult to ascertain which treatment was more likely to manifest in NEC like phenotype.
@@
 #### Aim:
+
 -   Repeat Exp 2 with greater sample sizes, adding the effect of formula feeding.
@@
 #### Conclusion:
+
 -   Combining MAT with formula feeding resulted in more pronounced signs of inflammation than MAT alone.
@@
 #### Detailed description of Exp 3.
-*Antimals and experimental setup*
-- To increase the number of pups available we doubled the number of mice for breeding, 36 females and 12 males. 
+
+*Antimals and experimental setup* - To increase the number of pups available we doubled the number of mice for breeding, 36 females and 12 males.
 - Pregnant females were again given either Vancomycin, Gentamicin (1 g/l) or sterile water on gestational day 15.
-- `r nrow(preexp_3` pups were born and separated into breastfeeding or formula feeding, for a total of 6 groups:
-- VANCO-BF, VANCO-FF, GENTA-BF, GENTA-FF, CON-BF and CON-FF.
+- `r nrow(preexp_3` pups were born and separated into breastfeeding or formula feeding, for a total of 6 groups: - VANCO-BF, VANCO-FF, GENTA-BF, GENTA-FF, CON-BF and CON-FF.
 
-*Intestinal permeability - FITC-Dextran administration* 
-- We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
+*Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
 - Formula-fed pups consistently showed higher levels of Serum FITC concentration compared to breastfed pups, irrespective of MAT status.
@@
 
-*Cytokine expression*
-- We found significant differences in concentrations of XXSignificantSCYTOKINESXX.
+*Cytokine expression* - We found significant differences in concentrations of XXSignificantSCYTOKINESXX.
 Linear regression analysis indicated the biggest determinant of differences in outcomes was formula feeding, with the type of antibiotic being of lesser importance compared.
@@
 #### Commentary and interpretation experimental results Exp 3:
+
 -   Intestinal permeability was by far the greatest in VANCO-FF, with levels of FITC-dex close to 10-fold increased in this group.
@@
 
-*Adjusting for weight of tissue*
-- In the previous experiments, cytokine concentrations were adjusted for the weight of the tissue analyzed, to control for biopsy sizes. 
+*Adjusting for weight of tissue* - In the previous experiments, cytokine concentrations were adjusted for the weight of the tissue analyzed, to control for biopsy sizes.
 - We did so because the average weight of tissue samples was not meaningfully different between groups.
@@
 
-*Combining results from exp 2 and exp 3*
-- The protocols for the breastfed groups were identical across experiment 2 and 3, allowing for pooling of these results.
+*Combining results from exp 2 and exp 3* - The protocols for the breastfed groups were identical across experiment 2 and 3, allowing for pooling of these results.
 - While analyzing the results from both groups together increased the group sizes and degrees of statistical freedom, it did not change the clinical conclusions.
@@
 -   Previous experiments had shown that MAT in combination with breastfeeding had no effect on inflammatory markers and VANCO-BF thus acted as a healthy control to both formula-fed treatment groups.
- **(The problem here is that formula is the only thing with an effect. A breastfed control only shows the effect of the formula)**
+**(The problem here is that formula is the only thing with an effect.
+A breastfed control only shows the effect of the formula)**
 
 #### Aim:
+
 -   To test whether signs of inflammation are reduced in pups treated with an FVT.
@@
 #### Conclusion:
+
 -   FVT treated pups showed similar intestinal permeability and pro-inflammatory cytokine concentrations as non-FVT-treated controls.
@@
 #### Detailed description of Exp 4.
+
 -   For the final experiment we repeated the previous breeding setup, assigning 36 females and 12 males for breeding.
+
 -   Only 5 of the 36 females became pregnant, and all were assigned to vancomycin (1 g/l) from pregnancy-day 15 until delivery, giving birth to `r nrow(preexp_4` pups on Saturday the XXth.
+
 -   2 days after birth pups were separated into breastfeeding or formula feeding.
+
 -   The formula-fed pups were further divided into a treatment and control group, receiving either 30 µl fecal virome filtrate (FVT) (0.01 mg farces/µL) or an equal amount of SM buffer (control) 1 hour prior to the first feeding.
+
 -   We repeated the FVT administration 24 hours later (09:00 day 2) for a total dose of 0.6 mg feces material per pup.
@@
 
-*Intestinal permeability - FITC-Dextran administration* 
-- We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
+*Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
 - The FVT and FVT-control (SM) had similar permeability scores.
@@
 
-*Cytokine expression*
-- We found significant differences in concentrations of TNFa in Exp 4, with greater level in the FVT-treated group (XXX), compared to both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively). 
+*Cytokine expression* - We found significant differences in concentrations of TNFa in Exp 4, with greater level in the FVT-treated group (XXX), compared to both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively).
 - This is in contrast experiment 3, where differences between VANCO-BF and VANCO-FF found significant differences for XXXCYtokinesXXX.
@@
 
-*Gene expression*
-- qPCR analyzis of intestinal samples found statistically significant differences in the level of XXXSIGNIFICANT_GENESXXX expression between groups with the delta-delta method (reformulate).
+*Gene expression* - qPCR analyzis of intestinal samples found statistically significant differences in the level of XXXSIGNIFICANT_GENESXXX expression between groups with the delta-delta method (reformulate).
 - However, a difference in expression of the reference gene between the FVT and breastfed control was observed.
-this difference was statistically significant, invalidating the assumption of *equal baseline expression.*(reformulate)
-- Comparing the groups on the basis of (SOMEHTING ELSE THAN DELTA DELTA) found a `r fold_increase_reg3g`-fold increased expression of the `r REG3G` gene in the FVT groups.
+this difference was statistically significant, invalidating the assumption of *equal baseline expression.*(reformulate) - Comparing the groups on the basis of (SOMEHTING ELSE THAN DELTA DELTA) found a `r fold_increase_reg3g`-fold increased expression of the `r REG3G` gene in the FVT groups.
 
 #### Commentary and interpretation
+
 -   FVT-treatmed did not seem to change the intestinal permeability.
+
 -   It is possible that something went wrong with the VANCO-BF group, given FITC-concentration was `r mean(fitc_level_exp4$bf`, which is 4-5 times higher than VANCO-BF groups on identical protocols in Exp 2 `r mean(fitc_level_exp2$bf` and 3 `r mean(fitc_level_exp3$bf`.
@@
 **FIGURE of EXP4 with DIFFERENT CONTROL GROUP**
+
 -   The level of pro-inflammatory cytokines were the same in the FVT and SM groups, again indicating a lack of efficacy of the FVT treatment.
@@
 
-### Post experimental analyses and Discussion of results
+### \--\>Post experimental analyses and Discussion of results
 
 -   Histological image analysis was performed after all experiments had finished.
+
 -   We inspected the morphology of the ileum and colon, and visually assessed thickness of the mucin layers.
+
 -   Analyzing AB-PAS stains for mucin-stained tissue to the total area, showed no differences between any treatment groups.
+
 -   Formula-fed and breastfed pups were slightly different, irrespective of antibiotics treatment, whereas there were no observable differences between antibiotic treatments.
+
 -   These findings supports the visual inspections of intestines made during sampling, where no signs of inflammation were detected.
+
 -   *Images of intestines*
 
-* Gut microbiome composition
+-   Gut microbiome composition
 

### 2022-11-04 submission/results backup copy.Rmd
Created thesis structure

---
title: ''
output: pdf_document
---
# Results

```{r open data and libraries, include = F}
library(knitr)  # for dynamic report generation
library(kableExtra) # to build complex HTML or 'LaTex' tables
library(readxl)
library(tidyverse)
library(broom)
library(ggpubr)
library(patchwork)
```

```{r exp1-weight, include=FALSE}
#____Load overview of sample IDs______ 
pups <- read_excel("../data/processed/weight_development.xlsx")

#____ Separate into experimental rounds ____#
preexp_1 <- pups %>% 
  filter(exp_number == "1")

exp1_survived <- preexp_1 %>% 
  filter(!is.na(bodyweight_baseline_plus_48h)) %>% 
  nrow()
#____ ____#
```

```{r exp1-groups}
#____Separate into groups ______ 
AB_FF <- nrow(preexp_1 %>% 
  filter(group2 == "AB-FF"))

AB_BF <- nrow(preexp_1 %>% 
  filter(group2 == "AB-BF"))

CON_FF <- nrow(preexp_1 %>% 
  filter(group2 == "CON-FF"))

CON_BF <- nrow(preexp_1 %>% 
  filter(group2 == "CON-BF")) 
#____ ____#

```
The main goal of this thesis project was to create a NEC mouse model by combining Maternal Antibiotics Treatment with Formula-Feeding.
This hypothesis was tested in a series of experiments (Exp 1-4), each successive experiment building on the findings from the former.
The following is a description of each experiment.
An abridged version of each experiment is provided, followed by a detailed description of experimental setup and results.

```{r path-to-folder plots fixed size, echo = FALSE, out.height="40%", fig.show='hold', fig.align="center",  fig.cap="Experimental protocols"}

include_graphics("../results/round_1/Table cytokines round 1.png")
```

\newpage

## Exp 1: Model feasibility

### Rationale:

-   *Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
-   XXX *[Find paper]* found formula-feeding to be damaging to the intestine.
-   One of the primary causal factors of NEC is formula-feeding (paper on what is thought to cause NEC).
-   We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}

### Aim:

-   To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
-   Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
-   Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum

### Conclusion:

-   Survival rate: `r exp1_survived` of `r nrow(preexp_1)`(`r (exp1_survived / nrow(preexp_1))*100`%)
-   Visual inspection: no clear signs of intestinal inflammation
-   Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.

### Detailed description of Exp 1. **Date of exp**

*Antimals and experimental setup* \n 
- The experimental setup has been described in detail in the "Animal experimental setup" methods section. Detials specific for Exp-1 are outlined below. \n
- Six male and 18 female C57BL/6N mice were co-housed for 24 hours in 6 cages (1 male + 3 female each).
- Two weeks after mating, pregnant females were separated evenly into maternal antibiotic treatment (MAT) and water-only controls (CON).
- Antibiotics (Gentamicin, Vancomycin, Neomycin, and Ampicillin (all 0.5 g/L)) were mixed into the drinking water of the MAT mice while controls received water only.
- Antibiotic treatment commenced on gestational day 15 and continued until delivery.
- Each mother was housed with her offspring in separate cages during the first 2 days after birth.
-   On day 3, mice from water or antibiotics treated mothers were separated and assigned to either formula feeding (FF) or breastfeeding (BF) forming four experimental groups: AB + formula-feeding (AB-FF, N = `r AB_FF` ) AB + breastfeeding (AB-BF, N = `r AB_BF` ), Water + formula-feeding (CON-FF, N = `r CON_FF` ), and CON + breastfeeding (CON-BF, N = `r CON_BF` ).
-   Formula-feeding was performed as per METHODS AND MATERIALs.
    In Exp-1 we administered 0.08 ml formula for pups \>2g, and 0.06 for pups ≤ 2g every 3 hours throughout the entire study.
  In later experiments this dose was changed, and feeding intervals were increased to every 4 hours between midnight and 08 AM. \n
- `r exp1_survived` of `r nrow(preexp_1)` pups survived the treatments.
- Those that did not survive were euthanize because of esophageal perforation during feeding.
- After fine-tuning of the feeding methods we were able to feed the pups without further harm.
- We never had to euthanize a pup that reached our humane endpoints, since all the surviving pups were in generally good shape.
- While significantly lighter than their breastfed controls, all pups increased their bodyweight during the experiment, attesting to the efficacy of our feeding regiment.
- The pups were physically active and responded to changes in light and physical stimuli.
- Skin color remained bright purple and fur development was visible throughout the experiment.
- No signs of intestinal inflammation were evident from the visual inspection of intestines during sampling. Our subjective observation of the intestines were small degrees of bloating in formula-fed pups.

```{r exp1-cytokines, include=FALSE}
#____Load cytokine data______ 
cytokines <- read_excel("../data/processed/weight_development.xlsx")

#____ Separate into experimental rounds ____#
preexp_1 <- pups %>% 
  filter(exp_number == "1")

exp1_survived <- preexp_1 %>% 
  filter(!is.na(bodyweight_baseline_plus_48h)) %>% 
  nrow()
#____ ____#
```

<!-- *Cytokines* \n -->
<!-- `r nrow(cytokines_1)` samples were available for analysis of the proinflammatory cytokines `r names_of_cytokines`. -->
<!-- Concentrations were measured in 1-cm sections of proximal/distal ileum tissue using MSD Mesoscale Kits. -->
<!-- - Directionality of change varied, with alternating over- and under-expression of various cytokines. -->
<!-- As can be seen in [*Cytokine plot of signif results* ] pups tended to group based on feeding-scheme, with both formula-fed groups being either suppressed or over-expressed together, compared with breastfed groups, irrespective of antibiotics treatment. -->
<!-- Statistically significant changes in the expression of `r names_of_cytokines`. -->

<!-- -   This observation is supported by a a multiple linear regression analysis, evaluating the effects of MAT (AB or CON) and type of feed (BF of FF) separately and interactively. The analysis found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val\$IL-6. -->

<!-- #### Commentary and interpretation of results - Exp 1. -->

<!-- -   In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers. -->
<!-- -   This is surprising given the findings of Chen et al., who used an identical protocol to our AB-BF vs. CON-BF (Maternal antibiotics treatment and weening with the mother for 7 days). -->
<!-- -   They claim that MAT during pregnancy resulted in significant alterations in the intestinal microbiota. lower body weight at birth and on day 7, visible intestinal injury and inflammation and disruption of the intestinal barrier. -->
<!-- -   Our initial analysis of weight and cytokine expression did not indicate a similar effect of MAT (evaluations of gut permeability, intestinal gene expression, histological analysis and gut microbial composition are conducted in later experiments). -->
<!-- -   One possible explanation is an effect of dose. While Chen et al., used a concentration of 1 g/l, we, unfortunately, tested 0.5 g/l, based on experience from previous studies, making a direct comparison difficult. -->
<!-- -   We did find significant over-expression of cytokines relevant in inflammation, indicating some effect of the treatments. -->
<!-- -   Since these biochemical signals were not strong enough to manifest phenotypically, we reasoned that our treatment wasn't sufficiently injurious. -->
<!-- -   We therefore decided to increase the dosage to 1 g/l for subsequent experiments, as well as to increase the osmolality of the formula for the next feeding experiment (Exp 3). -->
<!-- -   With assurance in our experimental setup we decided to include further measures of intestinal dysfunction. -->
<!-- -   We also decided to measure the birth weight of all pups, in addition to body weight at the beginning of the experiment (day 3) and end (day 5). -->

<!-- ### Experiment 2: Effects of various types of antibiotics - (Microbiome dysbiosis) -->

<!-- #### Rationale: -->

<!-- -   Exp 1 demonstrated that the type of feed (breastmilk or formula) had a greater effect on cytokine expression than maternal antibiotics treatment. -->
<!-- -   Central to our hypothesis of NEC development is an intestinal microbiome dysibiosis, and a broad-spectrered antibiotic treatment potentially resulted in an overall depletion rather than the manifisation of a dysbiosis. -->
<!-- -   We hypothesized that targeting specific strains of bacteria in the mother would lead to a more dysbiotic microbiome and allow for greater subsequent dysbiosis in the offspring. -->

<!-- #### Aim: -->

<!-- -   To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics, without formula feeding on intestinal permeability and cytokine expression. -->
<!-- -   To see whether a larger dose of antibiotics (1 g/l) causes decreased body weight at birth and at day 5 compared with non-treated controls. -->

<!-- #### Conclusion: -->

<!-- -   Birth weight: `r mean(bodyweight_birth_con` g (CON) vs `r mean(bodyweight_birth_mat` g (MAT) -->
<!-- -   Visual inspection: no clear signs of intestinal inflammation -->
<!-- -   Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES. -->

<!-- #### Detailed description of Exp 2. -->

<!-- *Antimals and experimental setup* - The breeding setup in Exp 2 was identical to Exp 1. -->
<!-- - On gestational day 15 vancomycin, gentamicin or control (tab water) was added to the drinking water pregnant females, forming 3 experimental groups; VANCO, GENTA and CON. -->

<!-- *Weight development* - We checked the cages twice daily from gestational day 18 and measured the weight of newly-born pups. -->
<!-- - VANCO and GENTA mothers gave birth on the 25th of february (getational day 20) with mean offspring body weights of `r mean(bodyweight_birth_vanco` (N = `r nrow(bodyweight_birth_vanco`) and `r mean(bodyweight_birth_genta` (N = `r nrow(bodyweight_birth_vanco`) respectively. -->
<!-- - The CON mother gave birth on the 26th of february (gestational day 21) with mean offspring body weight of `r mean(bodyweight_birth_con` (N = `r nrow(bodyweight_birth_vanco`). -->
<!-- - This difference in body weight at birth was statistically significant (p = `r lm_weight_exp2$p.value`), but adjusting for length of gestation erased the statistical significance (p = `r lm_weight_exp2_adj$p.value`). -->
<!-- - On day 3 and 5 this difference was in body weight was no longer detectable. -->
<!-- *(Figure of body weights in exp 2)* -->

<!-- *Intestinal permeability - FITC-Dextran administration* - On the day of euthanasia pups were separated from the mothers at 07:00 (halting breastfeeding). -->
<!-- - At 10:00, 50 µL FITC-dextran (0.045 mg/µL),for a final dose of 750 mg/kg, was administered orally using a PICC line and a 1 mL syringe. -->
<!-- - Two hours later euthanasia and sample collection began (12:00). -->
<!-- - Pups were euthanized by cervical dislocation and blood samples collected and spun in Eppendorf tubes. -->
<!-- - FITC molecules were detectable in all groups. -->
<!-- - The highest concentration was found in the GENTA group (`r fitc_exp2_genta` µg/ml) and the lowest in the VANCO group (`r fitc_exp2_vanco` µg/ml), respectively, with the CON group in between (`r fitc_exp2_vanco` µg/ml). -->
<!-- - This difference was statistically significant (p = `r anova_fitc_exp2`, one-way analysis of variance). -->
<!-- *(Figure of FITC in exp 2)* -->

<!-- *Cytokine expression* - Tissue weight-adjusted results of cytokine concentrations showed significant differences between the 3 groups for `r signif_cyto_exp2`. -->
<!-- ### Rerun these experiments. -->

<!-- #### Commentary and interpretation experimental results - Exp 2. -->

<!-- -   Weight development was now similar to the findings of Chen et al. -->

<!-- -   Whether it was a result of growth restriction or shorter gestational age is difficult to determine. -->

<!-- -   Chen et al. did not report whether dams were from the same day. -->

<!-- -   Intestinal permeability response was different depending of type of antibiotics. -->

<!-- -   Vancomycin decreased intestinal permeability whereas gentamicin increased it. -->

<!--     -   This has been shown in Winstar rats before (Thulstrup 2015). -->

<!-- -   Cytokine expression was most affected in the VANCO group. -->

<!-- -   In summary: -->

<!-- -   In experiment 2, Maternal antibiotics treatment did not have an effect. -->

<!-- -   This indicates that the greater effect is of formula-feeding, similar to the findings in exp 1. -->

<!-- -   The intestinal permeability assay is interesting in that it supports the findings of Thulstrup et al. 2015. -->

<!-- ### Experiment 3 - Interaction effects between antibiotics treatment and formula feeding -->

<!-- #### Rationale: -->

<!-- Based on the results from Exp 2 it was difficult to ascertain which treatment was more likely to manifest in NEC like phenotype. -->
<!-- - Gentamicin had increased intestinal permeability the most. -->
<!-- - However, the greatest effect of cytokine expression was found with the vancomycin treatment. -->
<!-- - In addition to this divergence, the interaction effect found in Exp 1 suggested it was likely that combining MAT with formula feeding would yield different results. -->
<!-- - Therefore, we decided to increase the breeding population to allow for an experiment that combined both types of antibiotics with either formula-feeding or breastfeeding, for a total of 6 experimental groups. -->

<!-- #### Aim: -->

<!-- -   Repeat Exp 2 with greater sample sizes, adding the effect of formula feeding. -->
<!-- -   To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics in combination with formula feeding on intestinal permeability and cytokine expression. -->
<!-- -    -->

<!-- #### Conclusion: -->

<!-- -   Combining MAT with formula feeding resulted in more pronounced signs of inflammation than MAT alone. -->
<!-- -   VANCO-FF and VANCO-BF had the highest and lowest intestinal permeability scores, respectively. -->
<!-- -   Significantly altered expression of XXSignificantSCYTOKINESXX. -->
<!-- -   Greatest determinant of variation was formula-feeding. -->

<!-- #### Detailed description of Exp 3. -->

<!-- *Antimals and experimental setup* - To increase the number of pups available we doubled the number of mice for breeding, 36 females and 12 males. -->
<!-- - Pregnant females were again given either Vancomycin, Gentamicin (1 g/l) or sterile water on gestational day 15. -->
<!-- - `r nrow(preexp_3` pups were born and separated into breastfeeding or formula feeding, for a total of 6 groups: - VANCO-BF, VANCO-FF, GENTA-BF, GENTA-FF, CON-BF and CON-FF. -->

<!-- *Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00. -->
<!-- - Formula-fed pups consistently showed higher levels of Serum FITC concentration compared to breastfed pups, irrespective of MAT status. -->
<!-- - VANCO-FF and VANCO-BF had the highest and lowest levels of FITC, respectively. -->

<!-- *Cytokine expression* - We found significant differences in concentrations of XXSignificantSCYTOKINESXX. -->
<!-- Linear regression analysis indicated the biggest determinant of differences in outcomes was formula feeding, with the type of antibiotic being of lesser importance compared. -->
<!-- - Directionally the trend was the same for all cytokines. -->
<!-- - IL-1b: Non-significant decrease in all antibiotics-treated offspring, irrespective of feeding. -->
<!-- - IL-2: significantly lower concentrations in all formula-fed pups compared with CON-BF. -->
<!-- - IL-6: Significantly greater levels in all formula-fed pups compared with CON-BF. -->
<!-- - INFg: significantly lower concentrations in all formula-fed pups compared with CON-BF. -->
<!-- - KCGRO: Significantly greater levels in all formula-fed pups compared with CON-BF. -->

<!-- #### Commentary and interpretation experimental results Exp 3: -->

<!-- -   Intestinal permeability was by far the greatest in VANCO-FF, with levels of FITC-dex close to 10-fold increased in this group. -->
<!-- -   In contrast, VANCO-BF was the group with the lowest measure of intestinal permeability, similar to the finding in exp 2. -->
<!-- -   Looking at **FIGURE CYTOKINES ROUND 3** shows how formula-fed groups systematically over- or underexpress compared to breastfed analimals, irrespective of type of antibiotic, indicating that the biggest determinant of cytokine variation was again formula feeding. -->

<!-- *Adjusting for weight of tissue* - In the previous experiments, cytokine concentrations were adjusted for the weight of the tissue analyzed, to control for biopsy sizes. -->
<!-- - We did so because the average weight of tissue samples was not meaningfully different between groups. -->
<!-- - This assumption wasn't met in Exp 3, where samples from formula-fed pups weighted on average XX%XX more (p = XXp.valueXX) than breastfed controls. -->
<!-- - Vanco-FF had the heaviest ileum tissue samples of all **mean + P (if its true)**. -->
<!-- - Tissue edema could well be indicative of inflammation, and hence desirable, however it introduces a bias against the formula-fed pups if cytokine concentrations were normalized to weight. -->
<!-- - The reported findings are no longer significant **Table of significant cytokines** when adjusting for weight. -->

<!-- -   This phenomenon is a common challenge in animal disease models, requiring measure of normalization other that tissue weight, such as total protein content. -->
<!-- -   Unfortunately this discovery was done after tissue homogenization and analysis, making it too late for further analysis. -->

<!-- *Combining results from exp 2 and exp 3* - The protocols for the breastfed groups were identical across experiment 2 and 3, allowing for pooling of these results. -->
<!-- - While analyzing the results from both groups together increased the group sizes and degrees of statistical freedom, it did not change the clinical conclusions. -->

<!-- ### Experiment 4 - FVT Trial -->

<!-- #### Rationale: **What was the actual rationale?** -->

<!-- -   Central to the development of NEC is an assumption of a gut microbial dysbiosis, which we tried to mimick by administering antibiotics prenatally. -->
<!-- -   Results from exp 1-3 indicated that maternal antibiotics treatment had little effect in the context of breastfeeding, but when combined with formula-feeding was able to induce mild symptoms of intestinal inflammation. -->
<!-- -   It was unclear whether this was solely an effect of the formula, or whether an interaction with the maternal antibiotics treatment existed. -->
<!-- -   If our findings were partly due to a successful manipulation of the gut microbiome, then eliminating the induced disbyosis should also partly mitigate the inflammatory signatures. -->
<!-- -   Relative bacterial abundance / bacterial composition is partly a function of viral (bacteriophage) abundance, and sevaral studies have shown how a FVT can alter the bacterial composition. -->
<!-- -   We therefore decided to conduct a final experiment, in which we administered an FVT to a vancomycing-treated, formula-fed group (since this combination (VANCO-FF) had the biggest effects in terms of intestinal permeability and proinflammatory cytokine expression in Exp 3). -->
<!-- -   We hypothesized that any dysbiosis created by maternal antibiotics treatment could be mitigated with a follow-up administration of FVT prior to beginnig formula-feeding. -->
<!-- -   If successful, the FVT-treated group would show fever intestinal injury scores compared with a non-FVT treated group. -->
<!-- -   Previous experiments had shown that MAT in combination with breastfeeding had no effect on inflammatory markers and VANCO-BF thus acted as a healthy control to both formula-fed treatment groups. **(The problem here is that formula is the only thing with an effect. A breastfed control only shows the effect of the formula)** -->

<!-- #### Aim: -->

<!-- -   To test whether signs of inflammation are reduced in pups treated with an FVT. -->
<!-- -   To test whether any reductions in symptoms can be explained with a manipulation of the microbial composition. -->

<!-- #### Conclusion: -->

<!-- -   FVT treated pups showed similar intestinal permeability and pro-inflammatory cytokine concentrations as non-FVT-treated controls. -->
<!-- -   An `r fold_increase_reg3g`-fold increase in expression of the `r REG3G` gene was found in the FVT group. -->

<!-- #### Detailed description of Exp 4. -->

<!-- -   For the final experiment we repeated the previous breeding setup, assigning 36 females and 12 males for breeding. -->

<!-- -   Only 5 of the 36 females became pregnant, and all were assigned to vancomycin (1 g/l) from pregnancy-day 15 until delivery, giving birth to `r nrow(preexp_4` pups on Saturday the XXth. -->

<!-- -   2 days after birth pups were separated into breastfeeding or formula feeding. -->

<!-- -   The formula-fed pups were further divided into a treatment and control group, receiving either 30 µl fecal virome filtrate (FVT) (0.01 mg farces/µL) or an equal amount of SM buffer (control) 1 hour prior to the first feeding. -->

<!-- -   We repeated the FVT administration 24 hours later (09:00 day 2) for a total dose of 0.6 mg feces material per pup. -->

<!-- -   2 hours after FITC administtion pups were euthanized by decapitation and blood collected in eppendorfs for analysis of serum FITC concentration. -->
<!--     Intestinal samples were collected for gene expression assays, histological assessment, pro-inflammatory cytokine analysis and gut microbiome sequencing. -->

<!-- *Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00. -->
<!-- - The FVT and FVT-control (SM) had similar permeability scores. -->
<!-- - Compared with the vancomycin-treated formula-fed pups from experiment 3, there were no significant differences **Figure X**. -->
<!-- - The breastfed pups did, however, have significantly higher levels compared with all other breastfed pups. -->

<!-- *Cytokine expression* - We found significant differences in concentrations of TNFa in Exp 4, with greater level in the FVT-treated group (XXX), compared to both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively). -->
<!-- - This is in contrast experiment 3, where differences between VANCO-BF and VANCO-FF found significant differences for XXXCYtokinesXXX. -->
<!-- OR: - This is similar to the findings of experiment 3, where no differences between VANCO-BF and VANCO-FF were found either. -->

<!-- *Gene expression* - qPCR analyzis of intestinal samples found statistically significant differences in the level of XXXSIGNIFICANT_GENESXXX expression between groups with the delta-delta method (reformulate). -->
<!-- - However, a difference in expression of the reference gene between the FVT and breastfed control was observed. -->
<!-- this difference was statistically significant, invalidating the assumption of *equal baseline expression.*(reformulate) - Comparing the groups on the basis of (SOMEHTING ELSE THAN DELTA DELTA) found a `r fold_increase_reg3g`-fold increased expression of the `r REG3G` gene in the FVT groups. -->

<!-- #### Commentary and interpretation -->

<!-- -   FVT-treatmed did not seem to change the intestinal permeability. -->

<!-- -   It is possible that something went wrong with the VANCO-BF group, given FITC-concentration was `r mean(fitc_level_exp4$bf`, which is 4-5 times higher than VANCO-BF groups on identical protocols in Exp 2 `r mean(fitc_level_exp2$bf` and 3 `r mean(fitc_level_exp3$bf`. Compared with theses levels the development looks more biologically reasonable. **FIGURE of EXP4 with DIFFERENT CONTROL GROUP** -->

<!-- -   The level of pro-inflammatory cytokines were the same in the FVT and SM groups, again indicating a lack of efficacy of the FVT treatment. -->

<!-- -   Summary: The FVT treatment did not seem to have an effect. -->

<!-- ### \--\>Post experimental analyses and Discussion of results -->

<!-- -   Histological image analysis was performed after all experiments had finished. -->

<!-- -   We inspected the morphology of the ileum and colon, and visually assessed thickness of the mucin layers. -->

<!-- -   Analyzing AB-PAS stains for mucin-stained tissue to the total area, showed no differences between any treatment groups. -->

<!-- -   Formula-fed and breastfed pups were slightly different, irrespective of antibiotics treatment, whereas there were no observable differences between antibiotic treatments. -->

<!-- -   These findings supports the visual inspections of intestines made during sampling, where no signs of inflammation were detected. -->

<!-- -   *Images of intestines* -->

<!-- -   Gut microbiome composition -->

<!-- ## Discussion -->

### 2022-11-04 submission/results.Rmd
Created thesis structure

@@
 output: pdf_document
-editor_options: 
-  chunk_output_type: console
-  markdown: 
-    wrap: sentence
 ---
+# Results
 
-<!--# Code -->
-
-<!--# Run scripts for plots on another file and save plots on img folder. Then use include_graphics to include plots where relevant -->
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = FALSE)
-```
-
-```{r loading libraries, message=FALSE, warning=FALSE}
+```{r load libraries, include = F}
+library(knitr)  # for dynamic report generation
+library(kableExtra) # to build complex HTML or 'LaTex' tables
 library(readxl)
-#library(tidyverse)
+library(tidyverse)
 library(broom)
 library(ggpubr)
-#library(AICcmodavg)
-#library(visdat)
-#library(rstatix)
-#library(gt)
-#library(gtsummary)
 library(patchwork)
-#library(jtools)
 ```
@@
 #____Load overview of sample IDs______ 
-pups <- read_excel("data/processed/weight_development.xlsx")
+pups <- read_excel("../data/processed/weight_development.xlsx")
 
@@
 
-<!-- Manuscript -->
-
-The main goal of this thesis project was to create a NEC mouse model by combining Maternal Antibiotics Treatment with Formula-Feeding.
-This hypothesis was tested in a series of experiments (Exp 1-4), each successive experiment building on the findings from the former.
-The following is a description of each experiment.
-An abridged version of each experiment is provided, followed by a detailed description of experimental setup and results.
-
-### Exp 1: Model feasibility
-
-#### Rationale:
-
--   *Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
--   XXX *[Find paper]* found formula-feeding to be damaging to the intestine.
--   One of the primary causal factors of NEC is formula-feeding (paper on what is thought to cause NEC).
--   We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
-
-#### Aim:
-
--   To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
--   Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
--   Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum
-
-#### Conclusion:
-
--   Survival rate: `r`exp1_survived`of`r nrow(preexp_1)`(`r (exp1_survived / nrow(preexp_1))\*100\`%)
--   Visual inspection: no clear signs of intestinal inflammation
--   Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
-
-#### Detailed description of Exp 1. **Date of exp**
-
-*Antimals and experimental setup* See "Animal experimental setup" in methods.
-- The experimental setup has been described in detail in the methods section.
-- Detials specific for Exp-1 are outlined below.
-- Briefly, six male and 18 female C57BL/6N mice were co-housed for 24 hours in 6 cages (1 male + 3 female each).
-- Two weeks after mating, pregnant females were separated evenly into maternal antibiotic treatment (MAT) and water-only controls (CON).
-- Antibiotics (Gentamicin, Vancomycin, Neomycin, and Ampicillin (all 0.5 g/L)) were mixed into the drinking water of the MAT mice while controls received water only.
-- Antibiotic treatment commenced on gestational day 15 and continued until delivery.
-- Each mother was housed with her offspring in separate cages during the first 2 days after birth.
-
-```{r exp1-groups}
+```{r exp1-groups, include=FALSE}
 #____Separate into groups ______ 
@@
 ```
+The main goal of this thesis project was to create a NEC mouse model by combining Maternal Antibiotics Treatment with Formula-Feeding.
+This hypothesis was tested in a series of experiments (Exp 1-4), each successive experiment building on the findings from the former.
+The following is a description of each experiment.
+An abridged version of each experiment is provided, followed by a detailed description of experimental setup and results.
 
+```{r protocols, echo = FALSE, out.height="40%", fig.show='hold', fig.align="center",  fig.cap="Experimental protocols"}
+
+include_graphics("../results/round_1/Table cytokines round 1.png")
+```
+
+\newpage
+
+## Exp 1: Model feasibility
+
+### Rationale:
+
+-   *Chen et al.*, showed that treatment with antibiotics during pregnancy(Maternal Antibiotics Treatment, **MAT**) "resulted" in mild NEC-like symptoms in the offspring.
+-   XXX *[Find paper]* found formula-feeding to be damaging to the intestine.
+-   One of the primary causal factors of NEC is formula-feeding (paper on what is thought to cause NEC).
+-   We tested the hypothesis that MAT combined with formula feeding (**FF**) would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
+
+### Aim:
+
+-   To test whether it was possible to keep mice alive from postnatal day 3 and 48 hours onwards, while being separated from the mother.
+-   Whether combining a broad spectrum antibiotic with formula-feeding will result in NEC-like symptoms such as:
+-   Visible signs of intestinal inflammation and / or greater levels of pro-inflammatory cytokines expressed in tissue or serum
+
+### Conclusion:
+
+-   Survival rate: `r exp1_survived` of `r nrow(preexp_1)`(`r (exp1_survived / nrow(preexp_1))*100`%)
+-   Visual inspection: no clear signs of intestinal inflammation
+-   Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
+
+### Detailed description of Exp 1. **Date of exp**
+
+*Antimals and experimental setup* \\
+- The experimental setup has been described in detail in the "Animal experimental setup" methods section. Detials specific for Exp-1 are outlined below. \\
+- Six male and 18 female C57BL/6N mice were co-housed for 24 hours in 6 cages (1 male + 3 female each).
+- Two weeks after mating, pregnant females were separated evenly into maternal antibiotic treatment (MAT) and water-only controls (CON).
+- Antibiotics (Gentamicin, Vancomycin, Neomycin, and Ampicillin (all 0.5 g/L)) were mixed into the drinking water of the MAT mice while controls received water only.
+- Antibiotic treatment commenced on gestational day 15 and continued until delivery.
+- Each mother was housed with her offspring in separate cages during the first 2 days after birth.
 -   On day 3, mice from water or antibiotics treated mothers were separated and assigned to either formula feeding (FF) or breastfeeding (BF) forming four experimental groups: AB + formula-feeding (AB-FF, N = `r AB_FF` ) AB + breastfeeding (AB-BF, N = `r AB_BF` ), Water + formula-feeding (CON-FF, N = `r CON_FF` ), and CON + breastfeeding (CON-BF, N = `r CON_BF` ).
-
 -   Formula-feeding was performed as per METHODS AND MATERIALs.
-    In Exp-1 we administered 0.08 ml formula for pups \>2g, and 0.06 for pups ≤ 2g every 3 hours throughout the entire study.
-    In later experiments this dose was changed, and feeding intervals were increased to every 4 hours between midnight and 08 AM.
-
+    In Exp-1 we administered 0.08 ml formula for pups over 2g, and 0.06 for pups less 2g every 3 hours throughout the entire study.
+  In later experiments this dose was changed, and feeding intervals were increased to every 4 hours between midnight and 08 AM. \\
 - `r exp1_survived` of `r nrow(preexp_1)` pups survived the treatments.
-
 - Those that did not survive were euthanize because of esophageal perforation during feeding.
-
 - After fine-tuning of the feeding methods we were able to feed the pups without further harm.
-
 - We never had to euthanize a pup that reached our humane endpoints, since all the surviving pups were in generally good shape.
-
 - While significantly lighter than their breastfed controls, all pups increased their bodyweight during the experiment, attesting to the efficacy of our feeding regiment.
-
 - The pups were physically active and responded to changes in light and physical stimuli.
-
 - Skin color remained bright purple and fur development was visible throughout the experiment.
-
 - No signs of intestinal inflammation were evident from the visual inspection of intestines during sampling.
@@
 
-**Cytokines**
+```{r exp1-cytokines, include=FALSE}
+#____Load cytokine data______ 
+cytokines <- read_excel("../data/processed/weight_development.xlsx")
 
-`r nrow(cytokines_1)` samples were available for analysis of the proinflammatory cytokines `r names_of_cytokines`.
-Concentrations were measured in 1-cm sections of proximal/distal ileum tissue using MSD Mesoscale Kits.
-Statistically significant changes in the expression of `r names_of_cytokines`.
-- Directionality of change varied, with alternating over- and under-expression of various cytokines.
-As can be seen in [*Cytokine plot of signif results* ] pups tended to group based on feeding-scheme, with both formula-fed groups being either suppressed or over-expressed together, compared with breastfed groups, irrespective of antibiotics treatment.
+#____ Separate into experimental rounds ____#
+preexp_1 <- pups %>% 
+  filter(exp_number == "1")
 
--   This observation is supported by a a multiple linear regression analysis, evaluating the effects of MAT (AB or CON) and type of feed (BF of FF) separately and interactively.
-The analysis found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val\$IL-6.
-
-#### Commentary and interpretation of results - Exp 1.
-
--   In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers.
--   This is surprising given the findings of Chen et al., who used an identical protocol to our AB-BF vs. CON-BF (Maternal antibiotics treatment and weening with the mother for 7 days).
--   They claim that MAT during pregnancy resulted in significant alterations in the intestinal microbiota.
-lower body weight at birth and on day 7, visible intestinal injury and inflammation and disruption of the intestinal barrier.
--   Our initial analysis of weight and cytokine expression did not indicate a similar effect of MAT (evaluations of gut permeability, intestinal gene expression, histological analysis and gut microbial composition are conducted in later experiments).
--   One possible explanation is an effect of dose.
-While Chen et al., used a concentration of 1 g/l, we, unfortunately, tested 0.5 g/l, based on experience from previous studies, making a direct comparison difficult.
--   We did find significant over-expression of cytokines relevant in inflammation, indicating some effect of the treatments.
--   Since these biochemical signals were not strong enough to manifest phenotypically, we reasoned that our treatment wasn't sufficiently injurious.
--   We therefore decided to increase the dosage to 1 g/l for subsequent experiments, as well as to increase the osmolality of the formula for the next feeding experiment (Exp 3).
--   With assurance in our experimental setup we decided to include further measures of intestinal dysfunction.
--   We also decided to measure the birth weight of all pups, in addition to body weight at the beginning of the experiment (day 3) and end (day 5).
-
-### Experiment 2: Effects of various types of antibiotics - (Microbiome dysbiosis)
-
-#### Rationale:
-
--   Exp 1 demonstrated that the type of feed (breastmilk or formula) had a greater effect on cytokine expression than maternal antibiotics treatment.
--   Central to our hypothesis of NEC development is an intestinal microbiome dysibiosis, and a broad-spectrered antibiotic treatment potentially resulted in an overall depletion rather than the manifisation of a dysbiosis.
--   We hypothesized that targeting specific strains of bacteria in the mother would lead to a more dysbiotic microbiome and allow for greater subsequent dysbiosis in the offspring.
-
-#### Aim:
-
--   To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics, without formula feeding on intestinal permeability and cytokine expression.
--   To see whether a larger dose of antibiotics (1 g/l) causes decreased body weight at birth and at day 5 compared with non-treated controls.
-
-#### Conclusion:
-
--   Birth weight: `r mean(bodyweight_birth_con` g (CON) vs `r mean(bodyweight_birth_mat` g (MAT)
--   Visual inspection: no clear signs of intestinal inflammation
--   Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
-
-#### Detailed description of Exp 2.
-
-*Antimals and experimental setup* - The breeding setup in Exp 2 was identical to Exp 1.
-- On gestational day 15 vancomycin, gentamicin or control (tab water) was added to the drinking water pregnant females, forming 3 experimental groups; VANCO, GENTA and CON.
-
-*Weight development* - We checked the cages twice daily from gestational day 18 and measured the weight of newly-born pups.
-- VANCO and GENTA mothers gave birth on the 25th of february (getational day 20) with mean offspring body weights of `r mean(bodyweight_birth_vanco` (N = `r nrow(bodyweight_birth_vanco`) and `r mean(bodyweight_birth_genta` (N = `r nrow(bodyweight_birth_vanco`) respectively.
-- The CON mother gave birth on the 26th of february (gestational day 21) with mean offspring body weight of `r mean(bodyweight_birth_con` (N = `r nrow(bodyweight_birth_vanco`).
-- This difference in body weight at birth was statistically significant (p = `r lm_weight_exp2$p.value`), but adjusting for length of gestation erased the statistical significance (p = `r lm_weight_exp2_adj$p.value`).
-- On day 3 and 5 this difference was in body weight was no longer detectable.
-*(Figure of body weights in exp 2)*
-
-*Intestinal permeability - FITC-Dextran administration* - On the day of euthanasia pups were separated from the mothers at 07:00 (halting breastfeeding).
-- At 10:00, 50 µL FITC-dextran (0.045 mg/µL),for a final dose of 750 mg/kg, was administered orally using a PICC line and a 1 mL syringe.
-- Two hours later euthanasia and sample collection began (12:00).
-- Pups were euthanized by cervical dislocation and blood samples collected and spun in Eppendorf tubes.
-- FITC molecules were detectable in all groups.
-- The highest concentration was found in the GENTA group (`r fitc_exp2_genta` µg/ml) and the lowest in the VANCO group (`r fitc_exp2_vanco` µg/ml), respectively, with the CON group in between (`r fitc_exp2_vanco` µg/ml).
-- This difference was statistically significant (p = `r anova_fitc_exp2`, one-way analysis of variance).
-*(Figure of FITC in exp 2)*
-
-*Cytokine expression* - Tissue weight-adjusted results of cytokine concentrations showed significant differences between the 3 groups for `r signif_cyto_exp2`.
-### Rerun these experiments.
-
-#### Commentary and interpretation experimental results - Exp 2.
-
--   Weight development was now similar to the findings of Chen et al.
-
--   Whether it was a result of growth restriction or shorter gestational age is difficult to determine.
-
--   Chen et al. did not report whether dams were from the same day.
-
--   Intestinal permeability response was different depending of type of antibiotics.
-
--   Vancomycin decreased intestinal permeability whereas gentamicin increased it.
-
-    -   This has been shown in Winstar rats before (Thulstrup 2015).
-
--   Cytokine expression was most affected in the VANCO group.
-
--   In summary:
-
--   In experiment 2, Maternal antibiotics treatment did not have an effect.
-
--   This indicates that the greater effect is of formula-feeding, similar to the findings in exp 1.
-
--   The intestinal permeability assay is interesting in that it supports the findings of Thulstrup et al. 2015.
-
-### Experiment 3 - Interaction effects between antibiotics treatment and formula feeding
-
-#### Rationale:
-
-Based on the results from Exp 2 it was difficult to ascertain which treatment was more likely to manifest in NEC like phenotype.
-- Gentamicin had increased intestinal permeability the most.
-- However, the greatest effect of cytokine expression was found with the vancomycin treatment.
-- In addition to this divergence, the interaction effect found in Exp 1 suggested it was likely that combining MAT with formula feeding would yield different results.
-- Therefore, we decided to increase the breeding population to allow for an experiment that combined both types of antibiotics with either formula-feeding or breastfeeding, for a total of 6 experimental groups.
-
-#### Aim:
-
--   Repeat Exp 2 with greater sample sizes, adding the effect of formula feeding.
--   To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics in combination with formula feeding on intestinal permeability and cytokine expression.
--
-
-#### Conclusion:
-
--   Combining MAT with formula feeding resulted in more pronounced signs of inflammation than MAT alone.
--   VANCO-FF and VANCO-BF had the highest and lowest intestinal permeability scores, respectively.
--   Significantly altered expression of XXSignificantSCYTOKINESXX.
--   Greatest determinant of variation was formula-feeding.
-
-#### Detailed description of Exp 3.
-
-*Antimals and experimental setup* - To increase the number of pups available we doubled the number of mice for breeding, 36 females and 12 males.
-- Pregnant females were again given either Vancomycin, Gentamicin (1 g/l) or sterile water on gestational day 15.
-- `r nrow(preexp_3` pups were born and separated into breastfeeding or formula feeding, for a total of 6 groups: - VANCO-BF, VANCO-FF, GENTA-BF, GENTA-FF, CON-BF and CON-FF.
-
-*Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
-- Formula-fed pups consistently showed higher levels of Serum FITC concentration compared to breastfed pups, irrespective of MAT status.
-- VANCO-FF and VANCO-BF had the highest and lowest levels of FITC, respectively.
-
-*Cytokine expression* - We found significant differences in concentrations of XXSignificantSCYTOKINESXX.
-Linear regression analysis indicated the biggest determinant of differences in outcomes was formula feeding, with the type of antibiotic being of lesser importance compared.
-- Directionally the trend was the same for all cytokines.
-- IL-1b: Non-significant decrease in all antibiotics-treated offspring, irrespective of feeding.
-- IL-2: significantly lower concentrations in all formula-fed pups compared with CON-BF.
-- IL-6: Significantly greater levels in all formula-fed pups compared with CON-BF.
-- INFg: significantly lower concentrations in all formula-fed pups compared with CON-BF.
-- KCGRO: Significantly greater levels in all formula-fed pups compared with CON-BF.
-
-#### Commentary and interpretation experimental results Exp 3:
-
--   Intestinal permeability was by far the greatest in VANCO-FF, with levels of FITC-dex close to 10-fold increased in this group.
--   In contrast, VANCO-BF was the group with the lowest measure of intestinal permeability, similar to the finding in exp 2.
--   Looking at **FIGURE CYTOKINES ROUND 3** shows how formula-fed groups systematically over- or underexpress compared to breastfed analimals, irrespective of type of antibiotic, indicating that the biggest determinant of cytokine variation was again formula feeding.
-
-*Adjusting for weight of tissue* - In the previous experiments, cytokine concentrations were adjusted for the weight of the tissue analyzed, to control for biopsy sizes.
-- We did so because the average weight of tissue samples was not meaningfully different between groups.
-- This assumption wasn't met in Exp 3, where samples from formula-fed pups weighted on average XX%XX more (p = XXp.valueXX) than breastfed controls.
-- Vanco-FF had the heaviest ileum tissue samples of all **mean + P (if its true)**.
-- Tissue edema could well be indicative of inflammation, and hence desirable, however it introduces a bias against the formula-fed pups if cytokine concentrations were normalized to weight.
-- The reported findings are no longer significant **Table of significant cytokines** when adjusting for weight.
-
--   This phenomenon is a common challenge in animal disease models, requiring measure of normalization other that tissue weight, such as total protein content.
--   Unfortunately this discovery was done after tissue homogenization and analysis, making it too late for further analysis.
-
-*Combining results from exp 2 and exp 3* - The protocols for the breastfed groups were identical across experiment 2 and 3, allowing for pooling of these results.
-- While analyzing the results from both groups together increased the group sizes and degrees of statistical freedom, it did not change the clinical conclusions.
-
-### Experiment 4 - FVT Trial
-
-#### Rationale: **What was the actual rationale?**
-
--   Central to the development of NEC is an assumption of a gut microbial dysbiosis, which we tried to mimick by administering antibiotics prenatally.
--   Results from exp 1-3 indicated that maternal antibiotics treatment had little effect in the context of breastfeeding, but when combined with formula-feeding was able to induce mild symptoms of intestinal inflammation.
--   It was unclear whether this was solely an effect of the formula, or whether an interaction with the maternal antibiotics treatment existed.
--   If our findings were partly due to a successful manipulation of the gut microbiome, then eliminating the induced disbyosis should also partly mitigate the inflammatory signatures.
--   Relative bacterial abundance / bacterial composition is partly a function of viral (bacteriophage) abundance, and sevaral studies have shown how a FVT can alter the bacterial composition.
--   We therefore decided to conduct a final experiment, in which we administered an FVT to a vancomycing-treated, formula-fed group (since this combination (VANCO-FF) had the biggest effects in terms of intestinal permeability and proinflammatory cytokine expression in Exp 3).
--   We hypothesized that any dysbiosis created by maternal antibiotics treatment could be mitigated with a follow-up administration of FVT prior to beginnig formula-feeding.
--   If successful, the FVT-treated group would show fever intestinal injury scores compared with a non-FVT treated group.
--   Previous experiments had shown that MAT in combination with breastfeeding had no effect on inflammatory markers and VANCO-BF thus acted as a healthy control to both formula-fed treatment groups.
-**(The problem here is that formula is the only thing with an effect.
-A breastfed control only shows the effect of the formula)**
-
-#### Aim:
-
--   To test whether signs of inflammation are reduced in pups treated with an FVT.
--   To test whether any reductions in symptoms can be explained with a manipulation of the microbial composition.
-
-#### Conclusion:
-
--   FVT treated pups showed similar intestinal permeability and pro-inflammatory cytokine concentrations as non-FVT-treated controls.
--   An `r fold_increase_reg3g`-fold increase in expression of the `r REG3G` gene was found in the FVT group.
-
-#### Detailed description of Exp 4.
-
--   For the final experiment we repeated the previous breeding setup, assigning 36 females and 12 males for breeding.
-
--   Only 5 of the 36 females became pregnant, and all were assigned to vancomycin (1 g/l) from pregnancy-day 15 until delivery, giving birth to `r nrow(preexp_4` pups on Saturday the XXth.
-
--   2 days after birth pups were separated into breastfeeding or formula feeding.
-
--   The formula-fed pups were further divided into a treatment and control group, receiving either 30 µl fecal virome filtrate (FVT) (0.01 mg farces/µL) or an equal amount of SM buffer (control) 1 hour prior to the first feeding.
-
--   We repeated the FVT administration 24 hours later (09:00 day 2) for a total dose of 0.6 mg feces material per pup.
-
--   2 hours after FITC administtion pups were euthanized by decapitation and blood collected in eppendorfs for analysis of serum FITC concentration.
-    Intestinal samples were collected for gene expression assays, histological assessment, pro-inflammatory cytokine analysis and gut microbiome sequencing.
-
-*Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
-- The FVT and FVT-control (SM) had similar permeability scores.
-- Compared with the vancomycin-treated formula-fed pups from experiment 3, there were no significant differences **Figure X**.
-- The breastfed pups did, however, have significantly higher levels compared with all other breastfed pups.
-
-*Cytokine expression* - We found significant differences in concentrations of TNFa in Exp 4, with greater level in the FVT-treated group (XXX), compared to both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively).
-- This is in contrast experiment 3, where differences between VANCO-BF and VANCO-FF found significant differences for XXXCYtokinesXXX.
-OR: - This is similar to the findings of experiment 3, where no differences between VANCO-BF and VANCO-FF were found either.
-
-*Gene expression* - qPCR analyzis of intestinal samples found statistically significant differences in the level of XXXSIGNIFICANT_GENESXXX expression between groups with the delta-delta method (reformulate).
-- However, a difference in expression of the reference gene between the FVT and breastfed control was observed.
-this difference was statistically significant, invalidating the assumption of *equal baseline expression.*(reformulate) - Comparing the groups on the basis of (SOMEHTING ELSE THAN DELTA DELTA) found a `r fold_increase_reg3g`-fold increased expression of the `r REG3G` gene in the FVT groups.
-
-#### Commentary and interpretation
-
--   FVT-treatmed did not seem to change the intestinal permeability.
-
--   It is possible that something went wrong with the VANCO-BF group, given FITC-concentration was `r mean(fitc_level_exp4$bf`, which is 4-5 times higher than VANCO-BF groups on identical protocols in Exp 2 `r mean(fitc_level_exp2$bf` and 3 `r mean(fitc_level_exp3$bf`.
-Compared with theses levels the development looks more biologically reasonable.
-**FIGURE of EXP4 with DIFFERENT CONTROL GROUP**
-
--   The level of pro-inflammatory cytokines were the same in the FVT and SM groups, again indicating a lack of efficacy of the FVT treatment.
-
--   Summary: The FVT treatment did not seem to have an effect.
-
-### \--\>Post experimental analyses and Discussion of results
-
--   Histological image analysis was performed after all experiments had finished.
-
--   We inspected the morphology of the ileum and colon, and visually assessed thickness of the mucin layers.
-
--   Analyzing AB-PAS stains for mucin-stained tissue to the total area, showed no differences between any treatment groups.
-
--   Formula-fed and breastfed pups were slightly different, irrespective of antibiotics treatment, whereas there were no observable differences between antibiotic treatments.
-
--   These findings supports the visual inspections of intestines made during sampling, where no signs of inflammation were detected.
-
--   *Images of intestines*
-
--   Gut microbiome composition
-
-## Discussion
+exp1_survived <- preexp_1 %>% 
+  filter(!is.na(bodyweight_baseline_plus_48h)) %>% 
+  nrow()
+#____ ____#
+```

### 2022-11-04 submission/thesis.Rmd
Created thesis structure

---
title: " "
output:
  pdf_document:
    number_sections: TRUE
geometry: "left = 2.5cm, right = 2cm, top = 2cm, bottom = 2cm"
fontsize: 12pt
mainfont: Arial 
header-includes:
  - \usepackage{float}
  - \usepackage{sectsty}
  - \usepackage{paralist}
  - \usepackage{setspace}\spacing{1.5}
  - \usepackage{fancyhdr}
  - \usepackage{lastpage}
  - \usepackage{dcolumn}
  - \usepackage{natbib}\bibliographystyle{agsm}
  - \usepackage[nottoc, numbib]{tocbibind}
  #- \usepackage[english, sund, titlepage, hyperref]{ku-frontpage} (Consider using  #extra_dependencies)
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(tinytex.verbose = TRUE)
```

\allsectionsfont{\centering}
\subsectionfont{\raggedright}
\subsubsectionfont{\raggedright}

\pagenumbering{gobble}

```{r uni_logo, echo=F, out.width="95%"}
knitr::include_graphics("img/sund.pdf")
```

\newpage

\begin{centering}

\vspace{1cm}

\Large
{\bf The University Of Edinburgh}

\Large
{\bf School Of Geosciences}

\vspace{1cm}

\Large

\doublespacing
{\bf COMPARISON OF TOP-DOWN AND BOTTOM-UP APPROACHES ON SPECIFIC LEAF AREA PATTERNS, \\AT GLOBAL, LATITUDINAL, AND BIOME SCALES}

\vspace{1 cm}

\normalsize
\singlespacing
By

\vspace{0.5 cm}

\Large

{\bf ANNA CHIRUMBOLO}

\vspace{1.5 cm}

in partial fulfilment of the requirement \\for the degree of BSc with Honours \\in Ecological and Environmental Sciences

\vspace{1.5 cm}

\normalsize
mm yy

\end{centering}

\newpage

\pagenumbering{gobble}

\begin{centering}

{\bf Abstract}

\end{centering}

\spacing{1.5}

(the spacing is set to 1.5)

no more than 250 words for the abstract

- a description of the research question/knowledge gap – what we know and what we don’t know
- how your research has attempted to fill this gap
- a brief description of the methods
- brief results
- key conclusions that put the research into a larger context

\pagenumbering{roman}

\newpage

\centering
\raggedright
\newpage
\tableofcontents

\newpage

```{r acknowledgments, child='acknowledgments.Rmd'}
```

\newpage

```{r intro, child = 'introduction.Rmd'}
```
\pagenumbering{arabic}

\newpage

```{r methods, child = 'methods.Rmd'}
```

\newpage

```{r results, child= 'results.Rmd'}
```

\newpage

```{r discussion, child = 'discussion.Rmd'}
```

\newpage

```{r conclusion, child = 'conclusion.Rmd'}
```

\newpage

```{r biblio, child = 'bibliography.Rmd'}
```

\newpage

```{r appendix, child = 'appendix.Rmd'}
```

### 2022-11-07 code/Round 1/Cytokines adjusted.Rmd
Progress

@@
 date: "11th October 2022"    
-output: html_document
+output: pdf_document
+editor_options: 
+  chunk_output_type: console
 ---
@@
 # Import dataset
-cytokines <- read_excel("../../Data/RawData/cytokines_round_1_r.xlsm")
+cytokines <- read_excel("../../data/processed/cytokines.xlsx")
 glimpse(cytokines)
 
-cytokines <- cytokines %>% 
+cytokines_adjusted <- cytokines %>% 
   pivot_longer(.,
@@
   ) %>% 
-  drop_na(
-    values) %>% 
-  mutate(MAT = ifelse(str_detect(group,'AB'),'AB',
-                                  ifelse(str_detect(group,'CON'),'CON', "miss")),
-         FEED = ifelse(str_detect(group,'FF'),'FF',
-                                  ifelse(str_detect(group,'BF'),'BF', "miss"))) %>% 
   mutate(
@@
       )) %>% 
-  mutate(values = if_else(tissue == "Illeum", values/tissue_weight_mg, values)) %>% 
-  pivot_wider(id_cols = c(sample,group,tissue,tissue_weight_mg, MAT,FEED), names_from = variables, values_from = values)
+  mutate(values = if_else(tissue == "ileum", values/tissue_weight_mg, values)) %>%
+  pivot_wider(
+    id_cols = c(
+      sample_id,exp_number, group, received_antibiotics,maternal_treatment, tissue,tissue_weight_mg),
+    names_from = variables, 
+    values_from = values)
+
+#____Pivot longer____
+cytokines_long <- cytokines_adjusted %>% 
+  pivot_longer(.,
+               cols = INFg:TNFa ,
+               names_to = "variables",
+               values_to = "values"
+  )
+#pivot_wider(id_cols = c(sample_id,exp_number, group, received_antibiotics,maternal_treatment, #tissue,tissue_weight_mg, adjusted_values), names_from = variables, values_from = values)
+```
+
+Check for data distribution.
+
+```{r visualize data, eval=FALSE, include=FALSE}
+#normality_check 
+shapiro_failed <- cytokines_long %>% 
+  filter(exp_number == 1) %>% 
+  filter(tissue == "ileum") %>%
+  filter(variables != "INFg") %>% 
+  drop_na(values) %>% 
+  group_by(variables, group) %>% 
+  shapiro_test(values) %>% 
+  filter(p < 0.05)
+
+log10_shapiro_failed <- cytokines_long %>% 
+  filter(exp_number == 1) %>% 
+  filter(tissue == "ileum") %>%
+  filter(variables != "INFg") %>% 
+  drop_na(values) %>% 
+  mutate(values = log10(values)) %>% 
+  group_by(variables, group) %>% 
+  shapiro_test(values) %>% 
+  filter(p < 0.05)
+
+#Density plots. Works
+histogram <- cytokines_long %>% 
+  filter(exp_number == 1) %>% 
+  filter(tissue == "ileum") %>%
+  filter(variables != "INFg") %>% 
+  drop_na(adjusted_values) %>% 
+  ggplot(aes(x = adjusted_values, fill = group)) + 
+  geom_density() +
+  facet_wrap(~ variables,
+             scales = "free")
+
+log10_histogram <- cytokines_long %>% 
+  filter(exp_number == 1) %>% 
+  filter(tissue == "ileum") %>%
+  filter(variables != "INFg") %>% 
+  drop_na(adjusted_values) %>% 
+  ggplot(aes(x = log10(adjusted_values), fill = group)) + 
+  geom_density() +
+  facet_wrap(~ variables,
+             scales = "free")
+
+#Jitter plots. Works
+jitter_plot <- cytokines_long %>% 
+  filter(exp_number == 1) %>% 
+  filter(tissue == "ileum") %>%
+  #filter(variables != "INFg") %>% 
+  drop_na(adjusted_values) %>% 
+  ggplot(aes(x = group, y = adjusted_values, color = group)) +
+  geom_jitter() +
+  facet_wrap(~ variables,
+             scales = "free")
+
+log10_jitter_plot <- cytokines_long %>% 
+  filter(exp_number == 1) %>% 
+  filter(tissue == "ileum") %>%
+  filter(variables != "INFg") %>% 
+  drop_na(adjusted_values) %>% 
+  mutate(adjusted_values = log10(adjusted_values)) %>% 
+  ggplot(aes(x = group, y = adjusted_values, color = group)) +
+  geom_jitter() +
+  facet_wrap(~ variables,
+             scales = "free") 
 ```
@@
 #set tissue type
-sample_tissue <- "Illeum"
+sample_tissue <- "ileum"
 ```
@@
 An appropriate statistical test for differences will be performed (student-t for two, normally distributed groups, Anova for more than 2 groups, Kruskal wallis for non-normally distrubuted results.)
-```{r}
+
+```{r advanced-cytokine-table}
+#_____Currently non-normally distributed values are plotted as median (IQR)____# 
+ #change statistics with help from here: https://www.danieldsjoberg.com/gtsummary/reference/tests.html 
+#  add_p(everything() ~ "aov")  assuming equal variance
+
 tbl1 <- cytokines %>% 
-  filter(tissue == sample_tissue) %>% 
+  filter(exp_number == "1") %>% 
+  filter(tissue == "ileum") %>% 
   select(group,INFg:TNFa) %>% 
@@
               missing = "no",
-              include = !INFg) %>%
+              include = !"INFg",
+             # statistic = list(all_continuous() ~ "{mean} ({sd})")  #Un-comment this line to get mean+sd. 
+             #%>%
+statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
   add_p() %>% 
+  #modify_footnote("Exp. 1") %>% 
   modify_header(
-    label ~ "**Groups**")
+    label ~ "Cytokines") %>% 
+  modify_spanning_header((c("stat_1", "stat_2", "stat_3", "stat_4") ~ "Treatment groups (N total = {N})"))
 
-tbl1
-# Save and export files
-# tbl1 %>%  #Save as excel file.
-#   as_hux_table() %>%
-#   huxtable::quick_xlsx(., file = "../Results/Round 1/ileum-cytokines.xlsx")
-# 
-# tbl1 %>% #Save as image file. 
-# as_flex_table() %>%
-#   flextable::save_as_image(path = "../../Results/Round 1/ileum_cytokines_table.png",webshot = "webshot")
+  tbl1_gt <- tbl1 %>% 
+  as_gt() %>%
+  #tab_header(title = md("**Proinflammatory cytokines**")) %>%
+  opt_align_table_header("center") %>% 
+  # tab_options(heading.title.font.size = 20,
+  #             heading.background.color = "lightgrey",
+  #             ) %>% 
+  tab_source_note(
+    source_note = " * P-value indicates overall group differences"
+  ) %>%
+  tab_source_note(
+    source_note = md("**Matrinem round 1**")
+  ) %>%
+  cols_width(label ~ px(100),
+             stat_1 ~ px(150),
+             stat_2 ~ px(150),
+             stat_3 ~ px(150),
+             stat_4 ~ px(150)) 
+tbl1_gt
+
 ```
@@
 First the data must be transformed to a long format (pivot longer). Mutate any new columns  with group treatments needed for the regression model (in this case MAT and FEED (experimental factors)).
-```{r long format, echo=FALSE}
-cytokines_long <- cytokines %>% 
-  pivot_longer(.,
-               cols = INFg:TNFa ,
-               names_to = "variables",
-               values_to = "values"
-  ) %>% 
-  drop_na(
-    values) %>% 
-  mutate(MAT = ifelse(str_detect(group,'AB'),'AB',
-                                  ifelse(str_detect(group,'CON'),'CON', "miss")),
-         FEED = ifelse(str_detect(group,'FF'),'FF',
-                                  ifelse(str_detect(group,'BF'),'BF', "miss"))) %>% 
-  mutate(
-    group = recode_factor(  #Change the order of the groups to desired plot order
-      group,
-      "CON-BF"= "CON-BF",
-      "CON-FF" = "CON-FF",
-      "AB-BF"= "AB-BF",
-      "AB-FF"= "AB-FF"
-      ))
-
-head(cytokines_long)
-
-```
 
@@
 ```{r plots all cytokines, warning=FALSE}
+
 plot_all_cytokines <- cytokines_long %>%
+  filter(exp_number == "1") %>% 
   filter(tissue == sample_tissue) %>% 
-  filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot
+  #filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot
 ggplot(aes(x = group, 
@@
              scales = 'free', 
-             nrow = 3,
-             ncol = 3) +
+             nrow = 5,
+             ncol = 2) +
   labs(title = "Levels of pro-inflammatory cytokines in the ileum",  # Change title, and axis names
@@
     axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
-    axis.text.y = element_blank(),
+    #axis.text.y = element_blank(),
     axis.ticks.x = element_blank(),  # Hide labels
@@
   scale_fill_brewer(palette="Set1")
+
+ggsave(filename = "exp1_cytokines.png",width = 2.14, height = 1.51, dpi = 300)
+
+ggsave('../../exports/exp1_cytokines.png',
+       plot_all_cytokines, scale =  width = 10, height = 16, units = "cm")
 

### 2022-11-07 submission/results backup copy.Rmd
Progress

@@
 ```
-The main goal of this thesis project was to create a NEC mouse model by combining Maternal Antibiotics Treatment with Formula-Feeding.
+The aim of this project was to establish a NEC mouse model by combining Maternal Antibiotics Treatment with Formula-Feeding.
 This hypothesis was tested in a series of experiments (Exp 1-4), each successive experiment building on the findings from the former.
@@
 
-*Antimals and experimental setup* \n 
-- The experimental setup has been described in detail in the "Animal experimental setup" methods section. Detials specific for Exp-1 are outlined below. \n
+*Animals and experimental setup* \n 
+The general experimental setup has been described in the "Animal experimental setup" methods section. Details specific for Exp-1 are outlined below. \n
 - Six male and 18 female C57BL/6N mice were co-housed for 24 hours in 6 cages (1 male + 3 female each).
 - Two weeks after mating, pregnant females were separated evenly into maternal antibiotic treatment (MAT) and water-only controls (CON).
-- Antibiotics (Gentamicin, Vancomycin, Neomycin, and Ampicillin (all 0.5 g/L)) were mixed into the drinking water of the MAT mice while controls received water only.
+- Antibiotics (Gentamicin, Vancomycin, Neomycin, and Ampicillin (all 0.5 g/L)) were mixed into the drinking water of the MAT mice while untreated controls received regular tab water.
 - Antibiotic treatment commenced on gestational day 15 and continued until delivery.
 - Each mother was housed with her offspring in separate cages during the first 2 days after birth.
--   On day 3, mice from water or antibiotics treated mothers were separated and assigned to either formula feeding (FF) or breastfeeding (BF) forming four experimental groups: AB + formula-feeding (AB-FF, N = `r AB_FF` ) AB + breastfeeding (AB-BF, N = `r AB_BF` ), Water + formula-feeding (CON-FF, N = `r CON_FF` ), and CON + breastfeeding (CON-BF, N = `r CON_BF` ).
--   Formula-feeding was performed as per METHODS AND MATERIALs.
-    In Exp-1 we administered 0.08 ml formula for pups \>2g, and 0.06 for pups ≤ 2g every 3 hours throughout the entire study.
-  In later experiments this dose was changed, and feeding intervals were increased to every 4 hours between midnight and 08 AM. \n
-- `r exp1_survived` of `r nrow(preexp_1)` pups survived the treatments.
-- Those that did not survive were euthanize because of esophageal perforation during feeding.
-- After fine-tuning of the feeding methods we were able to feed the pups without further harm.
-- We never had to euthanize a pup that reached our humane endpoints, since all the surviving pups were in generally good shape.
-- While significantly lighter than their breastfed controls, all pups increased their bodyweight during the experiment, attesting to the efficacy of our feeding regiment.
-- The pups were physically active and responded to changes in light and physical stimuli.
-- Skin color remained bright purple and fur development was visible throughout the experiment.
-- No signs of intestinal inflammation were evident from the visual inspection of intestines during sampling.
-Our subjective observation of the intestines were small degrees of bloating in formula-fed pups.
+- On day 3, mice from water or antibiotics treated mothers were separated and assigned to either formula feeding (FF) or breastfeeding (BF) forming four experimental groups:  
+-  AB + formula-feeding (AB-FF, N = `r AB_FF` )  
+-  AB + breastfeeding (AB-BF, N = `r AB_BF` )  
+-  Water + formula-feeding (CON-FF, N = `r CON_FF` )
+-  Water + breastfeeding (CON-BF, N = `r CON_BF` ).
+- Formula-feeding was performed as per (LINK to feeding in methods). In Exp-1 we administered 0.08 ml formula for pups \>2g, and 0.06 for pups ≤2g every 3 hours throughout the entire study.
+- In later experiments this dose was changed, and feeding intervals were increased to every 4 hours between midnight and 08 AM.
+
+*Survival and weight development*
+The experimental setup, including the feeding method, was a general success. 
+While we were fine-tuning the feeding method we lost `r (nrow(preexp_1))-exp1_survived` pups to esophageal perforation with the PICC line.
+After improving the feeding method we were able to maintain or increase body weight among the formula-fed pups, albeit significantly lighter than their breastfed controls (see *Exp-1 figure*).
+No pup required euthanasia (see *humane endpoints*).
+In fact pups were physically active and responded to changes in both light and physical stimuli.
+Skin color remained bright purple and fur development was visible throughout the experiment.
+In total, `r exp1_survived` of the `r nrow(preexp_1)`born survived the 48-hour treatment and were euthanised by cervical dislocation for sampling of blood and intestinal tissue.
 
@@
 #____ Separate into experimental rounds ____#
-preexp_1 <- pups %>% 
-  filter(exp_number == "1")
 
-exp1_survived <- preexp_1 %>% 
-  filter(!is.na(bodyweight_baseline_plus_48h)) %>% 
-  nrow()
 #____ ____#
@@
 
-<!-- *Cytokines* \n -->
-<!-- `r nrow(cytokines_1)` samples were available for analysis of the proinflammatory cytokines `r names_of_cytokines`. -->
-<!-- Concentrations were measured in 1-cm sections of proximal/distal ileum tissue using MSD Mesoscale Kits. -->
-<!-- - Directionality of change varied, with alternating over- and under-expression of various cytokines. -->
-<!-- As can be seen in [*Cytokine plot of signif results* ] pups tended to group based on feeding-scheme, with both formula-fed groups being either suppressed or over-expressed together, compared with breastfed groups, irrespective of antibiotics treatment.
--->
-<!-- Statistically significant changes in the expression of `r names_of_cytokines`. -->
+*Cytokines* \n
+No signs of intestinal inflammation were evident from  a visual inspection of intestines during sampling. 
+To investigate furhter we analysed 1-cm segments of the distal ilium for concentrations of the proinflammatory cytokines `r names_of_cytokines`.
+Although statistically significant between-group changes were observed (see *Exp 1 figure, B*), we did not see an overall increase in inflammatory markers in the treatment groups.
+Multiple linear regression analysis (MAT (AB or CON) and type of feed (BF of FF) separately and interactively found statistically significant effects of formula feeding for
 
-<!-- -   This observation is supported by a a multiple linear regression analysis, evaluating the effects of MAT (AB or CON) and type of feed (BF of FF) separately and interactively.
-The analysis found statistically significant effects of formula feeding for REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val\$IL-6.
--->
+Directionality of change tended to group on feeding-scheme (BF vs FF), with formula-fed groups either over- or under-expressing relative to breast-fed controls.
+We did not se a similar trend for antibiotics treatment.
 
-<!-- #### Commentary and interpretation of results - Exp 1. -->
+ REG_MODEL_CYTOKINES$TERM == TYPEOFFEED, with the greatest effect estimate of ESTIMATE$IL-6 , P-val\$IL-6.
 
-<!-- -   In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers. -->
-<!-- -   This is surprising given the findings of Chen et al., who used an identical protocol to our AB-BF vs. CON-BF (Maternal antibiotics treatment and weening with the mother for 7 days). -->
-<!-- -   They claim that MAT during pregnancy resulted in significant alterations in the intestinal microbiota.
+#### Commentary and interpretation of results - Exp 1.
+
+-   In summary, formula-feeding seemed to have greater effect on the inflammatory state of the pups than whether they were weened by antibiotics treated mothers.
+-   This is surprising given the findings of Chen et al., who used an identical protocol to our AB-BF vs. CON-BF (Maternal antibiotics treatment and weening with the mother for 7 days).
+-   They claim that MAT during pregnancy resulted in significant alterations in the intestinal microbiota.
 lower body weight at birth and on day 7, visible intestinal injury and inflammation and disruption of the intestinal barrier.
--->
-<!-- -   Our initial analysis of weight and cytokine expression did not indicate a similar effect of MAT (evaluations of gut permeability, intestinal gene expression, histological analysis and gut microbial composition are conducted in later experiments).
--->
-<!-- -   One possible explanation is an effect of dose.
+-   Our initial analysis of weight and cytokine expression did not indicate a similar effect of MAT (evaluations of gut permeability, intestinal gene expression, histological analysis and gut microbial composition are conducted in later experiments).
+-   One possible explanation is an effect of dose.
 While Chen et al., used a concentration of 1 g/l, we, unfortunately, tested 0.5 g/l, based on experience from previous studies, making a direct comparison difficult.
--->
-<!-- -   We did find significant over-expression of cytokines relevant in inflammation, indicating some effect of the treatments. -->
-<!-- -   Since these biochemical signals were not strong enough to manifest phenotypically, we reasoned that our treatment wasn't sufficiently injurious. -->
-<!-- -   We therefore decided to increase the dosage to 1 g/l for subsequent experiments, as well as to increase the osmolality of the formula for the next feeding experiment (Exp 3). -->
-<!-- -   With assurance in our experimental setup we decided to include further measures of intestinal dysfunction. -->
-<!-- -   We also decided to measure the birth weight of all pups, in addition to body weight at the beginning of the experiment (day 3) and end (day 5). -->
+-   We did find significant over-expression of cytokines relevant in inflammation, indicating some effect of the treatments.
+-   Since these biochemical signals were not strong enough to manifest phenotypically, we reasoned that our treatment wasn't sufficiently injurious.
+-   We therefore decided to increase the dosage to 1 g/l for subsequent experiments, as well as to increase the osmolality of the formula for the next feeding experiment (Exp 3).
+-   With assurance in our experimental setup we decided to include further measures of intestinal dysfunction.
+-   We also decided to measure the birth weight of all pups, in addition to body weight at the beginning of the experiment (day 3) and end (day 5).
 
-<!-- ### Experiment 2: Effects of various types of antibiotics - (Microbiome dysbiosis) -->
+-- Inlcude graphics: *Exp-1 figure*
 
-<!-- #### Rationale: -->
+### Experiment 2: Effects of various types of antibiotics - (Microbiome dysbiosis)
 
-<!-- -   Exp 1 demonstrated that the type of feed (breastmilk or formula) had a greater effect on cytokine expression than maternal antibiotics treatment. -->
-<!-- -   Central to our hypothesis of NEC development is an intestinal microbiome dysibiosis, and a broad-spectrered antibiotic treatment potentially resulted in an overall depletion rather than the manifisation of a dysbiosis.
--->
-<!-- -   We hypothesized that targeting specific strains of bacteria in the mother would lead to a more dysbiotic microbiome and allow for greater subsequent dysbiosis in the offspring. -->
+#### Rationale:
 
-<!-- #### Aim: -->
+-   Exp 1 demonstrated that the type of feed (breastmilk or formula) had a greater effect on cytokine expression than maternal antibiotics treatment.
+-   Central to our hypothesis of NEC development is an intestinal microbiome dysibiosis, and a broad-spectrered antibiotic treatment potentially resulted in an overall depletion rather than the manifisation of a dysbiosis.
+-   We hypothesized that targeting specific strains of bacteria in the mother would lead to a more dysbiotic microbiome and allow for greater subsequent dysbiosis in the offspring.
 
-<!-- -   To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics, without formula feeding on intestinal permeability and cytokine expression. -->
-<!-- -   To see whether a larger dose of antibiotics (1 g/l) causes decreased body weight at birth and at day 5 compared with non-treated controls. -->
+#### Aim:
 
-<!-- #### Conclusion: -->
+-   To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics, without formula feeding on intestinal permeability and cytokine expression.
+-   To see whether a larger dose of antibiotics (1 g/l) causes decreased body weight at birth and at day 5 compared with non-treated controls.
 
-<!-- -   Birth weight: `r mean(bodyweight_birth_con` g (CON) vs `r mean(bodyweight_birth_mat` g (MAT) -->
-<!-- -   Visual inspection: no clear signs of intestinal inflammation -->
-<!-- -   Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES. -->
+#### Conclusion:
 
-<!-- #### Detailed description of Exp 2. -->
+-   Birth weight: `r mean(bodyweight_birth_con` g (CON) vs `r mean(bodyweight_birth_mat` g (MAT)
+-   Visual inspection: no clear signs of intestinal inflammation
+-   Cytokine expression: Statistically significant differences in the level of STATISTICAL-SIGNIF-CYTOKINE-NAMES.
 
-<!-- *Antimals and experimental setup* - The breeding setup in Exp 2 was identical to Exp 1. -->
-<!-- - On gestational day 15 vancomycin, gentamicin or control (tab water) was added to the drinking water pregnant females, forming 3 experimental groups; VANCO, GENTA and CON. -->
+#### Detailed description of Exp 2.
 
-<!-- *Weight development* - We checked the cages twice daily from gestational day 18 and measured the weight of newly-born pups. -->
-<!-- - VANCO and GENTA mothers gave birth on the 25th of february (getational day 20) with mean offspring body weights of `r mean(bodyweight_birth_vanco` (N = `r nrow(bodyweight_birth_vanco`) and `r mean(bodyweight_birth_genta` (N = `r nrow(bodyweight_birth_vanco`) respectively.
--->
-<!-- - The CON mother gave birth on the 26th of february (gestational day 21) with mean offspring body weight of `r mean(bodyweight_birth_con` (N = `r nrow(bodyweight_birth_vanco`). -->
-<!-- - This difference in body weight at birth was statistically significant (p = `r lm_weight_exp2$p.value`), but adjusting for length of gestation erased the statistical significance (p = `r lm_weight_exp2_adj$p.value`).
--->
-<!-- - On day 3 and 5 this difference was in body weight was no longer detectable. -->
-<!-- *(Figure of body weights in exp 2)* -->
+*Antimals and experimental setup* - The breeding setup in Exp 2 was identical to Exp 1.
+- On gestational day 15 vancomycin, gentamicin or control (tab water) was added to the drinking water pregnant females, forming 3 experimental groups; VANCO, GENTA and CON.
 
-<!-- *Intestinal permeability - FITC-Dextran administration* - On the day of euthanasia pups were separated from the mothers at 07:00 (halting breastfeeding). -->
-<!-- - At 10:00, 50 µL FITC-dextran (0.045 mg/µL),for a final dose of 750 mg/kg, was administered orally using a PICC line and a 1 mL syringe. -->
-<!-- - Two hours later euthanasia and sample collection began (12:00). -->
-<!-- - Pups were euthanized by cervical dislocation and blood samples collected and spun in Eppendorf tubes. -->
-<!-- - FITC molecules were detectable in all groups. -->
-<!-- - The highest concentration was found in the GENTA group (`r fitc_exp2_genta` µg/ml) and the lowest in the VANCO group (`r fitc_exp2_vanco` µg/ml), respectively, with the CON group in between (`r fitc_exp2_vanco` µg/ml).
--->
-<!-- - This difference was statistically significant (p = `r anova_fitc_exp2`, one-way analysis of variance). -->
-<!-- *(Figure of FITC in exp 2)* -->
+*Weight development* - We checked the cages twice daily from gestational day 18 and measured the weight of newly-born pups.
+- VANCO and GENTA mothers gave birth on the 25th of february (getational day 20) with mean offspring body weights of `r mean(bodyweight_birth_vanco` (N = `r nrow(bodyweight_birth_vanco`) and `r mean(bodyweight_birth_genta` (N = `r nrow(bodyweight_birth_vanco`) respectively.
+- The CON mother gave birth on the 26th of february (gestational day 21) with mean offspring body weight of `r mean(bodyweight_birth_con` (N = `r nrow(bodyweight_birth_vanco`).
+- This difference in body weight at birth was statistically significant (p = `r lm_weight_exp2$p.value`), but adjusting for length of gestation erased the statistical significance (p = `r lm_weight_exp2_adj$p.value`).
+- On day 3 and 5 this difference was in body weight was no longer detectable.
+*(Figure of body weights in exp 2)*
 
-<!-- *Cytokine expression* - Tissue weight-adjusted results of cytokine concentrations showed significant differences between the 3 groups for `r signif_cyto_exp2`. -->
-<!-- ### Rerun these experiments. -->
+*Intestinal permeability - FITC-Dextran administration* - On the day of euthanasia pups were separated from the mothers at 07:00 (halting breastfeeding).
+- At 10:00, 50 µL FITC-dextran (0.045 mg/µL),for a final dose of 750 mg/kg, was administered orally using a PICC line and a 1 mL syringe.
+- Two hours later euthanasia and sample collection began (12:00).
+- Pups were euthanized by cervical dislocation and blood samples collected and spun in Eppendorf tubes.
+- FITC molecules were detectable in all groups.
+- The highest concentration was found in the GENTA group (`r fitc_exp2_genta` µg/ml) and the lowest in the VANCO group (`r fitc_exp2_vanco` µg/ml), respectively, with the CON group in between (`r fitc_exp2_vanco` µg/ml).
+- This difference was statistically significant (p = `r anova_fitc_exp2`, one-way analysis of variance).
+*(Figure of FITC in exp 2)*
 
-<!-- #### Commentary and interpretation experimental results - Exp 2. -->
+*Cytokine expression* - Tissue weight-adjusted results of cytokine concentrations showed significant differences between the 3 groups for `r signif_cyto_exp2`.
+### Rerun these experiments.
 
-<!-- -   Weight development was now similar to the findings of Chen et al. -->
+#### Commentary and interpretation experimental results - Exp 2.
 
-<!-- -   Whether it was a result of growth restriction or shorter gestational age is difficult to determine. -->
+-   Weight development was now similar to the findings of Chen et al.
 
-<!-- -   Chen et al. did not report whether dams were from the same day. -->
+-   Whether it was a result of growth restriction or shorter gestational age is difficult to determine.
 
-<!-- -   Intestinal permeability response was different depending of type of antibiotics. -->
+-   Chen et al. did not report whether dams were from the same day.
 
-<!-- -   Vancomycin decreased intestinal permeability whereas gentamicin increased it. -->
+-   Intestinal permeability response was different depending of type of antibiotics.
 
-<!--     -   This has been shown in Winstar rats before (Thulstrup 2015). -->
+-   Vancomycin decreased intestinal permeability whereas gentamicin increased it.
 
-<!-- -   Cytokine expression was most affected in the VANCO group. -->
+    -   This has been shown in Winstar rats before (Thulstrup 2015).
 
-<!-- -   In summary: -->
+-   Cytokine expression was most affected in the VANCO group.
 
-<!-- -   In experiment 2, Maternal antibiotics treatment did not have an effect. -->
+-   In summary:
 
-<!-- -   This indicates that the greater effect is of formula-feeding, similar to the findings in exp 1. -->
+-   In experiment 2, Maternal antibiotics treatment did not have an effect.
 
-<!-- -   The intestinal permeability assay is interesting in that it supports the findings of Thulstrup et al. 2015. -->
+-   This indicates that the greater effect is of formula-feeding, similar to the findings in exp 1.
 
-<!-- ### Experiment 3 - Interaction effects between antibiotics treatment and formula feeding -->
+-   The intestinal permeability assay is interesting in that it supports the findings of Thulstrup et al. 2015.
 
-<!-- #### Rationale: -->
+### Experiment 3 - Interaction effects between antibiotics treatment and formula feeding
 
-<!-- Based on the results from Exp 2 it was difficult to ascertain which treatment was more likely to manifest in NEC like phenotype. -->
-<!-- - Gentamicin had increased intestinal permeability the most. -->
-<!-- - However, the greatest effect of cytokine expression was found with the vancomycin treatment. -->
-<!-- - In addition to this divergence, the interaction effect found in Exp 1 suggested it was likely that combining MAT with formula feeding would yield different results. -->
-<!-- - Therefore, we decided to increase the breeding population to allow for an experiment that combined both types of antibiotics with either formula-feeding or breastfeeding, for a total of 6 experimental groups.
--->
+#### Rationale:
 
-<!-- #### Aim: -->
+Based on the results from Exp 2 it was difficult to ascertain which treatment was more likely to manifest in NEC like phenotype.
+- Gentamicin had increased intestinal permeability the most.
+- However, the greatest effect of cytokine expression was found with the vancomycin treatment.
+- In addition to this divergence, the interaction effect found in Exp 1 suggested it was likely that combining MAT with formula feeding would yield different results.
+- Therefore, we decided to increase the breeding population to allow for an experiment that combined both types of antibiotics with either formula-feeding or breastfeeding, for a total of 6 experimental groups.
 
-<!-- -   Repeat Exp 2 with greater sample sizes, adding the effect of formula feeding. -->
-<!-- -   To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics in combination with formula feeding on intestinal permeability and cytokine expression. -->
-<!-- -    -->
+#### Aim:
 
-<!-- #### Conclusion: -->
+-   Repeat Exp 2 with greater sample sizes, adding the effect of formula feeding.
+-   To test the effect of gram-positive-specific antibiotics versus gram-negative-specific antibiotics in combination with formula feeding on intestinal permeability and cytokine expression.
+-
 
-<!-- -   Combining MAT with formula feeding resulted in more pronounced signs of inflammation than MAT alone. -->
-<!-- -   VANCO-FF and VANCO-BF had the highest and lowest intestinal permeability scores, respectively. -->
-<!-- -   Significantly altered expression of XXSignificantSCYTOKINESXX. -->
-<!-- -   Greatest determinant of variation was formula-feeding. -->
+#### Conclusion:
 
-<!-- #### Detailed description of Exp 3. -->
+-   Combining MAT with formula feeding resulted in more pronounced signs of inflammation than MAT alone.
+-   VANCO-FF and VANCO-BF had the highest and lowest intestinal permeability scores, respectively.
+-   Significantly altered expression of XXSignificantSCYTOKINESXX.
+-   Greatest determinant of variation was formula-feeding.
 
-<!-- *Antimals and experimental setup* - To increase the number of pups available we doubled the number of mice for breeding, 36 females and 12 males. -->
-<!-- - Pregnant females were again given either Vancomycin, Gentamicin (1 g/l) or sterile water on gestational day 15. -->
-<!-- - `r nrow(preexp_3` pups were born and separated into breastfeeding or formula feeding, for a total of 6 groups: - VANCO-BF, VANCO-FF, GENTA-BF, GENTA-FF, CON-BF and CON-FF. -->
+#### Detailed description of Exp 3.
 
-<!-- *Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00. -->
-<!-- - Formula-fed pups consistently showed higher levels of Serum FITC concentration compared to breastfed pups, irrespective of MAT status. -->
-<!-- - VANCO-FF and VANCO-BF had the highest and lowest levels of FITC, respectively. -->
+*Antimals and experimental setup* - To increase the number of pups available we doubled the number of mice for breeding, 36 females and 12 males.
+- Pregnant females were again given either Vancomycin, Gentamicin (1 g/l) or sterile water on gestational day 15.
+- `r nrow(preexp_3` pups were born and separated into breastfeeding or formula feeding, for a total of 6 groups: - VANCO-BF, VANCO-FF, GENTA-BF, GENTA-FF, CON-BF and CON-FF.
 
-<!-- *Cytokine expression* - We found significant differences in concentrations of XXSignificantSCYTOKINESXX. -->
-<!-- Linear regression analysis indicated the biggest determinant of differences in outcomes was formula feeding, with the type of antibiotic being of lesser importance compared. -->
-<!-- - Directionally the trend was the same for all cytokines. -->
-<!-- - IL-1b: Non-significant decrease in all antibiotics-treated offspring, irrespective of feeding. -->
-<!-- - IL-2: significantly lower concentrations in all formula-fed pups compared with CON-BF. -->
-<!-- - IL-6: Significantly greater levels in all formula-fed pups compared with CON-BF. -->
-<!-- - INFg: significantly lower concentrations in all formula-fed pups compared with CON-BF. -->
-<!-- - KCGRO: Significantly greater levels in all formula-fed pups compared with CON-BF. -->
+*Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
+- Formula-fed pups consistently showed higher levels of Serum FITC concentration compared to breastfed pups, irrespective of MAT status.
+- VANCO-FF and VANCO-BF had the highest and lowest levels of FITC, respectively.
 
-<!-- #### Commentary and interpretation experimental results Exp 3: -->
+*Cytokine expression* - We found significant differences in concentrations of XXSignificantSCYTOKINESXX.
+Linear regression analysis indicated the biggest determinant of differences in outcomes was formula feeding, with the type of antibiotic being of lesser importance compared.
+- Directionally the trend was the same for all cytokines.
+- IL-1b: Non-significant decrease in all antibiotics-treated offspring, irrespective of feeding.
+- IL-2: significantly lower concentrations in all formula-fed pups compared with CON-BF.
+- IL-6: Significantly greater levels in all formula-fed pups compared with CON-BF.
+- INFg: significantly lower concentrations in all formula-fed pups compared with CON-BF.
+- KCGRO: Significantly greater levels in all formula-fed pups compared with CON-BF.
 
-<!-- -   Intestinal permeability was by far the greatest in VANCO-FF, with levels of FITC-dex close to 10-fold increased in this group. -->
-<!-- -   In contrast, VANCO-BF was the group with the lowest measure of intestinal permeability, similar to the finding in exp 2. -->
-<!-- -   Looking at **FIGURE CYTOKINES ROUND 3** shows how formula-fed groups systematically over- or underexpress compared to breastfed analimals, irrespective of type of antibiotic, indicating that the biggest determinant of cytokine variation was again formula feeding.
--->
+#### Commentary and interpretation experimental results Exp 3:
 
-<!-- *Adjusting for weight of tissue* - In the previous experiments, cytokine concentrations were adjusted for the weight of the tissue analyzed, to control for biopsy sizes. -->
-<!-- - We did so because the average weight of tissue samples was not meaningfully different between groups. -->
-<!-- - This assumption wasn't met in Exp 3, where samples from formula-fed pups weighted on average XX%XX more (p = XXp.valueXX) than breastfed controls. -->
-<!-- - Vanco-FF had the heaviest ileum tissue samples of all **mean + P (if its true)**. -->
-<!-- - Tissue edema could well be indicative of inflammation, and hence desirable, however it introduces a bias against the formula-fed pups if cytokine concentrations were normalized to weight. -->
-<!-- - The reported findings are no longer significant **Table of significant cytokines** when adjusting for weight. -->
+-   Intestinal permeability was by far the greatest in VANCO-FF, with levels of FITC-dex close to 10-fold increased in this group.
+-   In contrast, VANCO-BF was the group with the lowest measure of intestinal permeability, similar to the finding in exp 2.
+-   Looking at **FIGURE CYTOKINES ROUND 3** shows how formula-fed groups systematically over- or underexpress compared to breastfed analimals, irrespective of type of antibiotic, indicating that the biggest determinant of cytokine variation was again formula feeding.
 
-<!-- -   This phenomenon is a common challenge in animal disease models, requiring measure of normalization other that tissue weight, such as total protein content. -->
-<!-- -   Unfortunately this discovery was done after tissue homogenization and analysis, making it too late for further analysis. -->
+*Adjusting for weight of tissue* - In the previous experiments, cytokine concentrations were adjusted for the weight of the tissue analyzed, to control for biopsy sizes.
+- We did so because the average weight of tissue samples was not meaningfully different between groups.
+- This assumption wasn't met in Exp 3, where samples from formula-fed pups weighted on average XX%XX more (p = XXp.valueXX) than breastfed controls.
+- Vanco-FF had the heaviest ileum tissue samples of all **mean + P (if its true)**.
+- Tissue edema could well be indicative of inflammation, and hence desirable, however it introduces a bias against the formula-fed pups if cytokine concentrations were normalized to weight.
+- The reported findings are no longer significant **Table of significant cytokines** when adjusting for weight.
 
-<!-- *Combining results from exp 2 and exp 3* - The protocols for the breastfed groups were identical across experiment 2 and 3, allowing for pooling of these results. -->
-<!-- - While analyzing the results from both groups together increased the group sizes and degrees of statistical freedom, it did not change the clinical conclusions. -->
+-   This phenomenon is a common challenge in animal disease models, requiring measure of normalization other that tissue weight, such as total protein content.
+-   Unfortunately this discovery was done after tissue homogenization and analysis, making it too late for further analysis.
 
-<!-- ### Experiment 4 - FVT Trial -->
+*Combining results from exp 2 and exp 3* - The protocols for the breastfed groups were identical across experiment 2 and 3, allowing for pooling of these results.
+- While analyzing the results from both groups together increased the group sizes and degrees of statistical freedom, it did not change the clinical conclusions.
 
-<!-- #### Rationale: **What was the actual rationale?** -->
+### Experiment 4 - FVT Trial
 
-<!-- -   Central to the development of NEC is an assumption of a gut microbial dysbiosis, which we tried to mimick by administering antibiotics prenatally. -->
-<!-- -   Results from exp 1-3 indicated that maternal antibiotics treatment had little effect in the context of breastfeeding, but when combined with formula-feeding was able to induce mild symptoms of intestinal inflammation.
--->
-<!-- -   It was unclear whether this was solely an effect of the formula, or whether an interaction with the maternal antibiotics treatment existed. -->
-<!-- -   If our findings were partly due to a successful manipulation of the gut microbiome, then eliminating the induced disbyosis should also partly mitigate the inflammatory signatures. -->
-<!-- -   Relative bacterial abundance / bacterial composition is partly a function of viral (bacteriophage) abundance, and sevaral studies have shown how a FVT can alter the bacterial composition. -->
-<!-- -   We therefore decided to conduct a final experiment, in which we administered an FVT to a vancomycing-treated, formula-fed group (since this combination (VANCO-FF) had the biggest effects in terms of intestinal permeability and proinflammatory cytokine expression in Exp 3).
--->
-<!-- -   We hypothesized that any dysbiosis created by maternal antibiotics treatment could be mitigated with a follow-up administration of FVT prior to beginnig formula-feeding. -->
-<!-- -   If successful, the FVT-treated group would show fever intestinal injury scores compared with a non-FVT treated group. -->
-<!-- -   Previous experiments had shown that MAT in combination with breastfeeding had no effect on inflammatory markers and VANCO-BF thus acted as a healthy control to both formula-fed treatment groups.
+#### Rationale: **What was the actual rationale?**
+
+-   Central to the development of NEC is an assumption of a gut microbial dysbiosis, which we tried to mimick by administering antibiotics prenatally.
+-   Results from exp 1-3 indicated that maternal antibiotics treatment had little effect in the context of breastfeeding, but when combined with formula-feeding was able to induce mild symptoms of intestinal inflammation.
+-   It was unclear whether this was solely an effect of the formula, or whether an interaction with the maternal antibiotics treatment existed.
+-   If our findings were partly due to a successful manipulation of the gut microbiome, then eliminating the induced disbyosis should also partly mitigate the inflammatory signatures.
+-   Relative bacterial abundance / bacterial composition is partly a function of viral (bacteriophage) abundance, and sevaral studies have shown how a FVT can alter the bacterial composition.
+-   We therefore decided to conduct a final experiment, in which we administered an FVT to a vancomycing-treated, formula-fed group (since this combination (VANCO-FF) had the biggest effects in terms of intestinal permeability and proinflammatory cytokine expression in Exp 3).
+-   We hypothesized that any dysbiosis created by maternal antibiotics treatment could be mitigated with a follow-up administration of FVT prior to beginnig formula-feeding.
+-   If successful, the FVT-treated group would show fever intestinal injury scores compared with a non-FVT treated group.
+-   Previous experiments had shown that MAT in combination with breastfeeding had no effect on inflammatory markers and VANCO-BF thus acted as a healthy control to both formula-fed treatment groups.
 **(The problem here is that formula is the only thing with an effect.
-A breastfed control only shows the effect of the formula)** -->
+A breastfed control only shows the effect of the formula)**
 
-<!-- #### Aim: -->
+#### Aim:
 
-<!-- -   To test whether signs of inflammation are reduced in pups treated with an FVT. -->
-<!-- -   To test whether any reductions in symptoms can be explained with a manipulation of the microbial composition. -->
+-   To test whether signs of inflammation are reduced in pups treated with an FVT.
+-   To test whether any reductions in symptoms can be explained with a manipulation of the microbial composition.
 
-<!-- #### Conclusion: -->
+#### Conclusion:
 
-<!-- -   FVT treated pups showed similar intestinal permeability and pro-inflammatory cytokine concentrations as non-FVT-treated controls. -->
-<!-- -   An `r fold_increase_reg3g`-fold increase in expression of the `r REG3G` gene was found in the FVT group. -->
+-   FVT treated pups showed similar intestinal permeability and pro-inflammatory cytokine concentrations as non-FVT-treated controls.
+-   An `r fold_increase_reg3g`-fold increase in expression of the `r REG3G` gene was found in the FVT group.
 
-<!-- #### Detailed description of Exp 4. -->
+#### Detailed description of Exp 4.
 
-<!-- -   For the final experiment we repeated the previous breeding setup, assigning 36 females and 12 males for breeding. -->
+-   For the final experiment we repeated the previous breeding setup, assigning 36 females and 12 males for breeding.
 
-<!-- -   Only 5 of the 36 females became pregnant, and all were assigned to vancomycin (1 g/l) from pregnancy-day 15 until delivery, giving birth to `r nrow(preexp_4` pups on Saturday the XXth. -->
+-   Only 5 of the 36 females became pregnant, and all were assigned to vancomycin (1 g/l) from pregnancy-day 15 until delivery, giving birth to `r nrow(preexp_4` pups on Saturday the XXth.
 
-<!-- -   2 days after birth pups were separated into breastfeeding or formula feeding. -->
+-   2 days after birth pups were separated into breastfeeding or formula feeding.
 
-<!-- -   The formula-fed pups were further divided into a treatment and control group, receiving either 30 µl fecal virome filtrate (FVT) (0.01 mg farces/µL) or an equal amount of SM buffer (control) 1 hour prior to the first feeding.
--->
+-   The formula-fed pups were further divided into a treatment and control group, receiving either 30 µl fecal virome filtrate (FVT) (0.01 mg farces/µL) or an equal amount of SM buffer (control) 1 hour prior to the first feeding.
 
-<!-- -   We repeated the FVT administration 24 hours later (09:00 day 2) for a total dose of 0.6 mg feces material per pup. -->
+-   We repeated the FVT administration 24 hours later (09:00 day 2) for a total dose of 0.6 mg feces material per pup.
 
-<!-- -   2 hours after FITC administtion pups were euthanized by decapitation and blood collected in eppendorfs for analysis of serum FITC concentration. -->
-<!--     Intestinal samples were collected for gene expression assays, histological assessment, pro-inflammatory cytokine analysis and gut microbiome sequencing. -->
+-   2 hours after FITC administtion pups were euthanized by decapitation and blood collected in eppendorfs for analysis of serum FITC concentration.
+    Intestinal samples were collected for gene expression assays, histological assessment, pro-inflammatory cytokine analysis and gut microbiome sequencing.
 
-<!-- *Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00. -->
-<!-- - The FVT and FVT-control (SM) had similar permeability scores. -->
-<!-- - Compared with the vancomycin-treated formula-fed pups from experiment 3, there were no significant differences **Figure X**. -->
-<!-- - The breastfed pups did, however, have significantly higher levels compared with all other breastfed pups. -->
+*Intestinal permeability - FITC-Dextran administration* - We repeated the protocol for FITC-administration, separating pups from mothers at 08:00 and dosing at 11:00.
+- The FVT and FVT-control (SM) had similar permeability scores.
+- Compared with the vancomycin-treated formula-fed pups from experiment 3, there were no significant differences **Figure X**.
+- The breastfed pups did, however, have significantly higher levels compared with all other breastfed pups.
 
-<!-- *Cytokine expression* - We found significant differences in concentrations of TNFa in Exp 4, with greater level in the FVT-treated group (XXX), compared to both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively).
--->
-<!-- - This is in contrast experiment 3, where differences between VANCO-BF and VANCO-FF found significant differences for XXXCYtokinesXXX. -->
-<!-- OR: - This is similar to the findings of experiment 3, where no differences between VANCO-BF and VANCO-FF were found either. -->
+*Cytokine expression* - We found significant differences in concentrations of TNFa in Exp 4, with greater level in the FVT-treated group (XXX), compared to both VANCO-BF and VANCO-FF (SM) (XXX and XXXX respectively).
+- This is in contrast experiment 3, where differences between VANCO-BF and VANCO-FF found significant differences for XXXCYtokinesXXX.
+OR: - This is similar to the findings of experiment 3, where no differences between VANCO-BF and VANCO-FF were found either.
 
-<!-- *Gene expression* - qPCR analyzis of intestinal samples found statistically significant differences in the level of XXXSIGNIFICANT_GENESXXX expression between groups with the delta-delta method (reformulate).
--->
-<!-- - However, a difference in expression of the reference gene between the FVT and breastfed control was observed. -->
-<!-- this difference was statistically significant, invalidating the assumption of *equal baseline expression.*(reformulate) - Comparing the groups on the basis of (SOMEHTING ELSE THAN DELTA DELTA) found a `r fold_increase_reg3g`-fold increased expression of the `r REG3G` gene in the FVT groups.
--->
+*Gene expression* - qPCR analyzis of intestinal samples found statistically significant differences in the level of XXXSIGNIFICANT_GENESXXX expression between groups with the delta-delta method (reformulate).
+- However, a difference in expression of the reference gene between the FVT and breastfed control was observed.
+this difference was statistically significant, invalidating the assumption of *equal baseline expression.*(reformulate) - Comparing the groups on the basis of (SOMEHTING ELSE THAN DELTA DELTA) found a `r fold_increase_reg3g`-fold increased expression of the `r REG3G` gene in the FVT groups.
 
-<!-- #### Commentary and interpretation -->
+#### Commentary and interpretation
 
-<!-- -   FVT-treatmed did not seem to change the intestinal permeability. -->
+-   FVT-treatmed did not seem to change the intestinal permeability.
 
-<!-- -   It is possible that something went wrong with the VANCO-BF group, given FITC-concentration was `r mean(fitc_level_exp4$bf`, which is 4-5 times higher than VANCO-BF groups on identical protocols in Exp 2 `r mean(fitc_level_exp2$bf` and 3 `r mean(fitc_level_exp3$bf`.
+-   It is possible that something went wrong with the VANCO-BF group, given FITC-concentration was `r mean(fitc_level_exp4$bf`, which is 4-5 times higher than VANCO-BF groups on identical protocols in Exp 2 `r mean(fitc_level_exp2$bf` and 3 `r mean(fitc_level_exp3$bf`.
 Compared with theses levels the development looks more biologically reasonable.
-**FIGURE of EXP4 with DIFFERENT CONTROL GROUP** -->
+**FIGURE of EXP4 with DIFFERENT CONTROL GROUP**
 
-<!-- -   The level of pro-inflammatory cytokines were the same in the FVT and SM groups, again indicating a lack of efficacy of the FVT treatment. -->
+-   The level of pro-inflammatory cytokines were the same in the FVT and SM groups, again indicating a lack of efficacy of the FVT treatment.
 
-<!-- -   Summary: The FVT treatment did not seem to have an effect. -->
+-   Summary: The FVT treatment did not seem to have an effect.
 
-<!-- ### \--\>Post experimental analyses and Discussion of results -->
+### \--\>Post experimental analyses and Discussion of results
 
-<!-- -   Histological image analysis was performed after all experiments had finished. -->
+-   Histological image analysis was performed after all experiments had finished.
 
-<!-- -   We inspected the morphology of the ileum and colon, and visually assessed thickness of the mucin layers. -->
+-   We inspected the morphology of the ileum and colon, and visually assessed thickness of the mucin layers.
 
-<!-- -   Analyzing AB-PAS stains for mucin-stained tissue to the total area, showed no differences between any treatment groups. -->
+-   Analyzing AB-PAS stains for mucin-stained tissue to the total area, showed no differences between any treatment groups.
 
-<!-- -   Formula-fed and breastfed pups were slightly different, irrespective of antibiotics treatment, whereas there were no observable differences between antibiotic treatments. -->
+-   Formula-fed and breastfed pups were slightly different, irrespective of antibiotics treatment, whereas there were no observable differences between antibiotic treatments.
 
-<!-- -   These findings supports the visual inspections of intestines made during sampling, where no signs of inflammation were detected. -->
+-   These findings supports the visual inspections of intestines made during sampling, where no signs of inflammation were detected.
 
-<!-- -   *Images of intestines* -->
+-   *Images of intestines*
 
-<!-- -   Gut microbiome composition -->
+-   Gut microbiome composition
 
-<!-- ## Discussion -->
+## Discussion

### 2022-11-08 code/Matrinem_data_analysis.Rmd

@@
 
-# MATRINEM - Attempts at creating a NEC mouse model
-
-## The research questions
-
-Experiment no. 4 was the goal, but before we could test our overall hypothesis we needed to find out which combination of antibiotics and feeding scheme would results in the most NEC-like phenotype.
-
-These research questions were investigated in 3 pre-experiments (PreExp 1-3), each with their own research questions.
-
-### **Overview of all experiments**
-
-+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
-|           | PreExp 1                                                                      | PreExp 2 | PreExp 3 | NEC-FvT |
-+===========+===============================================================================+==========+==========+=========+
-| **Aim**   | To test feasibility of formula feeding with PICC line                         |          |          |         |
-+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
-| Protocols | Testing a broad spectrum antibiotics mix in combination with formula feeding. |          |          |         |
-|           |                                                                               |          |          |         |
-|           | 4 groups                                                                      |          |          |         |
-|           |                                                                               |          |          |         |
-|           | 1.  CON-BF                                                                    |          |          |         |
-|           | 2.  CON-FF                                                                    |          |          |         |
-|           | 3.  AB-BF                                                                     |          |          |         |
-|           | 4.  AB-FF                                                                     |          |          |         |
-+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
-| N         | 38 (32 survived)                                                              |          |          |         |
-+-----------+-------------------------------------------------------------------------------+----------+----------+---------+
-
 ### PreExp 1 - Feasibility Study
@@
 #Load in the data
+```{r exp1-weight, include=FALSE}
+#____Load overview of sample IDs______ 
+pups <- read_excel("data/processed/weight_development.xlsx")
 
-```{r include=FALSE}
-preexp_1 <- read_excel("data/processed/weight_development.xlsx") %>% 
+#____ Separate into experimental rounds ____#
+preexp_1 <- pups %>% 
   filter(exp_number == "1")
+
+exp1_survived <- preexp_1 %>% 
+  filter(!is.na(bodyweight_baseline_plus_48h)) %>% 
+  nrow()
+#____ ____#
 ```
+
+```{r exp1-groups}
+#____Separate into groups ______ 
+AB_FF <- nrow(preexp_1 %>% 
+  filter(group2 == "AB-FF"))
+
+AB_BF <- nrow(preexp_1 %>% 
+  filter(group2 == "AB-BF"))
+
+CON_FF <- nrow(preexp_1 %>% 
+  filter(group2 == "CON-FF"))
+
+CON_BF <- nrow(preexp_1 %>% 
+  filter(group2 == "CON-BF")) 
+#____ ____#
+
+```
+
+Survival rate: `r exp1_survived` of `r nrow(preexp_1)`(`r (exp1_survived / nrow(preexp_1))*100`%)
 
@@
 
-#### **Rationale:**
-
-*Chen et al.*, showed that Maternal Antibiotics Treatment (**MAT**) resulted in mild NEC-like symptoms of mice offspring.
-We believed that MAT combined with formula feeding (**FF**) of offspring would result in a worsened phenotype - [even more NEC-like than either treatment on it's own.]{.underline}
-
-#### **Aim:**
-
-To test if it was possible to keep mice separated from the mother alive from postnatal day 3 and 48 hours onwards.
-
-#### **Research Questions:**
-
-##### Can we keep offspring alive with formula feeding?
-
 `r nrow(preexp_1)` mice were born to 6 different mothers.
@@
   group_by(cytokines) %>% 
-  shapiro_test(cytokine_concentration) %>% 
-  select(cytokines, p)
+  shapiro_test(cytokine_concentration) 
 #___________________________________________________________

### 2022-11-08 code/Round 1/Cytokines adjusted.Rmd

@@
 # Import dataset
-cytokines <- read_excel("../../data/processed/cytokines.xlsx")
-glimpse(cytokines)
+cytokines <- read_excel("data/processed/cytokines.xlsx")
 
+tissue_weight_destribution <- cytokines %>%   # run normality test on tissue weight
+  filter(tissue == "ileum") %>% 
+  filter(exp_number == "1") %>% 
+  group_by(group) %>% 
+  shapiro_test(tissue_weight_mg)
+
+# group  variable         statistic      p
+#   <chr>  <chr>                <dbl>  <dbl>
+# 1 AB-BF  tissue_weight_mg     0.917 0.404 
+# 2 AB-FF  tissue_weight_mg     0.953 0.686 
+# 3 CON-BF tissue_weight_mg     0.771 0.0138
+# 4 CON-FF tissue_weight_mg     0.812 0.144
+
+#look at tissue weight distribution
+cytokines %>% 
+  filter(tissue == "ileum") %>% 
+  filter(exp_number == "1") %>% 
+  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
+  geom_boxplot() +
+  geom_point()
+```
+<!-- Looks fine. Go ahead and  adjust-->
+
+```{r adjust for tissue weight , echo=FALSE}
 cytokines_adjusted <- cytokines %>% 
@@
       )) %>% 
-  mutate(values = if_else(tissue == "ileum", values/tissue_weight_mg, values)) %>%
+  mutate(values = if_else(tissue == "ileum", values/tissue_weight_mg, values)) %>% #Adjust for tissue weight
   pivot_wider(
@@
   )
-#pivot_wider(id_cols = c(sample_id,exp_number, group, received_antibiotics,maternal_treatment, #tissue,tissue_weight_mg, adjusted_values), names_from = variables, values_from = values)
 ```
@@
   filter(variables != "INFg") %>% 
-  drop_na(adjusted_values) %>% 
-  ggplot(aes(x = adjusted_values, fill = group)) + 
+  drop_na(values) %>% 
+  ggplot(aes(x = values, fill = group)) + 
   geom_density() +
@@
   #filter(variables != "INFg") %>% 
-  drop_na(adjusted_values) %>% 
-  ggplot(aes(x = group, y = adjusted_values, color = group)) +
+  drop_na(values) %>% 
+  ggplot(aes(x = group, y = values, color = group)) +
   geom_jitter() +
@@
   filter(variables != "INFg") %>% 
-  drop_na(adjusted_values) %>% 
-  mutate(adjusted_values = log10(adjusted_values)) %>% 
-  ggplot(aes(x = group, y = adjusted_values, color = group)) +
+  drop_na(values) %>% 
+  mutate(adjusted_values = log10(values)) %>% 
+  ggplot(aes(x = group, y = values, color = group)) +
   geom_jitter() +
@@
   filter(exp_number == "1") %>% 
-  filter(tissue == "ileum") %>% 
+  filter(tissue == sample_tissue) %>% 
   select(group,INFg:TNFa) %>% 

### 2022-11-08 submission/appendix.Rmd

---
title: ''
output: pdf_document
---
# Appendix(ces)

## Appendix A: additional tables

\newpage

## Appendix B: additional figures

\newpage

## Appendix C: code

```{r get-labels, echo = FALSE}
labs = knitr::all_labels() # this is the new code chunk to add
labs = setdiff(labs, c("setup", "get-labels")) # this function excludes the code chunk with the label "setup" from being displayed.
```

```{r all-code, ref.label=labs, eval=FALSE}
# this code chunk displays all source code from your entire dissertation document (that you have written in .Rmd, not from your R script).
```

### 2022-11-08 submission/introduction.Rmd

---
title: " "
output: pdf_document
header-includes:
  - \usepackage{float}
  - \usepackage{sectsty}
---

# Introduction

- introduce the reader to the subject area and clarify the knowledge gap that the dissertation research will fill.
- set the context for the dissertation by reviewing the relevant literature.
- include relevant references to general (theoretical papers and reviews) and specific (specific to the particular question addressed) literature, to justify the research that has been undertaken and define the questions being addressed.
- state the primary research questions and hypotheses in the final paragraph.
- follow an ‘inverted triangle’ format, progressing from general scientific ideas and why they matter to the specific research questions addressed in the dissertation project.

*The introduction should not be just a ‘Literature Review’.*

\citep{breton_encounter_2006}

\citep*{breton_encounter_2006}

\citet{breton_encounter_2006}

\citet*{breton_encounter_2006}

\citep{martin_diet_1989, breton_encounter_2006}

\citeauthor{breton_encounter_2006}

\citeauthor*{breton_encounter_2006}

\citeyear{breton_encounter_2006}

\citeyearpar{breton_encounter_2006}

### 2022-11-08 submission/thesis.Rmd

@@
 
-```{r uni_logo, echo=F, out.width="95%"}
-knitr::include_graphics("img/sund.pdf")
+\begin{centering}
+```{r uni-logo-top, echo=F, out.width="80%"}
+knitr::include_graphics("img/sund.png")
 ```
 
-\newpage
-
-\begin{centering}
-
-\vspace{1cm}
+\Large
+{\bf The University Of Copenhagen}
 
 \Large
-{\bf The University Of Edinburgh}
-
-\Large
-{\bf School Of Geosciences}
+{\bf Factuly of Health}
 
@@
 \doublespacing
-{\bf COMPARISON OF TOP-DOWN AND BOTTOM-UP APPROACHES ON SPECIFIC LEAF AREA PATTERNS, \\AT GLOBAL, LATITUDINAL, AND BIOME SCALES}
+{\bf MATERNAL ANTIBIOTICS TREATMENT AND FORMULA FEEDING \\ATTEMPTS AT CREATING A NEC MOUSE MODEL}
 
@@
 
-{\bf ANNA CHIRUMBOLO}
+{\bf KAARE D. TRANÆS}
 
@@
 
-in partial fulfilment of the requirement \\for the degree of BSc with Honours \\in Ecological and Environmental Sciences
+Master thesis project \\MSc Human Biology \\SUND
 
@@
 mm yy
+
+```{r uni-logo-bottom, echo=F, out.width="60%"}
+knitr::include_graphics("img/absalon_watermark.png")
+```
 
@@
 
-```{r results, child= 'results.Rmd'}
-```
+#```{r results, child= 'results.Rmd'}
+#```
 

### 2022-11-08 code/Round 1/Cytokines adjusted.Rmd
Progress

@@
 
-#Density plots. Works
+#Density plots to show distribution of cytokines
 histogram <- cytokines_long %>% 
@@
   geom_density() +
-  facet_wrap(~ variables,
+  facet_wrap(~ variables, ncol = 2,
              scales = "free")
@@
   filter(variables != "INFg") %>% 
-  drop_na(adjusted_values) %>% 
-  ggplot(aes(x = log10(adjusted_values), fill = group)) + 
+  drop_na(values) %>% 
+  ggplot(aes(x = log10(values), fill = group)) + 
   geom_density() +
-  facet_wrap(~ variables,
+  facet_wrap(~ variables, ncol = 2,
              scales = "free")
@@
 
-ggsave(filename = "exp1_cytokines.png",width = 2.14, height = 1.51, dpi = 300)
-
-ggsave('../../exports/exp1_cytokines.png',
-       plot_all_cytokines, scale =  width = 10, height = 16, units = "cm")
+#ggsave('../../exports/exp1_cytokines.png',
+     #  plot_all_cytokines,  width = 400, height = 600, units = "px")
 
@@
 signif_cytokines <- cytokines_long %>%
+  filter(exp_number == "1") %>% 
   filter(tissue == sample_tissue) %>% 
-  filter(variables %in% tbl1_p$`**Groups**`) %>%  #Exclude any cytokines not wanted in the plot
+  filter(variables %in% tbl1_p$Cytokines) %>%  #Exclude any cytokines not wanted in the plot
 ggplot(aes(x = group, 
@@
   stat_compare_means(
-    aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
-    label.y.npc = 0.9,
-    label.x.npc = 0.5) +
+    aes(label = paste0("Global p = ", ..p.format..)),
+    label.y.npc = 0.75,
+    label.x.npc = 0.3) +
   facet_wrap(~variables, 
@@
 signif_cytokines_patch <- signif_cytokines + plot_annotation (
-  title = 'Cytokine expression',
-  subtitle = 'Statistically significant changes',
-  tag_levels = 'A',
+  #title = 'Cytokine expression',
+  #subtitle = 'Statistically significant changes',
+  #tag_levels = 'A',
   caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding. \n
-  Pairwise comparisons with CON-BF using wilcoxon test',
+  Pairwise comparisons against CON-BF, wilcoxon test',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),

### 2022-11-11 code/weight_analysis.Rmd
Including references

@@
 knitr::opts_chunk$set(echo = TRUE)
+library(rstatix)
+library(ggpubr)
+library(patchwork)
 ```
 
-# Analysis of the weight development in the MARTRINEM study 
+#___Analysis of the weight development in the MARTRINEM study___
 ```{r}
@@
   mutate(received_antibiotics = recode_factor(received_antibiotics, 
-                                       yes = "MAT",
-                                       no = "CON"))
+                                       Yes = "MAT",
+                                       No = "CON"))
   #%>%
   # #mutate(across(where(is_character),as_factor))
+```
 
+#_Birthweight of offspring, experiment 2_
+```{r}
 exp2.bw.density <- data %>% 
@@
   title = 'Effect of Maternal Antibiotics Treatment on birth weight of offspring',
-  subtitle = 'Experimental round 2',
+  subtitle = 'Experiment 2',
   tag_levels = 'A',

### 2022-11-11 code/Round 1/Cytokines adjusted.Rmd
Progressed

@@
 
-# Cytokines for experiment 1. - Tissue weight adjusted.
+# Cytokines for experiments  - Tissue weight adjusted.
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
+```
+
+```{r load packages, include=FALSE}
+library(readxl) 
+library(tidyverse)
+library(gt)
+library(gtsummary)
+library(broom)
+library(rstatix)
+library(knitr)
+library(patchwork)
+library(ggpubr)
+```
+
+# Introduction to script
+
+This script is useful for analyzing cytokine expression from MSD Mesoscale assays.
+It is currently optimized for the 'Pro-inflammatory Vplex 10 Mouse kit', analysing the cytokines INFg, IL_10, IL_12p70, IL_1b, IL_2, IL_4,IL_5, IL_6, KCGRO, and TNFa.
+
+Required data structure is a wide tibble (column headers = variables, samples and values = rows).
+
+This script is per default set to handle ileum and serum expression, although any tissue will work. This will just require resetting of the "tissue"-variable.
+
+# Cytokines for experiment 2.
 
@@
 First you load the data and check the structure. Make sure cytokine concentrations are all numeric (<dbl>). If not, mutate non-numeric columns with mutate_if(is_character).
+
 ```{r load data, echo=FALSE}
@@
 cytokines <- read_excel("data/processed/cytokines.xlsx")
+```
 
-tissue_weight_destribution <- cytokines %>%   # run normality test on tissue weight
-  filter(tissue == "ileum") %>% 
-  filter(exp_number == "1") %>% 
+Set the tissue type and experimental round here:
+
+```{r load data and set tissue type, include=FALSE}
+#set tissue type
+sample_tissue <- "ileum"
+study_round <- "2"
+```
+
+```{r load data, echo=FALSE}
+cytokines %>%   # run normality test on tissue weight
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
   group_by(group) %>% 
@@
  # <chr>    <chr>                <dbl>  <dbl>
-# 1 AB-BF  tissue_weight_mg     0.917 0.404 
-# 2 AB-FF  tissue_weight_mg     0.953 0.686 
-# 3 CON-BF tissue_weight_mg     0.771 0.0138
-# 4 CON-FF tissue_weight_mg     0.812 0.144
+# CON-BF   tissue_weight_mg     0.833 0.197 
+#GENTA-BF tissue_weight_mg     0.748 0.0120
+#VANCO-BF tissue_weight_mg     0.833 0.0861
+#look at tissue weight distribution
 
-#look at tissue weight distribution
-cytokines %>% 
-  filter(tissue == "ileum") %>% 
-  filter(exp_number == "1") %>% 
+cytokines %>%       #Weight of tissue samples
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
   ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
   geom_boxplot() +
-  geom_point()
+  geom_point() +
+  labs(
+    title = "Tissue weight of intestinal samples, experiment 2"
+  )
 ```
@@
     id_cols = c(
-      sample_id,exp_number, group, received_antibiotics,maternal_treatment, tissue,tissue_weight_mg),
+      sample_id,exp_number, group, received_antibiotics,maternal_treatment, type_of_feed, tissue,tissue_weight_mg),
     names_from = variables, 
@@
 ```{r visualize data, eval=FALSE, include=FALSE}
-#normality_check 
-shapiro_failed <- cytokines_long %>% 
-  filter(exp_number == 1) %>% 
-  filter(tissue == "ileum") %>%
-  filter(variables != "INFg") %>% 
+cytokines_long %>%                       #normality_check  of cytokine distributions
+  filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  #filter(variables != "INFg") %>% 
   drop_na(values) %>% 
@@
 
-log10_shapiro_failed <- cytokines_long %>% 
-  filter(exp_number == 1) %>% 
-  filter(tissue == "ileum") %>%
+#Non-normal distributions: 
+#  group    variables variable statistic      p
+#  <fct>    <chr>     <chr>        <dbl>  <dbl>
+# VANCO-BF IL-1b     values       0.795 0.0363
+# GENTA-BF IL-4      values       0.777 0.0240
+
+#log10 transformed shapiro_test - failed variables
+cytokines_long %>% 
+ filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
   filter(variables != "INFg") %>% 
@@
 
+#  group    variables variable statistic       p
+#  <fct>    <chr>     <chr>        <dbl>   <dbl>
+# VANCO-BF IL-1b     values       0.727 0.00728     # transformation helped.
+
 #Density plots to show distribution of cytokines
 histogram <- cytokines_long %>% 
-  filter(exp_number == 1) %>% 
-  filter(tissue == "ileum") %>%
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
   filter(variables != "INFg") %>% 
@@
 log10_histogram <- cytokines_long %>% 
-  filter(exp_number == 1) %>% 
-  filter(tissue == "ileum") %>%
-  filter(variables != "INFg") %>% 
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  #filter(variables != "INFg") %>% 
   drop_na(values) %>% 
@@
 jitter_plot <- cytokines_long %>% 
-  filter(exp_number == 1) %>% 
-  filter(tissue == "ileum") %>%
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
   #filter(variables != "INFg") %>% 
@@
 log10_jitter_plot <- cytokines_long %>% 
-  filter(exp_number == 1) %>% 
-  filter(tissue == "ileum") %>%
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
   filter(variables != "INFg") %>% 
@@
              scales = "free") 
-```
-
-Set the tissue type to whatever tissue your data are extracted from, or that you want to analyze first.
-```{r load data and set tissue type, include=FALSE}
-#set tissue type
-sample_tissue <- "ileum"
 ```
@@
   filter(tissue == sample_tissue) %>%
+  filter(exp_number == study_round) %>% 
   group_by(variables) %>% 
-  do(lm(data = ., values ~ MAT+FEED+MAT*FEED) %>% tidy)  %>% filter(term!='(Intercept)')
+  do(lm(data = ., values ~ maternal_treatment+type_of_feed +maternal_treatment*type_of_feed) %>% tidy)  %>% filter(term!='(Intercept)')
 
@@
   select(-statistic) %>% 
+  filter(p.value < 0.05) %>% 
   kable(., caption = 'Effect of treatment factors on cytokine expression levels', digits = 3)

### 2022-11-11 code/Round 2/Cytokines round 2.Rmd
Progressed

@@
 # Import dataset
-#cytokines <- read_excel("../../Data/RawData/cytokines_round_1_r.xlsm")
-cytokines <- read_excel("../../Data/CleanData/cytokines.xlsx")
-glimpse(cytokines)
+cytokines <- read_excel("data/processed/cytokines.xlsx")
 ```
 
-Set the tissue type to whatever tissue your data are extracted from, or that you want to analyze first.
+Set the tissue type and experimental round here:
 
@@
 study_round <- "2"
+```
+
+```{r load data, echo=FALSE}
+cytokines %>%   # run normality test on tissue weight
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  group_by(group) %>% 
+  shapiro_test(tissue_weight_mg)
+
+# group    variable         statistic      p
+ # <chr>    <chr>                <dbl>  <dbl>
+# CON-BF   tissue_weight_mg     0.833 0.197 
+#GENTA-BF tissue_weight_mg     0.748 0.0120
+#VANCO-BF tissue_weight_mg     0.833 0.0861
+#look at tissue weight distribution
+
+cytokines %>%       #Weight of tissue samples
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
+  geom_boxplot() +
+  geom_point() +
+  labs(
+    title = "Tissue weight of intestinal samples, experiment 2"
+  )
+```
+<!-- Looks fine. Go ahead and  adjust-->
+
+```{r adjust for tissue weight , echo=FALSE}
+cytokines_adjusted <- cytokines %>% 
+  pivot_longer(.,
+               cols = INFg:TNFa ,
+               names_to = "variables",
+               values_to = "values"
+  ) %>% 
+  mutate(
+    group = recode_factor(  #Change the order of the groups to desired plot order
+      group,
+      "CON-BF"= "CON-BF",
+      "CON-FF" = "CON-FF",
+      "AB-BF"= "AB-BF",
+      "AB-FF"= "AB-FF"
+      )) %>% 
+  mutate(values = if_else(tissue == "ileum", values/tissue_weight_mg, values)) %>% #Adjust for tissue weight
+  pivot_wider(
+    id_cols = c(
+      sample_id,exp_number, group, received_antibiotics,maternal_treatment, tissue,tissue_weight_mg),
+    names_from = variables, 
+    values_from = values)
+
+#____Pivot longer____
+cytokines_long <- cytokines_adjusted %>% 
+  pivot_longer(.,
+               cols = INFg:TNFa ,
+               names_to = "variables",
+               values_to = "values"
+  )
+```
+
+Check for data distribution.
+
+```{r visualize data, eval=FALSE, include=FALSE}
+cytokines_long %>%                       #normality_check  of cytokine distributions
+  filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  #filter(variables != "INFg") %>% 
+  drop_na(values) %>% 
+  group_by(variables, group) %>% 
+  shapiro_test(values) %>% 
+  filter(p < 0.05)
+
+#Non-normal distributions: 
+#  group    variables variable statistic      p
+#  <fct>    <chr>     <chr>        <dbl>  <dbl>
+# VANCO-BF IL-1b     values       0.795 0.0363
+# GENTA-BF IL-4      values       0.777 0.0240
+
+#log10 transformed shapiro_test - failed variables
+cytokines_long %>% 
+ filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  filter(variables != "INFg") %>% 
+  drop_na(values) %>% 
+  mutate(values = log10(values)) %>% 
+  group_by(variables, group) %>% 
+  shapiro_test(values) %>% 
+  filter(p < 0.05)
+
+#  group    variables variable statistic       p
+#  <fct>    <chr>     <chr>        <dbl>   <dbl>
+# VANCO-BF IL-1b     values       0.727 0.00728     # transformation helped.
+
+#Density plots to show distribution of cytokines
+histogram <- cytokines_long %>% 
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  filter(variables != "INFg") %>% 
+  drop_na(values) %>% 
+  ggplot(aes(x = values, fill = group)) + 
+  geom_density() +
+  facet_wrap(~ variables, ncol = 2,
+             scales = "free")
+
+log10_histogram <- cytokines_long %>% 
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  #filter(variables != "INFg") %>% 
+  drop_na(values) %>% 
+  ggplot(aes(x = log10(values), fill = group)) + 
+  geom_density() +
+  facet_wrap(~ variables, ncol = 2,
+             scales = "free")
+
+#Jitter plots. Works
+jitter_plot <- cytokines_long %>% 
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  #filter(variables != "INFg") %>% 
+  drop_na(values) %>% 
+  ggplot(aes(x = group, y = values, color = group)) +
+  geom_jitter() +
+  facet_wrap(~ variables,
+             scales = "free")
+
+log10_jitter_plot <- cytokines_long %>% 
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  filter(variables != "INFg") %>% 
+  drop_na(values) %>% 
+  mutate(adjusted_values = log10(values)) %>% 
+  ggplot(aes(x = group, y = values, color = group)) +
+  geom_jitter() +
+  facet_wrap(~ variables,
+             scales = "free") 
 ```
@@
 
-### Plots and regression models
-
-First the data must be transformed to a long format (pivot longer). Mutate any new columns with group treatments needed for the regression model (in this case MAT and FEED (experimental factors)).
-
-```{r long format, echo=FALSE}
-cytokines_long <- cytokines %>% 
-  filter(exp_number == study_round) %>% 
-  filter(tissue == sample_tissue) %>%
-  pivot_longer(.,
-               cols = INFg:TNFa ,
-               names_to = "variables",
-               values_to = "values"
-  ) %>% 
-  drop_na(
-    values) %>% 
-  # mutate(MAT = ifelse(str_detect(group,'AB'),'AB',
-  #                                 ifelse(str_detect(group,'CON'),'CON', "miss")),
-  #        FEED = ifelse(str_detect(group,'FF'),'FF',
-  #                                 ifelse(str_detect(group,'BF'),'BF', "miss"))) %>% 
-  mutate(
-    group = recode_factor(  #Change the order of the groups to desired plot order
-      group,
-      "CON-BF"= "CON-BF",
-      "CON-FF" = "CON-FF",
-      "AB-BF"= "AB-BF",
-      "AB-FF"= "AB-FF"
-      ))
-
-head(cytokines_long)
-
-```
-
 To see all the cytokine leves plotted run this code:
@@
 plot_all_cytokines <- cytokines_long %>%
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
   #filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot
@@
        x = "",
-       y = "Tissue conc.",
+       y = "pg/ml per mg tissue",
        fill = "Groups",
@@
 all_cytokines_patch <- plot_all_cytokines + plot_annotation (
-  title = 'Cytokine expression',
-   subtitle = 'All cytokines measured',
-  tag_levels = 'A',
+  #title = 'Cytokine expression',
+   #subtitle = 'All cytokines measured',
+  #tag_levels = 'A',
   caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin',

### 2022-11-11 code/Round 2/FITC round 2.Rmd
Progressed

@@
 # Import dataset
-fitc <- read_excel("../../Data/CleanData/fitc.xlsx")
+fitc <- read_excel("data/processed/fitc.xlsx")
 glimpse(fitc)
@@
 fitc<- fitc %>% 
-  rename("µg/ml" = `result_µg/ml_undiluted`)
+  rename(fitc_level = `result_µg/ml_undiluted`)
 ```
@@
 ```{r data distribution}
+
+fitc %>% 
+  filter(exp_number == "2") %>% 
+  drop_na(fitc_level) %>% 
+  group_by(group) %>% 
+  shapiro_test(fitc_level) 
+#%>% filter(p < 0.05)
+
+#Jitter plots. Works
+jitter_plot <- fitc %>%
+  filter(exp_number == 2) %>% 
+  #filter(variables != "INFg") %>% 
+  drop_na(fitc_level) %>% 
+  ggplot(aes(x = group, y = fitc_level, color = group)) +
+  geom_jitter()
+
 no_outliers <- fitc %>% 
@@
 
-The values will be displayed as mean + sd if normally distributed or median + IQR if not.
+The values will be displayed as median + IQR.
 An appropriate statistical test for differences will be performed (student-t for two, normally distributed groups, Anova for more than 2 groups, Kruskal wallis for non-normally distrubuted results.)
@@
   filter(exp_number == study_round) %>% 
-  select(group2, "µg/ml") %>% 
+  select(group2, fitc_level) %>% 
+  rename("µg/mL" = fitc_level) %>% 
   tbl_summary(by = group2,
@@
   filter(exp_number == study_round) %>% 
-  rename(values = "µg/ml") %>% 
+  rename(values = fitc_level) %>% 
   drop_na(

### 2022-11-11 code/weight_analysis.Rmd
Progressed

@@
 library(patchwork)
+library(tidyverse)
+library(readxl)
+library(broom)
 ```
@@
   # #mutate(across(where(is_character),as_factor))
+```
+
+#_bodyweight in all experimets._
+```{r}
+startweight <- data %>% 
+  filter(exp_number %in% c("1", "2", "3","4")) %>% 
+  filter(time %in% c("day_3", "day_5")) %>%
+  drop_na(bodyweight) %>% 
+  group_by(exp_number, type_of_feed, maternal_treatment, group, time) %>% 
+  summarise(startweight = mean(bodyweight),
+            sd = sd(bodyweight))
+
+startweight %>% 
+  filter(exp_number == "2") %>% 
+  arrange(time)
+```
+
+#_birthweight in exp 2_
+```{r}
+birthweight <- data %>% 
+  filter(exp_number %in% "2") %>% 
+  filter(time %in% c("day_0", "day_1", "day_2")) %>%
+  drop_na(bodyweight) %>% 
+  group_by(exp_number, maternal_treatment, time) %>% 
+  summarise(birthweight = mean(bodyweight),
+            sd = sd(bodyweight))
+
+lm_birthweight <- data %>%  #LM for effect of gestational age on offspring birthweight. 
+  filter(exp_number %in% "2") %>% 
+  filter(time %in% c("day_0", "day_1", "day_2")) %>%
+  drop_na(bodyweight) %>% 
+  filter(time == "day_0") %>% 
+  mutate(gestation_age = if_else(maternal_treatment == "CON", 21,20)) %>% 
+  do(lm(bodyweight ~ gestation_age,.) %>% tidy)
+lm_birthweight
+
+data %>% # ANOVA for effect of gestational age on offspring birthweight. 
+  filter(exp_number %in% "2") %>% 
+  filter(time %in% c("day_0", "day_1", "day_2")) %>%
+  drop_na(bodyweight) %>% 
+  filter(time == "day_0") %>% 
+  mutate(gestation_age = if_else(maternal_treatment == "CON", 21,20)) %>% 
+  rstatix::anova_test(., bodyweight ~ gestation_age)
+
 ```
@@
 
-#add gt table with group results
-
 ```
 
+#_Weight development during experiments_
 ```{r eval=FALSE, include=FALSE}
-### Weight development during experiments   ###########
-
 bw_development_feed <- data %>% 

### 2022-11-11 code/weight_development.R

@@ -1,0 +1,38 @@
+library(tidyverse)
+library(readxl)
+
+data <- read_excel("data/processed/weight_development.xlsx") %>% 
+  select(sample_id:cage, bodyweight_birthday_group_average, bodyweight_birthday_plus_24H_group_average, bodyweight_birthday_plus_48H_group_average, bodyweight_baseline, bodyweight_baseline_plus_24h, bodyweight_baseline_plus_48h) %>% 
+  rename(day_0 = "bodyweight_birthday_group_average",
+         day_1 = "bodyweight_birthday_plus_24H_group_average", 
+         day_2 = "bodyweight_birthday_plus_48H_group_average",
+         day_3 = bodyweight_baseline, 
+         day_4 = `bodyweight_baseline_plus_24h`, 
+         day_5 = `bodyweight_baseline_plus_48h`,) %>% 
+  pivot_longer(., cols = c(day_0:day_5),
+               names_to = "time",
+               values_to = "bodyweight") %>% 
+  mutate(received_antibiotics = recode_factor(received_antibiotics, 
+                                              yes = "MAT",
+                                              no = "CON"))
+startweight <- data %>% 
+  filter(exp_number %in% c("1","3","4")) %>% 
+  filter(time %in% c("day_3", "day_5")) %>%
+  drop_na(bodyweight) %>% 
+  group_by(exp_number, type_of_feed, time) %>% 
+  summarise(startweight = mean(bodyweight),
+            sd = sd(bodyweight))
+
+birthweight <- data %>% 
+  filter(exp_number %in% "2") %>% 
+  filter(time %in% c("day_0", "day_1", "day_2")) %>%
+  drop_na(bodyweight) %>% 
+  group_by(exp_number, maternal_treatment, time) %>% 
+  summarise(birthweight = mean(bodyweight),
+            sd = sd(bodyweight))
+
+birthweight %>% 
+  filter(time == "day_0") %>% 
+  knitr::kable()
+  
+

### 2022-11-12 code/Round 2/Cytokines round 2.Rmd
Progress again

@@
 sample_tissue <- "ileum"
-study_round <- "2"
+study_round <- "3"
 ```
@@
 signif_cytokines <- cytokines_long %>%
+    filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
   filter(variables %in% tbl1_p$Cytokines) %>%  #Exclude any cytokines not wanted in the plot

### 2022-11-12 code/Round 3/Cytokines round 3.Rmd
Progress again

@@
 #cytokines <- read_excel("../../Data/RawData/cytokines_round_1_r.xlsm")
-cytokines <- read_excel("../../Data/CleanData/cytokines.xlsx")
+cytokines <- read_excel("data/processed//cytokines.xlsx")
 glimpse(cytokines)
@@
 
-Set the tissue type to whatever tissue your data are extracted from, or that you want to analyze first.
+```{r load data and set tissue type, include=FALSE}
+#set tissue type
+sample_tissue <- "ileum"
+study_round <- "3"
+```
+
+```{r load data, echo=FALSE}
+cytokines %>%   # run normality test on tissue weight
+   filter(exp_number == 3) %>% 
+  filter(tissue == sample_tissue) %>%
+  group_by(type_of_feed) %>% 
+  shapiro_test(tissue_weight_mg)
+
+# group    variable         statistic      p
+ # <chr>    <chr>                <dbl>  <dbl>
+# CON-BF   tissue_weight_mg     0.833 0.197 
+#GENTA-BF tissue_weight_mg     0.748 0.0120
+#VANCO-BF tissue_weight_mg     0.833 0.0861
+#look at tissue weight distribution
+
+cytokines %>%       #Weight of tissue samples
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
+  geom_boxplot() +
+  geom_point() +
+  labs(
+    title = "Weight of intestinal samples",
+      x = "",
+       y = "Tissue weight (mg)",
+       fill = "Groups",
+       caption = "Matrinem round 3") +
+  theme_bw()
+
+cytokines %>%   # run normality test on tissue weight
+   filter(exp_number == 3) %>% 
+  filter(tissue == sample_tissue) %>%
+  group_by(type_of_feed) %>% 
+  summarise(mean = mean(tissue_weight_mg),
+            sd = sd(tissue_weight_mg))
+
+cytokines %>%   # run normality test on tissue weight
+   filter(exp_number == 3) %>% 
+  filter(tissue == sample_tissue) %>%
+  lm(tissue_weight_mg ~ type_of_feed,.) %>% 
+  summary()
+
+```
+<!-- Looks fine. Go ahead and  adjust-->
 
@@
   modify_header(
-    label ~ "Cytokines") %>% 
-  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Treatment groups (N total = {N})"))
+    label ~ "**Groups**:") %>% 
+  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Matrinem round 3 (N total = {N})"))
 
@@
     source_note = " * P-value indicates overall group differences"
-  ) %>%
-  tab_source_note(
-    source_note = md("**Matrinem round 3**")
   ) %>%
@@
 #tbl1_p contains cytokines with significant changes
+my_comparisons <- list( c("CON-BF", "CON-FF"), c("GENTA-BF", "GENTA-FF"))
 
 signif_cytokines <- cytokines_long %>%
-  filter(variables %in% tbl1_p$Cytokines) %>%  #Exclude any cytokines not wanted in the plot
+  filter(variables %in% c("IL-2","IL-6", "KCGRO")) %>% 
+  filter(values > 0.5) %>% #Exclude any cytokines not wanted in the plot
 ggplot(aes(x = group, 
@@
   geom_boxplot() +
-  stat_compare_means(aes(label = ..p.signif..),
-                   method = "wilcox.test", ref.group = "CON-BF",
-                   label.y.npc = 0.55) +
-  stat_compare_means(
-    aes(label = paste0("Global p = ", ..p.format.., "\n(Kruskal-Wallis)")),
-    label.y.npc = 0.85,
-    label.x.npc = 0.5) +
+  stat_compare_means(comparisons = my_comparisons, label.y.npc =  c(0.20, 0.15), method = "wilcox.test") +
   facet_wrap(~variables, 
-             scales = 'free', 
-             nrow = 3,
-             ncol = 3) +
+             scales = 'free') +
   labs(title = "Levels of pro-inflammatory cytokines in the ileum", # Change title, and axis names
@@
        x = "",
-       y = "Tissue conc.",
+       y = "Cytokine concentration (pg/ml)",
        fill = "Groups") +
@@
 signif_cytokines_patch <- signif_cytokines + plot_annotation (
-  title = 'Cytokine expression',
-  subtitle = 'Statistically significant changes',
-  tag_levels = 'A',
+  #title = 'Cytokine expression',
+  #subtitle = 'Statistically significant changes',
+  #tag_levels = 'A',
   caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding. \n
-  Pairwise comparisons with CON-BF using wilcoxon test. NS = Not Significant, * p<0.05, ** p<0.01, *** p<0.001',
+  Pairwise comparisons using wilcoxon test.',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
@@
 #ggsave('../../Results/Round 1/Significant Cytokines-boxplot.png',signif_cytokines_patch)
+```
+
+```{r}
+my_comparisons <- list( c("CON-BF", "CON-FF"), c("GENTA-BF", "GENTA-FF"), c("VANCO-BF", "VANCO-FF"), c("GENTA-FF", "VANCO-FF"))
+
+signif_cytokines_plot <- cytokines_long %>%    #____Normal values_____#
+filter(variables %in% tbl1_p$`**Groups**:`) %>% 
+  ggplot(aes(x = group, 
+           y = values, 
+           fill = group)) + 
+  geom_boxplot() +
+  geom_point() +
+  #coord_cartesian(ylim = c(1.5, 16)) + 
+  stat_compare_means(comparisons = my_comparisons, label.y.npc = c(0.30, 0.30, 0.30, 0.50), method = "wilcox.test")+
+  # stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
+  #   label.y.npc = 0.75,
+  #   label.x.npc = 0.55) +
+  labs( # Change title, and axis names
+       x = "",
+       y = "Cytokine concentration (pg/ml)",
+       fill = "Groups",
+       caption = "Matrinem round 3") +
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    #axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="lightgrey"),#Change legend box background color
+    plot.caption = element_text(color = "darkgrey", size=5)) + 
+  scale_fill_brewer(palette="Set1")
+
+cytokine_patch <- signif_cytokines_plot + plot_annotation (
+  title = 'Proinflammatory Cytokines',
+   #subtitle = 'All fitc measured',
+  #tag_levels = 'A',
+  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
+  Pariwise comparisons = Wilcoxon test.',
+  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
+                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
+                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))
+
+cytokine_patch
 ```

### 2022-11-12 code/Round 3/FITC round 2 and 3 copy.Rmd
Progress again

@@
 # Import dataset
-fitc <- read_excel("../../Data/CleanData/fitc.xlsx") %>% 
-  rename(values = `result_µg/ml_undiluted`) %>% 
+fitc <- read_excel("data/processed/fitc.xlsx") %>% 
+  rename(values = `result_µg/ml_undiluted`) %>%  #rename undiluted column.
   drop_na(values)
@@
 glimpse(fitc)
+```
+
+Set the experimental round here:
+
+```{r load data and set tissue type, include=FALSE}
+#set tissue type
+study_round <- c("2","3")
+#study_round <- 3
 ```
@@
 
-normality_check <- fitc %>% 
-  filter(exp_number == 2 | exp_number == 3) %>% 
+fitc %>% 
+  filter(exp_number == study_round) %>% 
+  group_by(exp_number, group2) %>% 
+  shapiro_test(values)
+#  group2   variable statistic           p
+#  <chr>    <chr>        <dbl>       <dbl>
+# CON-BF   values       0.975 0.946      
+# CON-FF   values       0.912 0.295      
+# GENTA-BF values       0.965 0.838      
+# GENTA-FF values       0.962 0.833      
+# VANCO-BF values       0.854 0.0155     
+# VANCO-FF values       0.526 0.000000232
+
+fitc %>%  # try log-tranformation
+  filter(exp_number == study_round) %>% 
   group_by(group2) %>% 
+  mutate(values = log10(values)) %>% 
   shapiro_test(values)
 
-normality_check %>% 
-  filter(p < 0.05)
-
-# fitc<- fitc %>% 
-#   mutate(values = log10(values))     # Log transforming doesn't help.
+#group2   variable statistic         p
+#  <chr>    <chr>        <dbl>     <dbl>
+# CON-BF   values       0.877 0.0640   
+# CON-FF   values       0.955 0.731    
+# GENTA-BF values       0.851 0.0445   
+# GENTA-FF values       0.943 0.636    
+# VANCO-BF values       0.833 0.00778  
+# VANCO-FF values       0.695 0.0000171   # Log transforming doesn't help.
 
@@
 fitc %>% 
-  filter(exp_number == 2 | exp_number == 3) %>% 
+  filter(exp_number == study_round) %>% 
   ggplot(aes(x = values, fill = group2)) + 
@@
 #Jitter plots. Works
-fitc %>% 
-  filter(exp_number == 2 | exp_number == 3) %>% 
+fitc_long %>% 
+  filter(exp_number == study_round) %>% 
   ggplot(aes(x = group2, y = values)) + 
@@
               missing = "no",
-              type = list(values ~ 'continuous2', Log10_values ~ 'continuous2'),
-              statistic = all_continuous() ~ c("{mean} ({sd})", "{median} ({p25}, {p75})"),
-              label = list(values ~ "µg/mL", Log10_values ~ "Log transformed")) %>%
+              statistic = all_continuous() ~ "{mean} ({sd})") %>%
   add_p(values ~ "aov") %>% 
@@
     stat_6 = "**VANCO-FF**  \n (n = {n})") %>% 
-  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Treatment groups (N total = {N})")) %>% 
-  bold_labels()
+  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Treatment groups (N total = {N})"))
 
@@
   tab_options(heading.title.font.size = 20,
-              heading.background.color = "lightgrey"
+              heading.background.color = "lightgrey",
               ) %>% 
@@
 tbl1_gt
-tbl1_gt %>% 
-  gtsave("../../Results/Round 3/FITC table round 3 TEST.png", expand = 50) #use .tex for latex files
+# Save and export files
+# tbl1 %>%  #Save as excel file.
+#   as_hux_table() %>%
+#   huxtable::quick_xlsx(., file = "../Results/Round 1/ileum-cytokines.xlsx")
+# 
+# tbl1 %>% #Save as image file.
+# as_flex_table() %>%
+#   flextable::save_as_image(path = "../../Results/Round 2/ileum_cytokines_table.png",webshot = "webshot")
 ```
@@
 fitc_long <- fitc %>% 
-  filter(exp_number == 2 | exp_number == 3)%>% 
   drop_na(
     values) %>% 
-    mutate(Log10_values = log10(values)) %>% 
   mutate(
@@
 
-Create a comparison vector to see pariwise comparisons.
+Create a comparison vector to see pariwise comparisons
+
 ```{r plots all fitc, warning=FALSE}
-my_comparisons <- list( c("CON-BF", "GENTA-BF"), c("CON-BF", "VANCO-BF"), c("GENTA-BF", "VANCO-BF"))
-```
+my_comparisons <- list( c("CON-BF", "CON-FF"), c("GENTA-BF", "GENTA-FF"), c("VANCO-BF", "VANCO-FF"), c("GENTA-FF", "VANCO-FF"))
 
-Boxplots with jitter - normal values
-```{r plots all fitc, warning=FALSE}
-plot_all_fitc <- fitc_long %>%
+plot_all_fitc <- fitc_long %>%    #____Normal values_____#
+  filter(exp_number == study_round) %>% 
 ggplot(aes(x = group2, 
@@
   #coord_cartesian(ylim = c(1.5, 16)) + 
-  # stat_compare_means(comparisons = my_comparisons, label.y = c(2.5, 2.2, 2), method = "wilcox.test")+
-  stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
-    label.y.npc = 0.75,
-    label.x.npc = 0.55) +
+  stat_compare_means(comparisons = my_comparisons, label.y = c(30, 30, 30, 50), method = "wilcox.test")+
+  # stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
+  #   label.y.npc = 0.75,
+  #   label.x.npc = 0.55) +
   labs( # Change title, and axis names
@@
        fill = "Groups",
-       caption = "Matrinem round 2 and 3") +
+       caption = "Matrinem round 3") +
   theme_bw()+
@@
 
-all_fitc_patch <- plot_all_fitc + plot_annotation (
-  title = 'Intestinal permeability',
-   #subtitle = 'All fitc measured',
-  #tag_levels = 'A',
-  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
-  Overall difference= Kruskal-Wallis.',
-  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
-                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
-                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))
-
-all_fitc_patch
-#ggsave('../../Results/Round 3/FITC boxplots.png',all_fitc_patch)
-```
-
-Log transformed values
-```{r plots all fitc, warning=FALSE}
-
-plot_all_fitc_log10 <- fitc_long %>%
+plot_all_fitc_log <- fitc_long %>%        #Log_transformed values_____#
+  filter(exp_number == study_round) %>% 
 ggplot(aes(x = group2, 
-           y = Log10_values, 
+           y = log10(values), 
            fill = group2)) + 
@@
   #coord_cartesian(ylim = c(1.5, 16)) + 
-  # stat_compare_means(comparisons = my_comparisons, label.y = c(2.5, 2.2, 2), method = "wilcox.test")+
-  stat_compare_means(aes(label = paste0("Anova \n p = ", ..p.format..)),
-    label.y.npc = 0.75,
-    label.x.npc = 0.55, method = "aov") +
+  stat_compare_means(comparisons = my_comparisons, label.y = c(1.5, 1.5, 1.5, 1.7), method = "wilcox.test")+
+  # stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
+  #   label.y.npc = 0.75,
+  #   label.x.npc = 0.55) +
   labs( # Change title, and axis names
        x = "",
-       y = "Serum FITC conc. log10 transformed",
+       y = "Log10 serum FITC conc. (µg/ml)",
        fill = "Groups",
-       caption = "Matrinem round 2 and 3") +
+       caption = "Matrinem round 3") +
   theme_bw()+
@@
 
-all_fitc_patch_log10 <- plot_all_fitc_log10 + plot_annotation (
+all_fitc_patch <- plot_all_fitc + plot_all_fitc_log + plot_annotation (
   title = 'Intestinal permeability',
    #subtitle = 'All fitc measured',
-  #tag_levels = 'A',
+  tag_levels = 'A',
   caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
-  Overall difference= Anova',
+  Pariwise comparisons = Wilcoxon test.',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
@@
 
-all_fitc_patch_log10
-ggsave('../../Results/Round 3/FITC boxplot log 10.png',all_fitc_patch)
+all_fitc_patch
+#ggsave('../../Results/Round 1/All fitc-boxplot.png',all_fitc_patch)
 ```
-
-```{r}
-FITC_PATCH <- all_fitc_patch + all_fitc_patch_log10
-```

### 2022-11-12 code/Round 3/FITC round 3.Rmd
Progress again

@@
 glimpse(fitc)
+```
+
+Set the experimental round here:
+
+```{r load data and set tissue type, include=FALSE}
+#set tissue type
+#study_round <- c("2","3")
+study_round <- 3
 ```
@@
 
-normality_check <- fitc %>% 
-  filter(exp_number == 3) %>% 
+fitc %>% 
+  filter(exp_number == study_round) %>% 
+  group_by(exp_number, group2) %>% 
+  shapiro_test(values)
+#  group2   variable statistic           p
+#  <chr>    <chr>        <dbl>       <dbl>
+# CON-BF   values       0.975 0.946      
+# CON-FF   values       0.912 0.295      
+# GENTA-BF values       0.965 0.838      
+# GENTA-FF values       0.962 0.833      
+# VANCO-BF values       0.854 0.0155     
+# VANCO-FF values       0.526 0.000000232
+
+fitc %>%  # try log-tranformation
+  filter(exp_number == study_round) %>% 
   group_by(group2) %>% 
+  mutate(values = log10(values)) %>% 
   shapiro_test(values)
 
-# fitc<- fitc %>% 
-#   mutate(values = log10(values))     # Log transforming doesn't help.
+#group2   variable statistic         p
+#  <chr>    <chr>        <dbl>     <dbl>
+# CON-BF   values       0.877 0.0640   
+# CON-FF   values       0.955 0.731    
+# GENTA-BF values       0.851 0.0445   
+# GENTA-FF values       0.943 0.636    
+# VANCO-BF values       0.833 0.00778  
+# VANCO-FF values       0.695 0.0000171   # Log transforming doesn't help.
 
@@
 tbl1 <- fitc %>% 
-  filter(exp_number == 2 | exp_number == 3) %>% 
-  mutate(Log10_values = log10(values)) %>% 
-  select(group2, values, Log10_values) %>% 
+  filter(exp_number == 3) %>% 
+  mutate("FITC log10(ug/ml)" = log10(values)) %>% 
+  mutate("FITC (ug/ml)" = values) %>% 
+  select(group2, `FITC (ug/ml)`, `FITC log10(ug/ml)`) %>% 
   tbl_summary(by = group2,
@@
               missing = "no",
-              statistic = all_continuous() ~ "{mean} ({sd})") %>%
-  add_p(values ~ "aov") %>% 
+              statistic = vars(`FITC log10(ug/ml)`) ~ "{mean} ({sd})") %>%
+  add_p(`FITC log10(ug/ml)` ~ "aov") %>% 
   modify_header(
-    label = "FITC Conc.",
+    label = "**Groups**:",
     stat_1 = "**CON-BF**  \n (n = {n})",
@@
     stat_6 = "**VANCO-FF**  \n (n = {n})") %>% 
-  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Treatment groups (N total = {N})"))
+  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Matrinem round 3 (N total = {N})"))
 
@@
   tab_source_note(
-    source_note = " * P-value indicates overall group differences"
-  ) %>%
-  tab_source_note(
-    source_note = md("**Matrinem round 3**")
+    source_note = " *Geometric mean and (sd) for log-transformed values"
   ) %>%
@@
 fitc_long <- fitc %>% 
-  filter(exp_number == study_round) %>% 
   drop_na(
@@
 ```{r plots all fitc, warning=FALSE}
-my_comparisons <- list( c("CON-BF", "GENTA-BF"), c("CON-BF", "VANCO-BF"), c("GENTA-BF", "VANCO-BF"))
+my_comparisons <- list( c("CON-BF", "CON-FF"), c("GENTA-BF", "GENTA-FF"), c("VANCO-BF", "VANCO-FF"), c("GENTA-FF", "VANCO-FF"))
 
-plot_all_fitc <- fitc_long %>%
+plot_all_fitc <- fitc_long %>%    #____Normal values_____#
+  filter(exp_number == study_round) %>% 
+  filter(values > 8) %>% 
 ggplot(aes(x = group2, 
@@
        fill = "Groups",
-       caption = "Matrinem round 2") +
+       caption = "Matrinem round 3") +
   theme_bw()+
@@
 
-all_fitc_patch <- plot_all_fitc + plot_annotation (
+plot_all_fitc_log <- fitc_long %>%        #Log_transformed values_____#
+  filter(exp_number == study_round) %>% 
+  filter(values > 8) %>% 
+ggplot(aes(x = group2, 
+           y = log10(values), 
+           fill = group2)) + 
+  geom_boxplot() +
+  geom_point() +
+  #coord_cartesian(ylim = c(1.5, 16)) + 
+  stat_compare_means(comparisons = my_comparisons, label.y = c(1.5, 1.5, 1.5, 1.7), method = "wilcox.test")+
+  # stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
+  #   label.y.npc = 0.75,
+  #   label.x.npc = 0.55) +
+  labs( # Change title, and axis names
+       x = "",
+       y = "Log10 serum FITC conc. (µg/ml)",
+       fill = "Groups",
+       caption = "Matrinem round 3") +
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    #axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="lightgrey"),#Change legend box background color
+    plot.caption = element_text(color = "darkgrey", size=5)) + 
+  scale_fill_brewer(palette="Set1")
+
+all_fitc_patch <- plot_all_fitc + plot_all_fitc_log + plot_annotation (
   title = 'Intestinal permeability',
    #subtitle = 'All fitc measured',
-  #tag_levels = 'A',
+  tag_levels = 'A',
   caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
-  Overall difference= Kruskal-Wallis \n
   Pariwise comparisons = Wilcoxon test.',

### 2022-11-13 code/Round 1/Cytokines adjusted.Rmd
Almost done with results

@@
 sample_tissue <- "ileum"
-study_round <- "2"
+study_round <- "1"
 ```
@@
   filter(tissue == sample_tissue) %>%
-  filter(variables != "INFg") %>% 
+  #filter(variables != "INFg") %>% 
   drop_na(values) %>% 
   ggplot(aes(x = values, fill = group)) + 
-  geom_density() +
+  geom_histogram() +
   facet_wrap(~ variables, ncol = 2,
@@
               include = !"INFg",
-             # statistic = list(all_continuous() ~ "{mean} ({sd})")  #Un-comment this line to get mean+sd. 
-             #%>%
-statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
+              #type = all_continuous() ~ "continuous2",
+              statistic = all_continuous() ~ #c("{mean} ({sd})",         
+                                     "{median} ({p25}, {p75})") %>%
   add_p() %>% 
@@
   modify_header(
-    label ~ "Cytokines") %>% 
-  modify_spanning_header((c("stat_1", "stat_2", "stat_3", "stat_4") ~ "Treatment groups (N total = {N})"))
+    label ~ "**Groups:**")
+  #modify_spanning_header((c("stat_1", "stat_2", "stat_3", "stat_4") ~ "Experiment 1 (N total = {N})")) %>%
 
@@
   as_gt() %>%
-  #tab_header(title = md("**Proinflammatory cytokines**")) %>%
+  tab_header(title = md("**Tissue concentraions of proinflammatory cytokines** *(pg/ml/mg tissue)*")) %>%
   opt_align_table_header("center") %>% 
-  # tab_options(heading.title.font.size = 20,
-  #             heading.background.color = "lightgrey",
-  #             ) %>% 
-  tab_source_note(
-    source_note = " * P-value indicates overall group differences"
+   tab_options(heading.title.font.size = 20,
+               heading.background.color = "lightgrey",
                ) %>% 
   tab_source_note(
-    source_note = md("**Matrinem round 1**")
+    source_note = " *P-value indicates overall group differences"
   ) %>%
-  cols_width(label ~ px(100),
+     tab_source_note(
+    source_note = "Matrinem Experiment 1"
+  ) %>%
+  cols_width(label ~ px(200),
              stat_1 ~ px(150),
-             stat_2 ~ px(150),
-             stat_3 ~ px(150),
-             stat_4 ~ px(150)) 
+             stat_2 ~ px(200),
+             stat_3 ~ px(200),
+             stat_4 ~ px(200)) 
 tbl1_gt
@@
   filter(tissue == sample_tissue) %>% 
-  #filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot
+  filter(!variables %in% c("INFg", "IL-12p70")) %>%  #Exclude any cytokines not wanted in the plot
 ggplot(aes(x = group, 
@@
              ncol = 2) +
-  labs(title = "Levels of pro-inflammatory cytokines in the ileum",  # Change title, and axis names
+  labs(title = "Levels of pro-inflammatory cytokines in serum",  # Change title, and axis names
        x = "",
-       y = "Tissue conc. (pg/mg)",
+       y = "Serum conc. (pg/ml)",
        fill = "Groups") +
@@
 all_cytokines_patch <- plot_all_cytokines + plot_annotation (
-  title = 'Cytokine expression',
-   subtitle = 'All cytokines measured',
-  tag_levels = 'A',
+  #title = 'Cytokine expression',
+   #subtitle = 'All cytokines measured',
+  #tag_levels = 'A',
   caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding',

### 2022-11-13 code/Round 2/FITC round 2.Rmd
Almost done with results

@@
 #set tissue type
-study_round <- "2"
+study_round <- "4"
 
@@
 fitc %>% 
-  filter(exp_number == "2") %>% 
+  filter(exp_number == study_round) %>% 
   drop_na(fitc_level) %>% 
@@
 #Jitter plots. Works
-jitter_plot <- fitc %>%
-  filter(exp_number == 2) %>% 
-  #filter(variables != "INFg") %>% 
+fitc %>%
+  filter(exp_number == study_round) %>% 
   drop_na(fitc_level) %>% 
@@
 no_outliers <- fitc %>% 
-  filter(exp_number == 2) %>% 
+  filter(exp_number == study_round) %>% 
   group_by(group2) %>% 

### 2022-11-13 code/Round 4/Cytokines round 4.Rmd
Almost done with results

@@
 #cytokines <- read_excel("../../Data/RawData/cytokines_round_1_r.xlsm")
-cytokines <- read_excel("../../Data/CleanData/cytokines.xlsx")
+cytokines <- read_excel("data/processed/cytokines.xlsx")
 glimpse(cytokines)
@@
 
-Below is a table of the mean levels of cytokines in `r sample_tissue` tissue for each of the experimental groups.
+```{r}
+cytokines %>%   # run normality test on tissue weight
+   filter(exp_number == 4) %>% 
+  filter(tissue == sample_tissue) %>%
+  group_by(type_of_feed) %>% 
+  shapiro_test(tissue_weight_mg)
 
-The values will be displayed as mean + sd if normally distributed or median + IQR if not.
-An appropriate statistical test for differences will be performed (student-t for two, normally distributed groups, Anova for more than 2 groups, Kruskal wallis for non-normally distrubuted results.)
+cytokines %>%       #Weight of tissue samples
+   filter(exp_number == study_round) %>% 
+  filter(tissue == sample_tissue) %>%
+  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
+  geom_boxplot() +
+  geom_point() +
+
+  labs(
+    title = "Weight of intestinal samples",
+      x = "",
+       y = "Tissue weight (mg)",
+       fill = "Groups",
+       caption = "Matrinem round 4") +
+  theme_bw()
+
+```
 
@@
     label ~ "Cytokines",
-    stat_1 = "**VANCO-BF (CON)**\n(n = {n})",
-    stat_2 = "**VANCO-FF (FVT)**\n(n = {n})",
-    stat_3 = "**VANCO-FF (SM)**\n(n = {n})") %>% 
+    stat_1 = "**VANCO-BF** \n(n = {n})",
+    stat_2 = "**VANCO-FVF** \n(n = {n})",
+    stat_3 = "**VANCO-SM** \n(n = {n})") %>% 
   modify_spanning_header((c("stat_1", "stat_2", "stat_3") ~ "Treatment groups (N total = {N})"))
@@
  as_gt() %>%
-  tab_header(title = md("**Proinflammatory cytokines (pg/mL)**")) %>%
+  tab_header(title = md("**Tissue conentration of proinflammatory cytokines** *(pg/mL)*")) %>%
   opt_align_table_header("center") %>% 
@@
               heading.background.color = "lightgrey",
-              ) %>% 
-  tab_source_note(
-    source_note = " * P-value indicates overall group differences"
               ) %>% 

### 2022-11-13 code/Round 4/FITC round 4.Rmd
Almost done with results

@@
 # Import dataset
-fitc <- read_excel("../../Data/CleanData/fitc.xlsx") %>% 
+fitc <- read_excel("data/processed/fitc.xlsx") %>% 
   rename(values = `result_µg/ml_undiluted`) %>% 
@@
 glimpse(fitc)
+```
+
+Set the tissue type and experimental round here:
+
+```{r load data and set tissue type, include=FALSE}
+#set tissue type
+study_round <- "4"
 ```
@@
 
-normality_check <- fitc %>% 
-  filter(exp_number == 4) %>% 
+fitc %>% 
+  filter(exp_number == study_round) %>% 
   group_by(group) %>% 
-  shapiro_test(values)
-
-normality_check %>% 
+  shapiro_test(values) %>% 
   filter(p < 0.05)
@@
   filter(exp_number == 4) %>% 
-  mutate(Log10_values = log10(values)) %>% 
-  select(group, values, Log10_values) %>% 
+  #mutate(Log10_values = log10(values)) %>% 
+  select(group, values) %>% 
   tbl_summary(by = group,
@@
               #type = list(values ~ 'continuous2', Log10_values ~ 'continuous2'),
-              statistic = list(values ~ "{median} ({p25}, {p75})", Log10_values ~"{median} ({p25}, {p75})"),
-              label = list(values ~ "µg/mL", Log10_values ~ "Log transformed")) %>%
-  add_p(list(values ~ "kruskal.test", Log10_values ~ "kruskal.test")) %>% 
+              statistic = list(values ~ "{median} ({p25}, {p75})"),
+              label = values ~ "µg/mL") %>%
+
+  add_p() %>% 
   # modify_table_styling(
@@
     label = "FITC Conc.",
-    stat_1 = "**VANCO-BF (CON)**\n(n = {n})",
-    stat_2 = "**VANCO-FF (FVT)**\n(n = {n})",
-    stat_3 = "**VANCO-FF (SM)**\n(n = {n})") %>% 
+    stat_1 = "**VANCO-BF** \n
+    (n = {n})",
+    stat_2 = "**VANCO-FVF** \n
+    (n = {n})",
+    stat_3 = "**VANCO-SM** \n
+    (n = {n})") %>% 
   modify_spanning_header((c("stat_1", "stat_2", "stat_3") ~ "Treatment groups (N total = {N})")) %>%  
@@
   tab_footnote( 
-    footnote = md("Log transformation did not normalize distribution.  \n  Non-parametric test (Kruskall-Wallis) used for both datasets."),
+    footnote = md("Median (IQR), Kruskal-Wallis."),
     locations = cells_column_labels(columns = c(p.value))
@@
 
-This table can be exported with the following code either to word, excel, pdf, html or image files. See the syntax of huxtable and flextable (?flextable).
-
-```{r ileum cytokine export, echo=TRUE, warning=FALSE}
-tbl1_p <- tbl1 %>%
-  filter_p() %>% 
-  as_tibble()
-```
-
 ### Plots and regression models
@@
       "CON-BF"= "VANCO-BF",
-      "FVT-FORM" = "VANCO-FF (FVT)",
-      "SM-FORM" = "VANCO-FF (SM)"
+      "FVT-FORM" = "VANCO-FVF",
+      "SM-FORM" = "VANCO-SM"
       ))
@@
 ```{r comparison list, warning=FALSE}
-my_comparisons <- list(c("VANCO-BF", "VANCO-FF (FVT)"), c("VANCO-BF", "VANCO-FF (SM)"), c("VANCO-FF (FVT)", "VANCO-FF (SM)"))
+my_comparisons <- list(c("VANCO-FVF", "VANCO-SM"), c("VANCO-BF", "VANCO-FVF"))
 ```
@@
   geom_point() +
-  #coord_cartesian(ylim = c(1.5, 16)) + 
-  stat_compare_means(comparisons = my_comparisons, label.y = c(60, 50, 45), method = "wilcox.test") +
-  stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
-    label.y.npc = 0.75,
-    label.x.npc = 0.55) +
+  coord_cartesian(ylim = c(10,35)) + 
+  stat_compare_means(comparisons = my_comparisons, label.y = c(20, 25), method = "wilcox.test") +
+  #stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
+   # label.y.npc = 0.75,
+    #label.x.npc = 0.55) +
   labs( # Change title, and axis names
@@
        y = "Serum FITC conc. (µg/ml)",
-       fill = "Groups",
-       caption = "Matrinem round 4") +
+       fill = "Groups"
+       #caption = "Matrinem round 4"
+       ) +
   theme_bw()+
@@
   #tag_levels = 'A',
-  caption = 'VANCO = Vancomycin, BF = Breast-Feeding, FF = Formula-Feeding\n
-  FVT = Fecal Virome Transfer, SM = SM-buffer',
+  caption = 'FVF = Fecal Virome Filtrate, SM = SM-buffer  \n
+  Pairwise comparisions using wilcoxon', 
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),

### 2022-11-14 code/Round 4/Cytokines round 4.Rmd
progress on imac

@@
 plot_all_cytokines <- cytokines_long %>%
+  #mutate(values = values/tissue_weight_mg) %>%    #to run analysis on tissue-weight adjusted concentrations
   #filter(!variables %in% "INFg") %>%  #Exclude any cytokines not wanted in the plot

### 2022-11-14 code/Round 4/Gene expression.Rmd
progress on imac

@@
 
-barcharts
 ```{r}
-delta_gene_expr <- read_excel("../Data/CleanData/gene_expression_fold_changes.xlsx")
+delta_gene_expr <- read_excel("data/processed/gene_expression_fold_changes.xlsx")
 
@@
       "con-bf"= "CON-BF",
-      "fvt-form" = "VANCO-FF (FVT)",
+      "fvt-form" = "VANCO-FVF",
       "sm-form" = "VANCO-FF (SM)"
@@
       groups,
-      "con-bf"= "CON-BF",
-      "fvt-form" = "VANCO-FF (FVT)",
-      "sm-form" = "VANCO-FF (SM)"
+      "con-bf"= "VANCO-BF",
+      "fvt-form" = "VANCO-FVF",
+      "sm-form" = "VANCO-SM"
       )) %>% 
@@
 bar_charts_selected_genes
+
 ```

### 2022-11-16 code/Round 3/Cytokines adjusted.Rmd
progress

--- 2022-11-13 code/Round 1/Cytokines adjusted.Rmd
@@
 ---
-title: "Cytokine analysis script - round 1 adjusted"
+title: "Cytokine analysis script - round 3 adjusted"
 author: "Kaare"
@@
 
-```{r load packages, include=FALSE}
-library(readxl) 
-library(tidyverse)
-library(gt)
-library(gtsummary)
-library(broom)
-library(rstatix)
-library(knitr)
-library(patchwork)
-library(ggpubr)
-```
-
-# Introduction to script
-
-This script is useful for analyzing cytokine expression from MSD Mesoscale assays.
-It is currently optimized for the 'Pro-inflammatory Vplex 10 Mouse kit', analysing the cytokines INFg, IL_10, IL_12p70, IL_1b, IL_2, IL_4,IL_5, IL_6, KCGRO, and TNFa.
-
-Required data structure is a wide tibble (column headers = variables, samples and values = rows).
-
-This script is per default set to handle ileum and serum expression, although any tissue will work. This will just require resetting of the "tissue"-variable.
-
-# Cytokines for experiment 2.
-
-### Basic data structure and results table.
-
-First you load the data and check the structure. Make sure cytokine concentrations are all numeric (<dbl>). If not, mutate non-numeric columns with mutate_if(is_character).
-
 ```{r load data, echo=FALSE}
@@
 sample_tissue <- "ileum"
-study_round <- "1"
+study_round <- "3"
 ```
@@
   filter(tissue == sample_tissue) %>%
-  group_by(group) %>% 
+  filter(group != "VANCO-BF") %>% 
+  group_by(type_of_feed) %>% 
   shapiro_test(tissue_weight_mg)
@@
   labs(
-    title = "Tissue weight of intestinal samples, experiment 2"
+    title = "Tissue weight of intestinal samples, experiment 3"
   )
@@
   ) %>% 
-  mutate(
-    group = recode_factor(  #Change the order of the groups to desired plot order
-      group,
-      "CON-BF"= "CON-BF",
-      "CON-FF" = "CON-FF",
-      "AB-BF"= "AB-BF",
-      "AB-FF"= "AB-FF"
-      )) %>% 
-  mutate(values = if_else(tissue == "ileum", values/tissue_weight_mg, values)) %>% #Adjust for tissue weight
+  mutate(values = values/tissue_weight_mg) %>% #Adjust for tissue weight
   pivot_wider(
@@
 #____Pivot longer____
-cytokines_long <- cytokines_adjusted %>% 
+cytokines_long_adjusted <- cytokines_adjusted %>% 
   pivot_longer(.,
@@
   filter(tissue == sample_tissue) %>%
-  #filter(variables != "INFg") %>% 
+  filter(group != "VANCO-BF") %>% 
   drop_na(values) %>% 
@@
   geom_histogram() +
-  facet_wrap(~ variables, ncol = 2,
+  facet_wrap(~ variables, ncol = 5,
              scales = "free")
@@
 
-tbl1 <- cytokines %>% 
-  filter(exp_number == "1") %>% 
+tbl1 <- cytokines_adjusted %>% 
+  filter(exp_number == "3") %>% 
   filter(tissue == sample_tissue) %>% 
+  filter(group != "VANCO-BF") %>% 
   select(group,INFg:TNFa) %>% 
@@
               missing = "no",
-              include = !"INFg",
               #type = all_continuous() ~ "continuous2",
@@
   modify_header(
-    label ~ "**Groups:**")
+    label ~ "Cytokines")
   #modify_spanning_header((c("stat_1", "stat_2", "stat_3", "stat_4") ~ "Experiment 1 (N total = {N})")) %>%
@@
      tab_source_note(
-    source_note = "Matrinem Experiment 1"
+    source_note = "Matrinem Experiment 3"
   ) %>%
@@
              stat_3 ~ px(200),
-             stat_4 ~ px(200)) 
+             stat_4 ~ px(200),
+             stat_5 ~ px(200)) 
 tbl1_gt
@@
 
-plot_all_cytokines <- cytokines_long %>%
+plot_all_cytokines <- cytokines_long_adjusted %>%
   filter(exp_number == "1") %>% 

### 2022-11-16 code/Round 3/Cytokines round 3.Rmd
progress

@@
 ```
+
+```{r}
+cytokines_unadjusted <- cytokines %>% 
+  pivot_longer(.,
+               cols = INFg:TNFa ,
+               names_to = "variables",
+               values_to = "values"
+  ) %>% 
+  mutate(
+    group = recode_factor(  #Change the order of the groups to desired plot order
+      group,
+      "CON-BF"= "CON-BF",
+      "CON-FF" = "CON-FF",
+      "AB-BF"= "AB-BF",
+      "AB-FF"= "AB-FF"
+      )) %>% 
+  #mutate(adjusted_values = if_else(tissue == "ileum", values/tissue_weight_mg, values)) %>% #Adjust for tissue weight
+  pivot_wider(
+    id_cols = c(
+      sample_id,exp_number, group, received_antibiotics,maternal_treatment, tissue,tissue_weight_mg),
+    names_from = variables, 
+    values_from = values)
+
+#____Pivot longer____
+cytokines_long_unadjusted <- cytokines %>% 
+  pivot_longer(.,
+               cols = INFg:TNFa ,
+               names_to = "variables",
+               values_to = "values"
+  )
+
+```
+
 <!-- Looks fine. Go ahead and  adjust-->
+
+```{r}
+cytokines_adjusted <- cytokines %>% 
+  pivot_longer(.,
+               cols = INFg:TNFa ,
+               names_to = "variables",
+               values_to = "values"
+  ) %>% 
+  mutate(
+    group = recode_factor(  #Change the order of the groups to desired plot order
+      group,
+      "CON-BF"= "CON-BF",
+      "CON-FF" = "CON-FF",
+      "AB-BF"= "AB-BF",
+      "AB-FF"= "AB-FF"
+      )) %>% 
+  mutate(adjusted_values = if_else(tissue == "ileum", values/tissue_weight_mg, values)) %>% #Adjust for tissue weight
+  pivot_wider(
+    id_cols = c(
+      sample_id,exp_number, group, received_antibiotics,maternal_treatment, tissue,tissue_weight_mg),
+    names_from = variables, 
+    values_from = adjusted_values)
+
+#____Pivot longer____
+cytokines_long_adjusted <- cytokines %>% 
+  pivot_longer(.,
+               cols = INFg:TNFa ,
+               names_to = "variables",
+               values_to = "adjusted_values"
+  )
+
+```
 
@@
   modify_header(
-    label ~ "**Groups**:") %>% 
-  modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Matrinem round 3 (N total = {N})"))
+    label ~ "**Groups**:")
+  #modify_spanning_header((c("stat_1", "stat_2", "stat_3","stat_4", "stat_5", "stat_6") ~ "Matrinem round 3 (N total = {N})"))
 
@@
  as_gt() %>%
-  tab_header(title = md("**Proinflammatory cytokines**")) %>%
+  tab_header(title = md("**Tissue concentrations of pro-inflammatory cytokines** *(pg/ml)*")) %>%
   opt_align_table_header("center") %>% 
@@
     source_note = " * P-value indicates overall group differences"
+  ) %>%
+   tab_source_note(
+    source_note = "Matrinem round 3"
   ) %>%

### 2022-11-16 code/Round 3/Cytokines unadjusted.Rmd
progress

--- 2022-10-18 code/Round 1/Cytokines unadjusted.Rmd
@@
 ---
-title: "Cytokine analysis script - round 1"
+title: "Cytokine analysis script - round 3"
 author: "Kaare"
@@
 tbl1 <- cytokines %>% 
-  filter(exp_number == study_round) %>% 
-  filter(tissue == sample_tissue) %>% 
+  filter(exp_number == 3) %>% 
+  filter(tissue == "ileum") %>% 
   select(group,INFg:TNFa) %>% 
@@
               digits = everything() ~ c(2,1,1),
-              missing = "no",
-              include = !"INFg") %>%
+              missing = "no") %>%
   add_p() %>% 
   modify_header(
-    label ~ "Cytokines") %>% 
-  modify_spanning_header((c("stat_1", "stat_2", "stat_3", "stat_4") ~ "Treatment groups (N total = {N})"))
+    label ~ "Cytokines")
+  #modify_spanning_header((c("stat_1", "stat_2", "stat_3", "stat_4") ~ "Treatment groups (N total = {N})"))
 
@@
   as_gt() %>%
-  tab_header(title = md("**Proinflammatory cytokines**")) %>%
+  tab_header(title = md("**Tissue concentrations of pro-inflammatory cytokines** *")) %>%
   opt_align_table_header("center") %>% 
@@
   tab_source_note(
-    source_note = md("**Matrinem round 1**")
+    source_note = md("Matrinem round 3, unadjusted")
   ) %>%
@@
              stat_3 ~ px(150),
-             stat_4 ~ px(150)) 
+             stat_4 ~ px(150),
+             stat_5 ~ px(150),
+             stat_6 ~ px(150)) 
 tbl1_gt

### 2022-11-16 code/Round 3/FITC round 3.Rmd
progress

@@
 ```{r plots all fitc, warning=FALSE}
-my_comparisons <- list( c("CON-BF", "CON-FF"), c("GENTA-BF", "GENTA-FF"), c("VANCO-BF", "VANCO-FF"), c("GENTA-FF", "VANCO-FF"))
+my_comparisons <- list( c("CON-BF", "CON-FF"), c("GENTA-BF", "GENTA-FF"), c("VANCO-FF", "GENTA-FF"), c("CON-FF", "VANCO-FF"))
 
@@
   #coord_cartesian(ylim = c(1.5, 16)) + 
-  stat_compare_means(comparisons = my_comparisons, label.y = c(30, 30, 30, 50), method = "wilcox.test")+
+  stat_compare_means(comparisons = my_comparisons, label.y = c(20, 20, 30, 50), method = "wilcox.test")+
   # stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
@@
   #coord_cartesian(ylim = c(1.5, 16)) + 
-  stat_compare_means(comparisons = my_comparisons, label.y = c(1.5, 1.5, 1.5, 1.7), method = "wilcox.test")+
+  stat_compare_means(comparisons = my_comparisons, label.y = c(1.5, 1.5, 1.7, 1.8), method = "wilcox.test")+
   # stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),

### 2022-11-16 code/weight_analysis.Rmd
progress

@@
 
+birthweight %>% 
+  filter(time %in% "day_0") %>% 
+  knitr::kable()
+
 lm_birthweight <- data %>%  #LM for effect of gestational age on offspring birthweight. 
@@
   mutate(gestation_age = if_else(maternal_treatment == "CON", 21,20)) %>% 
-  do(lm(bodyweight ~ gestation_age,.) %>% tidy)
+  do(lm(bodyweight ~ gestation_age + maternal_treatment,.) %>% tidy)
 lm_birthweight

### 2022-11-16 code/weight_development.R

@@ -27,7 +27,7 @@ birthweight <- data %>%
   filter(exp_number %in% "2") %>% 
   filter(time %in% c("day_0", "day_1", "day_2")) %>%
   drop_na(bodyweight) %>% 
-  group_by(exp_number, maternal_treatment, time) %>% 
+  group_by(exp_number, received_antibiotics, time) %>% 
   summarise(birthweight = mean(bodyweight),
             sd = sd(bodyweight))
 

### 2022-11-17 code/Round 3/FITC round 3.Rmd
created cytokine script and deleted lots of old files

@@
 
-plot_all_fitc <- fitc_long %>%    #____Normal values_____#
-  filter(exp_number == study_round) %>% 
+Exp3_fitc <- fitc_long %>%        #Log_transformed values_____#
+  filter(exp_number == 3) %>% 
+  filter(group2 != "VANCO-BF") %>% 
   filter(values > 8) %>% 
 ggplot(aes(x = group2, 
-           y = values, 
+           y = log10(values), 
            fill = group2)) + 
-  geom_boxplot() +
+  geom_boxplot(outlier.shape = NA) +
   geom_point() +
-  #coord_cartesian(ylim = c(1.5, 16)) + 
-  stat_compare_means(comparisons = my_comparisons, label.y = c(20, 20, 30, 50), method = "wilcox.test")+
-  # stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
-  #   label.y.npc = 0.75,
-  #   label.x.npc = 0.55) +
+  #stat_compare_means(comparisons = my_comparisons, label.y = c(1.5, 1.5, 1.7, 1.8), method = "wilcox.test")+
+  stat_compare_means(aes(label = ..p.signif..),
+                   method = "wilcox.test", ref.group = "CON-BF",
+                   label.y.npc = 0.55) +
   labs( # Change title, and axis names
        x = "",
-       y = "Serum FITC conc. (µg/ml)",
+       y = "Log10 serum FITC conc. (µg/ml)",
        fill = "Groups",
@@
   theme_bw()+
-  theme(                   # Change the appearance of labels
+  theme(                   
     plot.title = element_text(color="darkgrey", size=12, face="bold"),
@@
     legend.position="bottom",
-    legend.background = element_rect(fill="lightgrey"),#Change legend box background color
+    legend.background = element_rect(fill="lightgrey"),
     plot.caption = element_text(color = "darkgrey", size=5)) + 
-  scale_fill_brewer(palette="Set1")
-
-plot_all_fitc_log <- fitc_long %>%        #Log_transformed values_____#
-  filter(exp_number == study_round) %>% 
-  filter(values > 8) %>% 
-ggplot(aes(x = group2, 
-           y = log10(values), 
-           fill = group2)) + 
-  geom_boxplot() +
-  geom_point() +
-  #coord_cartesian(ylim = c(1.5, 16)) + 
-  stat_compare_means(comparisons = my_comparisons, label.y = c(1.5, 1.5, 1.7, 1.8), method = "wilcox.test")+
-  # stat_compare_means(aes(label = paste0("Kruskal-Wallis \n p = ", ..p.format..)),
-  #   label.y.npc = 0.75,
-  #   label.x.npc = 0.55) +
-  labs( # Change title, and axis names
-       x = "",
-       y = "Log10 serum FITC conc. (µg/ml)",
-       fill = "Groups",
-       caption = "Matrinem round 3") +
-  theme_bw()+
-  theme(                   # Change the appearance of labels
-    plot.title = element_text(color="darkgrey", size=12, face="bold"),
-    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
-    #axis.text.y = element_blank(),
-    axis.ticks.x = element_blank(),  # Hide labels
-    axis.text.x = element_blank(),
-    legend.title = element_text(face="bold"),   
-    legend.position="bottom",
-    legend.background = element_rect(fill="lightgrey"),#Change legend box background color
-    plot.caption = element_text(color = "darkgrey", size=5)) + 
-  scale_fill_brewer(palette="Set1")
-
-all_fitc_patch <- plot_all_fitc + plot_all_fitc_log + plot_annotation (
+  scale_fill_brewer(palette="Set1") + plot_annotation (
   title = 'Intestinal permeability',
@@
   tag_levels = 'A',
-  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
-  Pariwise comparisons = Wilcoxon test.',
+  caption = 'CON = Water, BF = Breast-Feeding, FF = Formula-Feeding, GENTA = Gentamicin,  VANCO = Vancomycin,  \n
+  Pairwise comparison with CON-BF as reference. \n
+    **: p < 0.01, *: p < 0.05, ns: not significant."',
   theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
@@
 
-all_fitc_patch
-#ggsave('../../Results/Round 1/All fitc-boxplot.png',all_fitc_patch)
+Exp3_fitc %>% 
+ggsave(path = "exports/fitc/", filename = "Exp3-fitc.png")
 ```

### 2022-11-17 code/Round 4/Gene expression.Rmd
created cytokine script and deleted lots of old files

@@
 
-#### Gene expression
+# Gene expression
+## Plot
 
 ```{r}
-delta_gene_expr <- read_excel("data/processed/gene_expression_fold_changes.xlsx")
-
-bar_charts_all_genes <- delta_gene_expr %>% 
+delta_gene_expr <- read_excel("data/processed/gene_expression_fold_changes.xlsx") %>% 
    mutate(
@@
       groups,
-      "con-bf"= "CON-BF",
+      "con-bf"= "VANCO-BF",
       "fvt-form" = "VANCO-FVF",
-      "sm-form" = "VANCO-FF (SM)"
-      )) %>% 
+      "sm-form" = "VANCO-SM"
+      ))
+increased <- delta_gene_expr %>% 
+ group_by(genes, groups) %>% 
+  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
+  filter(mean_fold_change > 2)
+
+decreased <- delta_gene_expr %>% 
+ group_by(genes, groups) %>% 
+  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
+  filter(mean_fold_change < 0.6)
+
+bar_charts_no_change <- delta_gene_expr %>% 
+  filter(!genes %in% c(increased$genes, decreased$genes)) %>% 
   group_by(genes, groups) %>% 
@@
   facet_wrap(~genes,
-             ncol = 5, 
+             nrow = 1, 
              scales = "free") +
-  labs(title = "Level of gene expression in ileum tissue",
+  labs(
+       y = "Gene expression relative fold change",
+       x = "",
+       fill = "Groups")
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="lightgrey"), # Change legend box background color
+    plot.caption = element_text(color = "darkgrey", size=5)) + 
+  scale_fill_brewer(palette="Set1")
+bar_charts_no_change
+
+bar_charts_increased <- delta_gene_expr %>% 
+  filter(!genes %in% c(increased$genes, decreased$genes)) %>% 
+  group_by(genes, groups) %>% 
+  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
+  ggplot(
+    aes(x = groups,
+        y = mean_fold_change,
+        fill = groups)
+  ) + 
+  geom_col() +
+  facet_wrap(~genes,
+             nrow = 1, 
+             scales = "free") +
+  labs(
+    #title = "Rate of gene expression in ileum tissue",
        y = "Gene expression relative fold change",
@@
   scale_fill_brewer(palette="Set1")
-
-bar_charts_all_genes
+bar_charts_increased
 
@@
   facet_wrap(~genes,
-             ncol = 5, 
               scales = "free") +
@@
 ```
+
+```{r}
+delta_gene_expr %>% 
+  mutate(
+    groups = recode_factor(  #Change the order of the groups to desired plot order
+      groups,
+      "con-bf"= "CON-BF",
+      "fvt-form" = "VANCO-FVF",
+      "sm-form" = "VANCO-SM"
+      )) %>% 
+  group_by(genes, groups) %>% 
+  ggplot(
+    aes(x = groups,
+        y = `2^-(ddct)`,
+        fill = groups)
+  ) + 
+  geom_boxplot() +
+  facet_wrap(~genes,
+              scales = "free")
+
+```

### 2022-11-17 code/cytokines/Cytokines_serum.Rmd
created cytokine script and deleted lots of old files

---
title: "Cytokine analysis - Serum"
author: "Kaare"
date: "11th October 2022"    
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

```{r load packages, include=FALSE}
library(readxl) 
library(tidyverse)
library(gt)
library(gtsummary)
library(broom)
library(rstatix)
library(knitr)
library(patchwork)
library(ggpubr)
```
#README: 
Serum cytokines only

# Data import
```{r adjust_data, echo=TRUE}
#_______Import dataset and Transform to adjusted values_______#

cytokines <- read_excel("data/processed/cytokines.xlsx") %>% 
  filter(tissue != "ileum") %>% 
  pivot_longer(.,
               cols = INFg:TNFa ,
               names_to = "variables",
               values_to = "values"
  ) %>% 
  drop_na(values) %>% 
   pivot_wider(
    id_cols = c(
      sample_id,exp_number, group, received_antibiotics,maternal_treatment, type_of_feed, tissue,),
    names_from = variables, 
    values_from = values)

#____ Make long format _______#
cytokines_long <- cytokines %>%
  pivot_longer(.,
               cols = INFg:TNFa ,
               names_to = "variables",
               values_to = "values"
  )
```

# Exp1
## Tissue weight 
```{r exp1_inspect data, echo=FALSE}

#_____Weight of tissue samples_____#
Exp1_tissue_weight <- cytokines_long %>%       
   filter(exp_number == 1) %>%
  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
  geom_boxplot() +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.55) +
  stat_compare_means(method = "kruskal.test",
    aes(label = paste0("Kruskal Wallis Test. \n p = ", ..p.format..)),
    label.y.npc = 0.75,
    label.x.npc = 0.3) +
  geom_point() +
  labs(
    title = "Tissue weight of intestinal samples, experiment 1",
    y = "Tissue weight (mg)",
    x = "",
    caption = "Pairwise comparison with CON-BF as reference. \n
    **: p < 0.01, *: p < 0.05, NS: not significant."
  ) 
Exp1_tissue_weight
ggsave('exports/cytokines/Exp1-weights-of-tissue.png',Exp1_tissue_weight)
```

## Check for data distribution. 
```{r Exp1 visualize data, eval=FALSE, include=FALSE}
cytokines_long %>%                       #normality_check  of cytokine distributions
  filter(exp_number == 1) %>% 
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values)

#log10 transformed shapiro_test - failed variables
cytokines_long %>% 
 filter(exp_number == 1) %>% 
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(values = log10(values)) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values) %>% 
  filter(p < 0.05)

#  group    variables variable statistic       p
#  <fct>    <chr>     <chr>        <dbl>   <dbl>
# VANCO-BF IL-1b     values       0.727 0.00728     # transformation helped.

#Density plots to show distribution of cytokines
histogram <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = values, fill = group)) + 
  geom_histogram() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

log10_histogram <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = log10(values), fill = group)) + 
  geom_density() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

#Jitter plots. Works
jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free")

log10_jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(adjusted_values = log10(values)) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free") 
```

##Table

```{r Exp1-table}
#_____Currently non-normally distributed values are plotted as median (IQR)____# 
 #change statistics with help from here: https://www.danieldsjoberg.com/gtsummary/reference/tests.html

Exp1_tbl1 <- cytokines %>% 
  filter(exp_number == "1") %>%
  select(group, INFg:TNFa) %>% 
  tbl_summary(by = group,
              digits = everything() ~ c(2,2,2),
              missing = "no",
              #include = !"INFg",
              #type = all_continuous() ~ "continuous2",
              statistic = all_continuous() ~ #c("{mean} ({sd})",         
                                     "{median} ({p25}, {p75})") %>%
  modify_header(
    label ~ "**Groups:**") %>% 
   #modify_footnote(stat_1 ~ "Median (IQR)", stat_2 ~ "p-value on overall group differences,Kruskall-Wallis rank sum test ") %>% 
add_p() %>% 
  as_gt() %>%
  tab_header(title = md("**Serum concentraions of proinflammatory cytokines** *(pg/ml)*")) %>%
  opt_align_table_header("center") %>% 
   tab_options(heading.title.font.size = 15,
               heading.background.color = "lightgrey",
               ) %>% 
     tab_source_note(
    source_note = "Matrinem Experiment 1"
  )

Exp1_tbl1 %>%
  gtsave(path = "exports/cytokines/", filename = "Exp1-serum-adjusted.png")
```

## Plot
To see all the cytokine leves plotted run this code: 
```{r plots all cytokines, warning=FALSE}

Exp1_plot_all_cytokines <- cytokines_long %>%
  filter(exp_number == "1") %>% 
  #filter(!variables %in% c("INFg", "IL-12p70")) %>%
ggplot(aes(x = group, 
           y = values, 
           fill = group)) + 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.55) +
  facet_wrap(~variables, 
             scales = 'free', 
             nrow = 2,
             ncol = 5) +
  labs(title = "Concentraion of pro-inflammatory cytokines in serum",
       x = "",
       y = "Tissue conc. (pg/ml)",
       fill = "Groups") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    #axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1") + plot_annotation (
  #title = 'Cytokine expression',
   #subtitle = 'All cytokines measured',
  #tag_levels = 'A',
  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

Exp1_plot_all_cytokines
ggsave('exports/cytokines/Exp1-serum-adjusted-plot.png',Exp1_plot_all_cytokines)
```

### 2022-11-17 code/cytokines/Cytokines_tissue_adjusted.Rmd
created cytokine script and deleted lots of old files

---
title: "Cytokine analysis script - all rounds - adjusted for tissue weight"
author: "Kaare"
date: "11th October 2022"    
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
```

```{r load packages, include=FALSE}
library(readxl) 
library(tidyverse)
library(gt)
library(gtsummary)
library(broom)
library(rstatix)
library(knitr)
library(patchwork)
library(ggpubr)
```
#README: 
Cytokine expression profiles in Matrinem.
Below are 4 sections, each making a table and a plot for individual experimental rounds.

# Data import
```{r adjust_data, echo=TRUE}
#_______Import dataset and Transform to adjusted values_______#

cytokines <- read_excel("data/processed/cytokines.xlsx") %>% 
  filter(tissue == "ileum") %>% 
  pivot_longer(.,
               cols = INFg:TNFa ,
               names_to = "variables",
               values_to = "values"
  ) %>% 
   #mutate(values = values/(tissue_weight_mg/1000)) %>% #Adjust for tissue weight / 1000 (grams)
   mutate(values = values/tissue_weight_mg) %>% #Adjust for tissue weight (mg)
  pivot_wider(
    id_cols = c(
      sample_id,exp_number, group, received_antibiotics,maternal_treatment, type_of_feed, tissue,tissue_weight_mg),
    names_from = variables, 
    values_from = values)

#____ Make long format _______#
cytokines_long <- cytokines %>%
  pivot_longer(.,
               cols = INFg:TNFa ,
               names_to = "variables",
               values_to = "values"
  )
```

# Exp1
## Tissue weight 
```{r exp1_inspect data, echo=FALSE}

#___Inspect tissue weight distribution____#
cytokines_long %>%       
  filter(exp_number == 1) %>% 
  kruskal_test(tissue_weight_mg ~ group) %>% 
  knitr::kable(escape = TRUE)

cytokines_long %>%       
  filter(exp_number == 1) %>% 
  anova_test(tissue_weight_mg ~ group) %>% 
  knitr::kable(escape = TRUE)

#_____Weight of tissue samples_____#
Exp1_tissue_weight <- cytokines_long %>%       
   filter(exp_number == 1) %>%
  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
  geom_boxplot() +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.55) +
  stat_compare_means(method = "kruskal.test",
    aes(label = paste0("Kruskal Wallis Test. \n p = ", ..p.format..)),
    label.y.npc = 0.75,
    label.x.npc = 0.3) +
  geom_point() +
  labs(
    title = "Tissue weight of intestinal samples, experiment 1",
    y = "Tissue weight (mg)",
    x = "",
    caption = "Pairwise comparison with CON-BF as reference. \n
    **: p < 0.01, *: p < 0.05, NS: not significant."
  ) 
Exp1_tissue_weight
ggsave('exports/cytokines/Exp1-weights-of-tissue.png',Exp1_tissue_weight)
```

## Check for data distribution. 
```{r Exp1 visualize data, eval=FALSE, include=FALSE}
cytokines_long %>%                       #normality_check  of cytokine distributions
  filter(exp_number == 1) %>% 
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values)

#log10 transformed shapiro_test - failed variables
cytokines_long %>% 
 filter(exp_number == 1) %>% 
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(values = log10(values)) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values) %>% 
  filter(p < 0.05)

#  group    variables variable statistic       p
#  <fct>    <chr>     <chr>        <dbl>   <dbl>
# VANCO-BF IL-1b     values       0.727 0.00728     # transformation helped.

#Density plots to show distribution of cytokines
histogram <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = values, fill = group)) + 
  geom_histogram() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

log10_histogram <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = log10(values), fill = group)) + 
  geom_density() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

#Jitter plots. Works
jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free")

log10_jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(adjusted_values = log10(values)) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free") 
```

##Table

```{r Exp1-table}
#_____Currently non-normally distributed values are plotted as median (IQR)____# 
 #change statistics with help from here: https://www.danieldsjoberg.com/gtsummary/reference/tests.html

Exp1_tbl1 <- cytokines %>% 
  filter(exp_number == "1") %>%
  select(group, INFg:TNFa) %>% 
  tbl_summary(by = group,
              digits = everything() ~ c(2,2,2),
              missing = "no",
              include = !"INFg",
              #type = all_continuous() ~ "continuous2",
              statistic = all_continuous() ~ #c("{mean} ({sd})",         
                                     "{median} ({p25}, {p75})") %>%
  modify_header(
    label ~ "**Groups:**") %>% 
   #modify_footnote(stat_1 ~ "Median (IQR)", stat_2 ~ "p-value on overall group differences,Kruskall-Wallis rank sum test ") %>% 
add_p() %>% 
  as_gt() %>%
  tab_header(title = md("**Tissue concentraions of proinflammatory cytokines** *(pg/ml/mg tissue)*")) %>%
  opt_align_table_header("center") %>% 
   tab_options(heading.title.font.size = 15,
               heading.background.color = "lightgrey",
               ) %>% 
     tab_source_note(
    source_note = "Matrinem Experiment 1"
  ) %>%
  cols_width(label ~ px(80),
             stat_1 ~ px(140),
             stat_2 ~ px(140),
             stat_3 ~ px(140),
             stat_4 ~ px(140)) 
Exp1_tbl1 %>%
  gtsave(path = "exports/cytokines/", filename = "Exp1-tissue-adjusted.png", vwidth = 760, vheight = 450)

```

## Plot
To see all the cytokine leves plotted run this code: 
```{r plots all cytokines, warning=FALSE}

Exp1_plot_all_cytokines <- cytokines_long %>%
  filter(exp_number == "1") %>% 
  #filter(!variables %in% c("INFg", "IL-12p70")) %>%
ggplot(aes(x = group, 
           y = values, 
           fill = group)) + 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.55) +
  facet_wrap(~variables, 
             scales = 'free', 
             nrow = 2,
             ncol = 5) +
  labs(title = "Concentraion of pro-inflammatory cytokines in tissue",
       x = "",
       y = "Tissue conc. (pg/ml/mg tissue)",
       fill = "Groups") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    #axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1") + plot_annotation (
  #title = 'Cytokine expression',
   #subtitle = 'All cytokines measured',
  #tag_levels = 'A',
  caption = 'CON = Water, AB = Antibiotics, BF = Breast-Feeding, FF = Formula-Feeding',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

Exp1_plot_all_cytokines
ggsave('exports/cytokines/Exp1-tissue-adjusted-plot.png',Exp1_plot_all_cytokines)
```

# Exp2 exp_number : Matrinem Experiment 1 
## Tissue weight 
```{r Exp2_inspect data, echo=FALSE}

#___Inspect tissue weight distribution____#
cytokines_long %>%       
  filter(exp_number == 2) %>% 
  kruskal_test(tissue_weight_mg ~ maternal_treatment) %>% 
  knitr::kable(escape = TRUE)

#_____Weight of tissue samples_____#
Exp2_tissue_weight <- cytokines_long %>%       
   filter(exp_number == 2) %>%
  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
  geom_boxplot() +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.55) +
  stat_compare_means(method = "kruskal.test",
    aes(label = paste0("Kruskal Wallis Test. \n p = ", ..p.format..)),
    label.y.npc = 0.75,
    label.x.npc = 0.3) +
  geom_point() +
  labs(
    title = "Tissue weight of intestinal samples, experiment 2",
    y = "Tissue weight (mg)",
    x = "",
    caption = "Pairwise comparison with CON-BF as reference. \n
    **: p < 0.01, *: p < 0.05, NS: not significant."
  ) 
ggsave('exports/cytokines/Exp2-weights-of-tissue.png',Exp2_tissue_weight)

```

## Check for data distribution. 
```{r Exp2 visualize data, eval=FALSE, include=FALSE}
cytokines_long %>%                       #normality_check  of cytokine distributions
  filter(exp_number == 2) %>% 
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values)

#log10 transformed shapiro_test - failed variables
cytokines_long %>% 
 filter(exp_number == 2) %>% 
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(values = log10(values)) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values) %>% 
  filter(p < 0.05)

#  group    variables variable statistic       p
#  <fct>    <chr>     <chr>        <dbl>   <dbl>
# VANCO-BF IL-1b     values       0.727 0.00728     # transformation helped.

#Density plots to show distribution of cytokines
histogram <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = values, fill = group)) + 
  geom_histogram() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

log10_histogram <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = log10(values), fill = group)) + 
  geom_density() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

#Jitter plots. Works
jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free")

log10_jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(adjusted_values = log10(values)) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free") 
```

##Table
```{r Exp2-table}
#_____Currently non-normally distributed values are plotted as median (IQR)____# 
 #change statistics with help from here: https://www.danieldsjoberg.com/gtsummary/reference/tests.html

Exp2_tbl1 <- cytokines %>% 
  filter(exp_number == 2) %>%
  select(group, INFg:TNFa) %>% 
  tbl_summary(by = group,
              digits = everything() ~ c(2,2,2),
              missing = "no",
              include = !"INFg",
              #type = all_continuous() ~ "continuous2",
              statistic = all_continuous() ~ #c("{mean} ({sd})",         
                                     "{median} ({p25}, {p75})") %>%
  modify_header(
    label ~ "**Groups:**") %>% 
   #modify_footnote(stat_1 ~ "Median (IQR)", stat_2 ~ "p-value on overall group differences,Kruskall-Wallis rank sum test ") %>% 
add_p() %>% 
  as_gt() %>%
  tab_header(title = md("**Tissue concentraions of proinflammatory cytokines** *(pg/ml/mg tissue)*")) %>%
  opt_align_table_header("center") %>% 
   tab_options(heading.title.font.size = 15,
               heading.background.color = "lightgrey",
               ) %>% 
     tab_source_note(
    source_note = "Matrinem Experiment 2"
  ) 
Exp2_tbl1 %>%
  gtsave(path = "exports/cytokines/", filename = "Exp2-tissue-adjusted-table.png")

```

## Plot
To see all the cytokine leves plotted run this code: 
```{r Exp2 plot all cytokines, warning=FALSE}
Exp2_plot_all_cytokines <- cytokines_long %>%
  filter(exp_number == 2) %>% 
  #filter(!variables %in% c("INFg", "IL-12p70")) %>%
ggplot(aes(x = group, 
           y = values, 
           fill = group)) + 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.75) +
  facet_wrap(~variables, 
             scales = 'free', 
             nrow = 2,
             ncol = 5) +
  labs(title = "Concentraion of pro-inflammatory cytokines in tissue",
       x = "",
       y = "Tissue conc. (pg/ml/mg tissue)",
       fill = "Groups") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    #axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1") + plot_annotation (
  #title = 'Cytokine expression',
   #subtitle = 'All cytokines measured',
  #tag_levels = 'A',
  caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin, VANCO = Vancomycin,',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

Exp2_plot_all_cytokines
ggsave('exports/cytokines/Exp2-tissue-adjusted-plot.png',Exp2_plot_all_cytokines)
```

# Exp3 exp_number : Matrinem Experiment 2 
## Tissue weight 
```{r Exp3_inspect data, echo=FALSE}

#___Inspect tissue weight distribution____#
cytokines_long %>%       
  filter(exp_number == 3) %>% 
  kruskal_test(tissue_weight_mg ~ maternal_treatment) %>% 
  knitr::kable(escape = TRUE)

#_____Weight of tissue samples_____#
Exp3_tissue_weight <- cytokines_long %>%       
   filter(exp_number == 3) %>%
  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
  geom_boxplot() +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.55) +
  stat_compare_means(method = "kruskal.test",
    aes(label = paste0("Kruskal Wallis Test. \n p = ", ..p.format..)),
    label.y.npc = 0.75,
    label.x.npc = 0.3) +
  geom_point() +
  labs(
    title = "Tissue weight of intestinal samples, experiment 3",
    y = "Tissue weight (mg)",
    x = "",
    caption = "Pairwise comparison with CON-BF as reference. \n
    **: p < 0.01, *: p < 0.05, NS: not significant."
  ) 
ggsave('exports/cytokines/Exp3-weights-of-tissue.png',Exp3_tissue_weight)

```

## Check for data distribution. 
```{r Exp3 visualize data, eval=FALSE, include=FALSE}
cytokines_long %>%                       #normality_check  of cytokine distributions
  filter(exp_number == 3) %>% 
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values)

#log10 transformed shapiro_test - failed variables
cytokines_long %>% 
 filter(exp_number == 3) %>% 
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(values = log10(values)) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values) %>% 
  filter(p < 0.05)

#  group    variables variable statistic       p
#  <fct>    <chr>     <chr>        <dbl>   <dbl>
# VANCO-BF IL-1b     values       0.727 0.00728     # transformation helped.

#Density plots to show distribution of cytokines
histogram <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = values, fill = group)) + 
  geom_histogram() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

log10_histogram <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = log10(values), fill = group)) + 
  geom_density() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

#Jitter plots. Works
jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free")

log10_jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(adjusted_values = log10(values)) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free") 
```

##Table
```{r Exp3-table}
#_____Currently non-normally distributed values are plotted as median (IQR)____# 
 #change statistics with help from here: https://www.danieldsjoberg.com/gtsummary/reference/tests.html

Exp3_tbl1 <- cytokines %>% 
  filter(exp_number == 3) %>%
  filter(group != "VANCO-BF") %>% 
  select(group, INFg:TNFa) %>% 
  tbl_summary(by = group,
              digits = everything() ~ c(2,2,2),
              missing = "no",
              include = !"INFg",
              #type = all_continuous() ~ "continuous2",
              statistic = all_continuous() ~ #c("{mean} ({sd})",         
                                     "{median} ({p25}, {p75})") %>%
  modify_header(
    label ~ "**Groups:**") %>% 
   #modify_footnote(stat_1 ~ "Median (IQR)", stat_2 ~ "p-value on overall group differences,Kruskall-Wallis rank sum test ") %>% 
add_p() %>% 
  as_gt() %>%
  tab_header(title = md("**Tissue concentraions of proinflammatory cytokines** *(pg/ml/mg tissue)*")) %>%
  opt_align_table_header("center") %>% 
   tab_options(heading.title.font.size = 15,
               heading.background.color = "lightgrey",
               ) %>% 
     tab_source_note(
    source_note = "Matrinem Experiment 3"
  ) 
Exp3_tbl1 %>%
  gtsave(path = "exports/cytokines/", filename = "Exp3-tissue-adjusted-table.png")
```

## Plot
To see all the cytokine leves plotted run this code: 
```{r Exp3 plot all cytokines, warning=FALSE}
Exp3_plot_all_cytokines <- cytokines_long %>%
  filter(exp_number == 3) %>% 
  filter(group != "VANCO-BF") %>% 
  #filter(!variables %in% c("INFg", "IL-12p70")) %>%
ggplot(aes(x = group, 
           y = values, 
           fill = group)) + 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "CON-BF",
                   label.y.npc = 0.75) +
  facet_wrap(~variables, 
             scales = 'free', 
             nrow = 2,
             ncol = 5) +
  labs(title = "Concentraion of pro-inflammatory cytokines in tissue",
       x = "",
       y = "Tissue conc. (pg/ml/mg tissue)",
       fill = "Groups") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    #axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1") + plot_annotation (
  #title = 'Cytokine expression',
   #subtitle = 'All cytokines measured',
  #tag_levels = 'A',
  caption = 'CON = Water, BF = Breast-Feeding, FF = Formula-Feeding, GENTA = Gentamicin, VANCO = Vancomycin,',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

Exp3_plot_all_cytokines
ggsave('exports/cytokines/Exp3-tissue-adjusted-plot.png',Exp3_plot_all_cytokines)
```

# Exp4 : exp_number : Matrinem Experiment 3 
## Tissue weight 
```{r Exp4_inspect data, echo=FALSE}

#___Inspect tissue weight distribution____#
cytokines_long %>%       
  filter(exp_number == 4) %>% 
  kruskal_test(tissue_weight_mg ~ group) %>% 
  knitr::kable(escape = TRUE)

#_____Weight of tissue samples_____#
Exp4_tissue_weight <- cytokines_long %>%       
   filter(exp_number == 4) %>%
  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
  geom_boxplot() +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "VANCO-BF",
                   label.y.npc = 0.55) +
  stat_compare_means(method = "kruskal.test",
    aes(label = paste0("Kruskal Wallis Test. \n p = ", ..p.format..)),
    label.y.npc = 0.75,
    label.x.npc = 0.3) +
  geom_point() +
  labs(
    title = "Tissue weight of intestinal samples, experiment 4",
    y = "Tissue weight (mg)",
    x = "",
    caption = "Pairwise comparison with VANCO-BF as reference. \n
    ****: p < 0.0005, ***: p < 0.001, NS: not significant."
  ) 
ggsave('exports/cytokines/Exp4-weights-of-tissue.png',Exp4_tissue_weight)

```

## Check for data distribution. 
```{r Exp4 visualize data, eval=FALSE, include=FALSE}
cytokines_long %>%                       #normality_check  of cytokine distributions
  filter(exp_number == 4) %>% 
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values)

#log10 transformed shapiro_test - failed variables
cytokines_long %>% 
 filter(exp_number == 4) %>% 
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(values = log10(values)) %>% 
  group_by(variables, group) %>% 
  shapiro_test(values) %>% 
  filter(p < 0.05)

#  group    variables variable statistic       p
#  <fct>    <chr>     <chr>        <dbl>   <dbl>
# VANCO-BF IL-1b     values       0.727 0.00728     # transformation helped.

#Density plots to show distribution of cytokines
histogram <- cytokines_long %>% 
   filter(exp_number == 4) %>%

  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = values, fill = group)) + 
  geom_histogram() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

log10_histogram <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = log10(values), fill = group)) + 
  geom_density() +
  facet_wrap(~ variables, ncol = 2,
             scales = "free")

#Jitter plots. Works
jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  #filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free")

log10_jitter_plot <- cytokines_long %>% 
   filter(exp_number == study_round) %>% 
  filter(tissue == sample_tissue) %>%
  filter(variables != "INFg") %>% 
  drop_na(values) %>% 
  mutate(adjusted_values = log10(values)) %>% 
  ggplot(aes(x = group, y = values, color = group)) +
  geom_jitter() +
  facet_wrap(~ variables,
             scales = "free") 
```

##Table
```{r Exp4-table}
#_____Currently non-normally distributed values are plotted as median (IQR)____# 
 #change statistics with help from here: https://www.danieldsjoberg.com/gtsummary/reference/tests.html

Exp4_tbl1 <- cytokines %>% 
  filter(exp_number == 4) %>%
  #filter(group != "VANCO-BF") %>% 
  select(group, INFg:TNFa) %>% 
  tbl_summary(by = group,
              digits = everything() ~ c(2,2,2),
              missing = "no",
              include = !"INFg",
              #type = all_continuous() ~ "continuous2",
              statistic = all_continuous() ~ #c("{mean} ({sd})",         
                                     "{median} ({p25}, {p75})") %>%
  modify_header(
    label ~ "**Groups:**") %>% 
   #modify_footnote(stat_1 ~ "Median (IQR)", stat_2 ~ "p-value on overall group differences,Kruskall-Wallis rank sum test ") %>% 
add_p() %>% 
  as_gt() %>%
  tab_header(title = md("**Tissue concentraions of proinflammatory cytokines** *(pg/ml/mg tissue)*")) %>%
  opt_align_table_header("center") %>% 
   tab_options(heading.title.font.size = 15,
               heading.background.color = "lightgrey",
               ) %>% 
     tab_source_note(
    source_note = "Matrinem Experiment 4"
  ) 
Exp4_tbl1 %>%
  gtsave(path = "exports/cytokines/", filename = "Exp4-tissue-adjusted-table.png")
```

## Plot
To see all the cytokine leves plotted run this code: 
```{r Exp4 plot all cytokines, warning=FALSE}
Exp4_plot_all_cytokines <- cytokines_long %>%
  filter(exp_number == 4) %>% 
  #filter(group != "VANCO-BF") %>% 
  #filter(!variables %in% c("INFg", "IL-12p70")) %>%
ggplot(aes(x = group, 
           y = values, 
           fill = group)) + 
  geom_boxplot(outlier.shape = NA) +
  stat_compare_means(aes(label = ..p.signif..),
                   method = "wilcox.test", ref.group = "VANCO-BF",
                   label.y.npc = 0.75) +
  facet_wrap(~variables, 
             scales = 'free', 
             nrow = 2,
             ncol = 5) +
  labs(title = "Concentraion of pro-inflammatory cytokines in tissue",
       x = "",
       y = "Tissue conc. (pg/ml/mg tissue)",
       fill = "Groups") +
  theme_bw()+
  theme(                   # Change the appearance of labels
    plot.title = element_text(color="darkgrey", size=12, face="bold"),
    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
    #axis.text.y = element_blank(),
    axis.ticks.x = element_blank(),  # Hide labels
    axis.text.x = element_blank(),
    legend.title = element_text(face="bold"),   
    legend.position="bottom",
    legend.background = element_rect(fill="white")) + # Change legend box background color
  scale_fill_brewer(palette="Set1") + plot_annotation (
  #title = 'Cytokine expression',
   #subtitle = 'All cytokines measured',
  #tag_levels = 'A',
  caption = 'BF = Breast-Feeding, FF = Formula-Feeding, VANCO = Vancomycin,',
  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold"),
                plot.subtitle = element_text(hjust = 0.5, face = "bold.italic"),
                plot.caption = element_text(hjust = 0.5, size=7,face = "bold")))

Exp4_plot_all_cytokines
ggsave('exports/cytokines/Exp4-tissue-adjusted-plot.png',Exp4_plot_all_cytokines)
```

### 2022-11-17 code/cytokines/Cytokines_tissue_un-adjusted.Rmd
created cytokine script and deleted lots of old files

--- 2022-11-17 code/cytokines/Cytokines_tissue_adjusted.Rmd
@@
 ---
-title: "Cytokine analysis script - all rounds - adjusted for tissue weight"
+title: "Cytokine analysis script - all rounds - un-adjusted for tissue weight"
 author: "Kaare"
@@
 # Data import
-```{r adjust_data, echo=TRUE}
-#_______Import dataset and Transform to adjusted values_______#
+```{r un-adjust_data, echo=TRUE}
+#_______Import dataset and Transform to un-adjusted values_______#
 
@@
    #mutate(values = values/(tissue_weight_mg/1000)) %>% #Adjust for tissue weight / 1000 (grams)
-   mutate(values = values/tissue_weight_mg) %>% #Adjust for tissue weight (mg)
+   #mutate(values = values/tissue_weight_mg) %>% #Adjust for tissue weight (mg)
   pivot_wider(
@@
 Exp1_tissue_weight
-ggsave('exports/cytokines/Exp1-weights-of-tissue.png',Exp1_tissue_weight)
+ggsave('exports/cytokines_unadjusted/Exp1-weights-unadjusted-of-tissue.png',Exp1_tissue_weight)
 ```
@@
 Exp1_tbl1 %>%
-  gtsave(path = "exports/cytokines/", filename = "Exp1-tissue-adjusted.png", vwidth = 760, vheight = 450)
+  gtsave(path = "exports/cytokines_unadjusted/", filename = "Exp1-tissue-un-adjusted.png", vwidth = 760, vheight = 450)
 
@@
 Exp1_plot_all_cytokines
-ggsave('exports/cytokines/Exp1-tissue-adjusted-plot.png',Exp1_plot_all_cytokines)
+ggsave('exports/cytokines_unadjusted/Exp1-tissue-un-adjusted-plot.png',Exp1_plot_all_cytokines)
 ```
@@
   ) 
-ggsave('exports/cytokines/Exp2-weights-of-tissue.png',Exp2_tissue_weight)
+ggsave('exports/cytokines_unadjusted/Exp2-weights-unadjusted-of-tissue.png',Exp2_tissue_weight)
 
@@
   drop_na(values) %>% 
-  mutate(adjusted_values = log10(values)) %>% 
+  mutate(un-adjusted_values = log10(values)) %>% 
   ggplot(aes(x = group, y = values, color = group)) +
@@
 Exp2_tbl1 %>%
-  gtsave(path = "exports/cytokines/", filename = "Exp2-tissue-adjusted-table.png")
+  gtsave(path = "exports/cytokines_unadjusted/", filename = "Exp2-tissue-un-adjusted-table.png")
 
@@
                    method = "wilcox.test", ref.group = "CON-BF",
-                   label.y.npc = 0.75) +
+                   label.y.npc = 0.98) +
   facet_wrap(~variables, 
@@
 Exp2_plot_all_cytokines
-ggsave('exports/cytokines/Exp2-tissue-adjusted-plot.png',Exp2_plot_all_cytokines)
+ggsave('exports/cytokines_unadjusted/Exp2-tissue-un-adjusted-plot.png',Exp2_plot_all_cytokines)
 ```
@@
   ) 
-ggsave('exports/cytokines/Exp3-weights-of-tissue.png',Exp3_tissue_weight)
+ggsave('exports/cytokines_unadjusted/Exp3-weights-unadjusted-of-tissue.png',Exp3_tissue_weight)
 
@@
 Exp3_tbl1 %>%
-  gtsave(path = "exports/cytokines/", filename = "Exp3-tissue-adjusted-table.png")
+  gtsave(path = "exports/cytokines_unadjusted/", filename = "Exp3-tissue-un-adjusted-table.png")
 ```
@@
 Exp3_plot_all_cytokines
-ggsave('exports/cytokines/Exp3-tissue-adjusted-plot.png',Exp3_plot_all_cytokines)
+ggsave('exports/cytokines_unadjusted/Exp3-tissue-un-adjusted-plot.png',Exp3_plot_all_cytokines)
 ```
@@
   ) 
-ggsave('exports/cytokines/Exp4-weights-of-tissue.png',Exp4_tissue_weight)
+ggsave('exports/cytokines_unadjusted/Exp4-weights-unadjusted-of-tissue.png',Exp4_tissue_weight)
 
@@
 Exp4_tbl1 %>%
-  gtsave(path = "exports/cytokines/", filename = "Exp4-tissue-adjusted-table.png")
+  gtsave(path = "exports/cytokines_unadjusted/", filename = "Exp4-tissue-un-adjusted-table.png")
 ```
@@
                    method = "wilcox.test", ref.group = "VANCO-BF",
-                   label.y.npc = 0.75) +
+                   label.y.npc = 0.98) +
   facet_wrap(~variables, 
@@
 Exp4_plot_all_cytokines
-ggsave('exports/cytokines/Exp4-tissue-adjusted-plot.png',Exp4_plot_all_cytokines)
+ggsave('exports/cytokines_unadjusted/Exp4-tissue-un-adjusted-plot.png',Exp4_plot_all_cytokines)
 ```

### 2022-11-19 code/Round 4/Gene expression.Rmd
progress

--- 2022-11-14 code/Round 4/Gene expression.Rmd
@@
 
-#### Gene expression
+# Gene expression in Matrinem Experiment 4
 
 ```{r}
-delta_gene_expr <- read_excel("data/processed/gene_expression_fold_changes.xlsx")
-
-bar_charts_all_genes <- delta_gene_expr %>% 
+delta_gene_expr <- read_excel("data/processed/gene_expression_fold_changes.xlsx") %>%
    mutate(
@@
       groups,
-      "con-bf"= "CON-BF",
-      "fvt-form" = "VANCO-FVF",
-      "sm-form" = "VANCO-FF (SM)"
-      )) %>% 
+      "con-bf"= "VANCO-BF",
+      "sm-form" = "VANCO-SM",
+      "fvt-form" = "VANCO-FVF"
+
+      ))
+```
+
+## Increased
+```{r increased}
+increased <- delta_gene_expr %>% 
+ group_by(genes, groups) %>% 
+  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
+  filter(mean_fold_change >= 1.75)
+```
+
+## Increased plot 
+```{r increased plot}
+increased_plot <- delta_gene_expr %>% 
+  filter(genes %in% increased$genes) %>% 
+  mutate(across(genes, factor, levels=c("TNFa","MUC1","IL1b","REG3G"))) %>%
   group_by(genes, groups) %>% 
@@
   facet_wrap(~genes,
-             ncol = 5, 
+             nrow = 1, 
              scales = "free") + 
-  labs(title = "Level of gene expression in ileum tissue",
-       y = "Gene expression relative fold change",
+  labs(
+       y = "Fold Δ",
        x = "",
-       fill = "Groups",
-       caption = "Matrinem round 4") +
+       fill = "Groups") +
   theme_bw()+
-  theme(                   # Change the appearance of labels
-    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+  theme(                   
     axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
@@
     axis.text.x = element_blank(),
-    legend.title = element_text(face="bold"),   
-    legend.position="bottom",
-    legend.background = element_rect(fill="lightgrey"), # Change legend box background color
-    plot.caption = element_text(color = "darkgrey", size=5)) + 
+    legend.position = "none") + 
     scale_fill_brewer(palette="Set1")
+  increased_plot
+```
 
-bar_charts_all_genes
+## Decreased
+```{r decreased}
+decreased <- delta_gene_expr %>% 
+ group_by(genes, groups) %>% 
+  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
+  filter(mean_fold_change < 0.6)
 
-bar_charts_selected_genes <- delta_gene_expr %>% 
-  filter(genes %in% c("TNFa","IL1b", "REG3G")) %>% 
-  mutate(
-    groups = recode_factor(  #Change the order of the groups to desired plot order
-      groups,
-      "con-bf"= "VANCO-BF",
-      "fvt-form" = "VANCO-FVF",
-      "sm-form" = "VANCO-SM"
-      )) %>% 
+```
+
+## Decreased plot
+```{r decreased plot}
+decreased_plot <- delta_gene_expr %>% 
+  filter(genes %in% decreased$genes) %>% 
+  #mutate(across(genes, factor, levels=c("MUC1","IL1b","REG3G"))) %>%
   group_by(genes, groups) %>% 
@@
   facet_wrap(~genes,
-             ncol = 5, 
+             nrow = 1, 
              scales = "free") + 
-  labs(title = "Level of gene expression in ileum tissue - significant changes",
-       y = "Gene expression relative fold change",
+  labs(
+       y = "Fold Δ",
        x = "",
-       fill = "Groups",
-       caption = "Matrinem round 4") +
+       fill = "Groups") +
+  theme_bw()+
+  theme(                   
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.position = "none") + 
+    scale_fill_brewer(palette="Set1")
+decreased_plot
+```
+
+## No change 
+```{r no change}
+no_change <- delta_gene_expr %>% 
+  filter(!genes %in% c(increased$genes, decreased$genes))
+```
+
+## No change plot 
+```{r plot}
+no_change_plot <- no_change %>% 
+  group_by(genes, groups) %>% 
+  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
+  ggplot(
+    aes(x = groups,
+        y = mean_fold_change,
+        fill = groups)
+  ) + 
+  geom_col() +
+  facet_wrap(~genes,
+             nrow = 1, 
+             scales = "free") +
+  labs(
+       y = "Fold Δ",
+       x = "",
+       fill = "Groups") +
   theme_bw()+
@@
     legend.position="bottom",
-    legend.background = element_rect(fill="lightgrey"), # Change legend box background color
+    legend.background = element_rect(fill="lightgrey"),
     plot.caption = element_text(color = "darkgrey", size=5)) + 
   scale_fill_brewer(palette="Set1")
+```
 
-bar_charts_selected_genes
+## Patchwork
+```{r}
+in_de_no_patch <- (decreased_plot + increased_plot) / no_change_plot + plot_annotation(
+  title = "Gene expression relative fold change",
+  tag_levels = "A",
+  tag_suffix = ')',
+  caption = "A) Decreased relative expression.     B) Increased relative expression.     C) No change in gene expression."
+) & 
+  theme(plot.tag = element_text(size = 12),
+        plot.caption = element_text(face = "bold"))
+in_de_no_patch
 

### 2022-11-19 code/cytokines/Cytokines_serum.Rmd
progress

@@
 # Exp1
-## Tissue weight 
-```{r exp1_inspect data, echo=FALSE}
-
-#_____Weight of tissue samples_____#
-Exp1_tissue_weight <- cytokines_long %>%       
-   filter(exp_number == 1) %>%
-  ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
-  geom_boxplot() +
-  stat_compare_means(aes(label = ..p.signif..),
-                   method = "wilcox.test", ref.group = "CON-BF",
-                   label.y.npc = 0.55) +
-  stat_compare_means(method = "kruskal.test",
-    aes(label = paste0("Kruskal Wallis Test. \n p = ", ..p.format..)),
-    label.y.npc = 0.75,
-    label.x.npc = 0.3) +
-  geom_point() +
-  labs(
-    title = "Tissue weight of intestinal samples, experiment 1",
-    y = "Tissue weight (mg)",
-    x = "",
-    caption = "Pairwise comparison with CON-BF as reference. \n
-    **: p < 0.01, *: p < 0.05, NS: not significant."
-  ) 
-Exp1_tissue_weight
-ggsave('exports/cytokines/Exp1-weights-of-tissue.png',Exp1_tissue_weight)
-```
 
@@
 Exp1_plot_all_cytokines
-ggsave('exports/cytokines/Exp1-serum-adjusted-plot.png',Exp1_plot_all_cytokines)
+ggsave(width = 170, height = 120, units = "mm",'exports/cytokines/Exp1-serum-adjusted-plot.png',Exp1_plot_all_cytokines)
 ```

### 2022-11-19 code/cytokines/Cytokines_tissue_adjusted.Rmd
progress

@@
 date: "11th October 2022"
-output: html_document
+output:
+  pdf_document: default
+  html_document: default
 editor_options:
@@
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE, warning=FALSE)
+knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
 ```
@@
 Exp1_tissue_weight
-ggsave('exports/cytokines/Exp1-weights-of-tissue.png',Exp1_tissue_weight)
+ggsave(width = 170, height = 120, units = "mm",'exports/cytokines/Exp1-weights-of-tissue.png',Exp1_tissue_weight)
 ```
@@
              stat_4 ~ px(140)) 
+Exp1_tbl1
 Exp1_tbl1 %>%
-  gtsave(path = "exports/cytokines/", filename = "Exp1-tissue-adjusted.png", vwidth = 760, vheight = 450)
+  gtsave(path = "exports/cytokines/", filename = "Exp1-tissue-adjusted-table.png")
 
@@
 Exp1_plot_all_cytokines
-ggsave('exports/cytokines/Exp1-tissue-adjusted-plot.png',Exp1_plot_all_cytokines)
+ggsave(width = 170, height = 120, units = "mm", 'exports/cytokines/Exp1-tissue-adjusted-plot.png',Exp1_plot_all_cytokines)
 ```
@@
   ) 
-ggsave('exports/cytokines/Exp2-weights-of-tissue.png',Exp2_tissue_weight)
+ggsave(width = 170, height = 120, units = "mm",'exports/cytokines/Exp2-weights-of-tissue.png',Exp2_tissue_weight)
 
@@
   ) 
+Exp2_tbl1
 Exp2_tbl1 %>%
@@
 Exp2_plot_all_cytokines
-ggsave('exports/cytokines/Exp2-tissue-adjusted-plot.png',Exp2_plot_all_cytokines)
+ggsave(width = 170, height = 120, units = "mm",'exports/cytokines/Exp2-tissue-adjusted-plot.png',Exp2_plot_all_cytokines)
 ```
@@
   ) 
-ggsave('exports/cytokines/Exp3-weights-of-tissue.png',Exp3_tissue_weight)
+ggsave(width = 170, height = 120, units = "mm",'exports/cytokines/Exp3-weights-of-tissue.png',Exp3_tissue_weight)
 
@@
   ) 
+Exp3_tbl1
 Exp3_tbl1 %>%
@@
 Exp3_plot_all_cytokines
-ggsave('exports/cytokines/Exp3-tissue-adjusted-plot.png',Exp3_plot_all_cytokines)
+ggsave(width = 200, height = 120, units = "mm",'exports/cytokines/Exp3-tissue-adjusted-plot.png',Exp3_plot_all_cytokines)
 ```
@@
   ) 
-ggsave('exports/cytokines/Exp4-weights-of-tissue.png',Exp4_tissue_weight)
+ggsave(width = 170, height = 120, units = "mm",'exports/cytokines/Exp4-weights-of-tissue.png',Exp4_tissue_weight)
 
@@
   ) 
+Exp4_tbl1
 Exp4_tbl1 %>%
@@
 Exp4_plot_all_cytokines
-ggsave('exports/cytokines/Exp4-tissue-adjusted-plot.png',Exp4_plot_all_cytokines)
+ggsave(width = 170, height = 120, units = "mm",'exports/cytokines/Exp4-tissue-adjusted-plot.png',Exp4_plot_all_cytokines)
 ```

### 2022-11-20 code/cytokines/Cytokines_tissue_adjusted.Rmd
Progress

@@
   )
+```
+
+# All experiments - tissue weight stats
+
+```{r}
+
+cytokines_long %>%       
+  filter(exp_number %in% c("1","3")) %>% 
+  group_by(exp_number) %>% 
+  kruskal_test(tissue_weight_mg ~ type_of_feed)
+
+#|exp_number |Effect       | DFn| DFd|       F|     p|p<.05 |   ges|
+#|:----------|:------------|---:|---:|-------:|-----:|:-----|-----:|
+#|1          |type_of_feed |   1| 298|   0.816| 0.367|      | 0.003|
+#|3          |type_of_feed |   1| 378| 373.425| 0.000|*     | 0.497|
+#|4          |type_of_feed |   1| 258|   9.863| 0.002|*     | 0.037|
+
 ```
@@
   filter(exp_number == 3) %>% 
-  kruskal_test(tissue_weight_mg ~ maternal_treatment) %>% 
-  knitr::kable(escape = TRUE)
+  kruskal_test(tissue_weight_mg ~ type_of_feed)
 
@@
   filter(exp_number == 4) %>% 
-  kruskal_test(tissue_weight_mg ~ group) %>% 
-  knitr::kable(escape = TRUE)
+  kruskal_test(tissue_weight_mg ~ type_of_feed)
+
+cytokines_long %>%       
+  filter(exp_number == 4) %>% 
+  group_by(type_of_feed) %>% 
+  summarize(mean = mean(tissue_weight_mg),
+            sd = sd(tissue_weight_mg),
+            median = median(tissue_weight_mg),
+            iqr = IQR(tissue_weight_mg))
+
+  anova_test(tissue_weight_mg ~ type_of_feed)
+
+cytokines_long %>%       
+  filter(exp_number == 4) %>% 
+  anova_test(tissue_weight_mg ~ type_of_feed)
 

### 2022-11-20 code/weight_analysis.Rmd
Progress

@@
   arrange(time)
+```
+
+## Weight at allocation - mat vs con
+```{r weight-at-allocation-mat-vs-con}
+get_box_stats <- function(y, upper_limit = max(data$bodyweight[data$time == "day_5"]) * 1.2) {
+  return(data.frame(
+    y =  0.9*upper_limit,
+    label = paste(
+      "n =", length(y), "\n",
+      "Mean =", round(mean(y), 2), "\n" #,
+      #"Median =", round(median(y), 2), "\n"
+    )
+  ))
+}
+
+my_comparisons <- list( c("CON", "MAT"))
+
+data %>% 
+  #filter(exp_number %in% c("1", "2", "3","4")) %>% 
+  filter(time %in% c("day_3")) %>%
+  drop_na(bodyweight) %>% 
+  ggplot(
+    aes(x = received_antibiotics,
+        y = bodyweight)
+  ) +
+  geom_boxplot() +
+  geom_dotplot(
+     binaxis = "y", stackdir = "center", dotsize = 0.5) +
+  facet_grid(~ exp_number) +
+  stat_summary(
+    fun.data = get_box_stats, geom = "text", hjust = 0.5, vjust = 0.9) +
+  labs(
+    #title = "Body weight of offspring from antibiotic-treated vs water-treated mice",
+   # subtitle = "Results shown from each of the 4 experimental rounds",
+    caption = "In Exp. 4 all mice received antibiotoics, hence no control.
+    Comparison method: Student T-test",
+    x = "Maternal Treatment",
+    y = "Bodyweight at end of study (g)"
+  ) +
+  theme_bw() + 
+  stat_compare_means(
+    method = "t.test", 
+    label.y = 3, 
+    comparisons = my_comparisons
+    )
+```
+
+## Weight at allocation - form vs BF
+```{r weight-at-allocation-ff-vs-bf}
+get_box_stats <- function(y, upper_limit = max(data$bodyweight[data$time == "day_5"]) * 1.2) {
+  return(data.frame(
+    y =  0.9*upper_limit,
+    label = paste(
+      "n =", length(y), "\n",
+      "Mean =", round(mean(y), 2), "\n" #,
+      #"Median =", round(median(y), 2), "\n"
+    )
+  ))
+}
+
+my_comparisons <- list( c("BF", "FF"))
+
+data %>% 
+  #filter(exp_number %in% c("1", "2", "3","4")) %>% 
+  filter(time %in% c("day_5")) %>%
+  drop_na(bodyweight) %>% 
+  ggplot(
+    aes(x = type_of_feed,
+        y = bodyweight)
+  ) +
+  geom_boxplot() +
+  geom_dotplot(
+     binaxis = "y", stackdir = "center", dotsize = 0.5) +
+  facet_grid(~ exp_number) +
+  stat_summary(
+    fun.data = get_box_stats, geom = "text", hjust = 0.5, vjust = 0.9) +
+  labs(
+    #title = "Body weight of offspring from antibiotic-treated vs water-treated mice",
+   # subtitle = "Results shown from each of the 4 experimental rounds",
+    caption = " Comparison method: Student T-test",
+    x = "Type of feed",
+    y = "Bodyweight at end of study (g)"
+  ) +
+  theme_bw() + 
+  stat_compare_means(
+    method = "t.test", 
+    label.y = 4.35, 
+    comparisons = my_comparisons
+    )
 ```
@@
 ```{r style baseline bodyweight}
-get_box_stats <- function(y, upper_limit = max(data$Bodyweight[data$Time == "Day2"]) * 1.0) {
+get_box_stats <- function(y, upper_limit = max(data$bodyweight[data$time == "day_5"]) * 1.0) {
   return(data.frame(
@@
 
-my_comparisons <- list( c("Water", "AB"))
+my_comparisons <- list( c("CON", "MAT"))
 
 data %>% 
-  filter(Time == "Day2") %>% 
+  filter(time %in% c("day_3", "day_5")) %>% 
   ggplot(aes(
-    received_antibiotic, Bodyweight
+    time, bodyweight
   )) + 
   geom_boxplot() +
-  facet_wrap(~experiment_no, 
+  facet_wrap(~ exp_number, 
              nrow = 1) + 
@@
   geom_boxplot() +
-  facet_wrap(~experiment_no, 
+  facet_wrap(~exp_number, 
              nrow = 1) +

### 2022-11-20 code/weight_development.R

@@ -15,14 +15,60 @@ data <- read_excel("data/processed/weight_development.xlsx") %>%
   mutate(received_antibiotics = recode_factor(received_antibiotics, 
                                               yes = "MAT",
                                               no = "CON"))
-startweight <- data %>% 
-  filter(exp_number %in% c("1","3","4")) %>% 
+
+### effect of FEED on startweight ###
+
+data %>% 
+  filter(exp_number %in% c("1","2", "3","4")) %>% 
   filter(time %in% c("day_3", "day_5")) %>%
   drop_na(bodyweight) %>% 
   group_by(exp_number, type_of_feed, time) %>% 
   summarise(startweight = mean(bodyweight),
-            sd = sd(bodyweight))
+            sd = sd(bodyweight)) %>% 
+  filter(time == "day_3") %>% 
+  knitr::kable()
 
+
+data %>% 
+  filter(exp_number %in% c("1","3", "4")) %>% 
+  filter(time %in% c("day_3")) %>%
+  drop_na(bodyweight) %>% 
+  group_by(exp_number) %>% 
+  t_test(bodyweight ~ type_of_feed,
+         alternative = "two.sided",
+         mu = 0) %>% 
+  knitr::kable(espace = TRUE)
+
+### effect of MAT on startweight ###
+
+data %>% 
+  filter(exp_number %in% c("3")) %>% 
+  filter(time %in% c("day_3", "day_5")) %>%
+  drop_na(bodyweight) %>% 
+  group_by(exp_number, received_antibiotics, time) %>% 
+  summarise(startweight = mean(bodyweight),
+            sd = sd(bodyweight)) %>% 
+  filter(time == "day_5") %>% 
+  knitr::kable()
+
+data %>% 
+  filter(exp_number %in% c("3")) %>% 
+  filter(time %in% c("day_5")) %>%
+  drop_na(bodyweight) %>% 
+  group_by(exp_number) %>% 
+  anova_test(bodyweight ~ received_antibiotics) %>% 
+  knitr::kable(espace = TRUE)
+
+data %>% 
+  filter(exp_number %in% c("1","2","3")) %>% 
+  filter(time %in% c("day_5")) %>%
+  drop_na(bodyweight) %>% 
+  group_by(exp_number) %>% 
+  anova_test(bodyweight ~ received_antibiotics) %>% 
+  knitr::kable(espace = TRUE)
+
+
+### effect of MAT on birthweight ###
[…]

### 2022-11-22 code/Gene expression.Rmd
Returning to Imac

--- 2022-11-19 code/Round 4/Gene expression.Rmd
@@
       ))
+```
+
+## Statistical analysis of dCT levels 
+```{r}
+anova <- delta_gene_expr %>% 
+  group_by(genes) %>% 
+  anova_test(dct ~ groups) %>% 
+  ungroup()
+
+significant <- delta_gene_expr %>% 
+  group_by(genes) %>% 
+  do(lm(dct ~ groups,.) %>%  tidy) %>% 
+  filter(term != "(Intercept)") %>% 
+  filter(p.value <= 0.05)
+```
+
+## Add house gene to list
+```{r}
+wide_genes <- delta_gene_expr %>% 
+  select(sample_id,ct_value_house_gene_HRPT, groups, genes, dct) %>% 
+  pivot_wider(id_cols = c("sample_id","groups","ct_value_house_gene_HRPT"), names_from = genes, values_from = dct) %>% 
+  mutate(HRPT = ct_value_house_gene_HRPT) 
+```
+
+## TAble
+```{r}
+Exp4_genes_table <- delta_gene_expr %>% 
+  select(sample_id,ct_value_house_gene_HRPT, groups, genes, dct) %>% 
+  pivot_wider(id_cols = c("sample_id","groups","ct_value_house_gene_HRPT"), 
+              names_from = genes, 
+              values_from = dct) %>% 
+  mutate(
+    HRPT = ct_value_house_gene_HRPT)  %>% 
+  select(
+    -sample_id, -ct_value_house_gene_HRPT) %>% 
+  tbl_summary(by = groups,
+              digits = everything() ~ c(1,1),
+              missing = "no",
+              #include = !"INFg",
+              #type = all_continuous() ~ "continuous2",
+              statistic = all_continuous() ~ ("{mean} ({sd})")) %>%
+  modify_header(
+    label ~ "**Groups:**") %>% 
+   #modify_footnote(stat_1 ~ "Median (IQR)", stat_2 ~ "p-value on overall group differences,Kruskall-Wallis rank sum test ") %>% 
+add_p(everything() ~ "aov") %>% 
+  as_gt() %>%
+  tab_header(title = md("**Gene expression** *(dCT values)*")) %>%
+  opt_align_table_header("center") %>% 
+   tab_options(heading.title.font.size = 15,
+               heading.background.color = "lightgrey",
+               ) %>% 
+     tab_source_note(
+    source_note = "HRPT served as reference gene."
+  )
+
+Exp4_genes_table %>% 
+  gtsave(path = "exports/genes/", filename = "Exp4-gene-dct-table-aov.png")
 ```

### 2022-11-22 code/Histology all rounds.Rmd
Returning to Imac

---
title: "HISTO1"
author: "Kaare"
date: "6/27/2022"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# MATRINEM histo-analysis. 
AB-PAS stains quantified as % of total tissue. Experiments 2-4 collected samples for histology.

Samples are of ileum (small intestine) and colon collected from mouse pups 6 days postnatally.

```{r load packages, include=FALSE}
# Packages required
library(readxl)
library(tidyverse)
library(broom)
library(ggpubr)
library(rstatix)
library(patchwork)
```

## Load data
```{r data load}
data <- read_excel("data/raw/Histo_analysis .xlsx")

allrounds <- read_excel("data/processed/weight_development.xlsx")

```

```{r}
clean_data <- data %>% 
  mutate(ID = str_remove(ID, "-C") %>% str_remove(.,"-I") %>% str_remove(.," I")) %>% 
  mutate(ID = as.numeric(ID))

filter_joined_df <- semi_join(allrounds, clean_data, by = c("sample_id" = "ID"))

histo <- filter_joined_df %>% 
  inner_join(clean_data, by = c("sample_id" = "ID"))

histo <- histo %>% 
  pivot_longer(., `PAS-RELATIVE-TO-TOTAL(%)`:`AB-RELATIVE-TO-TOTAL(%)`, 
               names_to = "AB-PAS",
               values_to = "values")

```

```{r}
histo %>%  ### Try to wrap histo longer to have all stain values in 1 column and facet-wrap by PAS + AB
  filter(TISSUE == "ileum") %>%
  filter('AB-PAS' == "PAS-RELATIVE-TO-TOTAL(%)") %>% 
  ggplot(
    aes(
    x = group2, 
    y = values,
    fill = group2)
    ) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~ exp_number,
             scales = "free") +
  labs(
    #title =  "Ileum AB-PAS staining % of total tissue", subtitle = "Faceted by experimental round (2-4)",
       y = "PAS stained as % of total tissue",
       x = "Maternal Treatment") + 
  theme_bw() +
  theme(
    axis.text.x = element_blank()
  )

```

```{r MAT vs CON}
histo %>% 
  filter(TISSUE == "colon") %>% 
   ggplot(aes(x = received_antibiotics, y = `% of total tissue stained`)) +
  geom_boxplot() + 
  facet_wrap(~ exp_number) +
  labs(title =  "COLON AB-PAS staining % of total tissue", subtitle = "Faceted by experimental round (2-4)",
       x = "Maternal Treatment") + 
  theme_bw()

```

```{r MAT vs CON}
histo %>% 
  filter(TISSUE == "colon") %>% 
   ggplot(aes(x = received_antibiotics, y = `% of total tissue stained`)) +
  geom_boxplot() + 
  facet_wrap(~ exp_number) +
  labs(title =  "COLON AB-PAS staining % of total tissue", subtitle = "Faceted by experimental round (2-4)",
       x = "Maternal Treatment") + 
  theme_bw()

```

### 2022-11-23 code/cleaning_raw_data NEW.Rmd
Updated raw data cleaning script to match project folder paths

@@
 ---
-title: "document_title"
+title: "Raw data cleaning"
 author: "Kaare D. Tranaes"
@@
 
-# Raw data cleaning document
-
-::: {#Project Description}
+#README
 This is a documentation of generating the working files from the raw data files. Every step of the process is explained below.  
-The raw data files can be accessed in the RawData folder associated with this project.  
-Once cleaned up, a new data file for further analysis is created and stored in CleanData folder.
-:::
+The raw data files can be accessed in the data/raw folder associated with this project.  
+Once cleaned up, a new data file for further analysis is created and stored in the data/processed folder.
 
@@
 ```{r load data}
-import_round1 <- read_excel("../Data/RawData/bodyweight_round_1.xlsx"
+import_round1 <- read_excel("data/raw/bodyweight_round_1.xlsx"
         #                     , col_types = c("text","text", "text", "text", 
@@
 
-import_round2 <- read_excel("../Data/RawData/bodyweight_round_2.xlsx"
+import_round2 <- read_excel("data/raw/bodyweight_round_2.xlsx"
         #                     ,col_types = c("text", "text", "text", 
@@
 
-import_round3 <- read_excel("../Data/RawData/bodyweight_round_3.xlsx"
+import_round3 <- read_excel("data/raw/bodyweight_round_3.xlsx"
         ) %>% 
@@
 
-import_round4 <- read_excel("../Data/RawData/bodyweight_round_4.xlsx"
+import_round4 <- read_excel("data/raw/bodyweight_round_4.xlsx"
         #                     ,col_types = c("text", "text", "text", 
@@
 View(weights_all_rounds)
-rio::export(weights_all_rounds, file = "../Data/CleanData/weight_development.xlsx", overwrite = TRUE)
+rio::export(weights_all_rounds, file = "data/processed/weight_development.xlsx", overwrite = TRUE)
 ```
@@
 
-rio::export(study_chars, file = "../Data/CleanData/study_id_info.xlsx", overwrite = T)
+rio::export(study_chars, file = "data/processed/study_id_info.xlsx", overwrite = T)
 
@@
 ```{r}
-study_chars <- read_excel("../Data/CleanData/study_id_info.xlsx")
+study_chars <- read_excel("data/processed/study_id_info.xlsx")
 
-mesoscale_round_1 <- read_excel("../Data/RawData/cytokines_round_1_r.xlsm")
+mesoscale_round_1 <- read_excel("data/raw/cytokines_round_1_r.xlsm")
 
-mesoscale_round_2_and_3 <- read_excel("../Data/RawData/cytokines_round_2_and_3.xlsx") %>% 
+mesoscale_round_2_and_3 <- read_excel("data/raw/cytokines_round_2_and_3.xlsx") %>% 
   rename("tissue_weight_mg" = "tissue_weight_(mg)", "exp_number" = "exp_no")
 
-mesoscale_round_4 <- read_excel("../Data/RawData/cytokines_round_4.xlsx") %>%
+mesoscale_round_4 <- read_excel("data/raw/cytokines_round_4.xlsx") %>%
   rename("group" = "Group", "exp_number" = "exp_no")
@@
 
-#rio::export(cytokines, file = "../Data/CleanData/cytokines.xlsx")
+#rio::export(cytokines, file = "data/processed/cytokines.xlsx")
 ```
@@
 ```{r}
-fitc_round_2_and_3 <- read_excel("../Data/RawData/FITC_round_2-3.xlsx") %>% 
+fitc_round_2_and_3 <- read_excel("data/raw/FITC_round_2-3.xlsx") %>% 
   rename(exp_number = exp_no) %>% 
@@
 
-fitc_round_4 <- read_excel("../Data/RawData/FITC_round_4.xlsx") %>% 
+fitc_round_4 <- read_excel("data/raw/FITC_round_4.xlsx") %>% 
   mutate("result_µg/ml_undiluted" = `result_µg/ml`*dilution_factor)
@@
 
-rio::export(fitc, file = "../Data/CleanData/fitc.xlsx")
+rio::export(fitc, file = "data/processed/fitc.xlsx")
 ```
@@
 ### Start here. Or do it in excel.
-study_chars <- read_excel("../Data/CleanData/study_id_info.xlsx")
-qpcr <- read_excel("../Data/RawData/all_genes_ct_values.xlsx")
+study_chars <- read_excel("data/processed/study_id_info.xlsx")
+qpcr <- read_excel("data/raw/all_genes_ct_values.xlsx")
 
@@
 
-gene_expr <- study_chars %>%   #### Maybe export this to excel and try to create double delta ct analysis there. 
+gene_expr <- study_chars %>%
   filter(sample_id %in% 89:112) %>% 
@@
                values_to = "ct_value") %>%
-  select(sample_id, groups, genes, ct_value, ct_value_house_gene_HRPT) %>% 
+    select(sample_id, group, genes, ct_value, ct_value_house_gene_HRPT) %>% 
     filter(!is.na(ct_value)) %>%
   mutate(dct = ct_value-ct_value_house_gene_HRPT) %>% 
-  group_by(genes, groups) %>% 
+  group_by(genes, group) %>% 
   mutate(avg_group_dct_val = mean(dct)) %>% 
   group_by(genes) %>% 
-  mutate(ddct = dct - (avg_group_dct_val[groups=="con-bf"])) %>% 
+  mutate(ddct = dct - (avg_group_dct_val[group=="CON-BF"])) %>% 
   mutate("2^-(ddct)" = 2^-ddct) %>% 
-  group_by(genes, groups) %>% 
-  mutate(avg_group_2_exp_ddct_val = mean(`2^-(ddct)`)) %>%
-  group_by(genes) %>% 
-  mutate(relative_fold_change = `2^-(ddct)`/(avg_group_2_exp_ddct_val[groups=="con-bf"]))
+  ungroup()
 
@@
 
-#rio::export(delta_gene_expr, file = "../Data/CleanData/gene_expression_fold_changes.xlsx")
+rio::export(delta_gene_expr, file = "data/processed/gene_expression_fold_changes.xlsx")
 
@@
 
-bar_charts <- delta_gene_expr %>% 
-  group_by(genes, groups) %>% 
-  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
+delta_gene_expr %>% 
+  group_by(genes, group) %>% 
+  summarise(mean_fold_change = mean(`2^-(ddct)`)) %>% 
   ggplot(
-    aes(x = groups,
+    aes(x = group,
         y = mean_fold_change,
-        fill = groups)
+        fill = group)
   ) + 
@@
   facet_wrap(~genes,
-             ncol = 5) +
+             ncol = 5,
+             scales = "free") +
   labs(title = "Gene expression fold change",
@@
       )
+
+delta_gene_expr %>% 
+  group_by(genes, group) %>% 
+  ggplot(
+    aes(x = group,
+        y = `2^-(ddct)`,
+        fill = group)
+  ) + 
+  geom_boxplot() +
+  facet_wrap(~genes,
+             ncol = 5,
+             scales = "free") +
+  labs(title = "Gene expression fold change",
+       y = "Gene expression fold change",
+       x = "Groups"
+      )
+
 ```

### 2022-11-25 code/Histology all rounds.Rmd
Progress on Imac

@@
 ```{r data load}
-data <- read_excel("data/raw/Histo_analysis .xlsx")
+data <- read_excel("data/raw/Histo_analysis .xlsx") %>% 
+  rename(pas_pct = `PAS-RELATIVE-TO-TOTAL(%)`,
+         ab_pct = `AB-RELATIVE-TO-TOTAL(%)`) %>% 
+  mutate(AB_PAS_pct = (ab_pct +pas_pct)/2) %>% 
+  mutate(tissue = if_else(TISSUE == "colon","Colon","Ileum"))
 
-allrounds <- read_excel("data/processed/weight_development.xlsx")
+allrounds <- read_excel("data/processed/weight_development.xlsx")  %>% 
+  select(-c(cage:comments))
 
@@
 histo <- filter_joined_df %>% 
-  inner_join(clean_data, by = c("sample_id" = "ID"))
+  inner_join(clean_data, by = c("sample_id" = "ID")) %>% 
+  rename(Group = group2)
+```
 
-histo <- histo %>% 
-  pivot_longer(., `PAS-RELATIVE-TO-TOTAL(%)`:`AB-RELATIVE-TO-TOTAL(%)`, 
-               names_to = "AB-PAS",
-               values_to = "values")
+# Exp2 
+```{r Exp2-ileum}
+Exp2_plot_histo <- histo %>% 
+  #filter(TISSUE == "ileum") %>%
+  filter(exp_number == 2) %>% 
+  ggplot(
+    aes(
+    x = Group, 
+    y = AB_PAS_pct,
+    fill = Group)
+    ) +
+  facet_wrap(~tissue,
+             scales = "free")+
+  geom_boxplot(outlier.shape = NA) +
+  geom_point() +
+  #coord_cartesian(ylim = c(0, 10)) +
+  stat_compare_means(aes(label = ..p.signif.., ),
+                   method = "wilcox.test", ref.group = "CON-BF",
+                   label.y.npc = 0.85) +
+  labs(
+    #title =  ,
+       y = "Mucin (% of total tissue)",
+       x = "") + 
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    #axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="white")) + # Change legend box background color
+  scale_fill_brewer(palette="Set1") + plot_annotation (
+  title = "Image analytical quantification of mycin",
+   #subtitle = 'All cytokines measured',
+  #tag_levels = 'A',
+  #caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin, VANCO = Vancomycin,',
+  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold")))
+
+Exp2_plot_histo
+ ggsave(width = 110, height = 110, units = "mm", 'exports/histo/Exp2-ab_pas_plot.png',Exp2_plot_histo)
 
@@
 
-```{r}
-histo %>%  ### Try to wrap histo longer to have all stain values in 1 column and facet-wrap by PAS + AB
-  filter(TISSUE == "ileum") %>%
-  filter('AB-PAS' == "PAS-RELATIVE-TO-TOTAL(%)") %>% 
+# Exp3 
+```{r Exp3-ileum}
+Exp3_plot_histo <- histo %>% 
+  filter(Group != "VANCO-BF") %>%
+  filter(exp_number == 3) %>% 
   ggplot(
     aes(
-    x = group2, 
-    y = values,
-    fill = group2)
+    x = Group, 
+    y = AB_PAS_pct,
+    fill = Group)
     ) +
+  facet_wrap(~tissue,
+             scales = "free")+
   geom_boxplot(outlier.shape = NA) +
-  facet_wrap(~ exp_number,
-             scales = "free") +
+  geom_point() +
+  #coord_cartesian(ylim = c(0, 10)) +
+  stat_compare_means(aes(label = ..p.signif.., ),
+                   method = "wilcox.test", ref.group = "CON-BF",
+                   label.y.npc = 0.85) +
   labs(
-    #title =  "Ileum AB-PAS staining % of total tissue", subtitle = "Faceted by experimental round (2-4)",
-       y = "PAS stained as % of total tissue",
-       x = "Maternal Treatment") + 
-  theme_bw() +
-  theme(
-    axis.text.x = element_blank()
-  )
+    #title =  ,
+       y = "Mucin (% of total tissue)",
+       x = "") + 
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    #axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="white")) + # Change legend box background color
+  scale_fill_brewer(palette="Set1") + plot_annotation (
+  title = "Image analytical quantification of mycin",
+   #subtitle = 'All cytokines measured',
+  #tag_levels = 'A',
+  #caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin, VANCO = Vancomycin,',
+  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold")))
+
+Exp3_plot_histo
+ ggsave(width = 170, height = 110, units = "mm", 'exports/histo/Exp3-ab_pas_plot.png',Exp3_plot_histo)
 
@@
 
-```{r MAT vs CON}
-histo %>% 
-  filter(TISSUE == "colon") %>% 
-   ggplot(aes(x = received_antibiotics, y = `% of total tissue stained`)) +
-  geom_boxplot() + 
-  facet_wrap(~ exp_number) +
-  labs(title =  "COLON AB-PAS staining % of total tissue", subtitle = "Faceted by experimental round (2-4)",
-       x = "Maternal Treatment") + 
-  theme_bw()
+# Exp4 
+```{r Exp4-ileum}
+Exp4_plot_histo <- histo %>% 
+  #filter(TISSUE == "ileum") %>%
+  filter(exp_number == 4) %>% 
+  ggplot(
+    aes(
+    x = Group, 
+    y = AB_PAS_pct,
+    fill = Group)
+    ) +
+  facet_wrap(~tissue,
+             scales = "free")+
+  geom_boxplot(outlier.shape = NA) +
+  geom_point() +
+  #coord_cartesian(ylim = c(0, 10)) +
+  stat_compare_means(aes(label = ..p.signif.., ),
+                   method = "wilcox.test", ref.group = "VANCO-SM",
+                   label.y.npc = 0.85) +
+  labs(
+    #title =  ,
+       y = "Mucin (% of total tissue)",
+       x = "") + 
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    #axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="white")) + # Change legend box background color
+  scale_fill_brewer(palette="Set1") + plot_annotation (
+  title = "Image analytical quantification of mycin",
+   #subtitle = 'All cytokines measured',
+  #tag_levels = 'A',
+  #caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin, VANCO = Vancomycin,',
+  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold")))
+
+Exp4_plot_histo
+ ggsave(width = 110, height = 110, units = "mm", 'exports/histo/Exp4-ab_pas_plot.png',Exp4_plot_histo)
 
 ```
+
+# Overall MAT
+```{r overall-mat}
+histo_overall_colon <- histo %>% 
+  filter(tissue == "Colon") %>%
+  filter(exp_number != 4) %>% 
+  ggplot(
+    aes(
+    x = received_antibiotics, 
+    y = AB_PAS_pct,
+    fill = received_antibiotics)
+    ) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_point() +
+  facet_wrap(~exp_number,
+             scales = "free") +
+  #coord_cartesian(ylim = c(0, 40)) +
+  stat_compare_means(aes(label = paste0("p = ", ..p.format..)),
+                   method = "wilcox.test", ref.group = "No",
+                   label.y.npc = 0.85) +
+  labs(
+    #title =  ,
+      y = "Mucin (% of total tissue)",
+       x = "",
+       fill='') + 
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    #axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_blank()) +  
+  #legend.background = element_rect(fill="white")) + # Change legend box background color
+  scale_fill_brewer(palette="Set1")
+
+histo_overall_colon
+
+histo_overall_ileum <- histo %>% 
+  filter(tissue == "Ileum") %>%
+  filter(exp_number != 4) %>% 
+  ggplot(
+    aes(
+    x = received_antibiotics, 
+    y = AB_PAS_pct,
+    fill = received_antibiotics)
+    ) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_point() +
+  facet_wrap(~exp_number,
+             scales = "free") +
+  #coord_cartesian(ylim = c(0, 40)) +
+  stat_compare_means(aes(label = paste0("p = ", ..p.format..)),
+                   method = "wilcox.test", ref.group = "No",
+                   label.y.npc = 0.85) +
+  labs(
+    #title =  ,
+      y = "Mucin (% of total tissue)",
+       x = "",
+      fill='Maternal Antibiotic Treatment') +
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    #axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="white")) + # Change legend box background color
+  scale_fill_brewer(palette="Set1")
+
+histo_overall_ileum
+
+histo_overall_both <- (histo_overall_colon/histo_overall_ileum) + plot_annotation (
+  title = "Effect of Maternal Antibiotics Treatment on intestinal mucin",
+   #subtitle = 'All cytokines measured',
+  tag_levels = 'A',
+  #caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin, VANCO = Vancomycin,',
+  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold")))
+
+histo_overall_both
+ ggsave(width = 110, height = 110, units = "mm", 'exports/histo/Hist_MAT',histo_overall_both)
+ ggsave('exports/histo/Hist_MAT.png', histo_overall_both)
+
+```
+
+# Overall FEED
+```{r overall-mat}
+histo_FEED_colon <- histo %>% 
+  filter(tissue == "Colon") %>%
+  filter(exp_number != 2) %>% 
+  ggplot(
+    aes(
+    x = type_of_feed, 
+    y = AB_PAS_pct,
+    fill = type_of_feed)
+    ) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_point() +
+  facet_wrap(~exp_number,
+             scales = "free") +
+  #coord_cartesian(ylim = c(0, 40)) +
+  stat_compare_means(aes(label = paste0("p = ", ..p.format..)),
+                   method = "wilcox.test", ref.group = "BF",
+                   label.y.npc = 0.85) +
+  labs(
+    #title =  ,
+      y = "Mucin (% of total tissue)",
+       x = "",
+       fill='') + 
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    #axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_blank()) +  
+  #legend.background = element_rect(fill="white")) + # Change legend box background color
+  scale_fill_brewer(palette="Set1")
+
+histo_FEED_colon
+
+histo_FEED_ileum <- histo %>% 
+  filter(tissue == "Ileum") %>%
+  filter(exp_number != 2) %>% 
+  ggplot(
+    aes(
+    x = type_of_feed, 
+    y = AB_PAS_pct,
+    fill = type_of_feed)
+    ) +
+  geom_boxplot(outlier.shape = NA) +
+  geom_point() +
+  facet_wrap(~exp_number,
+             scales = "free") +
+  #coord_cartesian(ylim = c(0, 40)) +
+  stat_compare_means(aes(label = paste0("p = ", ..p.format..)),
+                   method = "wilcox.test", ref.group = "BF",
+                   label.y.npc = 0.85) +
+  labs(
+    #title =  ,
+      y = "Mucin (% of total tissue)",
+       x = "",
+      fill='Maternal Antibiotic Treatment') +
+  theme_bw()+
+  theme(                   # Change the appearance of labels
+    plot.title = element_text(color="darkgrey", size=12, face="bold"),
+    axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    #axis.text.y = element_blank(),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.title = element_text(face="bold"),   
+    legend.position="bottom",
+    legend.background = element_rect(fill="white")) + # Change legend box background color
+  scale_fill_brewer(palette="Set1")
+
+histo_FEED_ileum
+
+histo_FEED_both <- (histo_FEED_colon/histo_FEED_ileum) + plot_annotation (
+  title = "Effect of formula or breastmilk on intestinal mucin",
+   #subtitle = 'All cytokines measured',
+  tag_levels = 'A',
+  #caption = 'CON = Water, BF = Breast-Feeding, GENTA = Gentamicin, VANCO = Vancomycin,',
+  theme = theme(plot.title = element_text(hjust = 0.5, size=14, face="bold")))
+
+histo_FEED_both
+ ggsave(width = 110, height = 110, units = "mm", 'exports/histo/Hist_MAT',histo_overall_both)
+ ggsave('exports/histo/Hist_FEED.png', histo_FEED_both)
+
+```

### 2022-11-26 code/Gene expression.Rmd
updated gene expression

@@
     groups = recode_factor(  #Change the order of the groups to desired plot order
-      groups,
-      "con-bf"= "VANCO-BF",
-      "sm-form" = "VANCO-SM",
-      "fvt-form" = "VANCO-FVF"
-
+      group,
+      "CON-BF"= "VANCO-BF",
+      "SM-FORM" = "VANCO-SM",
+      "FVT-FORM" = "VANCO-FVF"
       ))
@@
 
+```{r}
+delta_gene_expr %>% 
+  group_by(genes, groups) %>% 
+  summarise(mean_fold_change = mean(`2^-(ddct)`)) %>% 
+  ggplot(
+    aes(x = groups,
+        y = mean_fold_change,
+        fill = groups)
+  ) + 
+  geom_bar(position = "dodge", stat = "summary", fun = "mean") +
+  facet_wrap(~genes,
+             ncol = 5,
+             scales = "free") +
+  labs(title = "Gene expression fold change",
+       y = "Gene expression fold change",
+       x = "Groups"
+      )
+
+delta_gene_expr %>% 
+  group_by(genes, groups) %>% 
+  ggplot(
+    aes(x = groups,
+        y = `2^-(ddct)`,
+        fill = groups)
+  ) + 
+  geom_boxplot() +
+  facet_wrap(~genes,
+             ncol = 5,
+             scales = "free") +
+  labs(title = "Gene expression fold change",
+       y = "Gene expression fold change",
+       x = "Groups"
+      )
+
+```
+
+```{r}
+my_comparisons <- list(c("VANCO-FVF", "VANCO-SM"), c("VANCO-BF", "VANCO-FVF"))
+```
+
 ## Increased
@@
  group_by(genes, groups) %>% 
-  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
+  summarise(mean_fold_change = mean(`2^-(ddct)`)) %>% 
   filter(mean_fold_change >= 1.75)
@@
   filter(genes %in% increased$genes) %>% 
-  mutate(across(genes, factor, levels=c("TNFa","MUC1","IL1b","REG3G"))) %>%
+  mutate(across(genes, factor, levels=c("CXCL1","TNFa","MUC1","IL1b","REG3G"))) %>%
   group_by(genes, groups) %>% 
-  summarise(mean_fold_change = mean(relative_fold_change)) %>% 
   ggplot(
     aes(x = groups,
-        y = mean_fold_change,
+          y = `2^-(ddct)`,
         fill = groups)
   ) +
-  geom_col() +
+  #geom_boxplot() +
+  geom_bar(position = "dodge", stat = "summary", fun = "mean") +
   facet_wrap(~genes,
@@
     legend.position = "none") + 
-    scale_fill_brewer(palette="Set1")
+    scale_fill_brewer(palette="Set1") +
+  stat_compare_means(aes(label = ..p.signif..),
+                   method = "wilcox.test", ref.group = "VANCO-BF",
+                   label.y.npc = c(0.55,0.45),
+                   label.x = c(3,1))
+  #stat_compare_means(comparisons = my_comparisons, label.y.npc = c(0.3,0.5), method = "wilcox.test") 
   increased_plot

### 2022-11-26 code/cytokines/Cytokines_tissue_adjusted.Rmd
updated gene expression

@@
     y = "Tissue weight (mg)",
-    x = "",
-    caption = "Pairwise comparison with CON-BF as reference. \n
-    **: p < 0.01, *: p < 0.05, NS: not significant."
+    x = ""
+    #caption = "Pairwise comparison with CON-BF as reference. \n
+    #**: p < 0.01, *: p < 0.05, NS: not significant."
   ) 
@@
    filter(exp_number == 3) %>%
+  filter(group != "VANCO-BF") %>% 
   ggplot(aes(x = group, y = tissue_weight_mg, fill = group)) + 
@@
     label.y.npc = 0.75,
-    label.x.npc = 0.3) +
+    label.x.npc = 0.5) +
   geom_point() +
@@
     y = "Tissue weight (mg)",
-    x = "",
-    caption = "Pairwise comparison with VANCO-BF as reference. \n
-    ****: p < 0.0005, ***: p < 0.001, NS: not significant."
+    x = ""
+    #caption = "Pairwise comparison with VANCO-BF as reference. \n
+    #****: p < 0.0005, ***: p < 0.001, NS: not significant."
   ) 

### 2022-11-30 code/gene_expression/Gene expression.Rmd
Final commit for submission of project

--- 2022-11-26 code/Gene expression.Rmd
@@
 
+library(g)
+library(facetscales)
+scales_y <- list(
+  `4` = scale_y_continuous(limits = c(5, 25), breaks = seq(5, 25, 5)),
+  `f` = scale_y_continuous(limits = c(0, 40), breaks = seq(0, 40, 10)),
+  `r` = scale_y_continuous(limits = c(10, 20), breaks = seq(10, 20, 2))
+)
+ggplot(mpg, aes(displ, cty)) + 
+  geom_point() + 
+  facet_grid_sc(rows = vars(drv), scales = list(y = scales_y))
+
 ```{r}
@@
 ```{r}
-anova <- delta_gene_expr %>% 
-  group_by(genes) %>% 
-  anova_test(dct ~ groups) %>% 
-  ungroup()
-
 significant <- delta_gene_expr %>% 
@@
   filter(p.value <= 0.05)
+
+anova <- delta_gene_expr %>% 
+  group_by(genes) %>% 
+  rstatix::anova_test(dct ~ groups)
+
+tukey <- delta_gene_expr %>% 
+  group_by(genes) %>% 
+  rstatix::tukey_hsd(dct ~ groups) %>% 
+  filter(p.adj < 0.05) %>% 
+  flextable::as_flextable()
+
+delta_gene_expr$group
+
 ```
@@
       )
+```
 
-delta_gene_expr %>% 
+## decreased_fold_change
+```{r decreased-fold-change}
+decreased_fold_change <- delta_gene_expr %>% 
+  filter(genes %in% tukey$genes) %>% 
+  filter(!genes %in% c("IL1b","REG3G")) %>% 
+  mutate(across(genes, factor, levels=c("TGFb","cldn15","TJP1","VEGFa","Ocld1"))) %>%
   group_by(genes, groups) %>% 
@@
   ) + 
-  geom_boxplot() +
+  geom_boxplot(outlier.fill = NA,
+               outlier.shape = NA) +
   facet_wrap(~genes,
@@
              scales = "free") +
-  labs(title = "Gene expression fold change",
-       y = "Gene expression fold change",
-       x = "Groups"
-      )
+   stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label.y = c(1.2,1.4)) +
+  labs(
+       y = "Fold Δ",
+       x = "",
+       fill = "Groups") +
+  theme_bw()+
+  theme(                   
+    #axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.position = "none") + 
+    scale_fill_brewer(palette="Set1")
+  decreased_fold_change
+
+```
+
+## Increased fold change
+```{r increased-fold-change}
+increased_fold_change <- delta_gene_expr %>% 
+  #filter(genes %in% tukey$genes) %>% 
+  filter(genes %in% c("IL1b","REG3G")) %>% 
+ # mutate(across(genes, factor, levels=c("TGFb","cldn15","TJP1","VEGFa","Ocld1"))) %>%
+  group_by(genes, groups) %>% 
+  ggplot(
+    aes(x = groups,
+        y = `2^-(ddct)`,
+        fill = groups)
+  ) + 
+  geom_boxplot(outlier.fill = NA,
+               outlier.shape = NA) +
+  facet_wrap(~genes,
+             ncol = 2,
+             scales = "free") +
+   stat_compare_means(comparisons = my_comparisons, method = "wilcox.test", label.y.npc = c(0.5)) +
+  labs(
+       y = "Fold Δ",
+       x = "",
+       fill = "Groups") +
+  theme_bw()+
+  theme(                   
+    #axis.title.y = element_text(color="darkgrey", size=10, face="bold.italic"),
+    axis.ticks.x = element_blank(),  # Hide labels
+    axis.text.x = element_blank(),
+    legend.position = "none") + 
+    scale_fill_brewer(palette="Set1")
+  increased_fold_change
 
@@
 ```{r}
-my_comparisons <- list(c("VANCO-FVF", "VANCO-SM"), c("VANCO-BF", "VANCO-FVF"))
+my_comparisons <- list(c("VANCO-BF", "VANCO-FVF"), c("VANCO-SM", "VANCO-FVF"))
 ```

### 2023-03-13 manuscript/manus.Rmd
Publication setup
Created a folder for manuscript files. Next step is to write outline of matrinem story, and consider what to include.

---
output:
  bookdown::pdf_document2: 
    toc: no
    keep_tex: yes
  #bookdown::word_document2: default
fig_caption: yes
always_allow_html: yes
bibliography: 
  references.bib
#csl: biomed-central.csl #get Zotero running to work
# These are LaTex settings to take care of floating figures/tables, line spacing, etc
header-includes:
  - \usepackage{endfloat}    
  - \usepackage{setspace}\doublespacing
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage[utf8]{inputenc}
  #- \DeclareUnicodeCharacter{0301}{*******} #Used for finding Unicode Errors in PDF
---

```{r Setup..., message=FALSE, warning=FALSE, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'p') # Places figures on their own pages
knitr::opts_chunk$set(out.width = '100%', dpi=300)

# Load packages
library(bookdown); library(knitr); library(kableExtra); library(tidyverse); library(here); library(readxl)

```

# The great Matrinem Study

\vspace{20 mm}

Kaare D. Tranaes${^1}$, Anders Brunse${^2}$${^\dagger}$

\vspace{10 mm}

${^1}$ Department of Food Microbiology, UCPH  
${^2}$AEM, SUND

${^\dagger}$ To whom correspondance should be adressed

\newpage

## Abstract:

## Introduction

### About Necrotizing Entericolitis

Nescrotizing Enterocolitis (NEC) is a devastating inflammatory and necrotic bowel disease, representing the leading cause of death in premature infants [@Nio2016]. It is estimated that upwards of 10% of neonates born before week 33 of gestation are affected. The disease is largely incurable, and mortality ranges from 20-50% depending on disease severity [@Alsaied2020]. Those who survive often do so with severe, life-long complications [@Papillon2013]. There is no universal definition of NEC [@Patel2020], and entities use different protocols for diagnosing the disease. The most commonly used diagnostic methodology is a modified version of the 1978 Bell Stage [@BELL1978], which categorizes the disease on a severity scale from one to three, with one and two being medically treated and stage three requiring surgical intervention [@Sheikh2010]. The typical neonate with NEC is a thriving premature infant that suddenly presents with feeding intolerance, abdominal distension, bloody stools, and signs of sepsis (temperature-, blood pressure-, heart- and respiratory rate change) [@Sharma2013]. Early-stage treatment involves cessation of oral feeding in favor of parenteral nutrition, nasogastric tube suction, and antibiotic therapy. Metabolic acidosis and radiological detection of pneumatosis intestinalis - gas cysts in the intestinal wall - indicates further disease progression, and barring surgery, additional treatment is limited to metabolic stabilization. The final stage of the disease is characterized by respiratory and circulatory arrest, peritoneal fluid buildup (ascites), and pneumoperitoneum (gas within the peritoneal cavity), which requires laparotomy with subsequent excision of necrotic intestinal tissue [@Stoll2015]. Overall mortality for surgically treated NEC cases is high (20-50%) as patients experience recurrence, intestinal strictures, short bowel syndrome, growth delay, and neurodevelopmental impairment [@Papillon2013]. The exact etiology of NEC is still unclear, although dozens of predisposing risk factors have been identified. Those with the greatest risk include low gestational age, bacterial colonization, and administration of milk replacement formula [@Nio2016]. Gestational age is the primary risk factor associated with NEC [@Samuels2017], and infants that bypass week 38 of gestation are at exceedingly low risk of developing NEC. Interestingly, there appears to be an inverse relationship between the degree of prematurity and the delay of disease onset. Infants born around week 27 present with NEC on average 32 days from birth, whereas infants born after closer to week 37 present with NEC on average only seven days from delivery [@Stoll2015]. Bacterial colonization of the gut is still not fully understood but begins before birth and, with time, increases in both abundance and diversity [@Lawn2014]. Preterm delivery likely upsets the normal development, and perturbation of the gut microbiome (often called a *dysbiosis*) is strongly associated with NEC. The term *microbiome* encompasses the entirety of non-host genes found in the gastrointestinal tract and includes mainly bacteria and viruses (bacteriophages or *phages*) but also fungi, archaea, and protozoa [@Lynch2016]. In the case of premature infants, the underdeveloped host immune system is likely ill-equipped to deal with intestinal dysbiosis. Little is known about the exact mechanisms of disease development, but the loss of a functional mucus layer is believed to be a prerequisite in early-stage NEC development [@Liu2020]. In term infants, the intestinal epithelium is protected from bacteria by thick mucus layers of gel-forming glycoproteins called mucins[@Johansson2013]. Mucin-2 (Muc2) is the main secretory mucin in the small intestine [@Ermund2013], where a single mucus layer protects the enterocytes from pathogens and foodstuff while simultaneously allowing for the dynamic absorption of nutrients. If the mucus layer becomes permeable, bacterial infiltration of the apical-facing tissue will elicit the inflammatory immune response likely causative of intestinal permeability, inflammation and subsequent tissue necrosis characteristic of the disease. Support for this hypothesis comes from studies in Muc2-deficient mice, which have been shown to develop colitis when challenged with a colitis-inducing agent, dextran sulfate sodium (DSS) [@VanderSluis2006]. Further support comes in the form of a recently published study on human NEC tissue, finding severely altered levels of Muc2 in the tissue of NEC neonates but not in equally premature and NEC-free infants [@Liu2020]. A central pathway of this inflammatory response is mediated by Toll-like-receptor 4 (TLR-4) stimulation [@Hackam2014]. TLR-4 is a pattern recognition receptor sensitive to the lipopolysaccharide (LPS) residues on the surface of gram-negative bacteria. Upon ligand binding to the TLR-4 receptor, IL-8 (CXCL1 or KC/GRO in mice) and TNF~alpha~ are released from local macrophages and subsequently recruit effector cells to the site of infection [@Guang2010]. In turn, MUC1, a membrane-bound mucin on the surface of most epithelial cells, is upregulated and suppresses TLR-4 signaling, attenuating further inflammation [@Kyo2012]. Upregulated expression of both TLR-4 and MUC1 have been shown to increase susceptibility to NEC, highlighting the importance of the bilateral regulation of inflammation [@McAuley2007, @Ng2016]. With such a wide range of clinical and biochemical markers, it is unsurprising that animal models of NEC, each focusing on specific "NEC-like" symptoms (lesions, pneumatosis intestinalis, inflammation, or biomarkers of enterocyte damage), number in the several hundred [@Mendez2020].

### Animal models of NEC

There is no "gold standard" animal model for the study of NEC. Current models range from piglets to rabbits, rats, mice, quails, and even non-human primates. Of these, the rat model is most commonly used due to its cost-effectiveness and the robustness of rats to the stressors used to induce NEC [@Mendez2020]. This model relies on an initial insult to the intestines, typically bouts of oxygen deprivation, hypothermia, and DSS [@Sodhi2008]. For a long time, this approach was thought to mimic the etiology of human infant NEC. However, epidemiologic evidence suggests that the most common form of NEC is not triggered by a hypoxic-ischemic event [@Young2011], and several authors have questioned the translational value of some of the existing models [@Patel2020]. Despite the high costs and difficulty associated with the model, several models using piglets have been described in the literature [@Sodhi2008]. Successful models rely on prematurity (c-section) and formula-feeding to induce the disease, two of the main factors contributing to human NEC. Whether the development of NEC in association with formula feeding represents the presence of a harmful component in infant formula or the absence of the passive immunity awarded by agents only present in breast milk remains to be determined [@LeHurouLuron2010]. In a recent study, NEC was successfully mitigated in preterm c-section piglets on formula-replacement by administration of a fecal filtrate containing viral particles (phages) [@Brunse2021]. Phages have been used to treat bacterial infection diseases [@Ott2017] and likely mitigate their beneficial effect by manipulating the gut microbiome. The NEC-associated dysbiosis seen in the study's control animals was alleviated along with the subsequent disease. These findings helped pave the way for the preapproval of a human trial, PREPHAGE, in which premature neonates will receive virus-containing fecal filtrates to stabilize the suspected intestinal dysbiosis and thus prevent disease development.

### Maternal Antibiotics Treatment

Although highly successful as a model of the human neonate, there are disadvantages to working with pigs. Costs and long gestation means a lower scientific throughput, and a faster and cheaper alternative would speed up development. The use of mice has historically been discouraging because of the difficulty in sustaining a neonatal model (a mouse pup at term weighs approx. 1.5g compared with 14g for an equal term rat pup), making it difficult to work with very young mouse offspring. A recently published study [@Chen2021] investigating the effects of Maternal Antibiotics Treatment (MAT) during pregnancy found that a moderate dose of antibiotics administered in the drinking water of pregnant C57BL/6N mice on pregnancy day 15 resulted in alterations in the intestinal microbiome composition in the offspring. Furthermore, offspring from antibiotic-treated mice had lower birth weights, decreased expression of intestinal barrier proteins, and greater intestinal injury scores than offspring from non-antibiotic-treated mice. These effects were seen in breastfed mice, despite the passive immunity and maturating effects known to come with breastmilk [@Barlow1974]. These findings suggest that MAT potentially caused growth- and developmental retardation. Similar to the human babies, the offspring's slightly less mature immune system potentially rendered the pups more susceptible to intestinal damage from the reported microbiome dysbiosis. Although the authors did not categorize their findings as NEC-like, they share several characteristics.

### Aim

If developmental retardation could be achieved with MAT, we hypothesized that replacing breastmilk with formula feeding would halt intestinal maturation while simultaneously providing a barrier stressor. We, therefore, decided to test the viability of a model based on 3-day-old mice from maternally antibiotic-treated mothers, in which breastmilk was replaced with formula as soon as possible. This required oral feed administration of pups \<2g, which we were not sure would be technically feasible. Therefore, this project aimed to establish a mouse model of necrotizing enterocolitis with a similar etiology as that proposed in preterm human babies, i.e., intestinal immaturity and gut microbiome dysbiosis, by combining maternal antibiotic treatment with formula feeding.
\newpage

## Results
```{r results}
# Load weight data
data <- read_excel(here::here("data", "processed", "weight_development.xlsx")) %>% 
  select(sample_id:cage, bodyweight_birthday_group_average, bodyweight_birthday_plus_24H_group_average, bodyweight_birthday_plus_48H_group_average, bodyweight_baseline, bodyweight_baseline_plus_24h, bodyweight_baseline_plus_48h) %>% 
  rename(day_0 = "bodyweight_birthday_group_average",
         day_1 = "bodyweight_birthday_plus_24H_group_average", 
         day_2 = "bodyweight_birthday_plus_48H_group_average",
         day_3 = bodyweight_baseline, 
         day_4 = `bodyweight_baseline_plus_24h`, 
         day_5 = `bodyweight_baseline_plus_48h`,) %>% 
  pivot_longer(., cols = c(day_0:day_5),
                names_to = "time",
                values_to = "bodyweight") %>% 
  mutate(received_antibiotics = recode_factor(received_antibiotics, 
                                       Yes = "MAT",
                                       No = "CON"))
```

\newpage

## Discussion
\newpage

## Conclusion
\newpage

## Methods
\newpage
```{r figRef, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Figure caption here"}
## R code here, with plot output at end, for example using ggplot2

```

## References

### 2023-04-24 manuscript/manus.Rmd
Prepared for manuscript typesetting in R

@@
 output:
-  bookdown::pdf_document2: 
-    toc: no
-    keep_tex: yes
-  #bookdown::word_document2: default
+  #bookdown::pdf_document2: 
+    #toc: no
+    #keep_tex: yes
+  bookdown::word_document2: default
 fig_caption: yes
@@
 
-${^\dagger}$ To whom correspondance should be adressed
+${^\dagger}$ To whom correspondence should be adressed
 

### 2023-04-24 manuscript/manus.tex
Prepared for manuscript typesetting in R

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\usepackage{endfloat}
\usepackage{setspace}\doublespacing
\usepackage{lineno}
\linenumbers
\usepackage[utf8]{inputenc}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{\vspace{-2.5em}}

\begin{document}

\hypertarget{the-great-matrinem-study}{%
\section{The great Matrinem Study}\label{the-great-matrinem-study}}

\vspace{20 mm}

Kaare D. Tranaes\({^1}\), Anders Brunse\({^2}\)\({^\dagger}\)

\vspace{10 mm}

\({^1}\) Department of Food Microbiology, UCPH\\
\({^2}\)AEM, SUND

\({^\dagger}\) To whom correspondance should be adressed

\newpage

\hypertarget{abstract}{%
\subsection{Abstract:}\label{abstract}}

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

\hypertarget{about-necrotizing-entericolitis}{%
\subsubsection{About Necrotizing Entericolitis}\label{about-necrotizing-entericolitis}}

Nescrotizing Enterocolitis (NEC) is a devastating inflammatory and necrotic bowel disease, representing the leading cause of death in premature infants (Niño, Sodhi, and Hackam 2016). It is estimated that upwards of 10\% of neonates born before week 33 of gestation are affected. The disease is largely incurable, and mortality ranges from 20-50\% depending on disease severity (Alsaied, Islam, and Thalib 2020). Those who survive often do so with severe, life-long complications (Papillon et al. 2013). There is no universal definition of NEC (Patel et al. 2020), and entities use different protocols for diagnosing the disease. The most commonly used diagnostic methodology is a modified version of the 1978 Bell Stage (BELL et al. 1978), which categorizes the disease on a severity scale from one to three, with one and two being medically treated and stage three requiring surgical intervention (Sheikh and Gaillard 2010). The typical neonate with NEC is a thriving premature infant that suddenly presents with feeding intolerance, abdominal distension, bloody stools, and signs of sepsis (temperature-, blood pressure-, heart- and respiratory rate change) (Sharma and Hudak 2013). Early-stage treatment involves cessation of oral feeding in favor of parenteral nutrition, nasogastric tube suction, and antibiotic therapy. Metabolic acidosis and radiological detection of pneumatosis intestinalis - gas cysts in the intestinal wall - indicates further disease progression, and barring surgery, additional treatment is limited to metabolic stabilization. The final stage of the disease is characterized by respiratory and circulatory arrest, peritoneal fluid buildup (ascites), and pneumoperitoneum (gas within the peritoneal cavity), which requires laparotomy with subsequent excision of necrotic intestinal tissue (Stoll et al. 2015). Overall mortality for surgically treated NEC cases is high (20-50\%) as patients experience recurrence, intestinal strictures, short bowel syndrome, growth delay, and neurodevelopmental impairment (Papillon et al. 2013). The exact etiology of NEC is still unclear, although dozens of predisposing risk factors have been identified. Those with the greatest risk include low gestational age, bacterial colonization, and administration of milk replacement formula (Niño, Sodhi, and Hackam 2016). Gestational age is the primary risk factor associated with NEC (Samuels et al. 2017), and infants that bypass week 38 of gestation are at exceedingly low risk of developing NEC. Interestingly, there appears to be an inverse relationship between the degree of prematurity and the delay of disease onset. Infants born around week 27 present with NEC on average 32 days from birth, whereas infants born after closer to week 37 present with NEC on average only seven days from delivery (Stoll et al. 2015). Bacterial colonization of the gut is still not fully understood but begins before birth and, with time, increases in both abundance and diversity (Lawn et al. 2014). Preterm delivery likely upsets the normal development, and perturbation of the gut microbiome (often called a \emph{dysbiosis}) is strongly associated with NEC. The term \emph{microbiome} encompasses the entirety of non-host genes found in the gastrointestinal tract and includes mainly bacteria and viruses (bacteriophages or \emph{phages}) but also fungi, archaea, and protozoa (Lynch and Pedersen 2016). In the case of premature infants, the underdeveloped host immune system is likely ill-equipped to deal with intestinal dysbiosis. Little is known about the exact mechanisms of disease development, but the loss of a functional mucus layer is believed to be a prerequisite in early-stage NEC development (Liu et al. 2020). In term infants, the intestinal epithelium is protected from bacteria by thick mucus layers of gel-forming glycoproteins called mucins(Johansson, Sjövall, and Hansson 2013). Mucin-2 (Muc2) is the main secretory mucin in the small intestine (Ermund et al. 2013), where a single mucus layer protects the enterocytes from pathogens and foodstuff while simultaneously allowing for the dynamic absorption of nutrients. If the mucus layer becomes permeable, bacterial infiltration of the apical-facing tissue will elicit the inflammatory immune response likely causative of intestinal permeability, inflammation and subsequent tissue necrosis characteristic of the disease. Support for this hypothesis comes from studies in Muc2-deficient mice, which have been shown to develop colitis when challenged with a colitis-inducing agent, dextran sulfate sodium (DSS) (Sluis et al. 2006). Further support comes in the form of a recently published study on human NEC tissue, finding severely altered levels of Muc2 in the tissue of NEC neonates but not in equally premature and NEC-free infants (Liu et al. 2020). A central pathway of this inflammatory response is mediated by Toll-like-receptor 4 (TLR-4) stimulation (Lu, Sodhi, and Hackam 2014). TLR-4 is a pattern recognition receptor sensitive to the lipopolysaccharide (LPS) residues on the surface of gram-negative bacteria. Upon ligand binding to the TLR-4 receptor, IL-8 (CXCL1 or KC/GRO in mice) and TNF\textsubscript{alpha} are released from local macrophages and subsequently recruit effector cells to the site of infection (Guang et al. 2010). In turn, MUC1, a membrane-bound mucin on the surface of most epithelial cells, is upregulated and suppresses TLR-4 signaling, attenuating further inflammation (Kyo et al. 2012). Upregulated expression of both TLR-4 and MUC1 have been shown to increase susceptibility to NEC, highlighting the importance of the bilateral regulation of inflammation Ng and Sutton (2016). With such a wide range of clinical and biochemical markers, it is unsurprising that animal models of NEC, each focusing on specific ``NEC-like'' symptoms (lesions, pneumatosis intestinalis, inflammation, or biomarkers of enterocyte damage), number in the several hundred (Mendez et al. 2020).

\hypertarget{animal-models-of-nec}{%
\subsubsection{Animal models of NEC}\label{animal-models-of-nec}}

There is no ``gold standard'' animal model for the study of NEC. Current models range from piglets to rabbits, rats, mice, quails, and even non-human primates. Of these, the rat model is most commonly used due to its cost-effectiveness and the robustness of rats to the stressors used to induce NEC (Mendez et al. 2020). This model relies on an initial insult to the intestines, typically bouts of oxygen deprivation, hypothermia, and DSS (Sodhi et al. 2008). For a long time, this approach was thought to mimic the etiology of human infant NEC. However, epidemiologic evidence suggests that the most common form of NEC is not triggered by a hypoxic-ischemic event (Young, Kingma, and Neu 2011), and several authors have questioned the translational value of some of the existing models (Patel et al. 2020). Despite the high costs and difficulty associated with the model, several models using piglets have been described in the literature (Sodhi et al. 2008). Successful models rely on prematurity (c-section) and formula-feeding to induce the disease, two of the main factors contributing to human NEC. Whether the development of NEC in association with formula feeding represents the presence of a harmful component in infant formula or the absence of the passive immunity awarded by agents only present in breast milk remains to be determined (Huërou-Luron, Blat, and Boudry 2010). In a recent study, NEC was successfully mitigated in preterm c-section piglets on formula-replacement by administration of a fecal filtrate containing viral particles (phages) (Brunse et al. 2021). Phages have been used to treat bacterial infection diseases (Ott et al. 2017) and likely mitigate their beneficial effect by manipulating the gut microbiome. The NEC-associated dysbiosis seen in the study's control animals was alleviated along with the subsequent disease. These findings helped pave the way for the preapproval of a human trial, PREPHAGE, in which premature neonates will receive virus-containing fecal filtrates to stabilize the suspected intestinal dysbiosis and thus prevent disease development.

\hypertarget{maternal-antibiotics-treatment}{%
\subsubsection{Maternal Antibiotics Treatment}\label{maternal-antibiotics-treatment}}

Although highly successful as a model of the human neonate, there are disadvantages to working with pigs. Costs and long gestation means a lower scientific throughput, and a faster and cheaper alternative would speed up development. The use of mice has historically been discouraging because of the difficulty in sustaining a neonatal model (a mouse pup at term weighs approx. 1.5g compared with 14g for an equal term rat pup), making it difficult to work with very young mouse offspring. A recently published study (Chen, Chou, and Yang 2021) investigating the effects of Maternal Antibiotics Treatment (MAT) during pregnancy found that a moderate dose of antibiotics administered in the drinking water of pregnant C57BL/6N mice on pregnancy day 15 resulted in alterations in the intestinal microbiome composition in the offspring. Furthermore, offspring from antibiotic-treated mice had lower birth weights, decreased expression of intestinal barrier proteins, and greater intestinal injury scores than offspring from non-antibiotic-treated mice. These effects were seen in breastfed mice, despite the passive immunity and maturating effects known to come with breastmilk (Barlow et al. 1974). These findings suggest that MAT potentially caused growth- and developmental retardation. Similar to the human babies, the offspring's slightly less mature immune system potentially rendered the pups more susceptible to intestinal damage from the reported microbiome dysbiosis. Although the authors did not categorize their findings as NEC-like, they share several characteristics.

\hypertarget{aim}{%
\subsubsection{Aim}\label{aim}}

If developmental retardation could be achieved with MAT, we hypothesized that replacing breastmilk with formula feeding would halt intestinal maturation while simultaneously providing a barrier stressor. We, therefore, decided to test the viability of a model based on 3-day-old mice from maternally antibiotic-treated mothers, in which breastmilk was replaced with formula as soon as possible. This required oral feed administration of pups \textless2g, which we were not sure would be technically feasible. Therefore, this project aimed to establish a mouse model of necrotizing enterocolitis with a similar etiology as that proposed in preterm human babies, i.e., intestinal immaturity and gut microbiome dysbiosis, by combining maternal antibiotic treatment with formula feeding.
\newpage

\hypertarget{results}{%
\subsection{Results}\label{results}}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Load weight data}
\NormalTok{data }\OtherTok{\textless{}{-}} \FunctionTok{read\_excel}\NormalTok{(here}\SpecialCharTok{::}\FunctionTok{here}\NormalTok{(}\StringTok{"data"}\NormalTok{, }\StringTok{"processed"}\NormalTok{, }\StringTok{"weight\_development.xlsx"}\NormalTok{)) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{select}\NormalTok{(sample\_id}\SpecialCharTok{:}\NormalTok{cage, bodyweight\_birthday\_group\_average, bodyweight\_birthday\_plus\_24H\_group\_average, bodyweight\_birthday\_plus\_48H\_group\_average, bodyweight\_baseline, bodyweight\_baseline\_plus\_24h, bodyweight\_baseline\_plus\_48h) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{rename}\NormalTok{(}\AttributeTok{day\_0 =} \StringTok{"bodyweight\_birthday\_group\_average"}\NormalTok{,}
         \AttributeTok{day\_1 =} \StringTok{"bodyweight\_birthday\_plus\_24H\_group\_average"}\NormalTok{, }
         \AttributeTok{day\_2 =} \StringTok{"bodyweight\_birthday\_plus\_48H\_group\_average"}\NormalTok{,}
         \AttributeTok{day\_3 =}\NormalTok{ bodyweight\_baseline, }
         \AttributeTok{day\_4 =} \StringTok{\textasciigrave{}}\AttributeTok{bodyweight\_baseline\_plus\_24h}\StringTok{\textasciigrave{}}\NormalTok{, }
         \AttributeTok{day\_5 =} \StringTok{\textasciigrave{}}\AttributeTok{bodyweight\_baseline\_plus\_48h}\StringTok{\textasciigrave{}}\NormalTok{,) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{pivot\_longer}\NormalTok{(., }\AttributeTok{cols =} \FunctionTok{c}\NormalTok{(day\_0}\SpecialCharTok{:}\NormalTok{day\_5),}
                \AttributeTok{names\_to =} \StringTok{"time"}\NormalTok{,}
                \AttributeTok{values\_to =} \StringTok{"bodyweight"}\NormalTok{) }\SpecialCharTok{\%\textgreater{}\%} 
  \FunctionTok{mutate}\NormalTok{(}\AttributeTok{received\_antibiotics =} \FunctionTok{recode\_factor}\NormalTok{(received\_antibiotics, }
                                       \AttributeTok{Yes =} \StringTok{"MAT"}\NormalTok{,}
                                       \AttributeTok{No =} \StringTok{"CON"}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\newpage

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

\newpage

\hypertarget{conclusion}{%
\subsection{Conclusion}\label{conclusion}}

\newpage

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

\newpage

\hypertarget{references}{%
\subsection*{References}\label{references}}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-Alsaied2020}{}}%
Alsaied, Amer, Nazmul Islam, and Lukman Thalib. 2020. {``Global Incidence of Necrotizing Enterocolitis: A Systematic Review and Meta-Analysis.''} \emph{{BMC} Pediatrics} 20 (1). \url{https://doi.org/10.1186/s12887-020-02231-5}.

\leavevmode\vadjust pre{\hypertarget{ref-Barlow1974}{}}%
Barlow, Barbara, Thomas V. Santulli, William C. Heird, Jane Pitt, William A. Blanc, and John N. Schullinger. 1974. {``An Experimental Study of Acute Neonatal Enterocolitis{\textemdash}the Importance of Breast Milk.''} \emph{Journal of Pediatric Surgery} 9 (5): 587--95. \url{https://doi.org/10.1016/0022-3468(74)90093-1}.

\leavevmode\vadjust pre{\hypertarget{ref-BELL1978}{}}%
BELL, MARTIN J., JESSIE L. TERNBERG, RALPH D. FEIGIN, JAMES P. KEATING, RICHARD MARSHALL, LESLIE BARTON, and THOMAS BROTHERTON. 1978. {``Neonatal Necrotizing Enterocolitis.''} \emph{Annals of Surgery} 187 (1): 1--7. \url{https://doi.org/10.1097/00000658-197801000-00001}.

\leavevmode\vadjust pre{\hypertarget{ref-Brunse2021}{}}%
Brunse, Anders, Ling Deng, Xiaoyu Pan, Yan Hui, Josue L. Castro-Mejia, Witold Kot, Duc Ninh Nguyen, Jan Bojsen-Møller Secher, Dennis Sandris Nielsen, and Thomas Thymann. 2021. {``Fecal Filtrate Transplantation Protects Against Necrotizing Enterocolitis.''} \emph{The {ISME} Journal} 16 (3): 686--94. \url{https://doi.org/10.1038/s41396-021-01107-5}.

\leavevmode\vadjust pre{\hypertarget{ref-Chen2021}{}}%
Chen, Chung-Ming, Hsiu-Chu Chou, and Yu-Chen S. H. Yang. 2021. {``Maternal Antibiotic Treatment Disrupts the Intestinal Microbiota and Intestinal Development in Neonatal Mice.''} \emph{Frontiers in Microbiology} 12 (June). \url{https://doi.org/10.3389/fmicb.2021.684233}.

\leavevmode\vadjust pre{\hypertarget{ref-Ermund2013}{}}%
Ermund, Anna, André Schütte, Malin E. V. Johansson, Jenny K. Gustafsson, and Gunnar C. Hansson. 2013. {``Studies of Mucus in Mouse Stomach, Small Intestine, and Colon. I. Gastrointestinal Mucus Layers Have Different Properties Depending on Location as Well as over the Peyer{\textquotesingle}s Patches.''} \emph{American Journal of Physiology-Gastrointestinal and Liver Physiology} 305 (5): G341--47. \url{https://doi.org/10.1152/ajpgi.00046.2013}.

\leavevmode\vadjust pre{\hypertarget{ref-Guang2010}{}}%
Guang, Wei, Hua Ding, Steven J. Czinn, K. Chul Kim, Thomas G. Blanchard, and Erik P. Lillehoj. 2010. {``Muc1 Cell Surface Mucin Attenuates Epithelial Inflammation in Response to a Common Mucosal Pathogen.''} \emph{Journal of Biological Chemistry} 285 (27): 20547--57. \url{https://doi.org/10.1074/jbc.m110.121319}.

\leavevmode\vadjust pre{\hypertarget{ref-LeHurouLuron2010}{}}%
Huërou-Luron, Isabelle Le, Sophie Blat, and Gaëlle Boudry. 2010. {``Breast Vs. Formula-Feeding Formula-Feeding: Impacts on the Digestive Tract and Immediate and Long-Term Health Effects.''} \emph{Nutrition Research Reviews} 23 (1): 23--36. \url{https://doi.org/10.1017/s0954422410000065}.

\leavevmode\vadjust pre{\hypertarget{ref-Johansson2013}{}}%
Johansson, Malin E. V., Henrik Sjövall, and Gunnar C. Hansson. 2013. {``The Gastrointestinal Mucus System in Health and Disease.''} \emph{Nature Reviews Gastroenterology and Hepatology} 10 (6): 352--61. \url{https://doi.org/10.1038/nrgastro.2013.35}.

\leavevmode\vadjust pre{\hypertarget{ref-Kyo2012}{}}%
Kyo, Yoshiyuki, Kosuke Kato, Yong Sung Park, Sachin Gajhate, Tsuyoshi Umehara, Erik P. Lillehoj, Harumi Suzaki, and Kwang Chul Kim. 2012. {``Antiinflammatory Role of {MUC}1 Mucin During Infection with Nontypeable Haemophilus Influenzae.''} \emph{American Journal of Respiratory Cell and Molecular Biology} 46 (2): 149--56. \url{https://doi.org/10.1165/rcmb.2011-0142oc}.

\leavevmode\vadjust pre{\hypertarget{ref-Lawn2014}{}}%
Lawn, Joy E, Hannah Blencowe, Shefali Oza, Danzhen You, Anne CC Lee, Peter Waiswa, Marek Lalli, et al. 2014. {``Every Newborn: Progress, Priorities, and Potential Beyond Survival.''} \emph{The Lancet} 384 (9938): 189--205. \url{https://doi.org/10.1016/s0140-6736(14)60496-7}.

\leavevmode\vadjust pre{\hypertarget{ref-Liu2020}{}}%
Liu, Dong, Yanzhen Xu, Jinxing Feng, Jialin Yu, Jinjie Huang, and Zhiguang Li. 2020. {``Mucins and Tight Junctions Are Severely Altered in Necrotizing Enterocolitis Neonates.''} \emph{American Journal of Perinatology} 38 (11): 1174--80. \url{https://doi.org/10.1055/s-0040-1710558}.

\leavevmode\vadjust pre{\hypertarget{ref-Hackam2014}{}}%
Lu, Peng, Chhinder P. Sodhi, and David J. Hackam. 2014. {``Toll-Like Receptor Regulation of Intestinal Development and Inflammation in the Pathogenesis of Necrotizing Enterocolitis.''} \emph{Pathophysiology} 21 (1): 81--93. \url{https://doi.org/10.1016/j.pathophys.2013.11.007}.

\leavevmode\vadjust pre{\hypertarget{ref-Lynch2016}{}}%
Lynch, Susan V., and Oluf Pedersen. 2016. {``The Human Intestinal Microbiome in Health and Disease.''} Edited by Elizabeth G. Phimister. \emph{New England Journal of Medicine} 375 (24): 2369--79. \url{https://doi.org/10.1056/nejmra1600266}.

\leavevmode\vadjust pre{\hypertarget{ref-McAuley2007}{}}%
McAuley, Julie L., Sara K. Linden, Chin Wen Png, Rebecca M. King, Helen L. Pennington, Sandra J. Gendler, Timothy H. Florin, Geoff R. Hill, Victoria Korolik, and Michael A. McGuckin. 2007. {``{MUC}1 Cell Surface Mucin Is a Critical Element of the Mucosal Barrier to Infection.''} \emph{Journal of Clinical Investigation} 117 (8): 2313--24. \url{https://doi.org/10.1172/jci26705}.

\leavevmode\vadjust pre{\hypertarget{ref-Mendez2020}{}}%
Mendez, Yomara Stephanie, Faraz A Khan, Gregory Van Perrier, and Andrei Radulescu. 2020. {``Animal Models of Necrotizing Enterocolitis.''} \emph{World Journal of Pediatric Surgery} 3 (1): e000109. \url{https://doi.org/10.1136/wjps-2020-000109}.

\leavevmode\vadjust pre{\hypertarget{ref-Ng2016}{}}%
Ng, G Z, and P Sutton. 2016. {``The {MUC}1 Mucin Specifically Inhibits Activation of the {NLRP}3 Inflammasome.''} \emph{Genes and Immunity} 17 (3): 203--6. \url{https://doi.org/10.1038/gene.2016.10}.

\leavevmode\vadjust pre{\hypertarget{ref-Nio2016}{}}%
Niño, Diego F., Chhinder P. Sodhi, and David J. Hackam. 2016. {``Necrotizing Enterocolitis: New Insights into Pathogenesis and Mechanisms.''} \emph{Nature Reviews Gastroenterology and Hepatology} 13 (10): 590--600. \url{https://doi.org/10.1038/nrgastro.2016.119}.

\leavevmode\vadjust pre{\hypertarget{ref-Ott2017}{}}%
Ott, Stephan J., Georg H. Waetzig, Ateequr Rehman, Jacqueline Moltzau-Anderson, Richa Bharti, Juris A. Grasis, Liam Cassidy, et al. 2017. {``Efficacy of Sterile Fecal Filtrate Transfer for Treating Patients with Clostridium Difficile Infection.''} \emph{Gastroenterology} 152 (4): 799--811.e7. \url{https://doi.org/10.1053/j.gastro.2016.11.010}.

\leavevmode\vadjust pre{\hypertarget{ref-Papillon2013}{}}%
Papillon, Stephanie, Shannon L. Castle, Christopher P. Gayer, and Henri R. Ford. 2013. {``Necrotizing Enterocolitis.''} \emph{Advances in Pediatrics} 60 (1): 263--79. \url{https://doi.org/10.1016/j.yapd.2013.04.011}.

\leavevmode\vadjust pre{\hypertarget{ref-Patel2020}{}}%
Patel, Ravi Mangal, Joanne Ferguson, Steven J. McElroy, Minesh Khashu, and Michael S. Caplan. 2020. {``Defining Necrotizing Enterocolitis: Current Difficulties and Future Opportunities.''} \emph{Pediatric Research} 88 (S1): 10--15. \url{https://doi.org/10.1038/s41390-020-1074-4}.

\leavevmode\vadjust pre{\hypertarget{ref-Samuels2017}{}}%
Samuels, Noor, Rob A. van de Graaf, Rogier C. J. de Jonge, Irwin K. M. Reiss, and Marijn J. Vermeulen. 2017. {``Risk Factors for Necrotizing Enterocolitis in Neonates: A Systematic Review of Prognostic Studies.''} \emph{{BMC} Pediatrics} 17 (1). \url{https://doi.org/10.1186/s12887-017-0847-3}.

\leavevmode\vadjust pre{\hypertarget{ref-Sharma2013}{}}%
Sharma, Renu, and Mark Lawrence Hudak. 2013. {``A Clinical Perspective of Necrotizing Enterocolitis.''} \emph{Clinics in Perinatology} 40 (1): 27--51. \url{https://doi.org/10.1016/j.clp.2012.12.012}.

\leavevmode\vadjust pre{\hypertarget{ref-Sheikh2010}{}}%
Sheikh, Yusra, and Frank Gaillard. 2010. {``Necrotising Enterocolitis (Staging).''} Radiopaedia.org. \url{https://doi.org/10.53347/rid-8456}.

\leavevmode\vadjust pre{\hypertarget{ref-VanderSluis2006}{}}%
Sluis, Maria Van der, Barbara A. E. De Koning, Adrianus C. J. M. De Bruijn, Anna Velcich, Jules P. P. Meijerink, Johannes B. Van Goudoever, Hans A. Büller, et al. 2006. {``Muc2-Deficient Mice Spontaneously Develop Colitis, Indicating That {MUC}2 Is Critical for Colonic Protection.''} \emph{Gastroenterology} 131 (1): 117--29. \url{https://doi.org/10.1053/j.gastro.2006.04.020}.

\leavevmode\vadjust pre{\hypertarget{ref-Sodhi2008}{}}%
Sodhi, Chhinder, Ward Richardson, Steven Gribar, and David J. Hackam. 2008. {``The Development of Animal Models for the Study of Necrotizing Enterocolitis.''} \emph{Disease Models and Mechanisms} 1 (2-3): 94--98. \url{https://doi.org/10.1242/dmm.000315}.

\leavevmode\vadjust pre{\hypertarget{ref-Stoll2015}{}}%
Stoll, Barbara J., Nellie I. Hansen, Edward F. Bell, Michele C. Walsh, Waldemar A. Carlo, Seetha Shankaran, Abbot R. Laptook, et al. 2015. {``Trends in Care Practices, Morbidity, and Mortality of Extremely Preterm Neonates, 1993-2012.''} \emph{{JAMA}} 314 (10): 1039. \url{https://doi.org/10.1001/jama.2015.10244}.

\leavevmode\vadjust pre{\hypertarget{ref-Young2011}{}}%
Young, Christopher M., Sandra D. K. Kingma, and Josef Neu. 2011. {``Ischemia-Reperfusion and Neonatal Intestinal Injury.''} \emph{The Journal of Pediatrics} 158 (2): e25--28. \url{https://doi.org/10.1016/j.jpeds.2010.11.009}.

\end{CSLReferences}

\end{document}

### 2023-04-24 code/weight_analyses/weight_analysis.Rmd
Update weight_analysis.Rmd
Added here() for relative path building

--- 2022-11-20 code/weight_analysis.Rmd
@@
 library(broom)
+library(here)
 ```
@@
 ```{r}
-data <- read_excel("data/processed/weight_development.xlsx") %>% 
+data <- read_excel(here("data", "processed","weight_development.xlsx")) %>% 
   select(sample_id:cage, bodyweight_birthday_group_average, bodyweight_birthday_plus_24H_group_average, bodyweight_birthday_plus_48H_group_average, bodyweight_baseline, bodyweight_baseline_plus_24h, bodyweight_baseline_plus_48h) %>%
